### 2015-03-28 drafts/anderson-etal-fisheries-ensembles.Rmd
Add some notes

---
output: html_document
bibliography: ../notes/fisheries-ensembles.bib
csl: ../notes/cjfas.csl
---

# Combining data-limited fisheries models to derive robust estimates of stock status

Sean C. Anderson, Jamie Afflerbach, Andrew B. Cooper, Mark Dickey-Collas, Olaf P. Jensen, Kristin M. Kleisner, Catherine Longo, Cóilín Minto, Giacomo Chato Osio, Dan Ovando, Andrew A. Rosenberg, Elizabeth R. Selig, James T. Thorson, and potentially others...

# Introduction

@greene2006: probably the closest thing to what we're doing but with climate models (Bayesian multilevel ensemble)

# Methods

# Results

# Discussion

# Acknowledgements

# Figures

# References

### 2015-05-05 7-slope-intercept-response.R
Add in-progress script for new ensemble models

@@ -1,0 +1,166 @@
+library(dplyr)
+library(ggplot2)
+if (!file.exists("generated-data/cv-gbms/")) system("mkdir generated-data/cv-gbms")
+
+dsim <- readRDS("raw-data/batch1-results.rds")
+
+dsim <- dsim %>%
+  transform(stock_id =
+  paste0(stock_id, "_sigmaC_", sigmaC, "_sigmaR_", sigmaR, "_lh", LH, "_it_", iter)) %>%
+  arrange(stock_id, iter, year)
+
+if(!file.exists("generated-data/dsim_sum.rds")) {
+  dsim_sum <- dsim %>%
+    group_by(stock_id, method_id, LH) %>%
+    do({
+
+      years_mean <- 3
+      .n <- nrow(.)
+      i <- seq(.n-(years_mean-1), .n)
+
+      bbmsy_true_mean = mean(.$b_bmsy_true[i])
+      bbmsy_est_mean = mean(.$b_bmsy_est[i])
+
+      ytrue <- .$b_bmsy_true[i]
+      yest <- .$b_bmsy_est[i]
+      bbmsy_true_slope <- coef(mblm::mblm(ytrue ~ i))[[2]]
+      bbmsy_est_slope <- coef(mblm::mblm(yest ~ i))[[2]]
+
+      data.frame(bbmsy_true_mean, bbmsy_est_mean, bbmsy_true_slope, bbmsy_est_slope)
+    }) %>% as.data.frame
+    saveRDS(dsim_sum, file = "generated-data/dsim_sum.rds")
+} else {
+  dsim_sum <- readRDS("generated-data/dsim_sum.rds")
+}
+
+dsim_meta <- dsim %>%
+  group_by(stock_id) %>%
+  summarise(
+    max_catch = max(catch),
+    total_catch = sum(catch))
+
+dsim_sum <- inner_join(dsim_sum, dsim_meta)
+
+trues <- select(dsim_sum, stock_id, method_id, bbmsy_true_mean, bbmsy_true_slope)
+trues <- trues[!duplicated(trues), ]
+
+d_mean <- reshape2::dcast(dsim_sum, stock_id + LH ~ method_id,
+  value.var = "bbmsy_est_mean")  %>%
+  inner_join(select(trues, -bbmsy_true_slope))
+
+d_slope <- reshape2::dcast(dsim_sum, stock_id + LH ~ method_id,
+  value.var = "bbmsy_est_slope") %>%
+  inner_join(select(trues, -bbmsy_true_mean))
[…]

### 2015-05-19 4.5-add-spectral.R
Add spectral preditor

@@ -1,0 +1,35 @@
+# bring in spectral values based on Coilin's work:
+
+library("dplyr")
+
+train_spec_mat <- function(x, freq_vec = 1/c(2, 5, 10, 20)) {
+  # using AR as smoother, empirical didn't seem to confer much more benefit
+  sp <- spec.ar(x, plot = FALSE)
+  # approximate at fixed frequencies - necessary as series of different length
+  approx(x = sp$freq, y = sp$spec, xout = freq_vec) %>% as.data.frame
+}
+
+# raw data from:
+# ~/Dropbox/FisheriesWorkingGroupPhaseII/Decision_trees/harvest_dynamics_classification/StochasticSimFullFactorial_2013-06-21_no_status.RData
+
+load("raw-data/StochasticSimFullFactorial_2013-06-21_no_status.RData")
+
+spec <- simsStoch %>%
+  arrange(stock_id, year) %>%
+  group_by(stock_id, sigmaC, sigmaR, LH, iter, ED) %>%
+  do(train_spec_mat(.$catch)) %>%
+  rename(spec_freq = x, spec_dens = y) %>%
+  as.data.frame()
+
+library("ggplot2")
+p <- ggplot(spec, aes(spec_freq, log(spec_dens), group = spec_freq)) +
+  geom_boxplot() + facet_wrap(~ED)
+ggsave("figs/spectral-distributions.pdf", width = 8, height = 7)
+
+spec_wide <- spec %>%
+  mutate(spec_freq = paste0("spec_freq_", spec_freq)) %>%
+  reshape2::dcast(stock_id + sigmaC + sigmaR + LH + iter + ED ~ spec_freq,
+    value.var = "spec_dens")
+
+saveRDS(spec_wide, file = "generated-data/spec-frequencies-sim.rds")
+

### 2015-05-19 5-simulation-ensembles.R
Add spectral predictor to ensemble models

@@ -7,6 +7,8 @@ source("4-ensemble-functions.R")
 if (!file.exists("generated-data/cv-sim/")) system("mkdir generated-data/cv-sim")
 
 dsim <- readRDS("raw-data/batch1-results.rds")
+spec <- readRDS("generated-data/spec-frequencies-sim.rds")
+dsim <- suppressWarnings(left_join(dsim, spec)) # warnings on character-factor conversions
 dsim$method_id <- sub("COM.SIR", "COMSIR", dsim$method_id) # to match RAM fits
 dsim <- rename(dsim, stockid = stock_id, method = method_id) # to match RAM fits
 
@@ -35,9 +37,14 @@ dsim_meta <- dsim %>%
   summarise(
     cv_id = cv_id[1],
     LH = LH[1],
+    spec_freq_0.05 = spec_freq_0.05[1],
+    spec_freq_0.1 = spec_freq_0.1[1],
+    spec_freq_0.2 = spec_freq_0.2[1],
+    spec_freq_0.5 = spec_freq_0.5[1],
     max_catch = max(catch),
     total_catch = sum(catch))
 dsim_sum <- inner_join(dsim_sum, dsim_meta)
+dsim_sum$LH <- as.factor(dsim_sum$LH)
 
 # add a binary above/below B/Bmsy = 1 column for a potential response:
 dsim_sum <- dsim_sum %>% mutate(
@@ -49,10 +56,12 @@ trues <- select(dsim_sum, stockid, iter, above_bbmsy1_true,
 trues <- trues[!duplicated(trues), ] # one value per operating model stockid
 
 # switch from long to wide format for modelling:
-d_mean <- reshape2::dcast(dsim_sum, stockid + cv_id + iter + LH ~ method,
+d_mean <- reshape2::dcast(dsim_sum,
+  stockid + cv_id + iter + LH + max_catch + spec_freq_0.05 + spec_freq_0.2 ~ method,
   value.var = "bbmsy_est_mean")  %>%
   inner_join(select(trues, -bbmsy_true_slope))
-d_slope <- reshape2::dcast(dsim_sum, stockid + cv_id + iter + LH ~ method,
+d_slope <- reshape2::dcast(dsim_sum,
+  stockid + cv_id + iter + LH + max_catch + spec_freq_0.05 + spec_freq_0.2 ~ method,
[…]

### 2015-05-19 8-main-figs.R
Add motivating plot with an example stock

@@ -245,4 +245,91 @@ plyr::l_ply(c(1:4, 8, 7, 6, 5), function(m) {
  mtext(expression(B/B[MSY]), side = 1, line = 2.1, outer = TRUE, col = "grey20")
  mtext(expression(widehat(B/B[MSY])), side = 2, line = 1.6, outer = TRUE,
    las = 0, col = "grey20")
+# ----------------------------------------------
+# Example time series plot to motivate the study
+# ----------------------------------------------
+
+# try: SBWHITACIR
+ram <- read.csv("raw-data/RAM_bmsy_Ctousev4.csv", stringsAsFactors=FALSE) %>%
+  filter(stockid == "SBWHITACIR") %>%
+  select(tsyear, Bbmsy_toUse) %>%
+  rename(yr = tsyear, b_bmsy = Bbmsy_toUse)
+
+load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/CMSY/cmsy_rlegacy_output_v0.RData")
+cmsy <- cmsy.rlegacy.df0 %>%
+  filter(stock_id == "SBWHITACIR") %>%
+  select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
+  mutate(method = "CMSY")
+
+load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/COMSIR/comsir_rlegacy_results_table_v0.RData")
+comsir <- comsir.rlegacy.df0 %>%
+  filter(stock_id == "SBWHITACIR") %>%
+  select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
+  mutate(method = "COM-SIR")
+
+load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/SSCOM/sscom_rlegacy_results_table_v0.RData")
+sscom <- sscom.rlegacy.df0 %>%
+  filter(stock_id == "") %>%
+  select(yr, b_bmsy_iq25, b_bmsy, b_bmsy_iq75) %>%
+  rename(b_bmsyiq25 = b_bmsy_iq25, b_bmsyiq75 = b_bmsy_iq75) %>%
+  mutate(method = "SSCOM")
+
[…]

### 2015-05-19 8-main-figs.R
Edits to figs

@@ -3,7 +3,6 @@ library(dplyr)
 library(ggplot2)
 
 d <- readRDS("generated-data/cv_sim_long.rds") %>%
-  filter(type == "mean") %>%
   filter(method != "dummy") %>%
   as.data.frame()
 
@@ -14,27 +13,30 @@ clean_names <- dplyr::data_frame(
     "GBM Ensemble", "RF Ensemble", "LM Ensemble", "Mean Ensemble"),
   order = c(1, 2, 4, 3, 8, 7, 6, 5),
   label_fudge_x = c(
-    -0.01, -0.030, 0, -0.03,
-    0, 0, 0, -0.02),
+    0, -0.06,0,-0.06,
+    0, 0,0,0),
   label_fudge_y = c(
-    0, -0.005, 0.02, 0,
-    0, 0.025, 0.02, 0))
+    0, 0,0.03,0,
+    0, 0,0,0))
 
 d <- suppressWarnings(inner_join(d, clean_names))
 
-d2 <- d %>% group_by(test_iter, clean_method, order, label_fudge_x, label_fudge_y) %>%
+d2 <- d %>% group_by(test_iter, type, clean_method, order, label_fudge_x, label_fudge_y) %>%
   summarise(
     mare = median(abs(re)),
     mre = median(re),
     corr = cor(bbmsy_true_trans, bbmsy_est_trans, method = "spearman",
       use = "pairwise.complete.obs"))
 
+d2$clean_method <- as.factor(d2$clean_method)
+d2$clean_method <- reorder(d2$clean_method, d2$order)
+
+d2_slope <- filter(d2, type == "slope")
+d2 <- filter(d2, type == "mean")
+
 # ------------------------------------
 # Performance distributions simulation
 # ------------------------------------
-
-d2$clean_method <- as.factor(d2$clean_method)
-d2$clean_method <- reorder(d2$clean_method, d2$order)
 
 # ggplot(d2, aes(clean_method, corr)) + geom_violin()
 # ggplot(d2, aes(clean_method, mare)) + geom_violin()
@@ -139,14 +141,15 @@ text(
   m$corr + fudge_y$corr - 0.015,
   m$clean_method, cex = 0.90, pos = 4, col = m$text_col)
 box(col = "grey50")
[…]

### 2015-05-20 5-simulation-ensembles.R
Add code to apply sim ensembles to ram data

@@ -69,6 +69,7 @@ d_slope <- reshape2::dcast(dsim_sum,
 # create problems with some models, such as randomForest, otherwise
 d_mean <- na.omit(d_mean)
 d_slope <- na.omit(d_slope)
+saveRDS(d_mean, file = "generated-data/sim-mean-dat.rds")
 
 nvar <- 8L
 

### 2015-05-20 6.5-sim-to-ram.R

@@ -1,0 +1,84 @@
+# apply the simulation-trained ensemble model to the RAM dataset
+
+library("randomForest")
+library("dplyr")
+
+source("4-ensemble-functions.R")
+
+# prep the RAM data:
+ram <- readRDS("generated-data/ram_fits.rds")
+
+ram_sum <- ram %>%
+  group_by(stockid, method) %>%
+  do(mean_slope_bbmsy(.)) %>%
+  as.data.frame()
+
+ram_meta <- ram %>%
+  group_by(stockid) %>%
+  summarise(
+    habitat = habitat[1],
+    max_catch = max(catch),
+    total_catch = sum(catch))
+ram_sum <- inner_join(ram_sum, ram_meta)
+
+# join in spectral values:
+train_spec_mat <- function(x, freq_vec = 1/c(5, 20)) {
+  # using AR as smoother, empirical didn't seem to confer much more benefit
+  sp <- spec.ar(x, plot = FALSE)
+  # approximate at fixed frequencies - necessary as series of different length
+  approx(x = sp$freq, y = sp$spec, xout = freq_vec) %>% as.data.frame
+}
+spec_ram <- ram %>%
+  arrange(stockid, tsyear) %>%
+  group_by(stockid) %>%
+  do(train_spec_mat(.$catch)) %>%
+  rename(spec_freq = x, spec_dens = y) %>%
+  as.data.frame()
+
+spec_ram_wide <- spec_ram %>%
+  mutate(spec_freq = paste0("spec_freq_", spec_freq)) %>%
+  reshape2::dcast(stockid ~ spec_freq,
+    value.var = "spec_dens")
+
+# save a data frame of 'true' operating model values to merge in:
+trues <- select(ram_sum, stockid, bbmsy_true_mean, bbmsy_true_slope)
+trues <- trues[!duplicated(trues), ] # one value per operating model stockid
+
+# switch from long to wide format for modelling:
+d_mean <- reshape2::dcast(ram_sum, stockid + habitat + max_catch ~ method,
+  value.var = "bbmsy_est_mean")  %>%
+  inner_join(select(trues, stockid, bbmsy_true_mean)) %>%
+  inner_join(spec_ram_wide)
[…]

### 2015-05-20 5-simulation-ensembles.R
Tweak simulation ensembles script - nothing major

@@ -74,6 +74,10 @@ saveRDS(d_mean, file = "generated-data/sim-mean-dat.rds")
 nvar <- 8L
 
 # run a model on all the data to generate data for partial dependence plots:
+# m_rf <- randomForest::randomForest(
+#   log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM + LH +
+#   max_catch + spec_freq_0.05 + spec_freq_0.2, data = d_mean)
+
 m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM + LH +
   max_catch + spec_freq_0.05 + spec_freq_0.2,
   data = d_mean, n.trees = 10000L, interaction.depth = 2, shrinkage = 0.001)
@@ -87,7 +91,7 @@ partial <- plyr::ldply(seq_len(nvar), function(i) {
 
 # partial dependence plot:
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() +
-  facet_wrap(~predictor, scales = "free_x")
+  facet_wrap(~predictor, scales = "free_x") + ylim(0.3, 1.6)
 ggsave("figs/partial-sim.pdf", width = 7, height = 5)
 
 partial_2d <- plyr::ldply(1:nvar, function(x) plyr::ldply(1:nvar, function(y) {

### 2015-05-20 8-main-figs.R
Plot the sim ensembles applied to ram stocks

@@ -149,7 +149,6 @@ mtext("Across population correlation", side = 2, line = 2.2, col = "grey40", las
 #legend("bottomright", bty = "n", pch = c(21, 21, 21), bg = c("red", "red", "blue"),
   #legend = c(4, 0, -1))
 dev.off()
-1
 
 #mtext(LETTERS[ii], adj = 0.05, line = -1.5, col = "grey40", cex = 0.8)
 #text(0.5, 0.48, panel_labs[ii-4], col = label_col, pos = 4, cex = 1.05)
@@ -201,22 +200,32 @@ add_label <- function(xfrac, yfrac, label, pos = 4, ...) {
 
 plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   ylab = expression(widehat(B/B[MSY])), lims_hex = c(0, max(dat$bbmsy_est)),
-  xlim_plot = c(0, 2.5), ylim_plot = c(0, 2.7), axis_ticks = c(0, 1, 2)) {
+  xlim_plot = c(0, 2.5), ylim_plot = c(0, 2.7), axis_ticks = c(0, 1, 2),
+  add_hex = TRUE, alpha = 50) {
   par(mfrow = c(2, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
     oma = c(3.5, 3.5, .5, .5), cex = 0.8, mar = c(0, 0, 0, 0),
     xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
-  plyr::l_ply(c(1:4, 8, 7, 6, 5), function(m) {
-    xlim <- lims_hex
-    ylim <- lims_hex
-    bin <- hexbin::hexbin(
-      filter(dat, method == unique(d$method)[m])$bbmsy_true,
-      filter(dat, method == unique(d$method)[m])$bbmsy_est,
-      xbnds = xlim, ybnds = ylim, xbins = xbins)
-    dx <- hexbin::hcell2xy(bin)$x
-    dy <- hexbin::hcell2xy(bin)$y
-    dxy <- data.frame(x = dx, y = dy)
-    counts <- bin@count
-    counts <- round(log(counts*1.5))
+
+  #if (max(dat$order) == 8)
+    #panels <- c(1:4, 8, 7, 6, 5)
+  #else
+    panels <- seq_len(length(unique(dat$order)))
+
+  plyr::l_ply(panels, function(m) {
+
[…]

### 2015-05-20 analysis/2-run-catch-models.R
Clean up folder structure

@@ -28,7 +28,7 @@ for (i in seq_along(sig_r_vector)) {
   p <- dplyr::left_join(ram_cmsy, ramts[,c("year", "stockid", "stocklong", "bbmsy_ram"), ])
   saveRDS(p, file = paste0("generated-data/ram-cmsy-sigr-", sig_r_vector[i], ".rds"))
 
-  pdf(paste0("figs/cmsy-ram-sigr-", sig_r_vector[i], ".pdf"), width = 5, height = 4)
+  pdf(paste0("../figs/cmsy-ram-sigr-", sig_r_vector[i], ".pdf"), width = 5, height = 4)
   plyr::d_ply(p, "stocklong", function(x) {
     message(x$stocklong[1])
     q <- ggplot(x, aes(year, bbmsy_q50)) + geom_line()  +

### 2015-05-20 analysis/4.5-add-spectral.R

@@ -24,7 +24,7 @@ spec <- simsStoch %>%
 library("ggplot2")
 p <- ggplot(spec, aes(spec_freq, log(spec_dens), group = spec_freq)) +
   geom_boxplot() + facet_wrap(~ED)
-ggsave("figs/spectral-distributions.pdf", width = 8, height = 7)
+ggsave("../figs/spectral-distributions.pdf", width = 8, height = 7)
 
 spec_wide <- spec %>%
   mutate(spec_freq = paste0("spec_freq_", spec_freq)) %>%

### 2015-05-20 analysis/5-simulation-ensembles.R

@@ -92,7 +92,7 @@ partial <- plyr::ldply(seq_len(nvar), function(i) {
 # partial dependence plot:
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() +
   facet_wrap(~predictor, scales = "free_x") + ylim(0.3, 1.6)
-ggsave("figs/partial-sim.pdf", width = 7, height = 5)
+ggsave("../figs/partial-sim.pdf", width = 7, height = 5)
 
 partial_2d <- plyr::ldply(1:nvar, function(x) plyr::ldply(1:nvar, function(y) {
   dd <- gbm::plot.gbm(m, i.var = c(x, y), return.grid = TRUE, continuous.resolution = 20)
@@ -110,7 +110,7 @@ zlim <- c(min(partial_2d$z), max(partial_2d$z))
 pal <- rev(colorRampPalette(c("blue", "white", "red"))(17)[-c(13:17)])
 # white should line up with 1: (adjust the above line as needed)
 plot(seq(min(zlim), max(zlim), length.out = 12), 1:12, col = pal)
-pdf("figs/partial-sim-2d.pdf", width = 10, height = 10)
+pdf("../figs/partial-sim-2d.pdf", width = 10, height = 10)
 par(mfrow = c(nvar, nvar), mar = c(3,3,1,1), oma = c(4, 4, 1, 1), cex = 0.5)
 par(xpd = NA, mgp = c(1.5, 0.5, 0))
 plyr::d_ply(partial_2d, c("var1", "var2"), function(x) {
@@ -181,12 +181,12 @@ cv_sim_slope_long <- cv_sim_slope %>%
 p <- ggplot(cv_sim_mean_long, aes(bbmsy_true, bbmsy_est)) +
   geom_point(alpha = 0.01) +
     facet_wrap(~method) + ylim(0, 3) + xlim(0, 3)
-ggsave("figs/cv-sim-mean-scatter.png", width = 8, height = 8)
+ggsave("../figs/cv-sim-mean-scatter.png", width = 8, height = 8)
 
 p <- ggplot(cv_sim_slope_long, aes(bbmsy_true, bbmsy_est)) +
   geom_point(alpha = 0.01) +
     facet_wrap(~method) + xlim(-.5, .5) + ylim(-.5, .5)
-ggsave("figs/cv-sim-slope-scatter.png", width = 8, height = 8)
+ggsave("../figs/cv-sim-slope-scatter.png", width = 8, height = 8)
 
[…]

### 2015-05-20 analysis/6-ram-ensembles.R

@@ -54,7 +54,7 @@ partial <- plyr::ldply(seq_len(5), function(i) {
 })
 
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() + facet_wrap(~predictor) + xlim(0, 3)
-ggsave("figs/partial-ram.pdf", width = 7, height = 5)
+ggsave("../figs/partial-ram.pdf", width = 7, height = 5)
 
 # work through cross validation of ensemble models:
 library("doParallel")
@@ -105,12 +105,12 @@ cv_ram_slope_long <- reshape2::melt(select(cv_ram_slope, -above_bbmsy1_true),
 p <- ggplot(cv_ram_mean_long, aes(bbmsy_true, bbmsy_est)) +
   geom_point(alpha = 0.05) +
     facet_wrap(~method) + ylim(0, 3) + xlim(0, 3)
-ggsave("figs/cv-ram-mean-scatter.png", width = 8, height = 8)
+ggsave("../figs/cv-ram-mean-scatter.png", width = 8, height = 8)
 
 p <- ggplot(cv_ram_slope_long, aes(bbmsy_true, bbmsy_est)) +
   geom_point(alpha = 0.05) +
     facet_wrap(~method) + xlim(-.5, .5) + ylim(-.5, .5)
-ggsave("figs/cv-ram-slope-scatter.png", width = 8, height = 8)
+ggsave("../figs/cv-ram-slope-scatter.png", width = 8, height = 8)
 
 cv_ram_long <- suppressWarnings(
   dplyr::bind_rows(cv_ram_mean_long, cv_ram_slope_long))
@@ -149,4 +149,4 @@ p <- performance %>% reshape2::dcast(method + type ~ summary, value.var = c("per
   geom_text(aes(label = method), hjust = 0.2, size = 3) +
   facet_wrap(~type, scales = "free") +
   xlab("MARE within stocks") + ylab("Spearman correlation across stocks")
-ggsave("figs/performance-ram-scatter.pdf", width = 8, height = 4)
+ggsave("../figs/performance-ram-scatter.pdf", width = 8, height = 4)

### 2015-05-20 analysis/7-roc.R

@@ -44,7 +44,7 @@ p <- ggplot(rocs_sim, aes(spec, sens, colour = method)) + geom_line() + coord_eq
   theme(plot.background = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank())
-ggsave("figs/roc-sim.pdf", width = 7, height = 5)
+ggsave("../figs/roc-sim.pdf", width = 7, height = 5)
 
 p <- ggplot(rocs_ram, aes(spec, sens, colour = method)) + geom_line() + coord_equal() +
   geom_abline(intercept = 1, slope = 1, lty = 2, col = "darkgrey") +
@@ -53,4 +53,4 @@ p <- ggplot(rocs_ram, aes(spec, sens, colour = method)) + geom_line() + coord_eq
   theme(plot.background = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank())
-ggsave("figs/roc-ram.pdf", width = 7, height = 5)
+ggsave("../figs/roc-ram.pdf", width = 7, height = 5)

### 2015-05-20 analysis/8-main-figs.R

@@ -86,7 +86,7 @@ performance_panel <- function(dat, column_id, ylim_adj = c(-0.1, 0.1), xaxis = F
 
 }
 
-pdf("figs/performance-beanplots-sim.pdf", width = 5, height = 5)
+pdf("../figs/performance-beanplots-sim.pdf", width = 5, height = 5)
 par(mfrow = c(3, 1))
 par(mar = c(0, 3, 0, 0), cex = 0.8, oma = c(8, 3, 1.5, .5), tck = -0.03, mgp = c(2, 0.5, 0))
 performance_panel(d2, "corr", c(-0.1, 0.1), flip_yaxis = FALSE, yticks = c(0, 0.2, 0.4))
@@ -126,7 +126,7 @@ m$col <- pal$col[findInterval(m$mre, pal$mre)]
 xlim <- filter(d3, variable == "mare") %>% select(l, u) %>% range
 ylim <- filter(d3, variable == "corr") %>% select(l, u) %>% range
 
-pdf("figs/fig3.pdf", width = 4, height = 3.1)
+pdf("../figs/fig3.pdf", width = 4, height = 3.1)
 par(mfrow = c(1, 1), mgp = c(1.5, 0.4, 0), las = 1, tck = -0.012,
   oma = c(3, 3.5, .5, .5), cex = 0.8, mar = c(0, 0, 0, 0),
   xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
@@ -275,7 +275,7 @@ d_mean_plot <- filter(d, type == "mean")
 d_mean_plot$bbmsy_est[d_mean_plot$bbmsy_est > 10] <- NA
 d_mean_plot <- na.omit(d_mean_plot)
 
-pdf("figs/fig2.pdf", width = 8, height = 4)
+pdf("../figs/fig2.pdf", width = 8, height = 4)
 plot_hex_fig(d_mean_plot, xbins = 100L)
 dev.off()
 
@@ -283,7 +283,7 @@ d_slope_plot <- filter(d, type == "slope")
 d_slope_plot$bbmsy_est[d_slope_plot$bbmsy_est > 10] <- NA
 d_slope_plot <- na.omit(d_slope_plot)
 
-pdf("figs/hex-slope-sim.pdf", width = 8, height = 4)
+pdf("../figs/hex-slope-sim.pdf", width = 8, height = 4)
[…]

### 2015-05-20 analysis/6-ram-ensembles.R
Remove old RAM ensemble file; rename other

@@ -1,152 +1,0 @@
-# The main file to run ensemble models for the RAM data.
-# Right now this is largely copied from the simulation data version.
-# If we decide to keep this and do anything remotedly fancy then
-# I'll generalize the code into functions and call it for both
-# the simulation and RAM data.
-
-library("dplyr")
-library("ggplot2")
-source("4-ensemble-functions.R")
-
-if (!file.exists("generated-data/cv-ram/")) system("mkdir generated-data/cv-ram")
-
-ram <- readRDS("generated-data/ram_fits.rds")
-
-ram_sum <- ram %>%
-  group_by(stockid, method) %>%
-  do(mean_slope_bbmsy(.)) %>%
-  as.data.frame()
-
-ram_meta <- ram %>%
-  group_by(stockid) %>%
-  summarise(
-    habitat = habitat[1],
-    max_catch = max(catch),
-    total_catch = sum(catch))
-ram_sum <- inner_join(ram_sum, ram_meta)
-
-# add a binary above/below B/Bmsy = 1 column for a potential response:
-ram_sum <- ram_sum %>% mutate(
-  above_bbmsy1_true = ifelse(bbmsy_true_mean > 1, 1, 0))
-
-# save a data frame of 'true' operating model values to merge in:
-trues <- select(ram_sum, stockid, above_bbmsy1_true,
-  bbmsy_true_mean, bbmsy_true_slope)
-trues <- trues[!duplicated(trues), ] # one value per operating model stockid
-
-# switch from long to wide format for modelling:
-d_mean <- reshape2::dcast(ram_sum, stockid + habitat ~ method,
-  value.var = "bbmsy_est_mean")  %>%
-  inner_join(select(trues, -bbmsy_true_slope))
-d_slope <- reshape2::dcast(ram_sum, stockid + habitat ~ method,
-  value.var = "bbmsy_est_slope") %>%
-  inner_join(select(trues, -bbmsy_true_mean))
-
-# run a model on all the data to generate data for partial dependence plots:
-m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM + habitat,
-  data = d_mean, n.trees = 3000L, interaction.depth = 3, shrinkage = 0.001)
[…]

### 2015-05-27 analysis/8-main-figs.R
Delete old commented-out code

@@ -34,22 +34,6 @@ d2$clean_method <- reorder(d2$clean_method, d2$order)
 d2_slope <- filter(d2, type == "slope")
 d2 <- filter(d2, type == "mean")
 
-# ------------------------------------
-# Performance distributions simulation
-# ------------------------------------
-
-# ggplot(d2, aes(clean_method, corr)) + geom_violin()
-# ggplot(d2, aes(clean_method, mare)) + geom_violin()
-# ggplot(d2, aes(clean_method, mre)) + geom_violin()
-# d %>% filter(re < 4, re > -4) %>%
-#   ggplot(aes(re)) + geom_histogram() + facet_wrap(~clean_method) +
-#   geom_vline(xintercept = 0)
-
-#bg_plot <- function(colour = "#00000009") {
-  #rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4],
-    #col = colour, border = FALSE)
-#}
-
 performance_panel <- function(dat, column_id, ylim_adj = c(-0.1, 0.1), xaxis = FALSE,
   flip_yaxis = FALSE, yticks = NULL, add_zero_line = TRUE) {
 

### 2015-05-27 analysis/8-main-figs.R
Delete beanplots

@@ -34,55 +34,6 @@ d2$clean_method <- reorder(d2$clean_method, d2$order)
 d2_slope <- filter(d2, type == "slope")
 d2 <- filter(d2, type == "mean")
 
-performance_panel <- function(dat, column_id, ylim_adj = c(-0.1, 0.1), xaxis = FALSE,
-  flip_yaxis = FALSE, yticks = NULL, add_zero_line = TRUE) {
-
-  axis_col <- "grey50"
-  label_col <- "grey50"
-  cols_ii <- RColorBrewer::brewer.pal(8, "Blues")
-
-  ylim <- range(dat[,column_id][[1]]) + ylim_adj
-  if (flip_yaxis) ylim <- rev(ylim)
-
-  plot(1, 1, xlim = c(.8, nrow(clean_names)+0.2),
-    ylim = ylim, type = "n", axes = FALSE, ann = FALSE, yaxs = "i")
-  rect(4.5, ylim[1], 9.5, ylim[2], col = "#00000010", border = FALSE)
-  rect(5.5, ylim[1], 9.5, ylim[2], col = "#00000010", border = FALSE)
-  rect(6.5, ylim[1], 9.5, ylim[2], col = "#00000010", border = FALSE)
-  rect(7.5, ylim[1], 9.5, ylim[2], col = "#00000010", border = FALSE)
-  if(add_zero_line)
-    abline(h = 0, col = "grey70", lty = 2, lwd = 1)
-  beanplot::beanplot(as.formula(paste(column_id, "~ clean_method")),
-    data = dat, add = TRUE,
-    border = NA, axes = FALSE, col = cols_ii[c(5, 3, 4, 8)], what =
-    c(0, 1, 1, 0), log = FALSE)
-  points(jitter(as.numeric(dat$clean_method), amount = 0.09),
-    dat[,column_id][[1]], col = "#D0D0D0", pch = 20, cex = 0.37)
-  if (is.null(yticks))
-    axis(2, las = 1, col = axis_col, col.axis = axis_col)
-  else
-    axis(2, las = 1, at = yticks, col = axis_col, col.axis = axis_col)
-
-  if (xaxis)
-    axis(1, col = axis_col, col.axis = label_col, at = 1:length(unique(dat$clean_method)),
-      labels = levels(dat$clean_method), las = 3)
-  box(col = axis_col)
-
-}
-
[…]

### 2015-05-27 analysis/test-mprm.R
Add script to test mPRM from package vs. original fits

@@ -1,0 +1,36 @@
+# compare the datalimited::prm function to the original fits
+library("datalimited")
+library("dplyr")
+library("ggplot2")
+
+d <- dplyr::inner_join(datalimited::ram_ts, datalimited::spp_categories,
+  by = "scientificname")
+ram_prm_dat <- plyr::ddply(d, "stockid", function(x) {
+  datalimited::format_prm(year = x$year, catch = x$catch,
+    bbmsy = x$bbmsy_ram,
+    species_cat = x$spp_category[1L])
+})
+
+m <- datalimited::fit_prm(ram_prm_dat)
+d <- ram_prm_dat
+d <- cbind(d, datalimited::predict_prm(d, ci = TRUE))
+d <- dplyr::rename(d, package = fit)
+
+ram <- readRDS("generated-data/ram_fits.rds")
+qq <- subset(ram, method == "Costello")
+q <- dplyr::inner_join(select(qq, stockid, tsyear, b_bmsy_est),
+  select(d, stockid, year, package) %>% rename(tsyear = year))
+qlong <- reshape2::melt(q, id.vars = c("stockid", "tsyear"))
+
+ggplot(dplyr::filter(qlong, stockid %in% unique(qlong$stockid)[1:25]),
+  aes(tsyear, value)) + facet_wrap(~stockid) +
+  geom_line(aes(colour = variable))
+
+# CIs:
+ggplot(dplyr::filter(d, stockid %in% unique(d$stockid)[1:10]),
+  aes(year, package)) +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#FF000040") +
+  geom_line() +
+  ylab(expression(B/B[MSY])) +
+  facet_wrap(~stockid)
+

### 2015-05-28 analysis/5-simulation-ensembles.R
Remove max catch as predictor in simulation partial dependencies

@@ -72,7 +72,7 @@ saveRDS(d_mean, file = "generated-data/sim-mean-dat.rds")
 
 # Number of variables in the ensemble models
 # gets used to, for example, figure out how many panels to plot
-nvar <- 7L
+nvar <- 6L
 
 # run a model on all the data to generate data for partial dependence plots:
 # m_rf <- randomForest::randomForest(
@@ -103,7 +103,7 @@ plot(mm_gbm, metric = "Rsquared")
 dev.off()
 
 m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM +
-  max_catch + spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
+  spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
   data = d_mean, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
 
 partial <- plyr::ldply(seq_len(nvar), function(i) {
@@ -132,9 +132,9 @@ partial_2d <- plyr::ldply(1:nvar, function(x) plyr::ldply(1:nvar, function(y) {
 
 # check colour pallete:
 zlim <- c(min(partial_2d$z), max(partial_2d$z))
-pal <- rev(colorRampPalette(c("red", "white", "blue"))(15)[-c(13:15)])
+pal <- rev(colorRampPalette(c("red", "white", "blue"))(13)[-1]) #[-c(13:13)])
 # white should line up with 1: (adjust the above line as needed)
-#plot(seq(min(zlim), max(zlim), length.out = 12), 1:12, col = pal)
+plot(seq(min(zlim), max(zlim), length.out = 12), 1:12, col = pal);abline(v = 1)
 pdf("../figs/partial-sim-2d.pdf", width = 10, height = 10)
 par(mfrow = c(nvar, nvar), mar = c(3,3,1,1), oma = c(4, 4, 1, 1), cex = 0.5)
 par(xpd = NA, mgp = c(1.5, 0.5, 0))

### 2015-05-28 analysis/5-simulation-ensembles.R
Add proper above/below ensemble models with simulation data

@@ -172,17 +172,23 @@ cv_sim_slope <- plyr::ldply(seq_len(4), .parallel = TRUE,
 cv_sim_slope$cv_id <- NULL
 cv_sim_slope$dummy <- rnorm(nrow(cv_sim_slope), 0, 0.2)
 
-# TODO: update formula here or cut out:
-# cv_sim_binary <- plyr::ldply(seq_len(4), .parallel = TRUE,
-#   .fun = function(.n)
-#     cross_val_ensembles(.n = .n, dat = d_mean, geo_mean = TRUE,
-#       id = "sim-mean", distribution = "bernoulli",
-#       gbm_formula = "above_bbmsy1_true ~ CMSY + COMSIR + Costello + SSCOM + LH",
-#       glm_formula = "above_bbmsy1_true ~ (CMSY + COMSIR + Costello + SSCOM + LH)^2"))
-# cv_sim_binary$cv_id <- NULL
-# saveRDS(cv_sim_binary, file = "generated-data/cv_sim_binary.rds") # used in 5-roc.R
+cv_sim_binary <- plyr::ldply(seq_len(4), .parallel = TRUE,
+  .fun = function(.n)
+    cross_val_ensembles(.n = .n, dat = d_mean,
+      id = "sim-binary", distribution = "bernoulli",
+      gbm_formula = paste0("above_bbmsy1_true ~ ", eq),
+      glm_formula = paste0("above_bbmsy1_true ~ (", eq, ")^2")))
+cv_sim_binary$cv_id <- NULL
 
 # now switch to long format data, summarize, and compare:
+cv_sim_binary_long <- cv_sim_binary %>%
+  select(-max_catch, -spec_freq_0.05, -spec_freq_0.2, -bbmsy_true_mean) %>%
+  reshape2::melt(id.vars = c("stockid", "iter", "test_iter", "LH", "above_bbmsy1_true"),
+  variable.name = "method", value.name = "bbmsy_est") %>%
+  rename(bbmsy_true = above_bbmsy1_true) %>%
+  mutate(type = "binary")
+saveRDS(cv_sim_binary_long, file = "generated-data/cv_sim_binary.rds") # used in 5-roc.R
+
 cv_sim_mean_long <- cv_sim_mean %>%
   select(-max_catch, -spec_freq_0.05, -spec_freq_0.2, -above_bbmsy1_true) %>%
   reshape2::melt(id.vars = c("stockid", "iter", "test_iter", "LH", "bbmsy_true_mean"),

### 2015-05-28 analysis/6-sim-to-ram.R
Increase RAM CV iterations and reduce cores
Was failing sometimes on 4

@@ -175,8 +175,9 @@ cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
 }
 
 library("doParallel")
-registerDoParallel(cores = 4L)
-qq <- plyr::ldply(seq_len(16L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
+registerDoParallel(cores = 2L)
+set.seed(123)
+qq <- plyr::ldply(seq_len(40L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
   .parallel = TRUE)
 
 d_mean_long <- qq %>%

### 2015-05-28 analysis/8-main-figs.R
So many figure tweaks
This file needs some serious cleaning

@@ -13,11 +13,11 @@ clean_names <- dplyr::data_frame(
     "GBM Ensemble", "RF Ensemble", "LM Ensemble", "Mean Ensemble"),
   order = c(1, 2, 4, 3, 8, 7, 6, 5),
   label_fudge_x = c(
-    0, -0.06,0,-0.06,
-    0, 0,0,0),
+    0, -0.03,0,-0.04,
+    0, 0,0,-0.05),
   label_fudge_y = c(
-    0, 0,0.03,0,
-    0, 0,0,0))
+    0, -0.02,0.03,0.04,
+    0, 0.02,0,-0.02))
 
 d <- suppressWarnings(inner_join(d, clean_names))
 
@@ -58,31 +58,42 @@ pal <- data_frame(
   mre = seq(-max(m$mre)*1.05, max(m$mre)*1.05, length.out = 11))
 m$col <- pal$col[findInterval(m$mre, pal$mre)]
 
-xlim <- filter(d3, variable == "mare") %>% select(l, u) %>% range
-ylim <- filter(d3, variable == "corr") %>% select(l, u) %>% range
+xlim <- filter(d3, variable == "mare") %>% select(l, u) %>% range +
+  c(-0.01, 0.01)
+ylim <- filter(d3, variable == "corr") %>% select(l, u) %>% range +
+  c(-0.02, 0.01)
 
 pdf("../figs/fig3.pdf", width = 4, height = 3.1)
 par(mfrow = c(1, 1), mgp = c(1.5, 0.4, 0), las = 1, tck = -0.012,
-  oma = c(3, 3.5, .5, .5), cex = 0.8, mar = c(0, 0, 0, 0),
+  oma = c(2.7, 3.5, .5, .5), cex = 0.8, mar = c(0, 0, 0, 0),
   xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
 par(xpd = NA)
 plot(m$mare, m$corr, xlim = xlim, ylim = ylim, pch = 21, bg = m$col, col = "grey50",
   axes = FALSE, xlab = "", ylab = "", type = "n")
 segments(m$mare, l$corr, m$mare, u$corr, col = "#00000050", lwd = 1.4)
 segments(l$mare, m$corr, u$mare, m$corr, col = "#00000050", lwd = 1.4)
-points(m$mare, m$corr, pch = 21, bg = m$col, col = "grey50", cex = 1.5)
-text(
[…]

### 2015-06-04 analysis/8-main-figs.R
Add performance metrics plot for the slope model

@@ -274,3 +274,47 @@ par(xpd = NA)
 mtext("Rank-order correlation", cex = 1, side = 2, outer = TRUE, line = -0.9,
   col = "grey20", las = 0)
 dev.off()
+
+# error metrics for slope:
+d_slope_error <- d_slope_plot %>%
+  mutate(
+    re = bbmsy_est - bbmsy_true,
+    are = abs(re)) %>%
+  group_by(clean_method) %>%
+  summarise(
+    mre = median(re),
+    mare = median(are),
+    corr = cor(bbmsy_true, bbmsy_est, method = "spearman",
+      use = "pairwise.complete.obs")) %>%
+  as.data.frame
+
+pdf("../figs/performance-slope-sim.pdf", width = 5, height = 4)
+par(mgp = c(1.5, 0.5, 0), las = 1, tck = -0.015,
+  oma = c(2.8, 0.5, 1.5, .5), cex = 0.8, mar = c(0, 3, 0, 0),
+  xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
+par(family="serif")
+
+pal_df <- data_frame(mre = seq(-max(abs(d_slope_error$mre)+0.02),
+  max(abs(d_slope_error$mre)+0.02),
+  length.out = 100), col = pal(100))
+d_slope_error$col <- pal_df$col[findInterval(d_slope_error$mre, pal_df$mre)]
+perf(d_slope_error, label = "", xlim = c(0.02, 0.065), ylim = c(-0.05, 0.65))
+axis(1, at = seq(0, 0.1, 0.02), col = "grey50")
+# colour legend:
+blocks <- seq(0, 0.2, length.out = 100)
+leg_x <- 0.025
+for(i in 1:99) {
+  rect(leg_x, blocks[i], leg_x + 0.002, blocks[i+1]+0.001, border = NA, col = pal(100)[i])
+}
+tick1 <- findInterval(-0.02, pal_df$mre)
+tick3 <- findInterval(0.02, pal_df$mre)
+text(rep(leg_x + 0.002, 3),
+  blocks[c(tick1, 50, tick3)], labels = c("-0.02", "  0", "  0.02"), pos = 4,
+  col = "grey50", cex = 0.9)
+text(leg_x - 0.001, 0.22, "Bias (MRE)", col = "grey20", pos = 4)
[…]

### 2015-06-04 analysis/5-simulation-ensembles.R
Tweak the partial dependence plot

@@ -179,7 +179,9 @@ partial <- plyr::ldply(seq_len(nvar), function(i) {
 
 # partial dependence plot:
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() +
-  facet_wrap(~predictor, scales = "free_x") + ylim(0.3, 1.6)
+  facet_wrap(~predictor, scales = "free_x") + ylim(0.3, 1.6) + theme_bw() +
+  ylab("Expected B/B_MSY integrating out other predictors") +
+  xlab("Predictor value")
 ggsave("../figs/partial-sim.pdf", width = 7, height = 5)
 
 partial_2d <- plyr::ldply(1:nvar, function(x) plyr::ldply(1:nvar, function(y) {

### 2015-06-05 text/ms.Rmd
Switch main Markdown document to R Markdown
To import magic numbers

```{r, echo=FALSE}
# Values to use in the paper:
load("values.rda") # generated by `../analysis/9-output-values.R`

# Includes: mean_sim, mean_ram, slope_sim, auc_sim

# mean_sim, mean_ram, and slope_sim have the following list elements:
# > names(mean_sim)
# "mach_vs_ind_mare_fold" "ind_corr_range" "mach_corr_range" "ind_mare_range"  
# "mach_mare_range" "ensemble_mre_range" "ind_mre_range"

# auc_sim has the following list elements:
# > names(auc_sim)
# "mach_range" "ind_range"

# Many elements contain a numeric vector of length 2 indicating lower and 
# upper values.

# Insert one, with, for example: `r auc_sim$mach_range[1]`.
```

\noindent
Sean C. Anderson^1\*^, 
Jamie Afflerbach^2^, 
Mark Dickey-Collas^3^, 
Olaf P. Jensen^4^, 
Kristin M. Kleisner^5^, 
Catherine Longo^2^, 
C\'{o}il\'{i}n Minto^6^, 
Giacomo Chato Osio^7^, 
Dan Ovando^8^, 
Andrew A. Rosenberg^9^, 
Elizabeth R. Selig^10^, 
James T. Thorson^11^,
Jessica C. Walsh^1^,
Andrew B. Cooper^1^
*(order to be determined and others may be added)*

\bigskip

\noindent
^1^School of Resource and Environmental Management,
Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\noindent
^2^National Center for Ecological Analysis and Synthesis

\noindent
^3^International Council for the Exploration of the Sea

\noindent
^5^NOAA/NMFS/NEFSC, Ecosystem Assessment Program, 
Woods Hole, MA, 02543, USA

\noindent
^6^Galway-Mayo Institute of Technology, 
Marine and Freshwater Research Centre, Galway, 00000, Ireland

\noindent
^7^EC JRC, IPSC, MAU, Italy

\noindent
^8^University of California Santa Barbara, 
Bren School of Environmental Science and Management, 
Santa Barbara, CA, 93106-5131, USA

\noindent
^9^Union of Concerned Scientists, Cambridge, MA, USA

\noindent
^10^Conservation International, Arlington, VA, USA

\noindent
^11^National Marine Fisheries Service, 
National Oceanic and Atmospheric Administration, 
Fisheries Resource Assessment and Monitoring Division, 
Northwest Fisheries Science Center, Seattle, WA, 98112, USA

\noindent
^\*^Corresponding author: Sean C. Anderson,
School of Resource and Environmental Management,
Simon Fraser University,
Burnaby BC, V5A 1S6;
E-mail: sean_anderson@sfu.ca

\clearpage

# Abstract

What can we do when multiple ecological models give different impressions of
population status? Any single model is unlikely to perform best across all
species, exploitation dynamics, and statuses. Ensemble methods from the fields
of machine learning and climate science provide a powerful tool to optimally
leverage the information contained in multiple ecological models. Here we
evaluate the potential for ensemble methods to improve estimates of fish
population status. We fit four data-limited models of population and
exploitation dynamics and combine their estimates of the mean and slope of
\bbmsy\\ (biomass at maximum sustainable yield) with four ensemble models:
a simple average, a linear model, and two machine learning methods: random
forests and boosted regression trees. We build our ensemble models on a fully
factorial simulated dataset of 5760 fish stocks and test our models on
a cross-validated version of this dataset and on a global database of 166
assessed fish stocks. We find that ensemble models substantially improve
estimates of population status, trajectory, and classification. Machine
learning ensemble models performed the best (XX--XX% improved accuracy, XX--XX%
improved correlation, XX-XX% reduced bias compared to the individual models).
When the same ensembles were applied to a separate dataset of global fish
stocks and compared to status from full stock assessments, ensemble methods
still improved estimates of status and trajectory over individual models
(accuracy, correlation, and bias improved up to XX%, XX%, and XX%). Ensemble
models have great potential to improve estimates of population status and
trajectory across many taxa beyond fish. However, ensemble methods must be
carefully tested, formed from a diverse set of accurate models, and built on
a dataset sufficiently representative of the populations they are applied to.

# Introduction

We often have multiple models of ecological systems. For example, models might
incorporate different data types, assume alternate population dynamics, or make
contrasting assumptions about the starting state of a system. These models may
suggest different conclusions about population status and often times we do not
know which model is best. How can we reconcile multiple models to make robust
management decisions about ecological resources?

When models give conflicting estimates of a population's status, there are a
number of solutions. (1) We can pick a single model. However, it isn't always
obvious which model is best, and even objectively second-best models may contain
useful additional information. (2) We can maintain multiple models throughout
the decision-making process. However, this can be complicated (REFs)... (3) We
can combine the model outputs. This could be as simple as taking the average,
potentially weighting the models by some performance metric. This could be as
complicated as forming an additional 'hyper' model that draws on interactions
between model outputs and potentially incorporates additional information to
derive a best estimate.

Ensemble models are widely used in the climate sciences, where thousands of
models with varying structural combined across different structural ... The
idea behind ensemble models forms the backbone of many machine learning
methods. For example...

In fisheries science, a common problem is estimating the status of an exploited
fish population. For the majority of fish stocks, we have limited information to
go on and stock status isn't known. In recent years, a number of methods have
been proposed to derive population status based on limited information and a set
of assumptions. However, we know that these models can give conflicting output
and no one model is that accurate.

Motivation: (Fig. \ref{fig:motivate}).

Ensemble models combine estimates of individual models, potentially along with
additional covariates, in an additional 'ensemble' model
(Fig. \ref{fig:didactic}).

Exploit interactions between individual models and relationships between other
conditions of the data and the performance of the individual models.

Fitted to populations of known or assumed status and this ensemble model is
used to extrapolate to populations of interest.

Here, we develop ensemble models for data-limited exploited fish populations. We
explore a variety of ensemble approaches applied to both simulated and
real-world fish stocks and compare performance...

# Methods

We tested the ability of ensemble models to improve estimates of population
status when applied to both a large simulated dataset and a global database of
assessed stock status. Here we describe the datasets, individual models of
population status, and ensemble models to combine those estimates. We then
describe how we evaluated the ability of the various models to estimate
population status.

## Datasets

<!--TODO: github flr repo not available-->

We developed and tested ensemble models with both a fully factorial simulated
dataset [@rosenberg2014] and the RAM Legacy Stock Assessment Database
[@ricard2012]. @rosenberg2014 provide a full description of the simulation
model and the code to generate the simulations is available at
<https://github.com/flr/StockSims>. To summarize, the model was built in the
package FLR [@kell2007] for the statistical software \textsf{R} [@r2015]...
included all combinations of the following factors:

- Three life histories: small pelagic, demersal, large pelagic.
- Two time series lengths: 20 and 60 years.
- Three levels of initial depletion: 1, 0.7, and 0.4.
- Four harvest dynamics: (1) a constant harvest rate, (2) a harvest rate that
  is coupled with biomass to mimic an open-access single-species fishery, (3) a
  one-way trip where harvest rate increases five percent per year to 80% of the
  level at which the stock would crash, and (4) roller-coaster shaped where the
  harvest rate increases, maintains a steady level at 80% of the level at which
  the stock would crash, and then decreases to FMSY. 
- Two levels of multiplicative standard deviation around catch (0 and 0.2)
- Two levels of standard deviation of recruitment variability (0.2 and 0.6)
- Two levels of autoregressive correlation on recruitment variability (0 and 0.6)

\noindent @rosenberg2014 ran ten iterations for each combination of factors
adding stochastic draws of recruitment and catch-recording variability each
time to generate a total of 5760 stocks.

Our analysis of the RAM Legacy Stock Assessment Database was based on version
XX downloaded on XX. This version includes XX stocks from XX countries across
XX taxonomic orders.

## Individual models of population status

We fit four individual data-limited models to estimate \bbmsy\\. Three of the
models are mechanistic and based generally on Shaefer biomass dynamics
[@schaefer1954] of the form:

\begin{equation}
\hat{B}_{t+1} = B_t + r B_t \left(1 - B_t / B_0 \right) - C_t,
\end{equation}

\noindent where $B_t$ represents biomass at time $t$, $r$ represents the
intrinsic population growth rate, $B_0$ represents unfished biomass, and $C$
represents catch. One of the models is empirical instead of mechanistic.
@rosenberg2014 provide a full summary of these four methods and we provide some
further details in the Supporting Materials. Code to carry out all methods is
available in an accompanying package **datalimited** for the statistical
software \textsf{R}. In summary:

- *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
harvest-dynamics model developed by @vasconcellos2005. Biomass is assumed to
follow a Schaefer model and harvest dynamics are assumed to follow a logistic
model. The model is fit with a sample-importance-sampling algorithm developed
in @rosenberg2014.

- *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
dynamics [@martell2013]. It requires a prior distribution for $r$ and assigns
a prior to the relative proportion of biomass at the end compared to unfished
biomass (depletion) based on the percentage of maximum catch at the end of the
time series.

- *SSCOM* (state-space catch-only-model) ...

- *mPRM* (modified panel regression model) is a modified version of the
panel-regression model used in @costello2012. Unlike the other models
considered, mPRM is empirical and not mechanistic --- it fits a linear model to
a series of characteristics of the catch time series and stock with \bbmsy\\ as
the response. The model used in this paper and @rosenberg2014 is modified from
the original in that it simplifies the life-history categories to be applicable
to simulated data.

## Additional covariates

Ensemble methods allow us to incorporate additional covariates into our models
and potentially leverage interactions of how these covariates combine with the
main individual model estimates to estimate population status. Examples might
be life-history characteristics, additional information on exploitation
patterns, or additional statistical properties of the data going into the
individual models. For simplicity, and to allow us to apply models developed
with the simulated dataset to the real-world dataset, we added only one set of
additional covariates: spectral properties of the catch time series itself. We
fit spectral models to the catch time series and recorded representative
short-term and long-term spectral densities of 0.05 (corresponding to a 20-year
cycle) and 0.2 (corresponding to a 5-year cycle).

## Ensemble models

The individual models we seek to combine with ensemble models provide time
series of \bbmsy\\. Therefore, we can use ensemble models to estimate a number of
different attributes of the time series. Here, we focus on two attributes the
mean and slope of \bbmsy\\ in the last 3 years. We refer to these quantities
collectively as $\theta$. Together, these quantities address both the current
state and trajectory of status, both of which may be of management and
conservation interest. To avoid undue influence of the end points of the time
series on the calculated slope, we instead measured the median slope as the
Theil-Sen estimator [@theil1950].

An ensemble model can combine inference through anything from a simple average,
to a weighted average, to a complex non-linear model. Here we choose four
ensemble models to compare: an average, a linear model with two-way
interactions, a random forest, and boosted regression tree.

General form:

\begin{equation}
\hat{\theta} = \mathrm{argmin}_\theta(L(b \, | \, \hat{b}_{\mathrm{ensemble}}))
\end{equation}

\begin{equation}
\hat{b}_\mathrm{ensemble} = f(\hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, \hat{b}_\mathrm{SSCOM}, \hat{b}_\mathrm{mPRM}, \mathrm{other~covariates} \, | \, \theta)
\end{equation}

\noindent where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{COM-SIR}$ is the prediction from COM-SIR, etc., and $\theta$ is the set of parameters for the regression function.

*Linear model*: For populations $i$ $1$ through $n$, individual models and
covariates $j$ $1$ through $m$, and allowing $\theta$ to refer to both
quantities estimated from individual models and additional covariates, we fit
the linear model with two-way interactions as:

<!--\noindent-->
<!--The geometric mean is calculated as-->
<!--$$B/B_{\mathrm{MSY-ensemble}(i)} = \exp \left(\frac{\sum_{j=1}^{4} \log B/B_{\mathrm{MSY}(ji)}}{4}\right)$$-->

<!--$$-->
<!--\log B/B_{\mathrm{MSY-ensemble}(i)} = -->
  <!--\beta_0 + \beta_1 \log B/B_{\mathrm{MSY}(i)} -->
<!--$$-->

\begin{equation}
\theta_{\mathrm{ensemble}(i)} = 
  \beta_0 + 
  \sum_{j=1}^m \beta_j \theta_{j(i)} + 
  \sum_{j=1}^m \sum_{\substack{k =1 \\ j \neq k}}^m 
            \beta_{kj} \theta_{k(i)} \theta_{j(i)} +
  \epsilon_i, \quad
  \epsilon_i \sim \mathrm{Normal}(0, \sigma),
\end{equation}

\noindent
where the $\beta$ represent coefficients and $\epsilon$ represents normally
distributed residual variation. We also tried fitting linear models without
interactions but found better cross-validation predictive performance from all
performance measures evaluated using the procedure outlined below.

*Random forests* are a machine learning method that combines a series of
regression trees. Regression trees sequentially determine what value of
a predictor best splits the response data into two 'branches' based on
a specified loss function. In random forests, a series of regression trees are
built on a random subset of the data and random subset of the covariates of the
model. By combining many of these stochastically generated models, random
forests can provide good predictive performance and fit highly non-linear
relationships TODO. We fit random forest models with the **randomForest**
package [@liaw2002] for \textsf{R} with the default argument values.

*Boosted regression models* are a machine learning method that, like random
forests, combines a series of regression trees built on random subsets of the
data, but also add a boosting procedure. Boosting refers to fitting a series
of models with each subsequent model fit to the residuals from the previous
model; data points that are fit poorly in a given model are given more weight
in the next model [@elith2008]. Boosting can result in TODO and TODO and is
a common and powerful machine learning technique [@elith2008]. We fit boosted
regression tree models with the **GBM** package [@ridgeway2015] for \textsf{R}
with 10000 trees, an interaction depth of $2$, a learning rate (shrinkage
parameter) of $0.01$, and all other arguments at their default values.

For all ensemble models of $\overline{B/B_\mathrm{MSY}}$, which is bounded at
zero, we log transformed all $\overline{B/B_\mathrm{MSY}}_{(ji)}$ and
exponentiated the estimated $\log
\widehat{\overline{B/B_\mathrm{MSY}}}_\mathrm{ensemble}$ to derive a median
estimate of $\widehat{\overline{B/B_\mathrm{MSY}}}_\mathrm{ensemble}$. For the
estimates of \bbmsy\\ slope, which are not bounded at zero, we fit ensemble
models on the natural untransformed scale.

## Testing model performance

A critical component to any predictive modelling exercise is to evaluate the
predictive performance of a model on new data [@hastie2009]. Typically, data
are limited in availability, and so a common and effective tool is
cross-validation [@hastie2009]. We used repeated three-fold cross validation to
test predictive performance: we randomly divided the dataset into three sets,
built ensemble models on two-thirds of the data, and evaluated predictive
performance on the remaining third of the data. We repeated this across each of
the three splits and then repeated the whole procedure XX times to account for
bias that may result from any one set of validation splits.

Predictive performance can be evaluated with metrics that represent a variety
of modelling goals. For instance, a model may provide an unbiased but highly
imprecise prediction. For continuous response variables such as those
considered here performance metrics often measure some form of bias, precision,
or accuracy (a combination of bias and precision), or the ability to correctly
rank or correlate across populations. Here, we measure proportional error (PE;
also called 'relative error') defined as $(\hat{\theta} - \theta)/\theta$,
where $\theta$ represents the true parameter value and $\hat{\theta}$ the
estimate of that parameter ($\overline{B/B_\mathrm{BMSY}}$ or slope of
\bbmsy\\). We summarize median proportional error  to measure bias, median
absolute proportional error to measure accuracy, and Spearman's rank-order
correlation to measure the ability to correctly rank population status.

# Results

Individual model and ensemble model scatterplots:
Fig. \ref{fig:hexagon}

Accuracy, rank correlation, and bias:
Fig. \ref{fig:performance}

# Discussion

# Acknowledgements

Funding...

<!--# Citation notes-->

<!--
- @breiner2015 ensemble models of species distribution models for rare species

- @jones2015 ensemble models of species distribution models - globally for
  marine biodiversity

- [@greene2006] example similar to ours but with climate models (Bayesian
  multilevel ensemble)

- [@kell2007] FLR

- multiple learners book [@alpaydin2010]

- [@caruana2004] nice paper on ensemble models; prob of overfitting increases
  with more models, bagging important

- [@tebaldi2007] key paper: review of 'multi-model ensembles for climate projections'

- famous ensemble is DEMETER: Development of a European Multi-model Ensemble
  System for Seasonal to Interannual Prediction @hagedorn2005 is a good
  reference

- Examples of where ensemble models are shown to be better than any one:
  @thomson2006 (public health - malaria), @cantelaube2005 (agriculture - crop
  yield) (citations taken from @tebaldi2007)

- simple averages are used very commonly in climate science - e.g. IPCC 2001

- @tebaldi2007: weighting obviously makes sense, but how do we define a
  performance metric that is based on past observations that is relevant to the
  future?

- @tebaldi2007: model independence important

- for climate, weighted averages perform better than simple averages [@min2006],
  but are those same weights applicable to the future (or in our case other
  fisheries)?

- error cancellation is one but not the only reason for superiority of ensemble
  models [@hagedorn2005]

- ensemble model may be only marginally better than the best single model in
  any given case, but we don't usually know which is the best single model
  [@hagedorn2005]

- ensemble models useful for regional climate models too; as an example,
  @pierce2009 use 42 metrics to characterize model performance in regional
  climate model ensembles; found that ensemble models were superior to any one
  model --- especially when considering multiple metrics

- @knutti2009: key review paper on motivation and challenges of combining
  climate projection models; performance on testing / current data may only
  weakly relate to future / other datasets

- @murphy2004: Nature paper on ensembles of climate model simulations; weighted
  average better performance than unweighted average

- @dietterich2000: highly cited book chapter "Ensemble Methods in Machine
  Learning"; ensembles are often much more accurate than the "individual
  classifiers that make them up"; but components must be diverse and accurate

- @dietterich2000: a main justification for ensembles: they are
  representational --- no one model usually can contain all hypothetical
  functional forms, but many separate models can cover more hypotheses

- strong paper showing that ensemble models are most accurate when the various
  individual models make errors in uncorrelated ways [@ali1996]

- without substantial training-testing data, appropriate model weights can be
  very hard to deduce and can cause more harm than good (compared to equal
  weighting) [@weigel2010] (they give the example of seasonal forecasting where
  a ton of hind cast testing can be done, vs. long-term climate) (asymmetrical
  loss function)

- [@weigel2010] optimal weights are always as good or better than equal
  weights, but if you get the weights wrong, you can be better off just using
  equal weights; but averaging was almost always better than any one model

- @rykiel1996 "Testing ecological models: the meaning of validation" (see for
  performance criteria, theory on model testing and assessment)
-->

### 2015-06-05 analysis/8-main-figs.R
Add script to generate result values

@@ -48,6 +48,8 @@ d_sim_perf_summ$text_col <- ifelse(grepl("Ensemble", d_sim_perf_summ$clean_metho
 d_sim_perf_wide <- reshape2::dcast(d_sim_perf_summ, clean_method + order ~ variable,
   value.var = "m")
 
+saveRDS(d_sim_perf_wide, "generated-data/d_sim_perf_wide.rds")
+
 # Format basic simulation ensemble without covariates for plotting:
 d_sim_basic <- readRDS("generated-data/cv_sim_mean_basic_long.rds")
 d_sim_basic$bbmsy_est[d_sim_basic$bbmsy_est > 10] <- NA
@@ -71,6 +73,7 @@ re_ram_sum <- re_ram %>% group_by(clean_method) %>%
     mre = median(re),
     corr = cor(bbmsy_true, bbmsy_est, method = "spearman")) %>%
   as.data.frame()
+saveRDS(re_ram_sum, file = "generated-data/re_ram_sum.rds")
 
 # now make these data match the simulation data in format:
 re_ram_sum_long <- reshape2::melt(re_ram_sum, id.vars = "clean_method") %>%
@@ -102,6 +105,7 @@ auc_sim_mean <- rocs_sim_mean %>%
   summarise(auc = auc[1])
 rocs_sim_mean <- rocs_sim_mean %>% as.data.frame() %>%
   mutate(clean_method = paste(order, clean_method, sep = "-"))
+saveRDS(auc_sim_mean, file = "generated-data/auc_sim_mean.rds")
 
 # and do the same for the RAM data:
 d_ram_mean <- d_ram %>%
@@ -287,6 +291,7 @@ d_slope_error <- d_slope_plot %>%
     corr = cor(bbmsy_true, bbmsy_est, method = "spearman",
       use = "pairwise.complete.obs")) %>%
   as.data.frame
+saveRDS(d_slope_error, file = "generated-data/d_slope_error.rds")
 
 pdf("../figs/performance-slope-sim.pdf", width = 5, height = 4)
 par(mgp = c(1.5, 0.5, 0), las = 1, tck = -0.015,

### 2015-06-05 analysis/9-output-values.R

@@ -1,0 +1,55 @@
+# This file generates 'magic' numbers for input into the paper. These are,
+# for example, ranges of MARE or correlation or AUC values. They get saved to
+# `../text/values.rda` and are input through knitr in `../text/ms.Rmd`.
+
+get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
+  label = "") {
+
+  dat <- dat %>%
+    mutate(ensemble = ifelse(grepl("Ensemble", clean_method), TRUE, FALSE)) %>%
+    mutate(machine = ifelse(grepl("GBM|RF", clean_method), TRUE, FALSE))
+
+  x <- list()
+
+  mach <- dat %>% filter(ensemble == TRUE, machine == TRUE)
+  ensemble <- dat %>% filter(ensemble == TRUE)
+  ind <- dat %>% filter(ensemble == FALSE)
+
+  x$mach_vs_ind_mare_fold <- c(
+    round(min(ind$mare) / max(mach$mare), digits_fold),
+    round(max(ind$mare) / min(mach$mare), digits_fold)
+  )
+
+  x$ind_corr_range <- round(range(ind$corr), digits_raw)
+  x$mach_corr_range <- round(range(mach$corr), digits_raw)
+
+  x$ind_mare_range <- round(range(ind$mare), digits_raw)
+  x$mach_mare_range <- round(range(mach$mare), digits_raw)
+
+  x$ensemble_mre_range <- ensemble$mre %>% range %>% round(digits_raw)
+  x$ind_mre_range <- ind$mre %>% range %>% round(digits_raw)
+
+  # names(x) <- paste(names(x), label, sep = "_")
+
+  x
+}
+
+d_sim_perf_wide <- readRDS("generated-data/d_sim_perf_wide.rds")
+mean_sim <- get_performance_stats(d_sim_perf_wide)
+
+re_ram_sum <- readRDS("generated-data/re_ram_sum.rds")
+mean_ram <- get_performance_stats(re_ram_sum)
+
+d_slope_error <- readRDS("generated-data/d_slope_error.rds")
+slope_sim <- get_performance_stats(d_slope_error)
+
+auc_sim_mean <- readRDS("generated-data/auc_sim_mean.rds")
+auc_sim <- list()
+
+auc_sim$mach_range <- auc_sim_mean %>% filter(grepl("GBM|RF", clean_method)) %>%
+  select(auc) %>% range %>% round(2)
+
[…]

### 2015-06-05 text/ms.Rmd
Integrate results into abstract

@@
 
-# Includes: mean_sim, mean_ram, slope_sim, auc_sim
+# Includes: mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n
 
@@
 # "mach_range" "ind_range" 
-
-# Many elements contain a numeric vector of length 2 indicating lower and 
-# upper values.
-
-# Insert one, with, for example: `r auc_sim$mach_range[1]`.
 ```
@@
 
-What can we do when multiple ecological models give different impressions of
-population status? Any single model is unlikely to perform best across all
-species, exploitation dynamics, and statuses. Ensemble methods from the fields
-of machine learning and climate science provide a powerful tool to optimally
-leverage the information contained in multiple ecological models. Here we
-evaluate the potential for ensemble methods to improve estimates of fish
-population status. We fit four data-limited models of population and
-exploitation dynamics and combine their estimates of the mean and slope of
-\bbmsy\\ (biomass at maximum sustainable yield) with four ensemble models:
-a simple average, a linear model, and two machine learning methods: random
-forests and boosted regression trees. We build our ensemble models on a fully
-factorial simulated dataset of 5760 fish stocks and test our models on
-a cross-validated version of this dataset and on a global database of 166
-assessed fish stocks. We find that ensemble models substantially improve
-estimates of population status, trajectory, and classification. Machine
-learning ensemble models performed the best (XX--XX% improved accuracy, XX--XX%
-improved correlation, XX-XX% reduced bias compared to the individual models).
-When the same ensembles were applied to a separate dataset of global fish
-stocks and compared to status from full stock assessments, ensemble methods
-still improved estimates of status and trajectory over individual models
-(accuracy, correlation, and bias improved up to XX%, XX%, and XX%). Ensemble
-models have great potential to improve estimates of population status and
-trajectory across many taxa beyond fish. However, ensemble methods must be
-carefully tested, formed from a diverse set of accurate models, and built on
-a dataset sufficiently representative of the populations they are applied to.
+What can we do when multiple ecological models give different impressions of population status?
+Any single model is unlikely to perform best across all species, exploitation dynamics, and statuses.
+Ensemble methods from the fields climate science and machine learning provide a powerful tool to optimally leverage the information contained in multiple ecological models.
+Here we evaluate the potential for ensemble methods to improve estimates of fish population status.
+We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine learning methods---random forests and boosted regression trees.
+We build our ensemble models on a fully factorial simulated dataset of 5760 fish stocks and test our models on a cross-validated version of this dataset and a global database of `r ram_stocks_n` assessed fish stocks.
+We find that ensemble models substantially improve estimates of population status and trajectory.
+Machine learning ensemble models performed the best: `r mean_sim$mach_vs_ind_mare_fold`% improved accuracy (reduction in median absolute proportional error), rank-order correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements in performance when predicting the slope of status.
+When the same simulation-trained ensemble models were applied to a separate dataset of global fish stocks and compared to mean status from data-rich stock assessments, the best ensembles still improved all dimensions of performance (accuracy, correlation, and bias) over the next best individual model.
+Ensemble models have great potential to improve estimates of population status and trajectory across many taxa beyond fish.
+However, ensemble methods must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations they are applied to.
 

### 2015-06-05 analysis/9-output-values.R

@@ -3,7 +3,7 @@
 # `../text/values.rda` and are input through knitr in `../text/ms.Rmd`.
 
 get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
-  label = "") {
+  collapse = "--") {
 
   dat <- dat %>%
     mutate(ensemble = ifelse(grepl("Ensemble", clean_method), TRUE, FALSE)) %>%
@@ -18,7 +18,12 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   x$mach_vs_ind_mare_fold <- c(
     round(min(ind$mare) / max(mach$mare), digits_fold),
     round(max(ind$mare) / min(mach$mare), digits_fold)
-  )
+  ) * 100
+
+#   x$mach_vs_ind_bias <- c(
+#     (ind$mre %>% abs %>% max) / (mach$mre %>% abs %>% min) %>% round(digits_fold),
+#     (ind$mre %>% abs %>% min) / (mach$mre %>% abs %>% max) %>% round(digits_fold)
+#   ) * 100
 
   x$ind_corr_range <- round(range(ind$corr), digits_raw)
   x$mach_corr_range <- round(range(mach$corr), digits_raw)
@@ -27,11 +32,10 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   x$mach_mare_range <- round(range(mach$mare), digits_raw)
 
   x$ensemble_mre_range <- ensemble$mre %>% range %>% round(digits_raw)
+  x$mach_mre_range <- mach$mre %>% range %>% round(digits_raw)
   x$ind_mre_range <- ind$mre %>% range %>% round(digits_raw)
 
-  # names(x) <- paste(names(x), label, sep = "_")
-
-  x
+  lapply(x, paste, collapse = collapse)
 }
 
 d_sim_perf_wide <- readRDS("generated-data/d_sim_perf_wide.rds")
@@ -52,4 +56,7 @@ auc_sim$mach_range <- auc_sim_mean %>% filter(grepl("GBM|RF", clean_method)) %>%
 auc_sim$ind_range <- auc_sim_mean %>% filter(grepl("Ensemble", clean_method) == FALSE) %>%
   select(auc) %>% range %>% as.numeric() %>% round(2)
 
-save(mean_sim, mean_ram, slope_sim, auc_sim, file = "../text/values.rda")
[…]

### 2015-06-05 text/ms.Rmd
Shorten the abstract
Still too long, should be < 200 words

@@
 What can we do when multiple ecological models give different impressions of population status?
-Any single model is unlikely to perform best across all species, exploitation dynamics, and statuses.
-Ensemble methods from the fields climate science and machine learning provide a powerful tool to optimally leverage the information contained in multiple ecological models.
+Any single model is unlikely to perform best in all scenarios.
+However, ensemble methods from the fields climate science and machine learning provide a powerful tool to leverage information across multiple ecological models.
 Here we evaluate the potential for ensemble methods to improve estimates of fish population status.
-We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine learning methods---random forests and boosted regression trees.
-We build our ensemble models on a fully factorial simulated dataset of 5760 fish stocks and test our models on a cross-validated version of this dataset and a global database of `r ram_stocks_n` assessed fish stocks.
+We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees.
+We build our ensembles on a simulated dataset of 5760 fish stocks and test our models with cross-validation and a global database of `r ram_stocks_n` stock assessments.
 We find that ensemble models substantially improve estimates of population status and trajectory.
-Machine learning ensemble models performed the best: `r mean_sim$mach_vs_ind_mare_fold`% improved accuracy (reduction in median absolute proportional error), rank-order correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
-We found similar improvements in performance when predicting the slope of status.
-When the same simulation-trained ensemble models were applied to a separate dataset of global fish stocks and compared to mean status from data-rich stock assessments, the best ensembles still improved all dimensions of performance (accuracy, correlation, and bias) over the next best individual model.
-Ensemble models have great potential to improve estimates of population status and trajectory across many taxa beyond fish.
+Machine-learning ensembles performed the best: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements when predicting the slope of status.
+When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next best individual model.
+Ensemble models have potential to improve estimates of population status and trajectory across many taxa beyond fish.
 However, ensemble methods must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations they are applied to.

### 2015-06-08 text/ms.Rmd
Edit abstract; add todo

@@
 We found similar improvements when predicting the slope of status.
-When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next best individual model.
+When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next-best individual model.
 Ensemble models have potential to improve estimates of population status and trajectory across many taxa beyond fish.
@@
 model and the code to generate the simulations is available at
-<https://github.com/flr/StockSims>. To summarize, the model was built in the
+<https://github.com/flr/StockSims> (TODO: not there). To summarize, the models were built in the
 package FLR [@kell2007] for the statistical software \textsf{R} [@r2015]...

### 2015-06-09 text/ms.Rmd
Major rewrite of introduction and methods

@@
 
-What can we do when multiple ecological models give different impressions of population status?
-Any single model is unlikely to perform best in all scenarios.
+What can we do when multiple ecological models give different impressions of population status? Any single model is unlikely to perform best in all scenarios. 
 However, ensemble methods from the fields climate science and machine learning provide a powerful tool to leverage information across multiple ecological models. 
@@
 
-When models give conflicting estimates of a population's status, there are a
-number of solutions. (1) We can pick a single model. However, it isn't always
-obvious which model is best, and even objectively second-best models may contain
-useful additional information. (2) We can maintain multiple models throughout
-the decision-making process. However, this can be complicated (REFs)... (3) We
-can combine the model outputs. This could be as simple as taking the average,
-potentially weighting the models by some performance metric. This could be as
-complicated as forming an additional 'hyper' model that draws on interactions
-between model outputs and potentially incorporates additional information to
-derive a best estimate.
-
-Ensemble models are widely used in the climate sciences, where thousands of
-models with varying structural combined across different structural ... The
-idea behind ensemble models forms the backbone of many machine learning
-methods. For example...
+Ensemble models, which combine the output from multiple models, are one
+potential solution to this problem. A simple ensemble model might take the
+average of multiple model outputs, perhaps weighted by some goodness of fit
+(REFs). A complex ensemble model might also take into account the interactions
+between model outputs and additional covariates. Ensemble models are popular in
+many fields including public health [e.g. @thomson2006], agriculture [e.g.
+@cantelaube2005], climate science (e.g. REFs), and machine learning (e.g.
+REFs). We think they have tremendous opportunity in ecological resource
+management.
 
 In fisheries science, a common problem is estimating the status of an exploited
-fish population. For the majority of fish stocks, we have limited information to
-go on and stock status isn't known. In recent years, a number of methods have
-been proposed to derive population status based on limited information and a set
-of assumptions. However, we know that these models can give conflicting output
-and no one model is that accurate.
+fish population. For the majority of fish stocks, we have limited biological
+information and stock status isn't known (e.g. REFs).  In recent years, a
+number of methods have been proposed to derive population status based on the
+limited information that is frequently available: fisheries catch time series
+and basic knowledge about the productivity of the species. Recent work has
+investigated the ability for these models to estimate population status through
+a large-scale simulation model [@rosenberg2014]. A key finding was that these
+models frequently disagree on population status and any one model has
+relatively poor predictive ability (e.g. Fig. \ref{fig:motivate}).
 
-Motivation: (Fig. \ref{fig:motivate}).
-
-Ensemble models combine estimates of individual models, potentially along with
-additional covariates, in an additional 'ensemble' model
-(Fig. \ref{fig:didactic}).
-
+<!--
 Exploit interactions between individual models and relationships between other
 conditions of the data and the performance of the individual models.
-
 Fitted to populations of known or assumed status and this ensemble model is
 used to extrapolate to populations of interest.
+-->
 
-Here, we develop ensemble models for data-limited exploited fish populations. We
-explore a variety of ensemble approaches applied to both simulated and
-real-world fish stocks and compare performance...
+Here, we develop ensemble models to estimate the population status of exploited
+fish populations. We explore a variety of ensemble approaches applied to both
+simulated and real-world fish stocks and compare predictive performance. In
+developing these models we highlight the types of problems one encounters when
+forming and evaluating ensemble models. We find that ensembles generally
+provide improved predictive performance than any single model---especially when
+considered across multiple dimensions of predictive ability. Furthermore, the
+output from ensemble models can provide better mechanistic understanding of how
+invidual models perform.
 
@@
 
-We developed and tested ensemble models with both a fully factorial simulated
-dataset [@rosenberg2014] and the RAM Legacy Stock Assessment Database
-[@ricard2012]. @rosenberg2014 provide a full description of the simulation
-model and the code to generate the simulations is available at
-<https://github.com/flr/StockSims> (TODO: not there). To summarize, the models were built in the
-package FLR [@kell2007] for the statistical software \textsf{R} [@r2015]...
-included all combinations of the following factors:
+We first developed and tested ensemble models with a fully factorial simulated
+dataset [@rosenberg2014]. The simulation models included three life histories (small
+pelagic, demersal, large pelagic), three levels of how depleted biomass was at
+the start of the dataset (100%, 70%, and 40% of carrying capacity), four
+harvest dynamics (a constant harvest rate, a harvest rate that is coupled with
+biomass to mimic an open-access single-species fishery, a scenario where
+harvest rate increases to a fixed level, a 'roller-coaster' scenario where the
+harvest rate increases, maintains a steady level and decreases to a lower
+steady level). Process noise was introduced to the models through two levels of
+multiplicative recruitment variability ($N(0, 0.2)$ and $N(0, 0.6)$, two levels
+of first-order autoregressive correlation on recruitment variablity (0 and
+0.6). The simulation included a scenario without observation error and with
+multiplicative observation error around catch at $N(0, 0.2)$.
 
-- Three life histories: small pelagic, demersal, large pelagic.
-- Two time series lengths: 20 and 60 years.
-- Three levels of initial depletion: 1, 0.7, and 0.4.
-- Four harvest dynamics: (1) a constant harvest rate, (2) a harvest rate that
-  is coupled with biomass to mimic an open-access single-species fishery, (3) a
-  one-way trip where harvest rate increases five percent per year to 80% of the
-  level at which the stock would crash, and (4) roller-coaster shaped where the
-  harvest rate increases, maintains a steady level at 80% of the level at which
-  the stock would crash, and then decreases to FMSY. 
-- Two levels of multiplicative standard deviation around catch (0 and 0.2)
-- Two levels of standard deviation of recruitment variability (0.2 and 0.6)
-- Two levels of autoregressive correlation on recruitment variability (0 and 0.6)
+@rosenberg2014 ran ten iterations for each combination of factors adding
+stochastic draws of recruitment and catch-recording variability each time to
+generate a total of 5760 stocks. The simulation models were built in the
+package FLR [@kell2007] for the statistical software \textsf{R} [@r2015].
+@rosenberg2014 provide a full description of the simulation model and the code
+to generate the simulations is available at <https://github.com/flr/StockSims>
+(TODO: not there).
 
-\noindent @rosenberg2014 ran ten iterations for each combination of factors
-adding stochastic draws of recruitment and catch-recording variability each
-time to generate a total of 5760 stocks.
-
-Our analysis of the RAM Legacy Stock Assessment Database was based on version
-XX downloaded on XX. This version includes XX stocks from XX countries across
-XX taxonomic orders.
+We also tested our ensemble models on the RAM Legacy Stock Assessment Database
+[@ricard2012]. Our analysis of the stock-assessment database was based on
+version XX downloaded on XX. After removing stocks for which the models
+described below did not converge, this database included `r ram_stocks_n` from
+XX countries across XX taxonomic orders.
 
@@
 We fit four individual data-limited models to estimate \bbmsy\\. Three of the
-models are mechanistic and based generally on Shaefer biomass dynamics
-[@schaefer1954] of the form:
+models are mechanistic and based generally on @schaefer1954 biomass
+dynamics of the form:
 
@@
 
-\noindent where $B_t$ represents biomass at time $t$, $r$ represents the
-intrinsic population growth rate, $B_0$ represents unfished biomass, and $C$
-represents catch. One of the models is empirical instead of mechanistic.
-@rosenberg2014 provide a full summary of these four methods and we provide some
-further details in the Supporting Materials. Code to carry out all methods is
-available in an accompanying package **datalimited** for the statistical
-software \textsf{R}. In summary:
+\noindent where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus
+one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic
+population growth rate, $B_0$ represents unfished biomass or carrying capacity,
+and $C$ represents catch. @rosenberg2014 provide a full summary of these four
+methods and code to fit all the models is available in an accompanying package
+**datalimited** for the statistical software \textsf{R}. In summary:
 
@@
 to simulated data.
+<!--TODO: what are we doing with the RAM stocks again?-->
 
@@
 with the simulated dataset to the real-world dataset, we added only one set of
-additional covariates: spectral properties of the catch time series itself. We
-fit spectral models to the catch time series and recorded representative
-short-term and long-term spectral densities of 0.05 (corresponding to a 20-year
-cycle) and 0.2 (corresponding to a 5-year cycle).
+additional covariates: spectral properties of the catch time series itself.
+Spectral analysis decomposes a time series into the frequency domain and
+provides a useful means of describing the statistical properties of the catch
+series that is independent of time series length and magnitude of catch. We fit
+spectral models to the scaled catch time series (divided by maximum catch) with
+the `spec.ar` function in \textsf{R} and recorded representative short- and
+long-term spectral densities of 0.05 and 0.2, which correspond to 20- and
+5-year cycles.
 
@@
 The individual models we seek to combine with ensemble models provide time
-series of \bbmsy\\. Therefore, we can use ensemble models to estimate a number of
-different attributes of the time series. Here, we focus on two attributes the
-mean and slope of \bbmsy\\ in the last 3 years. We refer to these quantities
-collectively as $\theta$. Together, these quantities address both the current
+series of \bbmsy\\ ($\hat{b}_\mathrm{CMSY}$, $\hat{b}_\mathrm{COM-SIR}$, etc.).
+Therefore, we can use ensemble models to estimate multiple properties of these
+status time series. Here, we focus on two values: the mean and slope of
+\bbmsy\\ in the last 5 years. Below, we refer to these quantities collectively
+as $\hat{b}_\mathrm{ensemble}$. Together, these quantities address the current
 state and trajectory of status, both of which may be of management and
 conservation interest. To avoid undue influence of the end points of the time
-series on the calculated slope, we instead measured the median slope as the
-Theil-Sen estimator [@theil1950].
+series on the calculated slope, we measured the slope as the Theil-Sen
+estimator of median slope [@theil1950].
 
-An ensemble model can combine inference through anything from a simple average,
-to a weighted average, to a complex non-linear model. Here we choose four
-ensemble models to compare: an average, a linear model with two-way
-interactions, a random forest, and boosted regression tree.
+An ensemble model can combine models through anything from a simple average to
+a complex non-linear model. Here we choose four ensemble models to compare: an
+average, a linear model with two-way interactions, a random forest, and boosted
+regression tree. The general form of our models is as follows:
 
-General form:
+$$
+\hat{\theta} = \mathrm{argmin}_\theta(L(b \, | \, \hat{b}_{\mathrm{ensemble}}))
+$$
 
-\begin{equation}
-\hat{\theta} = \mathrm{argmin}_\theta(L(b \, | \, \hat{b}_{\mathrm{ensemble}}))
-\end{equation}
+$$
+\hat{b}_\mathrm{ensemble} = 
+f(\hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, 
+\hat{b}_\mathrm{SSCOM}, \hat{b}_\mathrm{mPRM}, 
+S(0.2), S(0.5), | \, \theta)
+$$
 
-\begin{equation}
-\hat{b}_\mathrm{ensemble} = f(\hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, \hat{b}_\mathrm{SSCOM}, \hat{b}_\mathrm{mPRM}, \mathrm{other~covariates} \, | \, \theta)
-\end{equation}
+\noindent where $L$ is a generalized loss function, $f$ is some generalized
+regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from COM-SIR,
+$\hat{b}_\mathrm{COM-SIR}$ etc., $b$ is the true value the model is fitted to,
+$S(n)$ represents the spectral density at frequency $n$ of the scaled catch
+series, and $\theta$ is the set of parameters for the regression function.
 
-\noindent where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{COM-SIR}$ is the prediction from COM-SIR, etc., and $\theta$ is the set of parameters for the regression function.
-
+<!--
 *Linear model*: For populations $i$ $1$ through $n$, individual models and
@@
 the linear model with two-way interactions as:
+-->
 
@@
 
+<!--
 \begin{equation}
@@
 performance measures evaluated using the procedure outlined below.
+-->
 
 *Random forests* are a machine learning method that combines a series of
-regression trees. Regression trees sequentially determine what value of
-a predictor best splits the response data into two 'branches' based on
-a specified loss function. In random forests, a series of regression trees are
+regression trees. Regression trees sequentially determine what value of a
+predictor best splits the response data into two 'branches' based on a
+specified loss function. In random forests, a series of regression trees are
 built on a random subset of the data and random subset of the covariates of the
 model. By combining many of these stochastically generated models, random
-forests can provide good predictive performance and fit highly non-linear
-relationships TODO. We fit random forest models with the **randomForest**
+forests can provide strong predictive performance and fit highly non-linear
+relationships (REFs). We fit random forest models with the **randomForest**
 package [@liaw2002] for \textsf{R} with the default argument values.
@@
 forests, combines a series of regression trees built on random subsets of the
-data, but also add a boosting procedure. Boosting refers to fitting a series
-of models with each subsequent model fit to the residuals from the previous
-model; data points that are fit poorly in a given model are given more weight
-in the next model [@elith2008]. Boosting can result in TODO and TODO and is
-a common and powerful machine learning technique [@elith2008]. We fit boosted
-regression tree models with the **GBM** package [@ridgeway2015] for \textsf{R}
-with 10000 trees, an interaction depth of $2$, a learning rate (shrinkage
-parameter) of $0.01$, and all other arguments at their default values.
+data, but also add a boosting procedure. Boosting refers to fitting a series of
+models with each subsequent model fit to the residuals from the previous model;
+data points that are fit poorly in a given model are given more weight in the
+next model [@elith2008]. Boosting can result in TODO and TODO and is a common
+and powerful machine learning technique [@elith2008]. We fit boosted regression
+tree models with the **GBM** package [@ridgeway2015] for \textsf{R}. Based on
+cross-validation with the **caret** package (Supporting Materials, Fig. SX), we
+fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate
+(shrinkage parameter) of $0.01$, and all other arguments at their default
+values.
 
-For all ensemble models of $\overline{B/B_\mathrm{MSY}}$, which is bounded at
-zero, we log transformed all $\overline{B/B_\mathrm{MSY}}_{(ji)}$ and
-exponentiated the estimated $\log
-\widehat{\overline{B/B_\mathrm{MSY}}}_\mathrm{ensemble}$ to derive a median
-estimate of $\widehat{\overline{B/B_\mathrm{MSY}}}_\mathrm{ensemble}$. For the
-estimates of \bbmsy\\ slope, which are not bounded at zero, we fit ensemble
-models on the natural untransformed scale.
+For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the
+models in log space and exponentiated the predictions to derive a median
+estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope, which are not
+bounded at zero, we fit ensemble models on the natural untransformed scale.
 
@@
 A critical component to any predictive modelling exercise is to evaluate the
-predictive performance of a model on new data [@hastie2009]. Typically, data
-are limited in availability, and so a common and effective tool is
-cross-validation [@hastie2009]. We used repeated three-fold cross validation to
-test predictive performance: we randomly divided the dataset into three sets,
-built ensemble models on two-thirds of the data, and evaluated predictive
-performance on the remaining third of the data. We repeated this across each of
-the three splits and then repeated the whole procedure XX times to account for
-bias that may result from any one set of validation splits.
+performance of a model new data [@hastie2009]. We used repeated three-fold
+cross validation to test predictive performance: we randomly divided the
+dataset into three sets, built ensemble models on two-thirds of the data, and
+evaluated predictive performance on the remaining third of the data. We
+repeated this across each of the three splits and then repeated the whole
+procedure XX times to account for bias that may result from any one set of
+validation splits. Since the dynamics of simulated populations differing only
+in the magnitude of observation error around catch, magnitude of recuitment
+variability, or stochastic draw of these values are likely similar, we grouped
+these characteristics together in the cross-validation process---we never
+trained and tested on models differing only in these characteristics.
 
 Predictive performance can be evaluated with metrics that represent a variety
-of modelling goals. For instance, a model may provide an unbiased but highly
-imprecise prediction. For continuous response variables such as those
-considered here performance metrics often measure some form of bias, precision,
-or accuracy (a combination of bias and precision), or the ability to correctly
-rank or correlate across populations. Here, we measure proportional error (PE;
-also called 'relative error') defined as $(\hat{\theta} - \theta)/\theta$,
-where $\theta$ represents the true parameter value and $\hat{\theta}$ the
-estimate of that parameter ($\overline{B/B_\mathrm{BMSY}}$ or slope of
-\bbmsy\\). We summarize median proportional error  to measure bias, median
-absolute proportional error to measure accuracy, and Spearman's rank-order
-correlation to measure the ability to correctly rank population status.
+of modelling goals. For continuous response variables such as the mean and
+slope of population status, peformance metrics often measure some form of bias,
+precision, accuracy (a combination of bias and precision), or the ability to
+correctly rank or correlate across populations. Here, we measure proportional
+error (PE; also called 'relative error') defined as $(\hat{b} - b)/b$. We
+summarize median proportional error to measure bias, median absolute
+proportional error to measure accuracy, and Spearman's rank-order correlation
+to measure the ability to correctly rank populations.
 

### 2015-06-09 text/ms.Rmd
More edits to methods

@@
 $$
-\hat{\theta} = \mathrm{argmin}_\theta(L(b \, | \, \hat{b}_{\mathrm{ensemble}}))
+\hat{\theta} = 
+\mathrm{argmin}_\theta \left( L(b \, | \, \hat{b}_{\mathrm{ensemble}}) \right)
 $$
-
 $$
 \hat{b}_\mathrm{ensemble} = 
-f(\hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, 
+f \left( \hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, 
 \hat{b}_\mathrm{SSCOM}, \hat{b}_\mathrm{mPRM}, 
-S(0.2), S(0.5), | \, \theta)
+S(0.2), S(0.5), | \, \theta \right)
 $$
@@
 \noindent where $L$ is a generalized loss function, $f$ is some generalized
-regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from COM-SIR,
-$\hat{b}_\mathrm{COM-SIR}$ etc., $b$ is the true value the model is fitted to,
-$S(n)$ represents the spectral density at frequency $n$ of the scaled catch
-series, and $\theta$ is the set of parameters for the regression function.
+regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from the CMSY
+model etc., $b$ is the true value the model is fitted to, $S(n)$ represents the
+spectral density at frequency $n$ of the scaled catch series, and $\theta$ is
+the set of parameters for the regression function. For all ensemble models of
+mean \bbmsy\\, which is bounded at zero, we fit the models in log space and
+exponentiated the predictions to derive a median estimate of \bbmsy\\. For the
+estimates of \bbmsy\\ slope we fit ensemble models on the natural untransformed
+scale.
 
-<!--
-*Linear model*: For populations $i$ $1$ through $n$, individual models and
-covariates $j$ $1$ through $m$, and allowing $\theta$ to refer to both
-quantities estimated from individual models and additional covariates, we fit
-the linear model with two-way interactions as:
--->
+We calculated the average ensemble models as the geometric mean of the
+individual model median \bbmsy\\ estimates and the mean of the individual model
+\bbmsy\\ slopes. We fit the linear model ensemble with second-order
+interactions.
 
-<!--\noindent-->
-<!--The geometric mean is calculated as-->
-<!--$$B/B_{\mathrm{MSY-ensemble}(i)} = \exp \left(\frac{\sum_{j=1}^{4} \log B/B_{\mathrm{MSY}(ji)}}{4}\right)$$-->
-
-<!--$$-->
-<!--\log B/B_{\mathrm{MSY-ensemble}(i)} = -->
-  <!--\beta_0 + \beta_1 \log B/B_{\mathrm{MSY}(i)} -->
-<!--$$-->
-
-<!--
-\begin{equation}
-\theta_{\mathrm{ensemble}(i)} = 
-  \beta_0 + 
-  \sum_{j=1}^m \beta_j \theta_{j(i)} + 
-  \sum_{j=1}^m \sum_{\substack{k =1 \\ j \neq k}}^m 
-            \beta_{kj} \theta_{k(i)} \theta_{j(i)} +
-  \epsilon_i, \quad
-  \epsilon_i \sim \mathrm{Normal}(0, \sigma),
-\end{equation}
-
-\noindent
-where the $\beta$ represent coefficients and $\epsilon$ represents normally
-distributed residual variation. We also tried fitting linear models without
-interactions but found better cross-validation predictive performance from all
-performance measures evaluated using the procedure outlined below.
--->
-
-*Random forests* are a machine learning method that combines a series of
-regression trees. Regression trees sequentially determine what value of a
-predictor best splits the response data into two 'branches' based on a
-specified loss function. In random forests, a series of regression trees are
-built on a random subset of the data and random subset of the covariates of the
-model. By combining many of these stochastically generated models, random
-forests can provide strong predictive performance and fit highly non-linear
-relationships (REFs). We fit random forest models with the **randomForest**
-package [@liaw2002] for \textsf{R} with the default argument values.
-
-*Boosted regression models* are a machine learning method that, like random
-forests, combines a series of regression trees built on random subsets of the
-data, but also add a boosting procedure. Boosting refers to fitting a series of
-models with each subsequent model fit to the residuals from the previous model;
-data points that are fit poorly in a given model are given more weight in the
-next model [@elith2008]. Boosting can result in TODO and TODO and is a common
-and powerful machine learning technique [@elith2008]. We fit boosted regression
-tree models with the **GBM** package [@ridgeway2015] for \textsf{R}. Based on
+We fit two machine learning ensemble models based on regression trees.
+Regression trees sequentially determine what value of a predictor best splits
+the response data into two 'branches' based on a specified loss function. In
+random forests, a series of regression trees are built on a random subset of
+the data and random subset of the covariates of the model (REF). In generalized
+boosted models (GBMs), each subsequent model is fit to the residuals from the
+previous model; data points that are fit poorly in a given model are given more
+weight in the next model [@elith2008]. Random forests and GBMs can provide
+strong predictive performance and fit highly non-linear relationships (REFs).
+We fit random forest models with the **randomForest** package [@liaw2002] for
+\textsf{R} with the default argument values. We fit boosted regression tree
+models with the **GBM** package [@ridgeway2015] for \textsf{R}. Based on
 cross-validation with the **caret** package (Supporting Materials, Fig. SX), we
@@
 values.
-
-For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the
-models in log space and exponentiated the predictions to derive a median
-estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope, which are not
-bounded at zero, we fit ensemble models on the natural untransformed scale.
 

### 2015-06-09 analysis/6-sim-to-ram.R
Add lm coefs supp plot

@@ -63,6 +63,20 @@ m_lm <- lm(
       spec_freq_0.05 + spec_freq_0.2)^2,
   data = d_mean_sim)
 
+
+pdf("../figs/lm-coefs.pdf", width = 5, height = 6)
+par(mfrow = c(1, 1))
+par(mar = c(4, 12, 1, 3), cex = 0.9)
+par(xpd = FALSE)
+#arm::coefplot(arm::standardize(m_lm0), main = "")
+#par(xpd = NA)
+#mtext("(a) Standardized coefficients\n(no interactions)", side = 3, line = 2)
+#par(xpd = FALSE)
+arm::coefplot(arm::standardize(m_lm), main = "")
+par(xpd = NA)
+mtext("Standardized regression coefficient", side = 3, line = 3)
+dev.off()
+
 library("mgcv")
 m_gam <- mgcv::gam(
   log(bbmsy_true_mean) ~ s(CMSY) + s(COMSIR) + s(Costello) + s(SSCOM) +

### 2015-06-09 analysis/5-simulation-ensembles.R
Switch 'Costello' to 'mPRM' throughout
Haven't tried re-running this yet

@@ -8,6 +8,7 @@ dsim <- readRDS("generated-data/dsim.rds")
 spec <- readRDS("generated-data/spec-frequencies-sim.rds")
 dsim <- suppressWarnings(left_join(dsim, spec)) # warnings on character-factor conversions
 dsim$method_id <- sub("COM.SIR", "COMSIR", dsim$method_id) # to match RAM fits
+dsim$method_id <- sub("Costello", "mPRM", dsim$method_id)
 dsim <- rename(dsim, stockid = stock_id, method = method_id) # to match RAM fits
 # extend the stockid to be truly unique per scenario:
 # (note that there are still multiple iterations with the same stockid)

### 2015-06-09 analysis/6-sim-to-ram.R

@@ -7,6 +7,7 @@ source("0-ensemble-functions.R")
 
 # prep the RAM data:
 ram <- readRDS("generated-data/ram_fits.rds")
+ram$method <- sub("Costello", "mPRM", ram$method)
 
 ram_sum <- ram %>%
   group_by(stockid, method) %>%
@@ -48,18 +49,23 @@ d_slope <- reshape2::dcast(ram_sum, stockid + scientificname ~ method,
 # bring in the simulation formatted data to build the simulation-trained models:
 d_mean_sim <- readRDS("generated-data/sim-mean-dat.rds")
 m_rf <- randomForest::randomForest(
-  log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM +
+  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
     spec_freq_0.05 + spec_freq_0.2,
   data = d_mean_sim, ntree = 1000L)
 
 m_gbm <- gbm::gbm(
-  log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM +
+  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
     spec_freq_0.05 + spec_freq_0.2,
   data = d_mean_sim, distribution = "gaussian",
   n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
 
+m_lm0 <- lm(
+  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
+      spec_freq_0.05 + spec_freq_0.2),
+  data = d_mean_sim)
+
 m_lm <- lm(
-  log(bbmsy_true_mean) ~ (CMSY + COMSIR + Costello + SSCOM +
+  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
       spec_freq_0.05 + spec_freq_0.2)^2,
   data = d_mean_sim)
 
@@ -79,13 +85,13 @@ dev.off()
 
 library("mgcv")
 m_gam <- mgcv::gam(
-  log(bbmsy_true_mean) ~ s(CMSY) + s(COMSIR) + s(Costello) + s(SSCOM) +
[…]

### 2015-06-09 text/ms.Rmd
Add to methods and results

@@
 
-\begin{equation}
-\hat{B}_{t+1} = B_t + r B_t \left(1 - B_t / B_0 \right) - C_t,
-\end{equation}
+$$\hat{B}_{t+1} = B_t + r B_t \left(1 - B_t / B_0 \right) - C_t,$$
 
@@
 <!--TODO: what are we doing with the RAM stocks again?-->
+
+<!--TODO: build in fig:didactic -->
 
@@
 \hat{b}_\mathrm{ensemble} = 
-f \left( \hat{b}_\mathrm{CMSY}, \hat{b}_\mathrm{COM-SIR}, 
-\hat{b}_\mathrm{SSCOM}, \hat{b}_\mathrm{mPRM}, 
-S(0.2), S(0.5), | \, \theta \right)
+f \left( \hat{b}_\mathrm{CMSY}, \, \hat{b}_\mathrm{COM-SIR}, \,
+\hat{b}_\mathrm{SSCOM}, \, \hat{b}_\mathrm{mPRM}, \,
+S(0.2), \, S(0.5), \, | \, \theta \right)
 $$
@@
 regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from the CMSY
-model etc., $b$ is the true value the model is fitted to, $S(n)$ represents the
-spectral density at frequency $n$ of the scaled catch series, and $\theta$ is
-the set of parameters for the regression function. For all ensemble models of
-mean \bbmsy\\, which is bounded at zero, we fit the models in log space and
-exponentiated the predictions to derive a median estimate of \bbmsy\\. For the
-estimates of \bbmsy\\ slope we fit ensemble models on the natural untransformed
-scale.
+model etc. (Table \ref{tab:predictors}), $b$ is the true value the model is
+fitted to, $S(n)$ represents the spectral density at frequency $n$ of the
+scaled catch series, and $\theta$ is the set of parameters for the regression
+function. For all ensemble models of mean \bbmsy\\, which is bounded at zero,
+we fit the models in log space and exponentiated the predictions to derive a
+median estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope we fit
+ensemble models on the natural untransformed scale.
 
@@
 \bbmsy\\ slopes. We fit the linear model ensemble with second-order
-interactions.
-
-We fit two machine learning ensemble models based on regression trees.
-Regression trees sequentially determine what value of a predictor best splits
-the response data into two 'branches' based on a specified loss function. In
-random forests, a series of regression trees are built on a random subset of
-the data and random subset of the covariates of the model (REF). In generalized
-boosted models (GBMs), each subsequent model is fit to the residuals from the
-previous model; data points that are fit poorly in a given model are given more
-weight in the next model [@elith2008]. Random forests and GBMs can provide
-strong predictive performance and fit highly non-linear relationships (REFs).
-We fit random forest models with the **randomForest** package [@liaw2002] for
-\textsf{R} with the default argument values. We fit boosted regression tree
-models with the **GBM** package [@ridgeway2015] for \textsf{R}. Based on
-cross-validation with the **caret** package (Supporting Materials, Fig. SX), we
-fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate
-(shrinkage parameter) of $0.01$, and all other arguments at their default
-values.
+interactions. We fit two machine learning ensemble models based on regression
+trees. Regression trees sequentially determine what value of a predictor best
+splits the response data into two 'branches' based on a specified loss
+function. In random forests, a series of regression trees are built on a random
+subset of the data and random subset of the covariates of the model (REF). In
+generalized boosted models (GBMs), each subsequent model is fit to the
+residuals from the previous model; data points that are fit poorly in a given
+model are given more weight in the next model [@elith2008]. Random forests and
+GBMs can provide strong predictive performance and fit highly non-linear
+relationships (REFs). We fit random forest models with the **randomForest**
+package [@liaw2002] for \textsf{R} with the default argument values. We fit
+boosted regression tree models with the **GBM** package [@ridgeway2015] for
+\textsf{R}. Based on cross-validation with the **caret** package (Supporting
+Materials, Fig. SX), we fit GBMs with 2000 trees, an interaction depth of $6$,
+a learning rate (shrinkage parameter) of $0.01$, and all other arguments at
+their default values.
 
@@
 
-Individual model and ensemble model scatterplots:
+* Individual model and ensemble model scatterplots:
 Fig. \ref{fig:hexagon}
 
-Accuracy, rank correlation, and bias:
+* Accuracy, rank correlation, and bias:
 Fig. \ref{fig:performance}
+
+* Regression coefficients from linear model ensemble:
+Fig. \ref{fig:lm-coefs}
+
+* Partial dependence plots of GBMs:
+Fig. \ref{fig:partial-sim}
+
+* Partial residuals version of GBMs:
+Fig. \ref{fig:partial-residuals-sim-gbm}
+
+* 2-dimension partial dependence plots:
+Fig. \ref{fig:partial-2d-sim}
+
+* ROC curves from simulation ensembles:
+Fig. \ref{fig:roc-sim}
+
+* Scatterplots of RAM fits:
+Fig. \ref{fig:hexagon-ram}
+
+* Scatterplots of basic ensembles without covariates:
+Fig. \ref{fig:hexagon-sim-basic}
+
+* Scatterplots estimating slope:
+Fig. \ref{fig:scatter-sim-slope}
+
+* Accuracy, rank correlation, and bias of slope:
+Fig. \ref{fig:performance-sim-slope}
 

### 2015-06-09 text/ms.Rmd
Add to text and bibliography

@@
 potential solution to this problem. A simple ensemble model might take the
-average of multiple model outputs, perhaps weighted by some goodness of fit
-(REFs). A complex ensemble model might also take into account the interactions
-between model outputs and additional covariates. Ensemble models are popular in
-many fields including public health [e.g. @thomson2006], agriculture [e.g.
-@cantelaube2005], climate science (e.g. REFs), and machine learning (e.g.
-REFs). We think they have tremendous opportunity in ecological resource
-management.
+average of multiple model outputs, perhaps weighted by some goodness of fit. A
+complex ensemble model might also take into account the interactions between
+model outputs and additional covariates. Ensemble models are popular in other
+fields including public health [e.g. @thomson2006], agriculture [e.g.
+@cantelaube2005], climate science [e.g. @murphy2004; @tebaldi2007;
+@pierce2009], and machine learning [e.g. @dietterich2000]. We think they have
+tremendous opportunity in ecological resource management.
 
@@
 fish population. For the majority of fish stocks, we have limited biological
-information and stock status isn't known (e.g. REFs).  In recent years, a
-number of methods have been proposed to derive population status based on the
-limited information that is frequently available: fisheries catch time series
-and basic knowledge about the productivity of the species. Recent work has
+information and stock status isn't known [@fao2014]. In recent years, a number
+of methods have been proposed to derive population status based on the limited
+information that is frequently available: fisheries catch time series and basic
+knowledge about the productivity of the species. Recently, @rosenberg2014
 investigated the ability for these models to estimate population status through
-a large-scale simulation model [@rosenberg2014]. A key finding was that these
-models frequently disagree on population status and any one model has
-relatively poor predictive ability (e.g. Fig. \ref{fig:motivate}).
+a large-scale simulation model. A key finding was that these models frequently
+disagree on population status and any one model has relatively poor predictive
+ability (e.g. Fig. \ref{fig:motivate}).
 
@@
 regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from the CMSY
-model etc. (Table \ref{tab:predictors}), $b$ is the true value the model is
-fitted to, $S(n)$ represents the spectral density at frequency $n$ of the
-scaled catch series, and $\theta$ is the set of parameters for the regression
-function. For all ensemble models of mean \bbmsy\\, which is bounded at zero,
-we fit the models in log space and exponentiated the predictions to derive a
-median estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope we fit
-ensemble models on the natural untransformed scale.
+model etc. (Table \ref{tab:predictors}), $S(n)$ represents the spectral density
+at frequency $n$ of the scaled catch series, $b$ is the true value the model is
+fitted to, and $\theta$ is the set of parameters for the regression function.
+For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the
+models in log space and exponentiated the predictions to derive a median
+estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope we fit ensemble
+models on the natural untransformed scale.
 
@@
 
-Funding...
+<!--Funding...-->
 

### 2015-06-09 analysis/0-ensemble-functions.R
'Costello' -> 'mPRM'

@@ -27,9 +27,9 @@ mean_slope_bbmsy <- function(dat, years_window = 5L) {
 
 # A general function for cross-validation testing ensemble models:
 cross_val_ensembles <- function(.n, dat, fraction_train = 0.5,
-  gbm_formula = "log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM + LH",
+  gbm_formula = "log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM + LH",
   geo_mean = TRUE,
-  individual_models = c("CMSY", "COMSIR", "Costello", "SSCOM"),
+  individual_models = c("CMSY", "COMSIR", "mPRM", "SSCOM"),
   id = "mean", cache_folder = "generated-data/cv-sim/", distribution = "gaussian",
   lm_formula = "", glm_formula = "", nfold = 3L, weighted = FALSE) {
 

### 2015-06-09 analysis/5-simulation-ensembles.R

@@ -73,13 +73,13 @@ nvar <- 6L
 
 # run a model on all the data to generate data for partial dependence plots:
 # m_rf <- randomForest::randomForest(
-#   log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM + LH +
+#   log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM + LH +
 #   spec_freq_0.05 + spec_freq_0.2, data = d_mean)
 
 # best.iter <- gbm::gbm.perf(m_gbm1, method="cv", oobag.curve = FALSE)
 # print(best.iter)
 # library("caret")
-# x <- dplyr::select(d_mean, CMSY, COMSIR, Costello, SSCOM, spec_freq_0.05, spec_freq_0.2)
+# x <- dplyr::select(d_mean, CMSY, COMSIR, mPRM, SSCOM, spec_freq_0.05, spec_freq_0.2)
 # mm_gbm <- train(x = x , y = log(d_mean_sim$bbmsy_true_mean), method = "gbm",
 #   trControl = trainControl(method = "cv", number = 3, repeats = 1),
 #   tuneGrid =
@@ -101,13 +101,13 @@ nvar <- 6L
 # # try and avoid positive bias at low bbmsy with weights:
 # library(randomForest)
 # library(gbm)
-# m1 <- randomForest(log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM +
+# m1 <- randomForest(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
 #     spec_freq_0.05 + spec_freq_0.2, data = d_mean)
 # d_mean$rf1 <- exp(predict(m1, n.trees = 1000))
 # ggplot(d_mean, aes(bbmsy_true_mean, rf1)) + geom_point() +
 #   geom_abline(slope = 1, intercept = 0, col = "red")
 #
-# m2 <- gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + Costello + SSCOM +
+# m2 <- gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
 #     spec_freq_0.05 + spec_freq_0.2, data = d_mean, distribution = "gaussian",
 #   n.trees = 1000L, interaction.depth = 4, shrinkage = 0.01)
 # d_mean$gbm2 <- exp(predict(m2, n.trees = 1000))
[…]

### 2015-06-09 analysis/8-main-figs.R

@@ -7,7 +7,7 @@ source("0-ensemble-functions.R")
 d_sim <- readRDS("generated-data/cv_sim_long.rds")
 
 clean_names <- dplyr::data_frame(
-  method = c("CMSY", "COMSIR", "Costello", "SSCOM",
+  method = c("CMSY", "COMSIR", "mPRM", "SSCOM",
     "gbm_ensemble", "rf_ensemble", "lm_ensemble", "mean_ensemble"),
   clean_method = c("CMSY", "COM-SIR", "mPRM", "SSCOM",
     "GBM Ensemble", "RF Ensemble", "LM Ensemble", "Mean Ensemble"),

### 2015-06-09 analysis/8-main-figs.R
MARE -> MAPE

@@ -261,7 +261,7 @@ tick3 <- findInterval(0.5, pal_df$mre)
 text(rep(leg_x + 0.025, 3),
   blocks[c(tick1, 50, tick3)], labels = c("-0.5", "  0", "  0.5"), pos = 4,
   col = "grey50", cex = 0.9)
-text(leg_x - 0.01, 0.22, "Bias (MRE)", col = "grey20", pos = 4)
+text(leg_x - 0.01, 0.22, "Bias (MPE)", col = "grey20", pos = 4)
 
 segments(
   x0 = rep(leg_x + 0.025, 3),
@@ -272,7 +272,7 @@ segments(
 
 perf(re_ram_sum, label = "(b) Stock assessment database")
 
-mtext("Inaccuracy (MARE)", side = 1, outer = TRUE, line = 1.5, col = "grey20",
+mtext("Inaccuracy (MAPE)", side = 1, outer = TRUE, line = 1.5, col = "grey20",
   cex = 1)
 par(xpd = NA)
 mtext("Rank-order correlation", cex = 1, side = 2, outer = TRUE, line = -0.9,
@@ -316,8 +316,8 @@ tick3 <- findInterval(0.02, pal_df$mre)
 text(rep(leg_x + 0.002, 3),
   blocks[c(tick1, 50, tick3)], labels = c("-0.02", "  0", "  0.02"), pos = 4,
   col = "grey50", cex = 0.9)
-text(leg_x - 0.001, 0.22, "Bias (MRE)", col = "grey20", pos = 4)
-mtext("Inaccuracy (MARE)", side = 1, outer = TRUE, line = 1.5, col = "grey20",
+text(leg_x - 0.001, 0.22, "Bias (MPE)", col = "grey20", pos = 4)
+mtext("Inaccuracy (MAPE)", side = 1, outer = TRUE, line = 1.5, col = "grey20",
   cex = 1)
 par(xpd = NA)
 mtext("Rank-order correlation", cex = 1, side = 2, outer = TRUE, line = -0.9,

### 2015-06-09 analysis/8-main-figs.R
Delete old plot

@@ -198,21 +198,6 @@ pdf("../figs/hex-mean-sim-basic-cv.pdf", width = 8, height = 4)
 plot_hex_fig(d_sim_basic, xbins = 100L)
 dev.off()
 
-## Performance panel figures:
-pdf("../figs/performance.pdf", width = 6, height = 3.1)
-par(mfrow = c(1, 2), mgp = c(1.5, 0.4, 0), las = 1, tck = -0.012,
-  oma = c(2.7, 3.5, .5, .5), cex = 0.8, mar = c(0, 0, 0, 0),
-  xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
-par(xpd = FALSE)
-performance_panel(d_sim_perf_summ, xlim = c(0.2, 0.6), ylim = c(0.1, 0.7))
-performance_panel(re_ram_sum_long, xlim =  c(0.48, 0.65), ylim = c(0.15, 0.45))
-mtext("Within population inaccuracy (MAPE)", side = 1, line = 1.4, col = "grey40", cex = 0.8, outer = TRUE)
-mtext("Across population correlation", side = 2, line = 2.2, col = "grey40", las = 0, cex = 0.8, outer = TRUE)
-#text(0.54, 0.5, "Bias (MPE)", col = "grey30", pos = 4)
-dev.off()
-
-# base:
-
 all <- bind_rows(d_sim_perf_wide, re_ram_sum)
 pal <- RColorBrewer::brewer.pal(11, "RdBu") %>% colorRampPalette
 pal_df <- data_frame(mre = seq(-max(abs(all$mre)+0.05), max(abs(all$mre)+0.05), length.out = 100),

### 2015-06-09 analysis/6-sim-to-ram.R
Add 3-way interaction coefficient plot

@@ -69,16 +69,26 @@ m_lm <- lm(
       spec_freq_0.05 + spec_freq_0.2)^2,
   data = d_mean_sim)
 
+m_lm3 <- lm(
+  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
+      spec_freq_0.05 + spec_freq_0.2)^3,
+  data = d_mean_sim)
+
 
 pdf("../figs/lm-coefs.pdf", width = 5, height = 6)
 par(mfrow = c(1, 1))
 par(mar = c(4, 12, 1, 3), cex = 0.9)
 par(xpd = FALSE)
-#arm::coefplot(arm::standardize(m_lm0), main = "")
-#par(xpd = NA)
-#mtext("(a) Standardized coefficients\n(no interactions)", side = 3, line = 2)
-#par(xpd = FALSE)
 arm::coefplot(arm::standardize(m_lm), main = "")
+par(xpd = NA)
+mtext("Standardized regression coefficient", side = 3, line = 3)
+dev.off()
+
+pdf("../figs/lm-coefs-3.pdf", width = 5.5, height = 6)
+par(mfrow = c(1, 1))
+par(mar = c(4, 16, 1, 3), cex = 0.9)
+par(xpd = FALSE)
+arm::coefplot(arm::standardize(m_lm3), main = "")
 par(xpd = NA)
 mtext("Standardized regression coefficient", side = 3, line = 3)
 dev.off()

### 2015-06-09 analysis/5-simulation-ensembles.R
Trivial tweaks to scripts

@@ -234,7 +234,7 @@ eq <- paste0("CMSY + COMSIR + mPRM + ",
 eq_basic <- paste0("CMSY + COMSIR + mPRM + SSCOM")
 
 # work through cross validation of ensemble models:
-cv_sim_mean <- plyr::ldply(seq_len(4), .parallel = TRUE,
+cv_sim_mean <- plyr::ldply(seq_len(4), .parallel = FALSE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_mean, geo_mean = TRUE, id = "sim-mean",
       gbm_formula = paste0("log(bbmsy_true_mean) ~ ", eq),
@@ -256,14 +256,14 @@ cv_sim_mean_basic <- cv_sim_mean_basic %>% mutate(
   lm_ensemble  = exp(lm_ensemble))
 cv_sim_mean_basic$cv_id <- NULL
 
-cv_sim_slope <- plyr::ldply(seq_len(2), .parallel = TRUE,
+cv_sim_slope <- plyr::ldply(seq_len(4), .parallel = TRUE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_slope, geo_mean = FALSE, id = "sim-slope",
       gbm_formula = paste0("bbmsy_true_slope ~ ", eq),
       lm_formula = paste0("bbmsy_true_slope ~ (", eq, ")^2")))
 cv_sim_slope$cv_id <- NULL
 
-cv_sim_binary <- plyr::ldply(seq_len(2), .parallel = TRUE,
+cv_sim_binary <- plyr::ldply(seq_len(1), .parallel = TRUE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_mean,
       id = "sim-binary", distribution = "bernoulli",

### 2015-06-09 analysis/6-sim-to-ram.R

@@ -203,7 +203,7 @@ cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
 library("doParallel")
 registerDoParallel(cores = 2L)
 set.seed(123)
-qq <- plyr::ldply(seq_len(30L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
+qq <- plyr::ldply(seq_len(40L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
   .parallel = TRUE)
 
 d_mean_long <- qq %>%

### 2015-06-09 analysis/8-main-figs.R

@@ -235,16 +235,17 @@ perf(d_sim_perf_wide, label = "(a) Simulation")
 
 # colour legend:
 blocks <- seq(0, 0.2, length.out = 100)
-leg_x <- 0.25
+leg_x <- 0.26
 
 for(i in 1:99) {
   rect(leg_x, blocks[i], leg_x + 0.03, blocks[i+1]+0.001, border = NA, col = pal(100)[i])
 }
-tick1 <- findInterval(-0.5, pal_df$mre)
-tick3 <- findInterval(0.5, pal_df$mre)
+if(0.4 >= max(pal_df$mre)) stop("Colour ticks are too wide for the observed MPE.")
+tick1 <- findInterval(-0.4, pal_df$mre)
+tick3 <- findInterval(0.4, pal_df$mre)
 
 text(rep(leg_x + 0.025, 3),
-  blocks[c(tick1, 50, tick3)], labels = c("-0.5", "  0", "  0.5"), pos = 4,
+  blocks[c(tick1, 50, tick3)], labels = c("-0.4", "  0", "  0.4"), pos = 4,
   col = "grey50", cex = 0.9)
 text(leg_x - 0.01, 0.22, "Bias (MPE)", col = "grey20", pos = 4)
 

### 2015-06-09 text/ms.Rmd
Quick copy editing of ms; mostly spelling

@@
 output from ensemble models can provide better mechanistic understanding of how
-invidual models perform.
+individual models perform.
 
@@
 We first developed and tested ensemble models with a fully factorial simulated
-dataset [@rosenberg2014]. The simulation models included three life histories (small
-pelagic, demersal, large pelagic), three levels of how depleted biomass was at
-the start of the dataset (100%, 70%, and 40% of carrying capacity), four
-harvest dynamics (a constant harvest rate, a harvest rate that is coupled with
+dataset [@rosenberg2014]. The simulation models included three life histories: small
+pelagic, demersal, large pelagic; three levels of how depleted biomass was at
+the start of the dataset: 100%, 70%, and 40% of carrying capacity; and four
+harvest dynamics: a constant harvest rate, a harvest rate that is coupled with
 biomass to mimic an open-access single-species fishery, a scenario where
@@
 harvest rate increases, maintains a steady level and decreases to a lower
-steady level). Process noise was introduced to the models through two levels of
-multiplicative recruitment variability ($N(0, 0.2)$ and $N(0, 0.6)$, two levels
-of first-order autoregressive correlation on recruitment variablity (0 and
-0.6). The simulation included a scenario without observation error and with
+steady level. Process noise was introduced to the models through two levels of
+multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, two levels
+of first-order autoregressive correlation on recruitment variability: 0 and
+0.6. The simulation included a scenario without observation error and with
 multiplicative observation error around catch at $N(0, 0.2)$.
@@
 validation splits. Since the dynamics of simulated populations differing only
-in the magnitude of observation error around catch, magnitude of recuitment
+in the magnitude of observation error around catch, magnitude of recruitment
 variability, or stochastic draw of these values are likely similar, we grouped
@@
 of modelling goals. For continuous response variables such as the mean and
-slope of population status, peformance metrics often measure some form of bias,
+slope of population status, performance metrics often measure some form of bias,
 precision, accuracy (a combination of bias and precision), or the ability to

### 2015-06-09 analysis/0-ensemble-functions.R
Tweak hex-bin scatter plot after skype call

@@ -120,10 +120,10 @@ add_label <- function(xfrac, yfrac, label, pos = 4, add_bg = FALSE, ...) {
 
 plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   ylab = expression(widehat(B/B[MSY])), lims_hex = c(0, max(dat$bbmsy_est)),
-  xlim_plot = c(0, 3.9), ylim_plot = c(0, 3.2), axis_ticks = c(0, 1, 2, 3),
+  xlim_plot = c(0, 3.9), ylim_plot = c(0, 3.3), axis_ticks = c(0, 1, 2, 3),
   add_hex = TRUE, alpha = 30, xbins3 = xbins, lims_hex3 = lims_hex,
   count_transform = 1.5, count_transform3 = 30, oma = c(3.5, 3.5, .5, .5),
-  bias = c(rep(4, 12))) {
+  bias = c(rep(3, 12))) {
 
   rows <- max(dat$order) / 4
   par(mfrow = c(rows, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
@@ -139,6 +139,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   hexcol3 <- RColorBrewer::brewer.pal(9, "Greens")
   hexcol4 <- RColorBrewer::brewer.pal(9, "Purples")
   hexcol_ensemble <- RColorBrewer::brewer.pal(9, "Greys")
+  hexcol_ensemble <- RColorBrewer::brewer.pal(9, "YlGnBu")
 
   hexcol1 <- hexcol2
   hexcol3 <- hexcol2
@@ -154,19 +155,20 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   # hexcol3 <- c(paste0(brewer.pal(4, "Dark2")[3], "FF"), paste0(brewer.pal(4, "Dark2")[3], "00")) %>% rev
   # hexcol4 <- c(paste0(brewer.pal(4, "Dark2")[4], "FF"), paste0(brewer.pal(4, "Dark2")[4], "00")) %>% rev
 
-  hexcol1 <- brewer.pal(9, "BuGn")
-  hexcol2 <- brewer.pal(9, "BuPu")
-  hexcol3 <- brewer.pal(9, "GnBu")
[…]

### 2015-06-09 analysis/8-main-figs.R

@@ -164,7 +164,7 @@ ram_third_row_lims <- filter(d_ram, grepl("Ensemble", clean_method)) %>%
 # dev.off()
 
 # Same as above but add the RAM ensembles as a third row:
-pdf("../figs/hex-mean-sim-ram-cv.pdf", width = 6.9, height = 4.2)
+pdf("../figs/hex-mean-sim-ram-cv.pdf", width = 6.6, height = 4.2)
 plot_hex_fig(d_mean_sim_and_ram, xbins = 100L, xbins3 = 25L,
   lims_hex3 = ram_third_row_lims, count_transform3 = 18,
   oma = c(3.5, 3.5, .5, 2.5), count_transform = 1,

### 2015-06-09 analysis/plot-sims.tex
Add plots comparing differences between simulation scenarios

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
  lmargin = 2.54cm, rmargin = 2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\begin{document}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-catch-by-sigmaC-iter1.pdf}
\caption{Time series of catch split by measurement error on catch. (All
  figures are for iteration 1 and demersal life history.) Seems like a
  judgement call on whether these are too similar. Hard to say. See the next
  plots that explore how these affect the model estimates.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-CMSY-by-sigmaC-iter1.pdf}
\caption{Time series of CMSY B/BMSY split by measurement error on catch.
  These, for the most part, are pretty similar.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-COMSIR-by-sigmaC-iter1.pdf}
\caption{Time series of COMSIR B/BMSY split by measurement error on catch.
  Again, fairly similar, but not sure if too similar.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-SSCOM-by-sigmaC-iter1.pdf}
\caption{Time series of SSCOM B/BMSY split by measurement error on catch.
  SSCOM is more variable here than I thought it would be.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-catch-by-AR-iter1.pdf}
\caption{Time series of catch split by recruitment deviation autocorrelation.
These look reasonably different in most cases.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-catch-by-TS-iter1.pdf}
\caption{Time series of catch split by time series length. I.e., I
    thought these should match exactly. Not sure what's happening here,
    although my guess is that one or more of the stochastic draws is based on
  a different seed.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-CMSY-by-TS-iter1.pdf}
\caption{Time series of CMSY B/BMSY split by time series length. These seem to
be deviating pretty strongly from each other, as we might expect having played
with these models before.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=\textwidth]{../figs/sim-catch-by-sigmaR-iter1.pdf}
\caption{Time series of catch split by magnitude of recruitment variability.
  Again, tough call.}
\end{center}
\end{figure}

\end{document}

### 2015-06-09 analysis/7-plot-sims.R

@@ -1,0 +1,84 @@
+library("dplyr")
+library("ggplot2")
+
+dsim <- readRDS("generated-data/dsim.rds")
+dsim$method_id <- sub("COM.SIR", "COMSIR", dsim$method_id) # to match RAM fits
+dsim$method_id <- sub("Costello", "mPRM", dsim$method_id)
+dsim <- rename(dsim, stockid = stock_id, method = method_id) # to match RAM fits
+dsim <- dsim %>%
+  mutate(cv_id = paste0(stockid, "_lh_", LH)) %>%
+  mutate(stockid =
+      paste0(stockid, "_sigmaR_", sigmaR, "_lh_", LH)) %>%
+  arrange(stockid, iter, year) # critical since not all in order
+
+simple <- theme_bw() +
+    theme(panel.grid.major = element_blank(),
+      panel.grid.minor = element_blank()) +
+    theme(panel.grid.major = element_blank(),
+      panel.grid.minor = element_blank(),
+      panel.background = element_blank(),
+      strip.background = element_blank()) +
+    theme(strip.text.x = element_text(size = 5))
+
+p <- dsim %>% filter(iter == 1, LH == "DE", method == "CMSY") %>%
+  ggplot(aes(year, catch, colour = as.factor(sigmaC), group = sigmaC)) +
+    geom_line() +
+    facet_wrap(~stockid, scales = "free_y", ncol = 9) + simple
+ggsave("../figs/sim-catch-by-sigmaC-iter1.pdf", width = 15, height = 13)
+
+p <- dsim %>% filter(iter == 1, LH == "DE", method == "CMSY") %>%
+  ggplot(aes(year, b_bmsy_est, colour = as.factor(sigmaC), group = sigmaC)) +
+    geom_line() +
+    facet_wrap(~stockid, scales = "free_y", ncol = 9) + simple
+ggsave("../figs/sim-CMSY-by-sigmaC-iter1.pdf", width = 15, height = 13)
+
+p <- dsim %>% filter(iter == 1, LH == "DE", method == "SSCOM") %>%
+  ggplot(aes(year, b_bmsy_est, colour = as.factor(sigmaC), group = sigmaC)) +
+    geom_line() +
+    facet_wrap(~stockid, scales = "free_y", ncol = 9) + simple
[…]

### 2015-06-12 analysis/0-ensemble-functions.R
Tweak hex-bin plot

@@ -128,7 +128,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   rows <- max(dat$order) / 4
   par(mfrow = c(rows, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
     oma = oma, cex = 0.8, mar = c(0, 0, 0, 0),
-    xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
+    xaxs = "i", yaxs = "i", col.axis = "grey60", col.lab = "grey50")
   par(family="serif")
 
   hexcol1 <- RColorBrewer::brewer.pal(9, "Blues")
@@ -167,9 +167,9 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   # hexcol_ensemble <- RColorBrewer::brewer.pal(9, "Spectral") %>% rev
   #hexcol_ensemble <- RColorBrewer::brewer.pal(9, "Greys")
 
-  hexcol_third_row <- RColorBrewer::brewer.pal(9, "Greys")
-   #hexcol_ensemble <- RColorBrewer::brewer.pal(9, "Spectral") %>% rev
-  # hexcol_third_row <- RColorBrewer::brewer.pal(9, "RdPu")
+  #hexcol_third_row <- RColorBrewer::brewer.pal(9, "Greys")
+   #hexcol_third_row <- RColorBrewer::brewer.pal(9, "Spectral") %>% rev
+   hexcol_third_row <- RColorBrewer::brewer.pal(9, "YlGnBu")
 
   panels <- seq_len(length(unique(dat$order)))
 
@@ -206,7 +206,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
     }
     #alternative power transformation:
     #counts <- round((counts)^0.30)
-    space = "rgb"
+    space = "Lab"
     if (m %in% 1)
       pal_function <- colorRampPalette(hexcol1, space = space, bias = bias[1])
     if (m %in% 2)
@@ -219,8 +219,8 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
       pal_function <- colorRampPalette(hexcol_ensemble, space = space, bias = bias[5])
     if (m %in% 9:12) {
[…]

### 2015-06-12 analysis/5-simulation-ensembles.R
Add predictor vs. predicted plots for GBM simulation ensembles

@@ -133,6 +133,27 @@ nvar <- 6L
 m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
     spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
   data = d_mean, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
+d_mean$gbm_pred <- exp(gbm::predict.gbm(m, n.trees = 2000L))
+
+p1 <- ggplot(d_mean, aes(CMSY, gbm_pred)) + geom_point()
+p2 <- ggplot(d_mean, aes(COMSIR, gbm_pred)) + geom_point()
+p3 <- ggplot(d_mean, aes(mPRM, gbm_pred)) + geom_point()
+p4 <- ggplot(d_mean, aes(SSCOM, gbm_pred)) + geom_point()
+p5 <- ggplot(d_mean, aes(spec_freq_0.05, gbm_pred)) + geom_point()
+p6 <- ggplot(d_mean, aes(spec_freq_0.2, gbm_pred)) + geom_point()
+pdf("../figs/predictor-vs-predicted-gbm.pdf", width = 10, height = 5)
+gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 2)
+dev.off()
+
+p1 <- ggplot(d_mean, aes(CMSY, bbmsy_true_mean)) + geom_point()
+p2 <- ggplot(d_mean, aes(COMSIR, bbmsy_true_mean)) + geom_point()
+p3 <- ggplot(d_mean, aes(mPRM, bbmsy_true_mean)) + geom_point()
+p4 <- ggplot(d_mean, aes(SSCOM, bbmsy_true_mean)) + geom_point()
+p5 <- ggplot(d_mean, aes(spec_freq_0.05, bbmsy_true_mean)) + geom_point()
+p6 <- ggplot(d_mean, aes(spec_freq_0.2, bbmsy_true_mean)) + geom_point()
+pdf("../figs/predictor-vs-truth-gbm.pdf", width = 10, height = 5)
+gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 2)
+dev.off()
 
 make_partial_resid_gbm <- function(var = "SSCOM") {
   d_temp <- d_mean

### 2015-06-12 analysis/5-simulation-ensembles.R
Switch to cross-validate grouping on only iterations

@@ -13,9 +13,10 @@ dsim <- rename(dsim, stockid = stock_id, method = method_id) # to match RAM fits
 # extend the stockid to be truly unique per scenario:
 # (note that there are still multiple iterations with the same stockid)
 # cv_id defines groups that should *all* be in the testing or training
-# datasets
+# datasets (note that stockid and cv_id are now essentially the same
+# these remain due to legacy code)
 dsim <- dsim %>%
-  mutate(cv_id = paste0(stockid, "_lh_", LH)) %>%
+  mutate(cv_id = paste0(stockid, "_sigC_", sigmaC, "_sigR_", sigmaR)) %>%
   mutate(stockid =
       paste0(stockid, "_sigmaC_", sigmaC, "_sigmaR_", sigmaR, "_lh_", LH)) %>%
   arrange(stockid, iter, year) # critical since not all in order

### 2015-06-12 analysis/5-simulation-ensembles.R
Dress up 2d partial dependence plot

@@ -200,9 +200,15 @@ partial <- plyr::ldply(seq_len(nvar), function(i) {
   dd
 })
 
+partial_names <- data_frame(
+  predictor = c("CMSY", "COMSIR", "mPRM", "SSCOM",
+    "spec_freq_0.05", "spec_freq_0.2"),
+  predictor_label = c("(a) CMSY", "(b) COM-SIR", "(d) mPRM", "(c) SSCOM",
+    "(e) Long-term spectral density", "(d) Short-term spectral density"))
+partial <- inner_join(partial, partial_names)
 # partial dependence plot:
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() +
-  facet_wrap(~predictor, scales = "free_x") + ylim(0.3, 1.6) + theme_bw() +
+  facet_wrap(~predictor_label, scales = "free_x") + ylim(0.3, 1.6) + theme_bw() +
   ylab("Expected B/B_MSY integrating out other predictors") +
   xlab("Predictor value")
 ggsave("../figs/partial-sim.pdf", width = 7, height = 5)
@@ -219,21 +225,22 @@ partial_2d <- plyr::ldply(1:nvar, function(x) plyr::ldply(1:nvar, function(y) {
 }))
 
 # check colour pallete:
-zlim <- c(min(partial_2d$z)+0.1, max(partial_2d$z))
-pal <- rev(colorRampPalette(c("red", "white", "blue"))(13))[-c(1:1)]
-# white should line up with 1: (adjust the above line as needed)
+zlim <- c(min(partial_2d$z)+0.11, max(partial_2d$z))
+pal <- rev(colorRampPalette(c("red", "white", "blue"), space = "Lab")(13))[-c(1:1)]
+# white should line up with 1: (adjust the above 2 lines as needed)
 
 pdf("../figs/partial-2d-col-check.pdf", width = 6, height = 4)
 plot(seq(min(zlim), max(zlim), length.out = 12), 1:12, bg = pal, pch = 21, cex = 3);abline(v = 1)
 dev.off()
 
 ii <<- 0
+lab <<- 0
 panels <- matrix(NA, ncol = nvar, nrow = nvar, byrow = TRUE) %>% upper.tri %>%
   t %>% as.vector
 
 pdf("../figs/partial-sim-2d.pdf", width = 10, height = 8.5)
[…]

### 2015-06-12 analysis/6-sim-to-ram.R
Remove mPRM species categories until I get the proper data

@@ -1,5 +1,6 @@
 # apply the simulation-trained ensemble model to the RAM dataset
 
+set.seed(123)
 library("randomForest")
 library("dplyr")
 
@@ -109,6 +110,9 @@ ts_dat <- plyr::ddply(ts_dat, "stockid", function(x) {
   datalimited::format_prm(year = x$year, catch = x$catch, bbmsy = x$bbmsy_ram,
     species_cat = x$spp_category[1L])
 })
+# until I get the proper 3-level species categories:
+ts_dat$species_cat <- NULL
+ts_dat$species_cat <- "ignore_me"
 
 # Apply the simulation ensembles to a cross-validated version of the RAM dataset:
 cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
@@ -195,15 +199,18 @@ cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
       d_test$gam_ensemble <- exp(predict(m_gam, newdata = d_test))
       d_test$.n = .n # for identification purposes
 
+      if (.n %% 10 == 0) write(.n, file = paste0(.n, ".txt"))
+
       d_test
+    } else {
+      print("skipped iteration", .n)
     }
   })
 }
 
 library("doParallel")
 registerDoParallel(cores = 2L)
-set.seed(123)
-qq <- plyr::ldply(seq_len(40L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
+qq <- plyr::ldply(seq_len(80L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
   .parallel = TRUE)
 
 d_mean_long <- qq %>%

### 2015-06-12 analysis/8-main-figs.R
Adjust RAM ensemble hex-bin colour parameters

@@ -166,9 +166,9 @@ ram_third_row_lims <- filter(d_ram, grepl("Ensemble", clean_method)) %>%
 # Same as above but add the RAM ensembles as a third row:
 pdf("../figs/hex-mean-sim-ram-cv.pdf", width = 6.6, height = 4.2)
 plot_hex_fig(d_mean_sim_and_ram, xbins = 100L, xbins3 = 25L,
-  lims_hex3 = ram_third_row_lims, count_transform3 = 18,
+  lims_hex3 = ram_third_row_lims, count_transform3 = 12,
   oma = c(3.5, 3.5, .5, 2.5), count_transform = 1,
-  alpha = 15)
+  alpha = 15, bias = c(rep(3, 8), rep(1.8, 4)))
 par(xpd = NA)
 mtext("Simulated data", side = 4, outer = TRUE, line = 0.4, las = 0, col = "grey20",
   adj = 0.8)

### 2015-06-12 text/ms.Rmd
Add some draft result notes

@@
 disagree on population status and any one model has relatively poor predictive
-ability (e.g. Fig. \ref{fig:motivate}).
+ability on average (e.g. Fig. \ref{fig:motivate}).
 
@@
 We first developed and tested ensemble models with a fully factorial simulated
-dataset [@rosenberg2014]. The simulation models included three life histories: small
-pelagic, demersal, large pelagic; three levels of how depleted biomass was at
-the start of the dataset: 100%, 70%, and 40% of carrying capacity; and four
-harvest dynamics: a constant harvest rate, a harvest rate that is coupled with
-biomass to mimic an open-access single-species fishery, a scenario where
-harvest rate increases to a fixed level, a 'roller-coaster' scenario where the
-harvest rate increases, maintains a steady level and decreases to a lower
-steady level. Process noise was introduced to the models through two levels of
-multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, two levels
-of first-order autoregressive correlation on recruitment variability: 0 and
-0.6. The simulation included a scenario without observation error and with
+dataset [@rosenberg2014]. In summary, the simulation models included three life
+histories: small pelagic, demersal, large pelagic; three levels of how depleted
+biomass was at the start of the dataset compared to carrying capacity: 100%,
+70%, and 40% of carrying capacity; and four harvest dynamics: a constant
+harvest rate, a harvest rate that is coupled with biomass to mimic an
+open-access single-species fishery, a scenario where harvest rate increases to
+a fixed level, a 'roller-coaster' scenario where the harvest rate increases,
+maintains a steady level and decreases to a lower steady level. Process noise
+was introduced to the models through two levels of multiplicative recruitment
+variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either
+uncorrelated or had a first-order autoregressive correlation of 0.6. The
+simulation included a scenario without observation error and with
 multiplicative observation error around catch at $N(0, 0.2)$.
@@
 
-- *SSCOM* (state-space catch-only-model) ...
+- *SSCOM* (state-space catch-only-model) TODO: Jim -- can you give a ~2 sentence summary?
 
@@
 
-* Individual model and ensemble model scatterplots:
-Fig. \ref{fig:hexagon}
+CMSY has bimodality (Fig. \ref{fig:hexagon}a)
 
-* Accuracy, rank correlation, and bias:
-Fig. \ref{fig:performance}
+Applied to the simulated data (Fig. \ref{fig:hexagon}e--h, Fig. \ref{fig:performance}a)...
 
-* Regression coefficients from linear model ensemble:
-Fig. \ref{fig:lm-coefs}
+machine learning models and linear models outperformed all individual models in terms of accuracy (MAPE) and Spearman's rank-order correlation across stocks
 
-* Partial dependence plots of GBMs:
-Fig. \ref{fig:partial-sim}
+ensembles generally had lower bias (MPE) than any individual model (Fig. \ref{fig:performance})
 
-* Partial residuals version of GBMs:
-Fig. \ref{fig:partial-residuals-sim-gbm}
+ensemble models generally had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY}$ = 1$ (Fig. \ref{fig:roc-sim})
 
-* 2-dimension partial dependence plots:
-Fig. \ref{fig:partial-2d-sim}
+ensembles also did better at estimating the slope of \bbmsy\\ in the last five years (Figs. \ref{fig:scatter-sim-slope}, \ref{fig:performance-sim-slope}).
 
-* ROC curves from simulation ensembles:
-Fig. \ref{fig:roc-sim}
+for slope: again, machine learning ensembles best by rank-order correlation and accuracy and had lowest bias (numbers) (Fig. \ref{fig:performance-sim-slope})
 
-* Scatterplots of RAM fits:
-Fig. \ref{fig:hexagon-ram}
+Although the additional spectral density covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
+~\ref{fig:hexagon} vs.
+Fig.
+\ref{fig:hexagon-sim-basic})
 
-* Scatterplots of basic ensembles without covariates:
-Fig. \ref{fig:hexagon-sim-basic}
+For example, the linear model ensemble suggests that if both mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ is also high (Fig. \ref{fig:lm-coefs})
 
-* Scatterplots estimating slope:
-Fig. \ref{fig:scatter-sim-slope}
+The GBM model, also indicates a similar relationship, but suggests a non-linear relationship between these individual models in predicting \bbmsy\\ (Fig. \ref{fig:partial-2d-sim}XX)
 
-* Accuracy, rank correlation, and bias of slope:
-Fig. \ref{fig:performance-sim-slope}
+Partial dependence plots illustrate the non-linear relationship between the
+various individual model predictions and the ensemble predictions (Fig.
+\ref{fig:partial-sim}). These integrate across other predictor values...
+
+Ensemble models can exploit the best performance characteristics of individual models. 
+For example, if SSCOM predicts a low \bbmsy\\ (less than about 0.X), true \bbmsy\\ is probably low (Fig. \ref{fig:hexagon}c).
+But, SSCOM provides less predictive ability on its own when it predicts higher \bbmsy\\.
+The machine learning ensemble models can exploit this (Fig.
+\ref{fig:partial-sim}XX).
+On its own, CMSY contributes the most to predicting \bbmsy\\ when it makes predictions of \bbmsy\\ over one.
+
+Ensemble models trained on the simulated dataset, outperformed any individual model across all performance dimensions when applied to a completely new dataset: the RAM Legacy Stock Assessment database.
+They even outperform the mPRM method which is trained on the stock assessment database itself (Fig.
+\ref{fig:hexagon-ram})
+
+Ensembles take advantage of not just the best aspects of individual models, but also the interactions between multiple individual models or the interaction between individual models and other covariates.
+For example, CMSY becomes more informative about mean \bbmsy\\ when combined with the spectral density measures (Fig.
+\ref{fig:partial-2d-sim}XX, XX).
+
+Across simulated data mean \bbmsy\\, stock assessment database \bbmsy\\ and
+simulated data slope, the ensembles consistently had the best, or among the best levels of accuracy, rank order correlation, and level of bias (Figs \ref{fig:performance}, \ref{fig:performance-sim-slope}).
+
+Some individual models performed nearly as well as the enembles in one or two performance dimensions: e.g.
+SSCOM does nearly as well as the ensembles at predicting the slope of \bbmsy\\ (Fig.
+\ref{fig:performance-sim-slope}).
+But no individual method is consistently nearly as good as the ensembles.
+SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for both the simulated and stock-assessment datasets (Fig.
+\ref{fig:performance).
 

### 2015-06-12 text/ms.Rmd
Edit introduction and methods; add to results

@@
 
-We often have multiple models of ecological systems. For example, models might
-incorporate different data types, assume alternate population dynamics, or make
-contrasting assumptions about the starting state of a system. These models may
-suggest different conclusions about population status and often times we do not
-know which model is best. How can we reconcile multiple models to make robust
+There are often multiple ecological models we can apply to a system to assess
+population status and trajectory. Models might incorporate different data
+types, assume alternate population dynamics, or make contrasting assumptions
+about the starting state of a system. These models may suggest different
+conclusions about population status and trajectory and we often do not know
+which model is best. How can we reconcile multiple models to make robust
 management decisions about ecological resources?
 
-Ensemble models, which combine the output from multiple models, are one
-potential solution to this problem. A simple ensemble model might take the
-average of multiple model outputs, perhaps weighted by some goodness of fit. A
-complex ensemble model might also take into account the interactions between
-model outputs and additional covariates. Ensemble models are popular in other
-fields including public health [e.g. @thomson2006], agriculture [e.g.
-@cantelaube2005], climate science [e.g. @murphy2004; @tebaldi2007;
-@pierce2009], and machine learning [e.g. @dietterich2000]. We think they have
-tremendous opportunity in ecological resource management.
+Ensemble methods, which combine the output from multiple models as predictors
+in a new 'ensemble' model, are one solution to this problem. A simple ensemble
+model might take the average of multiple model outputs, perhaps weighted by
+some goodness of fit. A complex ensemble model might also take into account the
+interactions between model outputs and additional covariates. Ensemble models
+are popular in other fields including public health [e.g. @thomson2006],
+agriculture [e.g. @cantelaube2005], climate science [e.g. @murphy2004;
+@tebaldi2007; @pierce2009], and machine learning [e.g. @dietterich2000]. We
+think they have tremendous opportunity in ecological resource management.
 
-In fisheries science, a common problem is estimating the status of an exploited
-fish population. For the majority of fish stocks, we have limited biological
-information and stock status isn't known [@fao2014]. In recent years, a number
-of methods have been proposed to derive population status based on the limited
-information that is frequently available: fisheries catch time series and basic
-knowledge about the productivity of the species. Recently, @rosenberg2014
-investigated the ability for these models to estimate population status through
-a large-scale simulation model. A key finding was that these models frequently
-disagree on population status and any one model has relatively poor predictive
-ability on average (e.g. Fig. \ref{fig:motivate}).
+In fisheries science, a common problem is estimating the status and trajectory
+of an exploited fish population. For the majority of fish stocks, we have
+limited biological information and stock status isn't known [@fao2014]. In
+recent years, a number of methods have been proposed to derive population
+status based on the limited information that is broadly available: fisheries
+catch time series and basic knowledge about the productivity of the species.
+Recently, @rosenberg2014 investigated the ability for these models to estimate
+population status through a large-scale simulation model. A key finding was
+that these models frequently disagree on population status and any one model
+has relatively poor predictive ability on average (e.g. Fig.
+\ref{fig:motivate}).
 
@@
 
-Here, we develop ensemble models to estimate the population status of exploited
-fish populations. We explore a variety of ensemble approaches applied to both
-simulated and real-world fish stocks and compare predictive performance. In
-developing these models we highlight the types of problems one encounters when
-forming and evaluating ensemble models. We find that ensembles generally
-provide improved predictive performance than any single model---especially when
-considered across multiple dimensions of predictive ability. Furthermore, the
-output from ensemble models can provide better mechanistic understanding of how
-individual models perform.
+Here, we develop ensemble models to estimate the population status and
+trajectory of exploited fish populations. We explore a variety of ensemble
+approaches applied to both simulated and real-world fish stocks and compare
+predictive performance. In developing these models we highlight the types of
+challenges one encounters when forming and evaluating ensemble models. We find
+that ensembles generally provide improved predictive performance than any
+single model---especially when considered across multiple dimensions of
+predictive ability and across multiple datasets. Furthermore, the output from
+ensemble models can provide better mechanistic understanding of how individual
+models of status and trajectory perform.
 
@@
 We tested the ability of ensemble models to improve estimates of population
-status when applied to both a large simulated dataset and a global database of
-assessed stock status. Here we describe the datasets, individual models of
-population status, and ensemble models to combine those estimates. We then
-describe how we evaluated the ability of the various models to estimate
-population status.
+status and trajectory when applied to both a large simulated dataset and a
+global database of assessed stock status. Here we describe the datasets,
+individual models of population status, and ensemble models to combine those
+estimates. We then describe how we evaluated the ability of the various models
+to estimate population status.
 
@@
 
-- *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
-dynamics [@martell2013]. It requires a prior distribution for $r$ and assigns
-a prior to the relative proportion of biomass at the end compared to unfished
-biomass (depletion) based on the percentage of maximum catch at the end of the
-time series.
+- *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
+  biomass dynamics [@martell2013]. It requires a prior distribution for $r$ and
+  assigns a prior to the relative proportion of biomass at the end compared to
+  unfished biomass (depletion) based on the percentage of maximum catch at the
+  end of the time series.
 
@@
   panel-regression model used in @costello2012. Unlike the other models
-considered, mPRM is empirical and not mechanistic --- it fits a linear model to
-a series of characteristics of the catch time series and stock with \bbmsy\\ as
-the response. The model used in this paper and @rosenberg2014 is modified from
-the original in that it simplifies the life-history categories to be applicable
-to simulated data.
+  considered, mPRM is empirical and not mechanistic --- it uses the RAM Legacy
+  Stock Assessment database to fit a linear model to a series of
+  characteristics of the catch time series and stock with stock-assessed
+  \bbmsy\\ as the response. The model used in this paper and @rosenberg2014 is
+  modified from the original in that it condenses the life-history categories
+  into three categories to match the simulated dataset. 
   <!--TODO: what are we doing with the RAM stocks again?-->
-
+  <!--TODO: bias adjustment?-->
   <!--TODO: build in fig:didactic -->
@@
 Ensemble methods allow us to incorporate additional covariates into our models
-and potentially leverage interactions of how these covariates combine with the
-main individual model estimates to estimate population status. Examples might
-be life-history characteristics, additional information on exploitation
-patterns, or additional statistical properties of the data going into the
-individual models. For simplicity, and to allow us to apply models developed
-with the simulated dataset to the real-world dataset, we added only one set of
-additional covariates: spectral properties of the catch time series itself.
+and potentially leverage interactions between these covariates and individual
+model predictions. Additional covariates could be, for example, life-history
+characteristics, information on exploitation patterns, or statistical
+properties of the data. For simplicity, and to allow us to apply models
+developed with the simulated dataset to the real-world dataset, we added only
+one set of additional covariates: spectral properties of the catch time series.
 Spectral analysis decomposes a time series into the frequency domain and
-provides a useful means of describing the statistical properties of the catch
-series that is independent of time series length and magnitude of catch. We fit
-spectral models to the scaled catch time series (divided by maximum catch) with
-the `spec.ar` function in \textsf{R} and recorded representative short- and
-long-term spectral densities of 0.05 and 0.2, which correspond to 20- and
-5-year cycles.
+provides a means of describing the statistical properties of the catch series
+that is independent of time series length and magnitude of catch. We fit
+spectral models to the scaled catch time series (catch divided by maximum
+catch) with the `spec.ar` function in \textsf{R} and recorded representative
+short- and long-term spectral densities of 0.05 and 0.2, which correspond to
+20- and 5-year cycles.
 
@@
 The individual models we seek to combine with ensemble models provide time
-series of \bbmsy\\ ($\hat{b}_\mathrm{CMSY}$, $\hat{b}_\mathrm{COM-SIR}$, etc.).
-Therefore, we can use ensemble models to estimate multiple properties of these
-status time series. Here, we focus on two values: the mean and slope of
-\bbmsy\\ in the last 5 years. Below, we refer to these quantities collectively
-as $\hat{b}_\mathrm{ensemble}$. Together, these quantities address the current
-state and trajectory of status, both of which may be of management and
-conservation interest. To avoid undue influence of the end points of the time
-series on the calculated slope, we measured the slope as the Theil-Sen
+series of \bbmsy\\. Therefore, we can use ensemble models to estimate multiple
+properties of these status time series. Here, we focus on two values: the mean
+and slope of \bbmsy\\ in the last 5 years. Together, these quantities address
+the recent state and trajectory of status, both of which may be of management
+and conservation interest. To avoid undue influence of the end points of the
+time series on the calculated slope, we measured the slope as the Theil-Sen
 estimator of median slope [@theil1950].
@@
 \hat{\theta} = 
-\mathrm{argmin}_\theta \left( L(b \, | \, \hat{b}_{\mathrm{ensemble}}) \right)
+\argmin_\theta \left( L(b \, | \, \hat{b}_{\mathrm{ensemble}}) \right)
 $$
@@
 regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from the CMSY
-model etc. (Table \ref{tab:predictors}), $S(n)$ represents the spectral density
-at frequency $n$ of the scaled catch series, $b$ is the true value the model is
-fitted to, and $\theta$ is the set of parameters for the regression function.
-For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the
-models in log space and exponentiated the predictions to derive a median
-estimate of \bbmsy\\. For the estimates of \bbmsy\\ slope we fit ensemble
-models on the natural untransformed scale.
+model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
+spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$
+is the true value the model is fitted to, and $\theta$ is the set of parameters
+for the regression function. For all ensemble models of mean \bbmsy\\, which is
+bounded at zero, we fit the models in log space and exponentiated the
+predictions to derive a median estimate of \bbmsy\\. For the estimates of
+\bbmsy\\ slope, we fit ensemble models on the natural untransformed scale.
 
 We calculated the average ensemble models as the geometric mean of the
-individual model median \bbmsy\\ estimates and the mean of the individual model
+individual model \bbmsy\\ estimates and the mean of the individual model
 \bbmsy\\ slopes. We fit the linear model ensemble with second-order
@@
 GBMs can provide strong predictive performance and fit highly non-linear
-relationships (REFs). We fit random forest models with the **randomForest**
-package [@liaw2002] for \textsf{R} with the default argument values. We fit
-boosted regression tree models with the **GBM** package [@ridgeway2015] for
-\textsf{R}. Based on cross-validation with the **caret** package (Supporting
-Materials, Fig. SX), we fit GBMs with 2000 trees, an interaction depth of $6$,
-a learning rate (shrinkage parameter) of $0.01$, and all other arguments at
-their default values.
+relationships [e.g. @elith2008; @hastie2009]. We fit random forest models with
+the **randomForest** package [@liaw2002] for \textsf{R} with the default
+argument values. We fit boosted regression tree models with the **GBM** package
+[@ridgeway2015] for \textsf{R}. Based on cross-validation with the **caret**
+\textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a
+learning rate (shrinkage parameter) of $0.01$, and all other arguments at their
+default values.
 
@@
 A critical component to any predictive modelling exercise is to evaluate the
-performance of a model new data [@hastie2009]. We used repeated three-fold
+performance of a model on new data [@hastie2009]. We used repeated three-fold
 cross validation to test predictive performance: we randomly divided the
 dataset into three sets, built ensemble models on two-thirds of the data, and
-evaluated predictive performance on the remaining third of the data. We
-repeated this across each of the three splits and then repeated the whole
-procedure XX times to account for bias that may result from any one set of
-validation splits. Since the dynamics of simulated populations differing only
-in the magnitude of observation error around catch, magnitude of recruitment
-variability, or stochastic draw of these values are likely similar, we grouped
-these characteristics together in the cross-validation process---we never
-trained and tested on models differing only in these characteristics.
+evaluated predictive performance on the remaining third. We repeated this
+across each of the three splits and then repeated the whole procedure 50 times
+to account for bias that may result from any one set of validation splits.
+Since the dynamics of simulated populations differing only in the stochastic
+draw of random values are likely similar, we grouped these stocks (10 per
+factorial cell) in the cross-validation process into either the training or
+testing split.
 
@@
 precision, accuracy (a combination of bias and precision), or the ability to
-correctly rank or correlate across populations. Here, we measure proportional
+correctly rank or correlate across populations (REFs). Here, we measure proportional
 error (PE; also called 'relative error') defined as $(\hat{b} - b)/b$. We
@@
 
-Applied to the simulated data (Fig. \ref{fig:hexagon}e--h, Fig. \ref{fig:performance}a)...
+Applied to the simulated data (Fig. \ref{fig:hexagon}e--h, Fig.
+\ref{fig:performance}a)...
 
-machine learning models and linear models outperformed all individual models in terms of accuracy (MAPE) and Spearman's rank-order correlation across stocks
+non-linear machine learning models were best able to predict higher \bbmsy\\ values (Fig.
+\ref{fig:hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
+\ref{fig:partial-2d-sim})
 
-ensembles generally had lower bias (MPE) than any individual model (Fig. \ref{fig:performance})
+machine learning and linear ensemble models outperformed all individual models
+in terms of accuracy (MAPE) and Spearman's rank-order correlation across stocks
+at predicting mean \bbmsy\\ with the simulated dataset of known status (Fig.
+\ref{fig:performance}a)
 
-ensemble models generally had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY}$ = 1$ (Fig. \ref{fig:roc-sim})
+the level of correlation across stocks and accuracy were both lower when
+predicting full-stock assessement \bbmsy\\ than when predicting true status in
+the simulated dataset (Fig. \ref{fig:performance}a vs. b)
 
-ensembles also did better at estimating the slope of \bbmsy\\ in the last five years (Figs. \ref{fig:scatter-sim-slope}, \ref{fig:performance-sim-slope}).
+ensembles generally had lower bias (MPE) than any individual model (Fig.
+\ref{fig:performance})
 
-for slope: again, machine learning ensembles best by rank-order correlation and accuracy and had lowest bias (numbers) (Fig. \ref{fig:performance-sim-slope})
+ensemble models generally had better ability to distinguish if simulated stocks
+were above or below $B/B_\mathrm{MSY}$ = 1$ (Fig. \ref{fig:roc-sim})
 
-Although the additional spectral density covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
-~\ref{fig:hexagon} vs.
-Fig.
-\ref{fig:hexagon-sim-basic})
+ensembles also did better at estimating the slope of \bbmsy\\ in the last five
+years (Figs. \ref{fig:scatter-sim-slope}, \ref{fig:performance-sim-slope}).
 
-For example, the linear model ensemble suggests that if both mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ is also high (Fig. \ref{fig:lm-coefs})
+for slope: again, machine learning ensembles best by rank-order correlation and
+accuracy and had lowest bias (numbers) (Fig. \ref{fig:performance-sim-slope})
 
-The GBM model, also indicates a similar relationship, but suggests a non-linear relationship between these individual models in predicting \bbmsy\\ (Fig. \ref{fig:partial-2d-sim}XX)
+Although the additional spectral density covariates improved ensemble fit,
+performance of the ensembles was only marginally degraded by removing these
+covariates (Fig. ~\ref{fig:hexagon} vs. Fig. \ref{fig:hexagon-sim-basic})
+
+For example, the linear model ensemble suggests that if both mPRM and SSCOM
+predict high \bbmsy\\, the true \bbmsy\\ is also high (Fig. \ref{fig:lm-coefs})
+The GBM model, also indicates a similar relationship, but suggests a non-linear
+relationship between these individual models in predicting \bbmsy\\ (Fig.
+\ref{fig:partial-2d-sim}l)
 
@@
 
-Ensemble models can exploit the best performance characteristics of individual models. 
-For example, if SSCOM predicts a low \bbmsy\\ (less than about 0.X), true \bbmsy\\ is probably low (Fig. \ref{fig:hexagon}c).
-But, SSCOM provides less predictive ability on its own when it predicts higher \bbmsy\\.
-The machine learning ensemble models can exploit this (Fig.
-\ref{fig:partial-sim}XX).
-On its own, CMSY contributes the most to predicting \bbmsy\\ when it makes predictions of \bbmsy\\ over one.
+Ensemble models can exploit the best performance characteristics of individual
+models. For example, if SSCOM predicts a low \bbmsy\\ (less than about 0.X),
+true \bbmsy\\ is probably low (Fig. \ref{fig:hexagon}c). But, SSCOM provides
+less predictive ability on its own when it predicts higher \bbmsy\\. The
+machine learning ensemble models can exploit this (Fig.
+\ref{fig:partial-sim}c). On its own, CMSY contributes the most to predicting
+\bbmsy\\ when it makes predictions of \bbmsy\\ between about 1.0 and 1.4 (Fig.
+\ref{fig:partial-sim}a)
 
-Ensemble models trained on the simulated dataset, outperformed any individual model across all performance dimensions when applied to a completely new dataset: the RAM Legacy Stock Assessment database.
-They even outperform the mPRM method which is trained on the stock assessment database itself (Fig.
+Ensemble models trained on the simulated dataset, outperformed any individual
+model across all performance dimensions when applied to a completely new
+dataset: the RAM Legacy Stock Assessment database. They even outperform the
+mPRM method which is trained on the stock assessment database itself (Fig.
 \ref{fig:hexagon-ram})
 
-Ensembles take advantage of not just the best aspects of individual models, but also the interactions between multiple individual models or the interaction between individual models and other covariates.
-For example, CMSY becomes more informative about mean \bbmsy\\ when combined with the spectral density measures (Fig.
-\ref{fig:partial-2d-sim}XX, XX).
+Ensembles take advantage of not just the best aspects of individual models, but
+also the interactions between multiple individual models or the interaction
+between individual models and other covariates. For example, CMSY becomes more
+informative about mean \bbmsy\\ when combined with the spectral density
+measures (Fig. \ref{fig:partial-2d-sim}XX, XX).
 
 Across simulated data mean \bbmsy\\, stock assessment database \bbmsy\\ and
-simulated data slope, the ensembles consistently had the best, or among the best levels of accuracy, rank order correlation, and level of bias (Figs \ref{fig:performance}, \ref{fig:performance-sim-slope}).
+simulated data slope, the ensembles consistently had the best, or among the
+best levels of accuracy, rank order correlation, and level of bias (Figs
+\ref{fig:performance}, \ref{fig:performance-sim-slope}).
 
-Some individual models performed nearly as well as the enembles in one or two performance dimensions: e.g.
-SSCOM does nearly as well as the ensembles at predicting the slope of \bbmsy\\ (Fig.
-\ref{fig:performance-sim-slope}).
-But no individual method is consistently nearly as good as the ensembles.
-SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for both the simulated and stock-assessment datasets (Fig.
-\ref{fig:performance).
+Some individual models performed nearly as well as the ensembles in one or two
+performance dimensions: e.g. SSCOM does nearly as well as the ensembles at
+predicting the slope of \bbmsy\\ (Fig. \ref{fig:performance-sim-slope}). But no
+individual method is consistently nearly as good as the ensembles. SSCOM does
+more poorly than the ensembles and other individual models in terms of
+accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for
+both the simulated and stock-assessment datasets (Fig. \ref{fig:performance).
 

### 2015-06-12 analysis/9-output-values.R
Out put more digits for slope MRE

@@ -3,7 +3,7 @@
 # `../text/values.rda` and are input through knitr in `../text/ms.Rmd`.
 
 get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
-  collapse = "--") {
+  collapse = "--", mre_digits = 2) {
 
   dat <- dat %>%
     mutate(ensemble = ifelse(grepl("Ensemble", clean_method), TRUE, FALSE)) %>%
@@ -31,9 +31,9 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   x$ind_mare_range <- round(range(ind$mare), digits_raw)
   x$mach_mare_range <- round(range(mach$mare), digits_raw)
 
-  x$ensemble_mre_range <- ensemble$mre %>% range %>% round(digits_raw)
-  x$mach_mre_range <- mach$mre %>% range %>% round(digits_raw)
-  x$ind_mre_range <- ind$mre %>% range %>% round(digits_raw)
+  x$ensemble_mre_range <- ensemble$mre %>% range %>% round(mre_digits)
+  x$mach_mre_range <- mach$mre %>% range %>% round(mre_digits)
+  x$ind_mre_range <- ind$mre %>% range %>% round(mre_digits)
 
   lapply(x, paste, collapse = collapse)
 }
@@ -45,7 +45,7 @@ re_ram_sum <- readRDS("generated-data/re_ram_sum.rds")
 mean_ram <- get_performance_stats(re_ram_sum)
 
 d_slope_error <- readRDS("generated-data/d_slope_error.rds")
-slope_sim <- get_performance_stats(d_slope_error)
+slope_sim <- get_performance_stats(d_slope_error, mre_digits = 3L)
 
 auc_sim_mean <- readRDS("generated-data/auc_sim_mean.rds")
 auc_sim <- list()

### 2015-06-12 text/ms.Rmd
First draft of results!

@@
 
-CMSY has bimodality (Fig. \ref{fig:hexagon}a)
+Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean and slope of \bbmsy\\ in the last five years (hereafter used interchangeably with 'status' and 'trajectory').
+All models exhibted a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{fig:hexagon}). 
+CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig. \ref{fig:performance}a).
+COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig. \ref{fig:hexagon}b, c). 
+mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{fig:hexagon}d). 
+There was generally little correlation between true and predicted trajectory for any of the individual models besides SSCOM (Figs \ref{fig:scatter-sim-slope}a--d, \ref{fig:performance-sim-slope}).
 
-Applied to the simulated data (Fig. \ref{fig:hexagon}e--h, Fig.
-\ref{fig:performance}a)...
+Ensemble models, and in particular the machine learning ensemble models (random forest and GBM), generally improved estimates of status and trajectory (Fig.
+\ref{fig:hexagon}e--h, Fig.
+\ref{fig:scatter-sim-slope}e--h).
+Compared to the individual models, machine learning ensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Figs \ref{fig:performance}a).
+These ensembles also had better ability to distinguish if simulated stocks
+were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{fig:roc-sim}).
+Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
+\ref{fig:performance-sim-slope}).
+The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating status (CMSY; Fig.
+\ref{fig:performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting slope (SSCOM; Fig.
+\ref{fig:performance-sim-slope}).
 
-non-linear machine learning models were best able to predict higher \bbmsy\\ values (Fig.
+The ensemble models were able to improve the predictions of status by exploiting the best properties of individual models, the covariance between individual models, and interactions with other covariates.
+For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (less than about 0.XX; Fig.
+\ref{fig:hexagon}c, Fig.
+\ref{fig:partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high bimodal prediction clusters (Fig.
+\ref{fig:partial-sim}).
+Ensembles also exploit the covariance between individual model predictions.
+For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ also tends to be high (Figs \ref{fig:lm-coefs}, \ref{fig:partial-2d-sim}l).
+The addition of spectral density covariates helped the ensemble models correctly predict higher \bbmsy\\ values (Fig.
 \ref{fig:hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
-\ref{fig:partial-2d-sim})
+\ref{fig:partial-2d-sim}).
+Although the additional covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
+\ref{fig:hexagon} vs.
+Fig.
+\ref{fig:hexagon-sim-basic}) TODO NEW FIG THAT QUANTIFIES THIS.
 
-machine learning and linear ensemble models outperformed all individual models
-in terms of accuracy (MAPE) and Spearman's rank-order correlation across stocks
-at predicting mean \bbmsy\\ with the simulated dataset of known status (Fig.
-\ref{fig:performance}a)
+When applied to the stock assessment database, the ensemble models---trained exclusively on the simulated dataset---generally performed as well or better than the best individual models.
+The mean, random forest, and GBM ensembles even outperformed the
+mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig. \ref{fig:performance}b, Fig. \ref{fig:hexagon-ram}).
+Compared to the individual models, the machine learning ensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
 
-the level of correlation across stocks and accuracy were both lower when
-predicting full-stock assessement \bbmsy\\ than when predicting true status in
-the simulated dataset (Fig. \ref{fig:performance}a vs. b)
-
-ensembles generally had lower bias (MPE) than any individual model (Fig.
-\ref{fig:performance})
-
-ensemble models generally had better ability to distinguish if simulated stocks
-were above or below $B/B_\mathrm{MSY}$ = 1$ (Fig. \ref{fig:roc-sim})
-
-ensembles also did better at estimating the slope of \bbmsy\\ in the last five
-years (Figs. \ref{fig:scatter-sim-slope}, \ref{fig:performance-sim-slope}).
-
-for slope: again, machine learning ensembles best by rank-order correlation and
-accuracy and had lowest bias (numbers) (Fig. \ref{fig:performance-sim-slope})
-
-Although the additional spectral density covariates improved ensemble fit,
-performance of the ensembles was only marginally degraded by removing these
-covariates (Fig. ~\ref{fig:hexagon} vs. Fig. \ref{fig:hexagon-sim-basic})
-
-For example, the linear model ensemble suggests that if both mPRM and SSCOM
-predict high \bbmsy\\, the true \bbmsy\\ is also high (Fig. \ref{fig:lm-coefs})
-The GBM model, also indicates a similar relationship, but suggests a non-linear
-relationship between these individual models in predicting \bbmsy\\ (Fig.
-\ref{fig:partial-2d-sim}l)
-
-Partial dependence plots illustrate the non-linear relationship between the
-various individual model predictions and the ensemble predictions (Fig.
-\ref{fig:partial-sim}). These integrate across other predictor values...
-
-Ensemble models can exploit the best performance characteristics of individual
-models. For example, if SSCOM predicts a low \bbmsy\\ (less than about 0.X),
-true \bbmsy\\ is probably low (Fig. \ref{fig:hexagon}c). But, SSCOM provides
-less predictive ability on its own when it predicts higher \bbmsy\\. The
-machine learning ensemble models can exploit this (Fig.
-\ref{fig:partial-sim}c). On its own, CMSY contributes the most to predicting
-\bbmsy\\ when it makes predictions of \bbmsy\\ between about 1.0 and 1.4 (Fig.
-\ref{fig:partial-sim}a)
-
-Ensemble models trained on the simulated dataset, outperformed any individual
-model across all performance dimensions when applied to a completely new
-dataset: the RAM Legacy Stock Assessment database. They even outperform the
-mPRM method which is trained on the stock assessment database itself (Fig.
-\ref{fig:hexagon-ram})
-
-Ensembles take advantage of not just the best aspects of individual models, but
-also the interactions between multiple individual models or the interaction
-between individual models and other covariates. For example, CMSY becomes more
-informative about mean \bbmsy\\ when combined with the spectral density
-measures (Fig. \ref{fig:partial-2d-sim}XX, XX).
-
-Across simulated data mean \bbmsy\\, stock assessment database \bbmsy\\ and
-simulated data slope, the ensembles consistently had the best, or among the
-best levels of accuracy, rank order correlation, and level of bias (Figs
-\ref{fig:performance}, \ref{fig:performance-sim-slope}).
-
+<!--
 Some individual models performed nearly as well as the ensembles in one or two
@@
 accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for
-both the simulated and stock-assessment datasets (Fig. \ref{fig:performance).
+both the simulated and stock-assessment datasets (Fig. \ref{fig:performance}).
+-->
 

### 2015-06-12 text/ms.Rmd
Remove fig: from all labels for brevity

@@
 has relatively poor predictive ability on average (e.g. Fig.
-\ref{fig:motivate}).
+\ref{motivate}).
 
@@
 Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean and slope of \bbmsy\\ in the last five years (hereafter used interchangeably with 'status' and 'trajectory').
-All models exhibted a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{fig:hexagon}). 
-CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig. \ref{fig:performance}a).
-COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig. \ref{fig:hexagon}b, c). 
-mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{fig:hexagon}d). 
-There was generally little correlation between true and predicted trajectory for any of the individual models besides SSCOM (Figs \ref{fig:scatter-sim-slope}a--d, \ref{fig:performance-sim-slope}).
+All models exhibted a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). 
+CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig. \ref{performance}a).
+COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig. \ref{hexagon}b, c). 
+mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{hexagon}d). 
+There was generally little correlation between true and predicted trajectory for any of the individual models besides SSCOM (Figs \ref{scatter-sim-slope}a--d, \ref{performance-sim-slope}).
 
 Ensemble models, and in particular the machine learning ensemble models (random forest and GBM), generally improved estimates of status and trajectory (Fig.
-\ref{fig:hexagon}e--h, Fig.
-\ref{fig:scatter-sim-slope}e--h).
-Compared to the individual models, machine learning ensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Figs \ref{fig:performance}a).
+\ref{hexagon}e--h, Fig.
+\ref{scatter-sim-slope}e--h).
+Compared to the individual models, machine learning ensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Figs \ref{performance}a).
 These ensembles also had better ability to distinguish if simulated stocks
-were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{fig:roc-sim}).
+were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{roc-sim}).
 Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
-\ref{fig:performance-sim-slope}).
+\ref{performance-sim-slope}).
 The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating status (CMSY; Fig.
-\ref{fig:performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting slope (SSCOM; Fig.
-\ref{fig:performance-sim-slope}).
+\ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting slope (SSCOM; Fig.
+\ref{performance-sim-slope}).
 
@@
 For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (less than about 0.XX; Fig.
-\ref{fig:hexagon}c, Fig.
-\ref{fig:partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high bimodal prediction clusters (Fig.
-\ref{fig:partial-sim}).
+\ref{hexagon}c, Fig.
+\ref{partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high bimodal prediction clusters (Fig.
+\ref{partial-sim}).
 Ensembles also exploit the covariance between individual model predictions.
-For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ also tends to be high (Figs \ref{fig:lm-coefs}, \ref{fig:partial-2d-sim}l).
+For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
 The addition of spectral density covariates helped the ensemble models correctly predict higher \bbmsy\\ values (Fig.
-\ref{fig:hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
-\ref{fig:partial-2d-sim}).
+\ref{hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
+\ref{partial-2d-sim}).
 Although the additional covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
-\ref{fig:hexagon} vs.
+\ref{hexagon} vs.
 Fig.
-\ref{fig:hexagon-sim-basic}) TODO NEW FIG THAT QUANTIFIES THIS.
+\ref{hexagon-sim-basic}) TODO NEW FIG THAT QUANTIFIES THIS.
 
@@
 The mean, random forest, and GBM ensembles even outperformed the
-mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig. \ref{fig:performance}b, Fig. \ref{fig:hexagon-ram}).
+mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
 Compared to the individual models, the machine learning ensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
@@
 performance dimensions: e.g. SSCOM does nearly as well as the ensembles at
-predicting the slope of \bbmsy\\ (Fig. \ref{fig:performance-sim-slope}). But no
+predicting the slope of \bbmsy\\ (Fig. \ref{performance-sim-slope}). But no
 individual method is consistently nearly as good as the ensembles. SSCOM does
@@
 accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for
-both the simulated and stock-assessment datasets (Fig. \ref{fig:performance}).
+both the simulated and stock-assessment datasets (Fig. \ref{performance}).
 -->

### 2015-06-12 text/ms.Rmd
Copy editing

@@
 factorial cell) in the cross-validation process into either the training or
-testing split.
+testing split. Since mPRM is built on the RAM Legacy Stock Assessment database,
+when testing model performance on the stock assessment database, we refit mPRM
+on each training split.
 
@@
 
-Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean and slope of \bbmsy\\ in the last five years (hereafter used interchangeably with 'status' and 'trajectory').
-All models exhibted a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). 
+Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean and slope of \bbmsy\\ in the last five years (hereafter used interchangeably with 'status' and 'trajectory', respectively).
+All models exhibited a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). 
 CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig. \ref{performance}a).
@@
 were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{roc-sim}).
-Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
+Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
 \ref{performance-sim-slope}).
@@
 The ensemble models were able to improve the predictions of status by exploiting the best properties of individual models, the covariance between individual models, and interactions with other covariates.
-For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (less than about 0.XX; Fig.
+For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (Fig.
 \ref{hexagon}c, Fig.
@@
 \ref{partial-sim}).
-Ensembles also exploit the covariance between individual model predictions.
-For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high \bbmsy\\, the true \bbmsy\\ also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
-The addition of spectral density covariates helped the ensemble models correctly predict higher \bbmsy\\ values (Fig.
+Ensembles also exploited the covariance between individual model predictions.
+For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
+The addition of spectral density covariates helped the ensemble models correctly predict higher status values (Fig.
 \ref{hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.

### 2015-06-17 text/ms.Rmd
Edit paper

@@
 
-What can we do when multiple ecological models give different impressions of population status? Any single model is unlikely to perform best in all scenarios. 
-However, ensemble methods from the fields climate science and machine learning provide a powerful tool to leverage information across multiple ecological models. 
-Here we evaluate the potential for ensemble methods to improve estimates of fish population status. 
+What can we do when multiple ecological models suggest different conclusions about population status?
+Any single model is unlikely to perform best under all scenarios.
+Ensemble methods, commonly used in fields such as climate science, provide a powerful tool to leverage information across multiple ecological models.
+Here we evaluate the potential for ensemble methods to improve estimates of fish population status and trajectory.
 We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees.
@@
 
-There are often multiple ecological models we can apply to a system to assess
-population status and trajectory. Models might incorporate different data
-types, assume alternate population dynamics, or make contrasting assumptions
-about the starting state of a system. These models may suggest different
-conclusions about population status and trajectory and we often do not know
-which model is best. How can we reconcile multiple models to make robust
-management decisions about ecological resources?
+There are often multiple models we can apply to an ecological system to assess population status and trajectory.
+Models might incorporate different data types, assume alternate population dynamics, or make contrasting assumptions about the starting state of a system.
+These models may suggest different conclusions about population status and trajectory and we often do not know which model is best.
+How can we reconcile multiple models to make robust management decisions about ecological resources?
 
-Ensemble methods, which combine the output from multiple models as predictors
-in a new 'ensemble' model, are one solution to this problem. A simple ensemble
-model might take the average of multiple model outputs, perhaps weighted by
-some goodness of fit. A complex ensemble model might also take into account the
-interactions between model outputs and additional covariates. Ensemble models
-are popular in other fields including public health [e.g. @thomson2006],
-agriculture [e.g. @cantelaube2005], climate science [e.g. @murphy2004;
-@tebaldi2007; @pierce2009], and machine learning [e.g. @dietterich2000]. We
-think they have tremendous opportunity in ecological resource management.
+Ensemble methods, which combine the output from multiple models as predictors in a new 'ensemble' model, are one solution to this problem.
+A simple ensemble model might take the average of multiple model outputs, perhaps weighted by some goodness of fit.
+A complex ensemble model might also take into account non-linear interactions between model outputs and additional covariates.
+Ensemble models are popular in other fields including public health [e.g.
+@thomson2006], agriculture [e.g.
+@cantelaube2005], climate science [e.g.
+@murphy2004; @tebaldi2007; @pierce2009], and machine learning [e.g.
+@dietterich2000].
+We think they have tremendous opportunity in ecological resource management.
 
-In fisheries science, a common problem is estimating the status and trajectory
-of an exploited fish population. For the majority of fish stocks, we have
-limited biological information and stock status isn't known [@fao2014]. In
-recent years, a number of methods have been proposed to derive population
-status based on the limited information that is broadly available: fisheries
-catch time series and basic knowledge about the productivity of the species.
-Recently, @rosenberg2014 investigated the ability for these models to estimate
-population status through a large-scale simulation model. A key finding was
-that these models frequently disagree on population status and any one model
-has relatively poor predictive ability on average (e.g. Fig.
+In fisheries science, a common task is estimating the status and trajectory of an exploited fish population.
+For the majority of fish stocks, we have limited biological information and stock status isn't known [@fao2014].
+In recent years, a number of methods have been proposed to derive population status based on the limited information that is broadly available: fisheries catch time series and basic knowledge about the productivity of the species.
+Recently, @rosenberg2014 investigated the ability for these models to estimate population status through a large-scale simulation model.
+A key finding was that these models frequently disagree on population status and any one model has relatively poor predictive ability on average (e.g.
+Fig.
 \ref{motivate}).
@@
 <!--
-Exploit interactions between individual models and relationships between other
-conditions of the data and the performance of the individual models.
-Fitted to populations of known or assumed status and this ensemble model is
-used to extrapolate to populations of interest.
+Exploit interactions between individual models and relationships between other conditions of the data and the performance of the individual models.
+Fitted to populations of known or assumed status and this ensemble model is used to extrapolate to populations of interest.
 -->
 
-Here, we develop ensemble models to estimate the population status and
-trajectory of exploited fish populations. We explore a variety of ensemble
-approaches applied to both simulated and real-world fish stocks and compare
-predictive performance. In developing these models we highlight the types of
-challenges one encounters when forming and evaluating ensemble models. We find
-that ensembles generally provide improved predictive performance than any
-single model---especially when considered across multiple dimensions of
-predictive ability and across multiple datasets. Furthermore, the output from
-ensemble models can provide better mechanistic understanding of how individual
-models of status and trajectory perform.
+Here, we develop ensemble models to estimate the population status and trajectory of exploited fish populations.
+We explore a variety of ensemble approaches applied to both simulated and real-world fish stocks and compare predictive performance.
+In developing these models we highlight the types of challenges one encounters when forming and evaluating ensemble models.
+We find that ensembles generally provide improved predictive performance than any single model---especially when considered across multiple dimensions of predictive ability and across multiple datasets.
+Furthermore, the output from ensemble models can provide better mechanistic understanding of how individual models of status and trajectory perform.
 
@@
 
-We tested the ability of ensemble models to improve estimates of population
-status and trajectory when applied to both a large simulated dataset and a
-global database of assessed stock status. Here we describe the datasets,
-individual models of population status, and ensemble models to combine those
-estimates. We then describe how we evaluated the ability of the various models
-to estimate population status.
+We tested the ability of ensemble models to improve estimates of population status and trajectory when applied to both a large simulated dataset and a global database of assessed stock status.
+Here we describe the datasets, individual models of population status, and ensemble models to combine those estimates.
+We then describe how we evaluated the ability of the various models to estimate population status.
 
@@
 
-<!--TODO: github flr repo not available-->
+We first developed and tested ensemble models with a fully factorial simulated dataset [@rosenberg2014].
+In summary, the simulation models included three life histories: small pelagic, demersal, large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation dynamics: a constant exploitation rate, a exploitation rate that is coupled with biomass to mimic an open-access single-species fishery, a scenario where exploitation rate increases to a fixed level, and a 'roller-coaster' scenario where the exploitation rate increases and then decreases.
+Process noise was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either uncorrelated or had a first-order autoregressive correlation of 0.6.
+The simulation included a scenario without observation error and with multiplicative observation error around catch at $N(0, 0.2)$.
 
-We first developed and tested ensemble models with a fully factorial simulated
-dataset [@rosenberg2014]. In summary, the simulation models included three life
-histories: small pelagic, demersal, large pelagic; three levels of how depleted
-biomass was at the start of the dataset compared to carrying capacity: 100%,
-70%, and 40% of carrying capacity; and four harvest dynamics: a constant
-harvest rate, a harvest rate that is coupled with biomass to mimic an
-open-access single-species fishery, a scenario where harvest rate increases to
-a fixed level, a 'roller-coaster' scenario where the harvest rate increases,
-maintains a steady level and decreases to a lower steady level. Process noise
-was introduced to the models through two levels of multiplicative recruitment
-variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either
-uncorrelated or had a first-order autoregressive correlation of 0.6. The
-simulation included a scenario without observation error and with
-multiplicative observation error around catch at $N(0, 0.2)$.
+@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
+The simulation models were built in the package FLR [@kell2007] for the statistical software \textsf{R} [@r2015].
+@rosenberg2014 provide a full description of the simulation model and the code to generate the simulations is available at <https://github.com/flr/StockSims> (TODO: not there).
 
-@rosenberg2014 ran ten iterations for each combination of factors adding
-stochastic draws of recruitment and catch-recording variability each time to
-generate a total of 5760 stocks. The simulation models were built in the
-package FLR [@kell2007] for the statistical software \textsf{R} [@r2015].
-@rosenberg2014 provide a full description of the simulation model and the code
-to generate the simulations is available at <https://github.com/flr/StockSims>
-(TODO: not there).
-
-We also tested our ensemble models on the RAM Legacy Stock Assessment Database
-[@ricard2012]. Our analysis of the stock-assessment database was based on
-version XX downloaded on XX. After removing stocks for which the models
-described below did not converge, this database included `r ram_stocks_n` from
-XX countries across XX taxonomic orders.
+We also tested our ensemble models on the RAM Legacy Stock Assessment Database [@ricard2012].
+Our analysis of the stock-assessment database was based on version XX downloaded on XX.
+After removing stocks for which the models described below did not converge, this database included `r ram_stocks_n` from XX countries.
 
@@
 
-We fit four individual data-limited models to estimate \bbmsy\\. Three of the
-models are mechanistic and based generally on @schaefer1954 biomass
-dynamics of the form:
+We fit four individual data-limited models to estimate \bbmsy\\. Three of the models are mechanistic and based generally on @schaefer1954 biomass dynamics of the form
 
@@
 
-\noindent where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus
-one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic
-population growth rate, $B_0$ represents unfished biomass or carrying capacity,
-and $C$ represents catch. @rosenberg2014 provide a full summary of these four
-methods and code to fit all the models is available in an accompanying package
-**datalimited** for the statistical software \textsf{R}. In summary:
+\noindent
+where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$ represents catch.
+@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package **datalimited** for the statistical software \textsf{R}.
+In summary:
 
-- *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
+* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
   harvest-dynamics model developed by @vasconcellos2005. Biomass is assumed to
@@
 
-- *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
   biomass dynamics [@martell2013]. It requires a prior distribution for $r$ and
@@
 
-- *SSCOM* (state-space catch-only-model) TODO: Jim -- can you give a ~2 sentence summary?
+* *SSCOM* (state-space catch-only-model) TODO: Jim -- can you give a ~2 sentence summary?
 
-- *mPRM* (modified panel regression model) is a modified version of the
+* *mPRM* (modified panel regression model) is a modified version of the
   panel-regression model used in @costello2012. Unlike the other models
@@
 
-Ensemble methods allow us to incorporate additional covariates into our models
-and potentially leverage interactions between these covariates and individual
-model predictions. Additional covariates could be, for example, life-history
-characteristics, information on exploitation patterns, or statistical
-properties of the data. For simplicity, and to allow us to apply models
-developed with the simulated dataset to the real-world dataset, we added only
-one set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and
-provides a means of describing the statistical properties of the catch series
-that is independent of time series length and magnitude of catch. We fit
-spectral models to the scaled catch time series (catch divided by maximum
-catch) with the `spec.ar` function in \textsf{R} and recorded representative
-short- and long-term spectral densities of 0.05 and 0.2, which correspond to
-20- and 5-year cycles.
+Ensemble methods allow us to incorporate additional covariates into our models and potentially leverage interactions between these covariates and individual model predictions.
+Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data.
+For simplicity, and to allow us to apply models developed with the simulated dataset to the real-world dataset, we added only one set of additional covariates: spectral properties of the catch time series.
+Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length and magnitude of catch.
+We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities of 0.05 and 0.2, which correspond to 20- and 5-year cycles.
 
@@
 
-The individual models we seek to combine with ensemble models provide time
-series of \bbmsy\\. Therefore, we can use ensemble models to estimate multiple
-properties of these status time series. Here, we focus on two values: the mean
-and slope of \bbmsy\\ in the last 5 years. Together, these quantities address
-the recent state and trajectory of status, both of which may be of management
-and conservation interest. To avoid undue influence of the end points of the
-time series on the calculated slope, we measured the slope as the Theil-Sen
-estimator of median slope [@theil1950].
+The individual models we seek to combine with ensemble models provide time series of \bbmsy\\.
+Therefore, we can use ensemble models to estimate multiple properties of these status time series.
+Here, we focus on two values: the mean and slope of \bbmsy\\ in the last 5 years.
+Together, these quantities address the recent state and trajectory of status, both of which may be of management and conservation interest.
+To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
 
-An ensemble model can combine models through anything from a simple average to
-a complex non-linear model. Here we choose four ensemble models to compare: an
-average, a linear model with two-way interactions, a random forest, and boosted
-regression tree. The general form of our models is as follows:
+An ensemble model can combine models through anything from a simple average to a complex non-linear model.
+Here we chose four ensemble models to compare: an average, a linear model with two-way interactions, a random forest, and boosted regression tree.
+The general form of our models was
 
@@
 
-\noindent where $L$ is a generalized loss function, $f$ is some generalized
-regression function, $\hat{b}_\mathrm{CMSY}$ is the prediction from the CMSY
-model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
-spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$
-is the true value the model is fitted to, and $\theta$ is the set of parameters
-for the regression function. For all ensemble models of mean \bbmsy\\, which is
-bounded at zero, we fit the models in log space and exponentiated the
-predictions to derive a median estimate of \bbmsy\\. For the estimates of
-\bbmsy\\ slope, we fit ensemble models on the natural untransformed scale.
+\noindent 
+where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.
+(Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$ is the mean or slope value the model is fitted to, and $\theta$ is the set of parameters for the regression function.
+For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the models in log space and exponentiated the predictions to derive a median estimate of \bbmsy\\.
+For the estimates of \bbmsy\\ slope, we fit ensemble models on the natural untransformed scale.
 
-We calculated the average ensemble models as the geometric mean of the
-individual model \bbmsy\\ estimates and the mean of the individual model
-\bbmsy\\ slopes. We fit the linear model ensemble with second-order
-interactions. We fit two machine learning ensemble models based on regression
-trees. Regression trees sequentially determine what value of a predictor best
-splits the response data into two 'branches' based on a specified loss
-function. In random forests, a series of regression trees are built on a random
-subset of the data and random subset of the covariates of the model (REF). In
-generalized boosted models (GBMs), each subsequent model is fit to the
-residuals from the previous model; data points that are fit poorly in a given
-model are given more weight in the next model [@elith2008]. Random forests and
-GBMs can provide strong predictive performance and fit highly non-linear
-relationships [e.g. @elith2008; @hastie2009]. We fit random forest models with
-the **randomForest** package [@liaw2002] for \textsf{R} with the default
-argument values. We fit boosted regression tree models with the **GBM** package
-[@ridgeway2015] for \textsf{R}. Based on cross-validation with the **caret**
-\textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a
-learning rate (shrinkage parameter) of $0.01$, and all other arguments at their
-default values.
+We calculated the average ensemble models as the geometric mean of the individual model \bbmsy\\ estimates and the mean of the individual model \bbmsy\\ slopes.
+We fit the linear model ensemble with second-order interactions.
+We fit two machine learning ensemble models based on regression trees.
+Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function.
+In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model (REF).
+In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008].
+Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [e.g.
+@elith2008; @hastie2009].
+We fit random forest models with the **randomForest** package [@liaw2002] for \textsf{R} with the default argument values.
+We fit boosted regression tree models with the **GBM** package [@ridgeway2015] for \textsf{R}.
+Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
 
@@
 
-A critical component to any predictive modelling exercise is to evaluate the
-performance of a model on new data [@hastie2009]. We used repeated three-fold
-cross validation to test predictive performance: we randomly divided the
-dataset into three sets, built ensemble models on two-thirds of the data, and
-evaluated predictive performance on the remaining third. We repeated this
-across each of the three splits and then repeated the whole procedure 50 times
-to account for bias that may result from any one set of validation splits.
-Since the dynamics of simulated populations differing only in the stochastic
-draw of random values are likely similar, we grouped these stocks (10 per
-factorial cell) in the cross-validation process into either the training or
-testing split. Since mPRM is built on the RAM Legacy Stock Assessment database,
-when testing model performance on the stock assessment database, we refit mPRM
-on each training split.
+A critical component to any predictive modelling exercise is to evaluate the performance of a model on new data [@hastie2009].
+We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built ensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third.
+We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits.
+Since the dynamics of simulated populations differing only in the stochastic draw of random values are likely similar, we grouped these stocks (10 per factorial cell) in the cross-validation process into either the training or testing split.
+Since mPRM is built on the RAM Legacy Stock Assessment database, when testing model performance on the stock assessment database, we refit mPRM on each training split.
 
-Predictive performance can be evaluated with metrics that represent a variety
-of modelling goals. For continuous response variables such as the mean and
-slope of population status, performance metrics often measure some form of bias,
-precision, accuracy (a combination of bias and precision), or the ability to
-correctly rank or correlate across populations (REFs). Here, we measure proportional
-error (PE; also called 'relative error') defined as $(\hat{b} - b)/b$. We
-summarize median proportional error to measure bias, median absolute
-proportional error to measure accuracy, and Spearman's rank-order correlation
-to measure the ability to correctly rank populations.
+Predictive performance can be evaluated with metrics that represent a variety of modelling goals.
+For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of bias, precision, accuracy (a combination of bias and precision), or the ability to correctly rank or correlate across populations (REFs).
+Here, we measure proportional error (PE; also called 'relative error') defined as $(\hat{b} - b)/b$.
+We summarize median proportional error to measure bias, median absolute proportional error to measure accuracy, and Spearman's rank-order correlation to measure the ability to correctly rank populations.
 
@@
 
-Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean and slope of \bbmsy\\ in the last five years (hereafter used interchangeably with 'status' and 'trajectory', respectively).
-All models exhibited a high degree of scatter around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). 
-CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig. \ref{performance}a).
-COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig. \ref{hexagon}b, c). 
-mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{hexagon}d). 
+Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of \bbmsy\\ in the last five years.
+All models exhibited a high degree of scatter around the 1:1 line of perfect status prediction (Fig.
+\ref{hexagon}).
+CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig.
+\ref{performance}a).
+COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig.
+\ref{hexagon}b, c).
+mPRM had relatively poor ability to predict status for the simulated dataset (Fig.
+\ref{hexagon}d).
 There was generally little correlation between true and predicted trajectory for any of the individual models besides SSCOM (Figs \ref{scatter-sim-slope}a--d, \ref{performance-sim-slope}).
@@
 Compared to the individual models, machine learning ensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Figs \ref{performance}a).
-These ensembles also had better ability to distinguish if simulated stocks
-were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{roc-sim}).
+These ensembles also had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
+\ref{roc-sim}).
 Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
@@
 \ref{hexagon}c, Fig.
-\ref{partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high bimodal prediction clusters (Fig.
+\ref{partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high prediction clusters (Fig.
 \ref{partial-sim}).
@@
 When applied to the stock assessment database, the ensemble models---trained exclusively on the simulated dataset---generally performed as well or better than the best individual models.
-The mean, random forest, and GBM ensembles even outperformed the
-mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
+The mean, random forest, and GBM ensembles even outperformed the mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig.
+\ref{performance}b, Fig.
+\ref{hexagon-ram}).
 Compared to the individual models, the machine learning ensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
@@
 <!--
-Some individual models performed nearly as well as the ensembles in one or two
-performance dimensions: e.g. SSCOM does nearly as well as the ensembles at
-predicting the slope of \bbmsy\\ (Fig. \ref{performance-sim-slope}). But no
-individual method is consistently nearly as good as the ensembles. SSCOM does
-more poorly than the ensembles and other individual models in terms of
-accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for
-both the simulated and stock-assessment datasets (Fig. \ref{performance}).
+Some individual models performed nearly as well as the ensembles in one or two performance dimensions: e.g.
+SSCOM does nearly as well as the ensembles at predicting the slope of \bbmsy\\ (Fig.
+\ref{performance-sim-slope}).
+But no individual method is consistently nearly as good as the ensembles.
+SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for both the simulated and stock-assessment datasets (Fig.
+\ref{performance}).
 -->

### 2015-06-20 text/ms.Rmd
Start developing discussion

@@
 
+Ensemble models did not always outperform the best individual models, but they were consistently the best or not far from the best across many performance dimensions, two response variables, two datasets, and multiple ensemble model types.
+Our results suggest choosing an ensemble model that allows for non-linear and non-parametric relationships, such as machine learning methods---these methods provide added insight into individual model behaviour and generally performed the best---but even a simple average of predictions across multiple models may be useful.
+We conclude that ensemble models provide a useful approach to situations where conservation and management decisions must be made on the basis of multiple, potentially contrasting estimates of status.
+These situations are common to many ecological settings beyond fisheries and so we suggest that ensemble models are currently a powerful and underused tool in ecology.
+
+Under what conditions will an ensemble model be most effective?
+First, ensemble models will be most effective when they are comprised of 'diverse' individual models [e.g.
+@ali1996; @tebaldi2007].
+By diverse we mean the individual models make substantially different assumptions about underlying dynamics, or make different but plausible assumptions about certain parameter values.
+Such models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance from each model.
+Second, ensemble models will be most effective when they are not over-fit to the training.
+The probability of over-fitting ensemble models generally increases with the number of individual models [e.g.
+@caruana2004].
+Cross-validation testing, information theoretic methods (e.g.
+AIC), and methods that are robust to over-fitting such as random forests (REF), are tools to help avoid over-fitting ensemble models.
+We note that our simplest ensemble model, a mean of the individual model predictions, performed approximately as well as the complex machine learning models when we applied our ensembles to an entirely new dataset (FIG).
+Finally, ensemble models will be most effective when they are trained on data that are representative of the dataset of interest.
+Cross-validation will provide a optimistically biased impression of predictive performance if the training dataset fundamentally differs from the dataset of interest.
+
+Ensemble models share similarities with coefficient averaging (Burnham and Anderson REF) but differ in many important way.
+At a high level, ensemble models and coefficient averaging share the same idea that no one model is likely correct and multiple models can contribute useful information for inference.
+A fundamental difference between coefficient averaging and ensemble models is that coefficient averaging focuses on a averaging coefficients whereas ensembles instead average predictions.
+By focussing on predictions, ensembles are a more general purpose tool: they don't require information theoretics, they can combine completely different types of models (e.g.
+combining parametric and non-parametric models or frequentist and Bayesian predictions), and they can combine the predictions in complex non-linear ways that also depend on additional covariates.
+
+A particular strength of ensemble models is that multiple ensemble models can be generated to predict multiple response variables.
+For example, we built separate models of mean \bbmsy\\ and the slope of \bbmsy\\.
+The same set of model weights or non-linear relationships need not hold across different response variables.
+Ensemble models can optimally combine the individual models depending on the response.
+Ensemble models could be combined to estimate any number of response variables of interest.
+For example, they could be combined to estimate a binary response such as whether status is above or below some threshold.
+
+* Coilin: How ensembles can help understand mechanistic underpinnings of individual 
+  models
+    * discover surprises: counterintuitive relationships and interactions
+    * can dig into surprising conditions and learn about why certain models
+    * fail; potentially learn how to improve models
+    * can learn under what conditions certain models perform well
+    * pick out an example
+
+* Jim: Other possible applications of ensemble models outside fisheries status and
+  trajectory
+    * other taxa examples
+    * other kinds of response variables
+
+<!--
+* Other possible topics:
+    * how much better are the ensembles really? translate the results into
+      some concrete examples in an absolute not relative sense
+    * other types of ensembles: assigning weights, incorporating uncertainty
+      around estimates, GAMs, ...
+    * when wouldn't you want to use ensembles? perhaps if models represent 
+      dichotomous assumptions where either one or the other is right and imply
+      different management actions? ...
+    * your ideas here...
+-->
+
+# References
+
 # Acknowledgements

### 2015-06-22 text/ms.Rmd
Edit discussion

--- 2015-06-17 text/ms.Rmd
@@
 
+Ensemble models provide a useful approach to situations where conservation and management decisions must be made on the basis of multiple, potentially contrasting estimates of status---a situation that is common to many ecological settings beyond fisheries.
+Compared to individual models of fish population status, ensemble models were consistently the best or among the best across three performance dimensions, two response variables, two datasets, and multiple ensemble model types.
+Our results suggest choosing an ensemble model that allows for non-parametric relationships, such as machine learning methods---these methods provide added insight into individual model behaviour and generally performed the best---but even a simple average of predictions across multiple models may be useful.
+
+Certain conditions will make some ensemble models more effective than others.
+First, ensemble models will be most effective when they are comprised of diverse individual models that make substantially different assumptions about underlying dynamics, or make different but plausible assumptions about certain parameter values [e.g.
+@ali1996; @tebaldi2007].
+Such models would be expected to perform well across different conditions and an ensemble model can exploit the best predictive performance of each model.
+Second, ensemble models will be most effective when they are not overfit to training data.
+The probability of overfitting ensemble models generally increases with the number of individual models [e.g.
+@caruana2004].
+Cross-validation testing, information theoretic methods (e.g.
+AIC), and methods that are robust to over-fitting such as random forests (REF), may help avoid overfitting ensemble models.
+We note that our simplest ensemble model, a mean of the individual model predictions, performed approximately as well as the complex machine learning models when we applied our ensembles to an entirely new dataset (FIG).
+Third, ensemble models will be most effective when they are trained on data that are representative of the dataset of interest.
+Cross-validation within a training dataset will provide a optimistically biased impression of predictive performance if the training dataset fundamentally differs from the dataset of interest.
+
+Ensemble models share similarities with coefficient averaging (Burnham and Anderson REF) but differ in many important way.
+At a high level, ensemble models and coefficient averaging share the same idea that no one model is correct and multiple models can contribute useful information for inference.
+A fundamental difference between coefficient averaging and ensemble models is that coefficient averaging focuses on a averaging *coefficients* whereas ensembles instead average *predictions*.
+By focussing on predictions, ensembles are a more general purpose tool: they don't require information theoretics, they can combine completely different types of models (e.g.
+combining parametric and non-parametric models or frequentist and Bayesian predictions), and they can combine predictions in complex non-linear ways that also depend on additional covariates.
+
+A strength of ensemble models is that they can be tailored or specific response variables.
+For example, we built separate models of mean \bbmsy\\ and the slope of \bbmsy\\.
+The same set of model weights or non-linear relationships need not hold across different response variables.
+Ensemble models can optimally combine the individual models depending on the response.
+Ensemble models could be combined to estimate any number of response variables of interest.
+For example, they could be combined to estimate a binary response such as whether status is above or below some threshold.
+
+* Coilin: How ensembles can help understand mechanistic underpinnings of individual 
+  models
+    * discover surprises: counterintuitive relationships and interactions
+    * can dig into surprising conditions and learn about why certain models fail; 
+      potentially learn how to improve models
+    * can learn under what conditions certain models perform well
+    * pick out an example
+
+* Jim: Other possible applications of ensemble models outside fisheries status and
+  trajectory
+    * other taxa examples
+    * other kinds of response variables
+
+<!--
+* Other possible topics:
+    * how much better are the ensembles really? translate the results into
+      some concrete examples in an absolute not relative sense
+    * other types of ensembles: assigning weights, incorporating uncertainty
+      around estimates, GAMs, ...
+    * when wouldn't you want to use ensembles? perhaps if models represent 
+      dichotomous assumptions where either one or the other is right and imply
+      different management actions? ...
+    * your ideas here...
+-->
+
+# References
+
 # Acknowledgements

### 2015-06-22 text/ms.Rmd
More discussion editing

@@
 Certain conditions will make some ensemble models more effective than others.
-First, ensemble models will be most effective when they are comprised of diverse individual models that make substantially different assumptions about underlying dynamics, or make different but plausible assumptions about certain parameter values [e.g.
+First, ensemble models will be most effective when they are comprised of diverse individual models that make different assumptions about underlying dynamics, or make contrasting but plausible assumptions about parameter values [e.g.
 @ali1996; @tebaldi2007].
-Such models would be expected to perform well across different conditions and an ensemble model can exploit the best predictive performance of each model.
-Second, ensemble models will be most effective when they are not overfit to training data.
+Such individual models would be expected to perform well across a variety of conditions and an ensemble model can exploit the best predictive performance of each.
+Second, ensemble models will be most effective when they are not overfit to the training dataset.
 The probability of overfitting ensemble models generally increases with the number of individual models [e.g.
@@
 
-Ensemble models share similarities with coefficient averaging (Burnham and Anderson REF) but differ in many important way.
-At a high level, ensemble models and coefficient averaging share the same idea that no one model is correct and multiple models can contribute useful information for inference.
+Ensemble models share similarities with coefficient averaging [@burnham2002] but differ other important ways.
+Ensemble models and coefficient averaging share the same notion that no one model is correct and multiple models can contribute useful information for inference.
 A fundamental difference between coefficient averaging and ensemble models is that coefficient averaging focuses on a averaging *coefficients* whereas ensembles instead average *predictions*.
-By focussing on predictions, ensembles are a more general purpose tool: they don't require information theoretics, they can combine completely different types of models (e.g.
-combining parametric and non-parametric models or frequentist and Bayesian predictions), and they can combine predictions in complex non-linear ways that also depend on additional covariates.
+By focussing on predictions, ensembles are a more general purpose tool: they do not require information theoretics, they can combine different types of models (e.g.
+combining parametric and non-parametric models or frequentist and Bayesian predictions), and they can combine predictions in complex non-linear ways that may depend on additional covariates.
 
@@
 For example, we built separate models of mean \bbmsy\\ and the slope of \bbmsy\\.
-The same set of model weights or non-linear relationships need not hold across different response variables.
-Ensemble models can optimally combine the individual models depending on the response.
+The same set of model weights or non-linear relationships need not hold across different response variables: ensemble models can optimally combine the individual models depending on the response.
 Ensemble models could be combined to estimate any number of response variables of interest.

### 2015-06-25 text/ms.Rmd
Add Jim's draft paragraph

@@
 
-* Jim: Other possible applications of ensemble models outside fisheries status and
-  trajectory
-    * other taxa examples
-    * other kinds of response variables
+Combining predictions from multiple models via ensemble methods is also broadly useful in other subfields of ecology and fisheries science.
+Predictions about extinction risk are widely used at the national (endangered species act in the US, citation, species at risk act in Canada, citation) and international (the IUCN red list, citation) levels.
+These risk assessments generally involve fitting regression models to outcomes for individual species along with predictors of species risk [@pinsky2011], or population dynamics models to data for individual species [@dfo2010].
+Both types of models are prone to error caused by model-misspecification and therefore results are sensitive to a priori decisions about model structure [@brooks2015].
+Although there are many potential options to account for potential model-misspecification in determination of species risk (e.g., model averaging, @burnham2002, or semi-parametric methods, @thorson2014), ensemble methods are a relatively easy way to combine predictions from multiple models in a way that is clearly explained (i.e., without complicated statistics).
+Similarly, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], or to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014].
 

### 2015-06-25 analysis/2-merge-ram-fits.R
Substitute locally fit version of mPRM
- Use recreated species categories
- Note the coefficients are not identical, but impossible to debug

@@ -57,4 +57,55 @@ ram_fits <- rename(ram_fits,
   b_bmsy_true = b2bmsy_true,
   b_bmsy_est = b2bmsy)
 
+# refit the mPRM fits with the 3-level species categories:
+sp_cat <- read.csv("raw-data/spp-3-lifehistory-categories.csv",
+  stringsAsFactors = FALSE, strip.white = TRUE, comment.char = "#") %>%
+  select(spname, LH2) %>%
+  unique() %>%
+  rename(scientificname = spname, spp_category = LH2) %>%
+  filter(!is.na(spp_category))
+stopifnot(sum(duplicated(sp_cat$spname)) == 0)
+
+# bring in the missing species I categorized:
+sp_cat2 <- read.csv("raw-data/missing-spp3.csv", strip.white = TRUE,
+  stringsAsFactors = FALSE, comment.char = "#") %>%
+  select(scientificname, species_cat) %>% unique() %>%
+  rename(spp_category = species_cat)
+stopifnot(sum(duplicated(sp_cat2$scientificname)) == 0)
+
+sp_cat <- bind_rows(sp_cat, sp_cat2)
+stopifnot(ncol(sp_cat) == 2)
+stopifnot(sum(is.na(sp_cat$spp_category)) == 0)
+
+ram2 <- select(ram, assessid, stockid, tsyear, stocklong, scientificname,
+  CtoUse, res, Bbmsy_toUse) %>%
+  filter(!is.na(CtoUse)) %>%
+  left_join(sp_cat, by = "scientificname")
+stopifnot(sum(is.na(ram2$spp_category)) == 0)
+
+ram_prm_dat <- plyr::ddply(ram2, "stockid", function(x) {
+  datalimited::format_prm(year = x$tsyear, catch = x$CtoUse, bbmsy = x$Bbmsy_toUse,
+    species_cat = x$spp_category[1L])
+})
+
+m <- datalimited::fit_prm(ram_prm_dat)
+
+mprm_ram <- plyr::ddply(ram_prm_dat, c("stockid"), function(x) {
+  out <- predict_prm(x, model = m, ci = TRUE, level = 0.95)
+  out$tsyear <- x$year
+  out
+})
+
+mprm_ram <- rename(mprm_ram, b2bmsy = fit) %>% mutate(method = "Costello") %>%
+  select(-lower, -upper)
+
+# do both have the same stocks?
+stopifnot(identical(setdiff(
[…]

### 2015-06-26 analysis/2-merge-ram-fits.R
Fix local mPRM fits

@@ -80,15 +80,33 @@ stopifnot(sum(is.na(sp_cat$spp_category)) == 0)
 ram2 <- select(ram, assessid, stockid, tsyear, stocklong, scientificname,
   CtoUse, res, Bbmsy_toUse) %>%
   filter(!is.na(CtoUse)) %>%
-  left_join(sp_cat, by = "scientificname")
+  left_join(sp_cat, by = "scientificname") %>%
+  unique() %>%
+  arrange(stockid, tsyear) %>%
+  filter(stockid %in% ram_fits$stockid)
 stopifnot(sum(is.na(ram2$spp_category)) == 0)
 
 ram_prm_dat <- plyr::ddply(ram2, "stockid", function(x) {
   datalimited::format_prm(year = x$tsyear, catch = x$CtoUse, bbmsy = x$Bbmsy_toUse,
     species_cat = x$spp_category[1L])
 })
+saveRDS(ram_prm_dat, "generated-data/ram_prm_dat.rds")
 
-m <- datalimited::fit_prm(ram_prm_dat)
+# this version drops maximum catch since it doesn't translate between simulated
+# and real datasets:
+m <- datalimited::fit_prm(ram_prm_dat,
+  eqn = log(bbmsy) ~
+    mean_scaled_catch +
+    scaled_catch +
+    scaled_catch1 +
+    scaled_catch2 +
+    scaled_catch3 +
+    scaled_catch4 +
+    species_cat +
+    catch_to_rolling_max +
+    time_to_max +
+    years_back +
+    initial_slope - 1)
 
 mprm_ram <- plyr::ddply(ram_prm_dat, c("stockid"), function(x) {
   out <- predict_prm(x, model = m, ci = TRUE, level = 0.95)
@@ -96,8 +114,13 @@ mprm_ram <- plyr::ddply(ram_prm_dat, c("stockid"), function(x) {
   out
 })
 
-mprm_ram <- rename(mprm_ram, b2bmsy = fit) %>% mutate(method = "Costello") %>%
+mprm_ram <- rename(mprm_ram, b_bmsy_est = fit) %>% mutate(method = "Costello") %>%
   select(-lower, -upper)
+
+# merge back in other stock data:
+mprm_ram <- filter(ram_fits, method == "Costello") %>% unique() %>%
+  select(-b_bmsy_est, -method) %>%
+  inner_join(mprm_ram, by = c("stockid", "tsyear"))
 
 # do both have the same stocks?
 stopifnot(identical(setdiff(

### 2015-06-26 analysis/3-load-sim-data.R
Tweak comments

@@ -1,11 +1,14 @@
 library("dplyr")
 
 # Bring in all results exept CMSY:
-# TODO: absolute file path:
+# TODO: local file path:
 load("../../../Dropbox/FisheriesWorkingGroupPhaseII/Decision_trees/CMSY_COMSIR_SSCOM_COSTELLO_STO_BATCH_ALL_RESULTS 2013-06-24 .Rdata")
 
+# bringing in corrected fits below
 d1 <- dplyr::filter(batch1results, method_id != "CMSY")
+
 rm(batch1results)
+
 d1$stock_id <- as.character(d1$stock_id)
 d1$ID <- as.numeric(as.character(d1$ID))
 d1$b_bmsyUpper <- as.numeric(as.character(d1$b_bmsyUpper))
@@ -16,6 +19,7 @@ d1$seed <- as.numeric(as.character(d1$seed))
 d1$n_iterations <- as.numeric(as.character(d1$n_iterations))
 d1$effective_sample_size <- as.numeric(as.character(d1$effective_sample_size))
 
+# TODO: data only local:
 load("raw-data/cmsy_fits_22052015.RData")
 d2 <- cmsynewfits
 rm(cmsynewfits)

### 2015-06-26 analysis/6-sim-to-ram.R
Bring in new full mPRM model to sim-ram ensembles

@@ -7,7 +7,10 @@ library("dplyr")
 source("0-ensemble-functions.R")
 
 # prep the RAM data:
-ram <- readRDS("generated-data/ram_fits.rds")
+ram <- readRDS("generated-data/ram_fits.rds") %>%
+  as.data.frame() %>%
+  filter(!is.na(b_bmsy_true))
+
 ram$method <- sub("Costello", "mPRM", ram$method)
 
 ram_sum <- ram %>%
@@ -75,7 +78,6 @@ m_lm3 <- lm(
       spec_freq_0.05 + spec_freq_0.2)^3,
   data = d_mean_sim)
 
-
 pdf("../figs/lm-coefs.pdf", width = 5, height = 6)
 par(mfrow = c(1, 1))
 par(mar = c(4, 12, 1, 3), cex = 0.9)
@@ -102,17 +104,8 @@ m_gam <- mgcv::gam(
     s(spec_freq_0.05) + s(spec_freq_0.2),
   data = d_mean_sim)
 
-# join in life-history data for mPRM method
-# spp_categories is in the datalimited package as data:
-ts_dat <- dplyr::left_join(datalimited::ram_ts, datalimited::spp_categories)
-# format before to save time, can be done once before cross-validation:
-ts_dat <- plyr::ddply(ts_dat, "stockid", function(x) {
-  datalimited::format_prm(year = x$year, catch = x$catch, bbmsy = x$bbmsy_ram,
-    species_cat = x$spp_category[1L])
-})
-# until I get the proper 3-level species categories:
-ts_dat$species_cat <- NULL
-ts_dat$species_cat <- "ignore_me"
+# load the RAM data formatted for mPRM:
+ram_prm_dat <- readRDS("generated-data/ram_prm_dat.rds")
 
 # Apply the simulation ensembles to a cross-validated version of the RAM dataset:
 cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
@@ -136,10 +129,22 @@ cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
     test_ids  <- ids_scrambled_cut[
       seq(chunk_starts[i], chunk_starts[i] + chunk_size)] %>% na.omit()
     train_ids <- ids_scrambled_cut[!ids_scrambled_cut %in% test_ids]
-    train_dat <- dplyr::filter(ts_dat, stockid %in% train_ids)
[…]

### 2015-06-26 analysis/5-simulation-ensembles.R
Increase CV iterations

@@ -278,7 +278,7 @@ eq <- paste0("CMSY + COMSIR + mPRM + ",
 eq_basic <- paste0("CMSY + COMSIR + mPRM + SSCOM")
 
 # work through cross validation of ensemble models:
-cv_sim_mean <- plyr::ldply(seq_len(4), .parallel = FALSE,
+cv_sim_mean <- plyr::ldply(seq_len(20), .parallel = TRUE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_mean, geo_mean = TRUE, id = "sim-mean",
       gbm_formula = paste0("log(bbmsy_true_mean) ~ ", eq),
@@ -289,7 +289,7 @@ cv_sim_mean <- cv_sim_mean %>% mutate(
   lm_ensemble  = exp(lm_ensemble))
 cv_sim_mean$cv_id <- NULL
 
-cv_sim_mean_basic <- plyr::ldply(seq_len(4), .parallel = TRUE,
+cv_sim_mean_basic <- plyr::ldply(seq_len(20), .parallel = TRUE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_mean, geo_mean = TRUE, id = "sim-mean",
       gbm_formula = paste0("log(bbmsy_true_mean) ~ ", eq_basic),
@@ -300,7 +300,7 @@ cv_sim_mean_basic <- cv_sim_mean_basic %>% mutate(
   lm_ensemble  = exp(lm_ensemble))
 cv_sim_mean_basic$cv_id <- NULL
 
-cv_sim_slope <- plyr::ldply(seq_len(4), .parallel = TRUE,
+cv_sim_slope <- plyr::ldply(seq_len(20), .parallel = TRUE,
   .fun = function(.n)
     cross_val_ensembles(.n = .n, dat = d_slope, geo_mean = FALSE, id = "sim-slope",
       gbm_formula = paste0("bbmsy_true_slope ~ ", eq),

### 2015-06-26 text/ms.Rmd
Major editing of paper, especially discussion

--- 2015-06-22 text/ms.Rmd
@@
 In developing these models we highlight the types of challenges one encounters when forming and evaluating ensemble models.
-We find that ensembles generally provide improved predictive performance than any single model---especially when considered across multiple dimensions of predictive ability and across multiple datasets.
+We find that ensembles generally provide improved predictive performance over any single model---especially when considered across multiple dimensions of predictive ability and across multiple datasets.
 Furthermore, the output from ensemble models can provide better mechanistic understanding of how individual models of status and trajectory perform.
@@
 
-A critical component to any predictive modelling exercise is to evaluate the performance of a model on new data [@hastie2009].
-We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built ensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third.
-We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits.
-Since the dynamics of simulated populations differing only in the stochastic draw of random values are likely similar, we grouped these stocks (10 per factorial cell) in the cross-validation process into either the training or testing split.
-Since mPRM is built on the RAM Legacy Stock Assessment database, when testing model performance on the stock assessment database, we refit mPRM on each training split.
+A critical component to any predictive modelling exercise is to evaluate the
+performance of a model on new data [@hastie2009]. We used repeated three-fold
+cross validation to test predictive performance: we randomly divided the
+dataset into three sets, built ensemble models on two-thirds of the data, and
+evaluated predictive performance on the remaining third. We repeated this
+across each of the three splits and then repeated the whole procedure 50 times
+to account for bias that may result from any one set of validation splits.
+Since the dynamics of simulated populations differing only in the stochastic
+draw of random values are likely similar, we grouped these stocks (10 per
+factorial cell) in the cross-validation process into either the training or
+testing split. Since mPRM is built on the RAM Legacy Stock Assessment database,
+when testing model performance on the stock assessment database, we refit mPRM
+on each training split.
 
-Predictive performance can be evaluated with metrics that represent a variety of modelling goals.
-For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of bias, precision, accuracy (a combination of bias and precision), or the ability to correctly rank or correlate across populations (REFs).
-Here, we measure proportional error (PE; also called 'relative error') defined as $(\hat{b} - b)/b$.
-We summarize median proportional error to measure bias, median absolute proportional error to measure accuracy, and Spearman's rank-order correlation to measure the ability to correctly rank populations.
+Predictive performance can be evaluated with metrics that represent a variety
+of modelling goals. For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of
+bias, precision, accuracy (a combination of bias and precision), or the ability
+to correctly rank or correlate across populations (REFs). Here, we measure
+proportional error (PE; also called 'relative error') defined as $(\hat{b} -
+b)/b$. We summarize median proportional error to measure bias, median absolute
+proportional error to measure accuracy, and Spearman's rank-order correlation
+to measure the ability to correctly rank populations.
 
@@
 
-Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of \bbmsy\\ in the last five years.
-All models exhibited a high degree of scatter around the 1:1 line of perfect status prediction (Fig.
-\ref{hexagon}).
-CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig.
-\ref{performance}a).
-COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig.
-\ref{hexagon}b, c).
-mPRM had relatively poor ability to predict status for the simulated dataset (Fig.
-\ref{hexagon}d).
-There was generally little correlation between true and predicted trajectory for any of the individual models besides SSCOM (Figs \ref{scatter-sim-slope}a--d, \ref{performance-sim-slope}).
+Applied to the simulated dataset of known status, the individual models had
+variable success at recovering the mean (status) and slope (trajectory) of
+\bbmsy\\ in the last five years. All models exhibited a high degree of scatter
+around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). CMSY
+exhibited bimodal predictions of status but had the best rank-order correlation
+and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly
+identified a number of low status stocks, but frequently predicted a high
+status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had
+relatively poor ability to predict status for the simulated dataset (Fig.
+\ref{hexagon}d). There was generally little correlation between true and
+predicted trajectory for any of the individual models besides SSCOM (Figs
+\ref{scatter-sim-slope}a--d, \ref{performance-sim-slope}).
 
-Ensemble models, and in particular the machine learning ensemble models (random forest and GBM), generally improved estimates of status and trajectory (Fig.
-\ref{hexagon}e--h, Fig.
-\ref{scatter-sim-slope}e--h).
-Compared to the individual models, machine learning ensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Figs \ref{performance}a).
-These ensembles also had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
-\ref{roc-sim}).
-Results were similar when predicting trajectory: compared to individual models, machine learning ensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
-\ref{performance-sim-slope}).
-The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating status (CMSY; Fig.
-\ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting slope (SSCOM; Fig.
+Ensemble models, and in particular the machine learning ensemble models (random
+forest and GBM), generally improved estimates of status and trajectory (Fig.
+\ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the
+individual models, machine learning ensembles improved accuracy (median
+absolute proportional error) by 
+`r mean_sim$mach_vs_ind_mare_fold`%, 
+increased rank-order correlation from `r mean_sim$ind_corr_range` 
+to `r mean_sim$mach_corr_range`, 
+and reduced bias (median proportional error) from 
+`r mean_sim$ind_mre_range` to 
+`r mean_sim$mach_mre_range` (Fig. \ref{performance}a). 
+These ensembles also had better ability to distinguish if
+simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
+\ref{roc-sim}). Results were similar when predicting trajectory: compared to
+individual models, machine learning ensembles improved accuracy by 
+`r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from 
+`r slope_sim$ind_corr_range` to 
+`r slope_sim$mach_corr_range`, and reduced bias
+from `r slope_sim$ind_mre_range` 
+to `r slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The
+ensemble models that simply took a mean of the individual models ranked
+slightly behind the best individual model for estimating status (CMSY; Fig.
+\ref{performance}a) and had slightly lower correlation but higher accuracy than
+the best individual model at predicting slope (SSCOM; Fig.
 \ref{performance-sim-slope}).
 
-The ensemble models were able to improve the predictions of status by exploiting the best properties of individual models, the covariance between individual models, and interactions with other covariates.
-For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (Fig.
-\ref{hexagon}c, Fig.
-\ref{partial-sim}c) and CMSY predictions were approximately linearly related to \bbmsy\\ within the low and high prediction clusters (Fig.
-\ref{partial-sim}).
-Ensembles also exploited the covariance between individual model predictions.
-For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
-The addition of spectral density covariates helped the ensemble models correctly predict higher status values (Fig.
-\ref{hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
-\ref{partial-2d-sim}).
-Although the additional covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
-\ref{hexagon} vs.
-Fig.
+The ensemble models were able to improve the predictions of status by
+exploiting the best properties of individual models, the covariance between
+individual models, and interactions with other covariates. For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (Fig.
+\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
+approximately linearly related to \bbmsy\\ within the low and high prediction
+clusters (Fig. \ref{partial-sim}). Ensembles also exploited the covariance
+between individual model predictions. For instance, both the linear model and
+GBM ensemble suggest that if mPRM and SSCOM predict high status, the true
+status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l). The
+addition of spectral density covariates helped the ensemble models correctly
+predict higher status values (Fig. \ref{hexagon}g, h) by exploiting
+interactions between the spectral density predictors and the individual model
+predictions (Fig. \ref{partial-2d-sim}). Although the additional covariates
+improved ensemble fit, performance of the ensembles was only marginally
+degraded by removing these covariates (Fig. \ref{hexagon} vs. Fig.
 \ref{hexagon-sim-basic}) TODO NEW FIG THAT QUANTIFIES THIS.
 
-When applied to the stock assessment database, the ensemble models---trained exclusively on the simulated dataset---generally performed as well or better than the best individual models.
-The mean, random forest, and GBM ensembles even outperformed the mPRM method which is trained on the RAM Legacy Stock Assessment database itself (Fig.
-\ref{performance}b, Fig.
-\ref{hexagon-ram}).
-Compared to the individual models, the machine learning ensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
+When applied to the stock assessment database, the ensemble models---trained
+exclusively on the simulated dataset---generally performed as well or better
+than the best individual models. The mean, random forest, and GBM ensembles
+even outperformed the mPRM method which is trained on the RAM Legacy Stock
+Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
+Compared to the individual models, the machine learning ensembles increased
+accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, 
+improved correlation from `r mean_ram$ind_corr_range` to 
+`r mean_ram$mach_corr_range`, and reduced bias from
+`r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
 
@@
 
-Ensemble models provide a useful approach to situations where conservation and management decisions must be made on the basis of multiple, potentially contrasting estimates of status---a situation that is common to many ecological settings beyond fisheries.
-Compared to individual models of fish population status, ensemble models were consistently the best or among the best across three performance dimensions, two response variables, two datasets, and multiple ensemble model types.
-Our results suggest choosing an ensemble model that allows for non-parametric relationships, such as machine learning methods---these methods provide added insight into individual model behaviour and generally performed the best---but even a simple average of predictions across multiple models may be useful.
+Ensemble methods provide a useful approach to situations where conservation and
+management decisions must be made on the basis of multiple, potentially
+contrasting estimates of status---a situation common to many ecological
+settings beyond fisheries. Compared to individual models of fish population
+status, ensemble models were consistently the best or among the best across
+three performance dimensions, two response variables, two datasets, and
+multiple ensemble model types. Our results suggest choosing an ensemble model
+that allows for non-parametric relationships, such as machine learning
+methods---these methods provide added insight into individual model behaviour
+and generally performed the best---but even a simple average of predictions
+across multiple models may be useful.
 
 Certain conditions will make some ensemble models more effective than others.
-First, ensemble models will be most effective when they are comprised of diverse individual models that make different assumptions about underlying dynamics, or make contrasting but plausible assumptions about parameter values [e.g.
-@ali1996; @tebaldi2007].
-Such individual models would be expected to perform well across a variety of conditions and an ensemble model can exploit the best predictive performance of each.
-Second, ensemble models will be most effective when they are not overfit to the training dataset.
-The probability of overfitting ensemble models generally increases with the number of individual models [e.g.
-@caruana2004].
-Cross-validation testing, information theoretic methods (e.g.
-AIC), and methods that are robust to over-fitting such as random forests (REF), may help avoid overfitting ensemble models.
-We note that our simplest ensemble model, a mean of the individual model predictions, performed approximately as well as the complex machine learning models when we applied our ensembles to an entirely new dataset (FIG).
-Third, ensemble models will be most effective when they are trained on data that are representative of the dataset of interest.
-Cross-validation within a training dataset will provide a optimistically biased impression of predictive performance if the training dataset fundamentally differs from the dataset of interest.
+First, ensembles will be most effective when they are comprised of diverse
+individual models that choose different structural model forms, explore
+contrasting but plausible ranges of parameter values, or make errors in
+uncorrelated ways [e.g. @ali1996; @dietterich2000; @tebaldi2007]. Such
+individual models would be expected to perform well in different conditions and
+an ensemble model can exploit the best predictive performance of each. Second,
+ensemble models will be most effective when they are not overfit to the
+training dataset. Cross-validation testing [e.g. @caruana2004; @hastie2009] and
+methods that are robust to over-fitting such as random forests [@breiman2001],
+may help avoid overfitting ensemble models. We note that our simplest ensemble
+model, an average of individual model predictions, performed approximately as
+well as complex machine learning models when we applied our ensembles to an
+entirely new dataset (Fig. \ref{performance}b) and averages are a common
+ensemble approach in climate science [@ipcc2013]. Third, ensemble
+models will be most effective when they are trained on data that are
+representative of the dataset of interest [e.g. @knutti2009; @weigel2010].
+Cross-validation within a training dataset will provide a optimistically biased
+impression of predictive performance if the training dataset fundamentally
+differs from the dataset of interest [@hastie2009].
 
-Ensemble models share similarities with coefficient averaging [@burnham2002] but differ other important ways.
-Ensemble models and coefficient averaging share the same notion that no one model is correct and multiple models can contribute useful information for inference.
-A fundamental difference between coefficient averaging and ensemble models is that coefficient averaging focuses on a averaging *coefficients* whereas ensembles instead average *predictions*.
-By focussing on predictions, ensembles are a more general purpose tool: they do not require information theoretics, they can combine different types of models (e.g.
-combining parametric and non-parametric models or frequentist and Bayesian predictions), and they can combine predictions in complex non-linear ways that may depend on additional covariates.
+Multi-model inference in the form of coefficient averaging weighted by
+information theoretics such as the Akaike Information Criteron (AIC) is a
+common analytical approach in ecology [e.g. @burnham2002; @grueber2011]. The
+ensemble methods described in this paper share similarities with coefficient
+averaging but differ in other important ways. Ensemble models and coefficient
+averaging share the long-held notion that multiple working hypotheses can
+contribute useful information for inference [@chamberlin1890]. A fundamental
+difference is that coefficient averaging focuses on a averaging *coefficients*
+whereas ensembles instead average *predictions*. By focussing on predictions,
+ensembles do not suffer from the muddled inference of coefficient averaging
+when coefficients covary [@cade2015] and are a more general purpose tool: they
+do not require information theoretics, they can combine different types of
+models (e.g. parametric and non-parametric models or frequentist and
+Bayesian predictions), and they can combine predictions in complex non-linear
+ways that may depend on additional covariates.
 
-A strength of ensemble models is that they can be tailored or specific response variables.
-For example, we built separate models of mean \bbmsy\\ and the slope of \bbmsy\\.
-The same set of model weights or non-linear relationships need not hold across different response variables: ensemble models can optimally combine the individual models depending on the response.
-Ensemble models could be combined to estimate any number of response variables of interest.
-For example, they could be combined to estimate a binary response such as whether status is above or below some threshold.
+A strength of ensemble models is that they can be tailored or specific response
+variables. For example, we built separate ensemble models of mean \bbmsy\\ and
+the slope of \bbmsy\\. The same set of model weights or non-linear
+relationships need not hold across different response variables. For example,
+SSCOM contributed little to the GBM ensemble estimate of status at higher
+levels of predicted \bbmsy\\, but contributed strongly to estimates of slope
+(TODO FIG). Formally, this fitting of ensemble models to specific quantities of
+interest provides an additional calibration step [@rykiel1996]. Individual
+models are typically calibrated (i.e. parameters are estimated) based on a
+single response variable; however, this model might then provide a biased or
+inaccurate estimate of quantities. For example, a data-limited fisheries model
+might calibrate parameters by maximizing likelihood with respect to observed
+and predicted catches but those parameters may then provide a biased or
+inaccurate estimate of the mean and slope of \bbmsy\\. Ensemble methods can
+provide a final calibration step to a quantity of interest.
 
@@
 
-* Jim: Other possible applications of ensemble models outside fisheries status and
-  trajectory
-    * other taxa examples
-    * other kinds of response variables
+Combining predictions from multiple models via ensemble methods is also broadly
+useful in other subfields of ecology and fisheries science. Predictions about
+extinction risk are widely used at the national (e.g. the US Endangered Species
+Act and the Canadian Species at Risk Act) and international [e.g. the IUCN Red
+List, @iucn2015] levels. These risk assessments generally involve fitting
+regression models to outcomes for individual species along with predictors of
+species risk [e.g. @anderson2011a;@pinsky2011], or population dynamics models to
+data for individual species [e.g. @dfo2010]. Both types of models are prone to
+error caused by model-misspecification and therefore results are sensitive to
+decisions about model structure [@brooks2015]. Although there are options to
+account for potential model-misspecification in determination of species risk
+[e.g., coefficient averaging, @burnham2002; or semi-parametric methods,
+@thorson2014], ensemble methods are a relatively simple way to combine
+predictions in a transparent manner. Beyond estimates of status and trajectory,
+ensemble methods could be used to increase the robustness of spatial
+predictions when designing networks of protected areas [@rassweiler2014], or to
+forecast potential spatial shifts in species distribution given climate impacts
+[@harsch2014].
 
@@
     * your ideas here...
+
+Why do ensembles work?
+
+- 'representational' no one model can contain all 'hypothetical functional
+  forms'... many models can cover many hypotheses @dietterich2000
+- error cancellation --- especially if multiple models make errors in
+  uncorrelated ways [@ali1996]
+- they provide an additional step of 'calibration' [@rykiel1996] where the
+  predictions are calibrated to data of known or assumed truth about a
+  particular aspect of interest in the data (e.g. mean status, slope of status)
+  whereas individual model may be calibrated (parameters estimated) with a
+  slightly different response in mind (e.g. maximizing likelihood with respect
+  to observed and predicted catches)
+- we usually do not know which individual model is 'best' [@hagedorn2005];
+  ensembles have the property of rarely being far from the best
+
 -->
-
-# References
 
@@
 
+Why do ensemble methods work?
+
 - error cancellation is one but not the only reason for superiority of ensemble
@@
   performance criteria, theory on model testing and assessment)
+
+model averaging AIC:
+f Chamberlin 1890).  multiple woring hypotheses
+  (e.g., Johnson and Omland 2004, Hobbs and Hilborn 2006, Burnham et al. 2011,
+  52 Grueber et al. 2011). 
 -->

### 2015-06-27 text/ms.Rmd
Editing throughout
Including switching to numbered references

@@
 
+<!--200 words-->
+
 What can we do when multiple ecological models suggest different conclusions about population status?
@@
 
+<!-- 3-6 keywords -->
+Keywords: data-limited fisheries, model averaging, population dynamics, ... *(up to 6)*
+
 # Introduction
 
-There are often multiple models we can apply to an ecological system to assess population status and trajectory.
-Models might incorporate different data types, assume alternate population dynamics, or make contrasting assumptions about the starting state of a system.
-These models may suggest different conclusions about population status and trajectory and we often do not know which model is best.
-How can we reconcile multiple models to make robust management decisions about ecological resources?
+There are often multiple models we can apply to an ecological system to assess
+population status and trajectory. Models might incorporate different data
+types, assume alternate population dynamics, or make contrasting assumptions
+about the starting state of a system. These models may suggest different
+conclusions about population status and trajectory and we often do not know
+which model is best. How can we reconcile multiple models to make robust
+management decisions about ecological resources?
 
-Ensemble methods, which combine the output from multiple models as predictors in a new 'ensemble' model, are one solution to this problem.
-A simple ensemble model might take the average of multiple model outputs, perhaps weighted by some goodness of fit.
-A complex ensemble model might also take into account non-linear interactions between model outputs and additional covariates.
-Ensemble models are popular in other fields including public health [e.g.
-@thomson2006], agriculture [e.g.
-@cantelaube2005], climate science [e.g.
-@murphy2004; @tebaldi2007; @pierce2009], and machine learning [e.g.
-@dietterich2000].
-We think they have tremendous opportunity in ecological resource management.
+Ensemble methods, which combine the output from multiple models as predictors
+in a new 'ensemble' model, are one solution to this problem. A simple ensemble
+model might take the average of multiple model outputs, perhaps weighted by
+some goodness of fit. A complex ensemble model might also take into account
+non-linear interactions between model outputs and additional covariates.
+Ensemble models are popular in other fields including public health
+[@thomson2006], agriculture [@cantelaube2005], climate science [@murphy2004;
+@tebaldi2007; @pierce2009], and machine learning [@dietterich2000]. We think
+they have tremendous opportunity in ecological resource management.
 
-In fisheries science, a common task is estimating the status and trajectory of an exploited fish population.
-For the majority of fish stocks, we have limited biological information and stock status isn't known [@fao2014].
-In recent years, a number of methods have been proposed to derive population status based on the limited information that is broadly available: fisheries catch time series and basic knowledge about the productivity of the species.
-Recently, @rosenberg2014 investigated the ability for these models to estimate population status through a large-scale simulation model.
-A key finding was that these models frequently disagree on population status and any one model has relatively poor predictive ability on average (e.g.
-Fig.
-\ref{motivate}).
+In fisheries science, a common task is estimating the status and trajectory of
+an exploited fish population. For the majority of fish stocks, we have limited
+fisheries independent data to estimate stock status [@fao2014]. In recent
+years, a number of methods have been proposed to derive population status based
+on the limited information that is broadly available: fisheries catch time
+series and basic knowledge about the productivity of the species (e.g.
+[@vasconcellos2005; @martell2013]). Recently, Rosenberg et al. [@rosenberg2014]
+investigated the ability for four such models to estimate population status through
+a large-scale simulation experiment. A key finding was that these models
+frequently disagree on population status and any one model has relatively poor
+predictive ability on average (e.g. Fig. \ref{motivate}).
 
@@
 
-Here, we develop ensemble models to estimate the population status and trajectory of exploited fish populations.
-We explore a variety of ensemble approaches applied to both simulated and real-world fish stocks and compare predictive performance.
-In developing these models we highlight the types of challenges one encounters when forming and evaluating ensemble models.
-We find that ensembles generally provide improved predictive performance over any single model---especially when considered across multiple dimensions of predictive ability and across multiple datasets.
-Furthermore, the output from ensemble models can provide better mechanistic understanding of how individual models of status and trajectory perform.
+Here, we develop ensemble models to estimate the population status and
+trajectory of exploited fish populations. We explore a variety of ensemble
+approaches applied to both simulated and real-world fish stocks and compare
+predictive performance. In developing these models we highlight the types of
+challenges one encounters when forming and evaluating ensemble models. We find
+that ensembles generally provide improved predictive performance over any
+single model---especially when considered across multiple dimensions of
+predictive ability and across multiple datasets. Furthermore, the output from
+ensemble models can provide better mechanistic understanding of how individual
+models of status and trajectory perform.
 
@@
 
-We tested the ability of ensemble models to improve estimates of population status and trajectory when applied to both a large simulated dataset and a global database of assessed stock status.
-Here we describe the datasets, individual models of population status, and ensemble models to combine those estimates.
-We then describe how we evaluated the ability of the various models to estimate population status.
+We tested the ability of ensemble models to improve estimates of population
+status and trajectory when applied to both a large simulated dataset and a
+global database of assessed stock status. Here we describe the datasets,
+individual models of population status, and ensemble methods to combine those
+estimates. We then describe how we evaluated the ability of the various models
+to estimate population status.
 
@@
 
-We first developed and tested ensemble models with a fully factorial simulated dataset [@rosenberg2014].
-In summary, the simulation models included three life histories: small pelagic, demersal, large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation dynamics: a constant exploitation rate, a exploitation rate that is coupled with biomass to mimic an open-access single-species fishery, a scenario where exploitation rate increases to a fixed level, and a 'roller-coaster' scenario where the exploitation rate increases and then decreases.
-Process noise was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either uncorrelated or had a first-order autoregressive correlation of 0.6.
-The simulation included a scenario without observation error and with multiplicative observation error around catch at $N(0, 0.2)$.
+We first developed and tested ensemble models with a fully factorial simulated
+dataset [@rosenberg2014]. In summary, the simulation models included three life
+histories: small pelagic, demersal, large pelagic; three levels of how depleted
+biomass was at the start of the dataset compared to carrying capacity: 100%,
+70%, and 40% of carrying capacity; and four exploitation dynamics: a constant
+exploitation rate, a exploitation rate that is coupled with biomass to mimic an
+open-access single-species fishery, a scenario where exploitation rate
+increases to a fixed level, and a 'roller-coaster' scenario where the
+exploitation rate increases and then decreases. Process noise was introduced to
+the models through two levels of multiplicative recruitment variability: $N(0,
+0.2)$ and $N(0, 0.6)$, and that process noise was either uncorrelated or had a
+first-order autoregressive correlation of 0.6. The simulation included a
+scenario without observation error and with multiplicative observation error
+around catch at $N(0, 0.2)$.
 
-@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-The simulation models were built in the package FLR [@kell2007] for the statistical software \textsf{R} [@r2015].
-@rosenberg2014 provide a full description of the simulation model and the code to generate the simulations is available at <https://github.com/flr/StockSims> (TODO: not there).
+Rosenberg et al. [@rosenberg2014] ran ten iterations for each combination of
+factors adding stochastic draws of recruitment and catch-recording variability
+each time to generate a total of 5760 stocks. The simulation models were built
+in the package `FLR` [@kell2007] for the statistical software \textsf{R}
+[@r2015]. The simulation model and the code to generate the simulations is
+available at <https://github.com/flr/StockSims> (TODO: not there).
 
-We also tested our ensemble models on the RAM Legacy Stock Assessment Database [@ricard2012].
-Our analysis of the stock-assessment database was based on version XX downloaded on XX.
-After removing stocks for which the models described below did not converge, this database included `r ram_stocks_n` from XX countries.
+We also tested our ensemble models on the RAM Legacy Stock Assessment Database
+[@ricard2012]. Our analysis of the stock-assessment database was based on
+version XX downloaded on XX. After removing stocks for which the models
+described below did not converge, this database included `r ram_stocks_n` from
+XX countries.
 
@@
 
-We fit four individual data-limited models to estimate \bbmsy\\. Three of the models are mechanistic and based generally on @schaefer1954 biomass dynamics of the form
+We fit four individual data-limited models to estimate \bbmsy\\. Three of the
+models are mechanistic and based generally on Schaefer biomass dynamics
+[@schaefer1954] of the form
 
@@
 \noindent
-where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$ represents catch.
-@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package **datalimited** for the statistical software \textsf{R}.
-In summary:
+where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year,
+$B_t$ represents biomass at time $t$, $r$ represents the intrinsic population
+growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$
+represents catch. @rosenberg2014 provide a full summary of these four methods
+and code to fit all the models is available in an accompanying package
+`datalimited` for the statistical software \textsf{R}. In summary:
 
 * *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
-  harvest-dynamics model developed by @vasconcellos2005. Biomass is assumed to
-  follow a Schaefer model and harvest dynamics are assumed to follow a logistic
-  model. The model is fit with a sample-importance-sampling algorithm developed
-  in @rosenberg2014.
+  harvest-dynamics model @vasconcellos2005. Biomass is assumed to follow a
+  Schaefer model and harvest dynamics are assumed to follow a logistic model.
+  The model is fit with a sample-importance-sampling algorithm developed in
+  @rosenberg2014.
 
@@
 
-* *SSCOM* (state-space catch-only-model) TODO: Jim -- can you give a ~2 sentence summary?
+* *SSCOM* (state-space catch-only model) TODO: Jim -- can you give a ~2 sentence summary?
 
 * *mPRM* (modified panel regression model) is a modified version of the
-  panel-regression model used in @costello2012. Unlike the other models
-  considered, mPRM is empirical and not mechanistic --- it uses the RAM Legacy
-  Stock Assessment database to fit a linear model to a series of
+  panel-regression model from Costello et al. [@costello2012]. Unlike the other
+  models considered, mPRM is empirical and not mechanistic---it uses the RAM
+  Legacy Stock Assessment database to fit a linear model to a series of
   characteristics of the catch time series and stock with stock-assessed
@@
   modified from the original in that it condenses the life-history categories
-  into three categories to match the simulated dataset.
-  <!--TODO: what are we doing with the RAM stocks again?-->
-  <!--TODO: bias adjustment?-->
+  into three categories to match the simulated dataset, removes the maximum
+  catch predictor since the absolute catch in the simulated dataset is
+  arbitrary, and does not implement the bias correction needed in Costello et
+  al. [@costello2012] for deriving aggregate estimates of mean status across
+  multiple stocks. (*TODO: can we avoid listing most of these modifications or
+  relegate them to the supplement?*)
+
   <!--TODO: build in fig:didactic -->
@@
 
-Ensemble methods allow us to incorporate additional covariates into our models and potentially leverage interactions between these covariates and individual model predictions.
-Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data.
-For simplicity, and to allow us to apply models developed with the simulated dataset to the real-world dataset, we added only one set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length and magnitude of catch.
-We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities of 0.05 and 0.2, which correspond to 20- and 5-year cycles.
+Ensemble methods allow us to incorporate additional covariates into our models
+and potentially leverage interactions between these covariates and individual
+model predictions. Additional covariates could be, for example, life-history
+characteristics, information on exploitation patterns, or statistical
+properties of the data. For simplicity, and to allow us to apply models
+developed with the simulated dataset to the real-world dataset, we added one
+set of additional covariates: spectral properties of the catch time series.
+Spectral analysis decomposes a time series into the frequency domain and
+provides a means of describing the statistical properties of the catch series
+that is independent of time series length and magnitude of catch. We fit
+spectral models to the scaled catch time series (catch divided by maximum
+catch) with the `spec.ar` function in \textsf{R} and recorded representative
+short- and long-term spectral densities at frequencies of 0.20 and 0.05, which
+correspond to 5- and 20-year cycles.
 
@@
 
-The individual models we seek to combine with ensemble models provide time series of \bbmsy\\.
-Therefore, we can use ensemble models to estimate multiple properties of these status time series.
-Here, we focus on two values: the mean and slope of \bbmsy\\ in the last 5 years.
-Together, these quantities address the recent state and trajectory of status, both of which may be of management and conservation interest.
-To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
+The individual models we seek to combine with ensemble models provide time
+series of \bbmsy\\. Therefore, we can use ensemble models to estimate multiple
+properties of these status time series. Here, we focus on two properties: the
+mean and slope of \bbmsy\\ in the last 5 years. Together, these quantities
+address the recent state and trajectory of status, which are both of management
+and conservation interest (e.g. [@mace2008]). To avoid undue influence of the
+end points of the time series on the calculated slope, we measured the slope as
+the Theil-Sen estimator of median slope [@theil1950].
 
-An ensemble model can combine models through anything from a simple average to a complex non-linear model.
-Here we chose four ensemble models to compare: an average, a linear model with two-way interactions, a random forest, and boosted regression tree.
-The general form of our models was
+An ensemble model can combine models through anything from a simple average to
+a complex non-linear model. Here we chose four ensemble models to compare: an
+average, a linear model with two-way interactions, a random forest, and boosted
+regression tree. The general form of our models was
 
@@
 \noindent 
-where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.
-(Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$ is the mean or slope value the model is fitted to, and $\theta$ is the set of parameters for the regression function.
-For all ensemble models of mean \bbmsy\\, which is bounded at zero, we fit the models in log space and exponentiated the predictions to derive a median estimate of \bbmsy\\.
-For the estimates of \bbmsy\\ slope, we fit ensemble models on the natural untransformed scale.
+where $L$ is a generalized loss function, $f$ is some generalized regression
+function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY
+model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
+spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$
+is the mean or slope value the model is fitted to, and $\theta$ is the set of
+parameters for the regression function. For all ensemble models of mean
+\bbmsy\\, which is bounded at zero, we fit the models in log space and
+exponentiated the predictions to derive a median estimate of \bbmsy\\. For the
+estimates of \bbmsy\\ slope, we fit ensemble models on the natural
+untransformed scale.
 
-We calculated the average ensemble models as the geometric mean of the individual model \bbmsy\\ estimates and the mean of the individual model \bbmsy\\ slopes.
-We fit the linear model ensemble with second-order interactions.
-We fit two machine learning ensemble models based on regression trees.
-Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function.
-In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model (REF).
-In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008].
-Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [e.g.
-@elith2008; @hastie2009].
-We fit random forest models with the **randomForest** package [@liaw2002] for \textsf{R} with the default argument values.
-We fit boosted regression tree models with the **GBM** package [@ridgeway2015] for \textsf{R}.
-Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
+We calculated the average ensemble models as the geometric mean of the
+individual model \bbmsy\\ estimates and the mean of the individual model
+\bbmsy\\ slopes. We fit the linear model ensemble with second-order
+interactions. We fit two machine learning ensemble models based on regression
+trees. Regression trees sequentially determine what value of a predictor best
+splits the response data into two 'branches' based on a specified loss
+function. In random forests, a series of regression trees are built on a random
+subset of the data and random subset of the covariates of the model (REF). In
+generalized boosted models (GBMs), each subsequent model is fit to the
+residuals from the previous model; data points that are fit poorly in a given
+model are given more weight in the next model [@elith2008]. Random forests and
+GBMs can provide strong predictive performance and fit highly non-linear
+relationships [@elith2008; @hastie2009]. We fit random forest models with the
+`randomForest` package [@liaw2002] for \textsf{R} with the default argument
+values. We fit boosted regression tree models with the `gbm` package
+[@ridgeway2015] for \textsf{R}. Based on cross-validation with the **caret**
+\textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a
+learning rate (shrinkage parameter) of $0.01$, and all other arguments at their
+default values.
 
@@
 Predictive performance can be evaluated with metrics that represent a variety
-of modelling goals. For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of
+of modelling goals. For continuous response variables such as the mean and
+slope of population status, performance metrics often measure some form of
 bias, precision, accuracy (a combination of bias and precision), or the ability
@@
 exploiting the best properties of individual models, the covariance between
-individual models, and interactions with other covariates. For example, SSCOM had strong predictive ability when it predicted low \bbmsy\\ (Fig.
+individual models, and interactions with other covariates. For example, SSCOM
+had strong predictive ability when it predicted low \bbmsy\\ (Fig.
 \ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
@@
 contrasting but plausible ranges of parameter values, or make errors in
-uncorrelated ways [e.g. @ali1996; @dietterich2000; @tebaldi2007]. Such
+uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007]. Such
 individual models would be expected to perform well in different conditions and
@@
 ensemble models will be most effective when they are not overfit to the
-training dataset. Cross-validation testing [e.g. @caruana2004; @hastie2009] and
+training dataset. Cross-validation testing [@caruana2004; @hastie2009] and
 methods that are robust to over-fitting such as random forests [@breiman2001],
@@
 models will be most effective when they are trained on data that are
-representative of the dataset of interest [e.g. @knutti2009; @weigel2010].
+representative of the dataset of interest [@knutti2009; @weigel2010].
 Cross-validation within a training dataset will provide a optimistically biased
@@
 information theoretics such as the Akaike Information Criteron (AIC) is a
-common analytical approach in ecology [e.g. @burnham2002; @grueber2011]. The
+common analytical approach in ecology [@burnham2002; @grueber2011]. The
 ensemble methods described in this paper share similarities with coefficient
@@
 extinction risk are widely used at the national (e.g. the US Endangered Species
-Act and the Canadian Species at Risk Act) and international [e.g. the IUCN Red
-List, @iucn2015] levels. These risk assessments generally involve fitting
+Act and the Canadian Species at Risk Act) and international (e.g. the IUCN Red
+List, [@iucn2015]) levels. These risk assessments generally involve fitting
 regression models to outcomes for individual species along with predictors of
-species risk [e.g. @anderson2011a;@pinsky2011], or population dynamics models to
-data for individual species [e.g. @dfo2010]. Both types of models are prone to
-error caused by model-misspecification and therefore results are sensitive to
-decisions about model structure [@brooks2015]. Although there are options to
+species risk (e.g. [@anderson2011a; @pinsky2011)], or population dynamics models
+to data for individual species (e.g. [@dfo2010]). Both types of models are prone
+to error caused by model-misspecification and therefore results are sensitive
+to decisions about model structure [@brooks2015]. Although there are options to
 account for potential model-misspecification in determination of species risk
-[e.g., coefficient averaging, @burnham2002; or semi-parametric methods,
-@thorson2014], ensemble methods are a relatively simple way to combine
+(e.g., coefficient averaging, [@burnham2002], or semi-parametric methods,
+[@thorson2014]), ensemble methods are a relatively simple way to combine
 predictions in a transparent manner. Beyond estimates of status and trajectory,
@@
 
+# Authors' contributions
+
+TODO *Example: AB carried out the molecular lab work, participated in data
+analysis, carried out sequence alignments, participated in the design of the
+study and drafted the manuscript; CD carried out the statistical analyses; EF
+collected field data; GH conceived of the study, designed the study,
+coordinated the study and helped draft the manuscript. All authors gave final
+approval for publication.*
+
 # Acknowledgements
+
+TODO
+
+# Funding
+
+TODO
 

### 2015-06-30 analysis/0-ensemble-functions.R
Extend 1:1 line

@@ -246,7 +246,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
     segments(1, 0, 1, 2.8, lty = "22", col = line_col, lwd = 1.3)
     abline(h = 1, lty = "22", col = line_col, lwd = 1.3)
     #abline(a = 0, b = 1, lty = "22", col = line_col, lwd = 1.3)
-    segments(0, 0, 2.8, 2.8, lty = "22", col = line_col, lwd = 1.3)
+    segments(-1, -1, 2.8, 2.8, lty = "22", col = line_col, lwd = 1.3)
     box(col = "grey60")
     add_label(-0.01, 0.08, paste0("(", letters[m], ") ", unique(filter(dat, order == m)$clean_method)),
       #col = "grey20", add_bg = ifelse(m %in% 1:4, FALSE, TRUE))

### 2015-06-30 analysis/8-main-figs.R
Expand ROC axis labels

@@ -126,7 +126,9 @@ make_roc_curves <- function(dat) {
     geom_line() + coord_equal() +
     geom_abline(intercept = 1, slope = 1, lty = 2, col = "darkgrey") +
     xlim(1, 0) +
-    xlab("Specificity") + ylab("Sensitivity") + theme_bw() +
+    xlab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as < 1)) +
+    ylab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as > 1)) +
+    theme_bw() +
     theme(plot.background = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank()) + facet_wrap(~Ensemble) +

### 2015-06-30 analysis/9-output-values.R
Maintain all "significant" digits

@@ -25,15 +25,15 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
 #     (ind$mre %>% abs %>% min) / (mach$mre %>% abs %>% max) %>% round(digits_fold)
 #   ) * 100
 
-  x$ind_corr_range <- round(range(ind$corr), digits_raw)
-  x$mach_corr_range <- round(range(mach$corr), digits_raw)
+  x$ind_corr_range <- round(range(ind$corr), digits_raw) %>% formatC(digits = digits_raw, format = "f")
+  x$mach_corr_range <- round(range(mach$corr), digits_raw) %>% formatC(digits = digits_raw, format = "f")
 
-  x$ind_mare_range <- round(range(ind$mare), digits_raw)
-  x$mach_mare_range <- round(range(mach$mare), digits_raw)
+  x$ind_mare_range <- round(range(ind$mare), digits_raw) %>% formatC(digits = digits_raw, format = "f")
+  x$mach_mare_range <- round(range(mach$mare), digits_raw) %>% formatC(digits = digits_raw, format = "f")
 
-  x$ensemble_mre_range <- ensemble$mre %>% range %>% round(mre_digits)
-  x$mach_mre_range <- mach$mre %>% range %>% round(mre_digits)
-  x$ind_mre_range <- ind$mre %>% range %>% round(mre_digits)
+  x$ensemble_mre_range <- ensemble$mre %>% range %>% round(mre_digits) %>% formatC(digits = mre_digits, format = "f")
+  x$mach_mre_range <- mach$mre %>% range %>% round(mre_digits) %>% formatC(digits = mre_digits, format = "f")
+  x$ind_mre_range <- ind$mre %>% range %>% round(mre_digits) %>% formatC(digits = mre_digits, format = "f")
 
   lapply(x, paste, collapse = collapse)
 }

### 2015-06-30 text/ms.Rmd
Edit the whole paper

@@
 
-There are often multiple models we can apply to an ecological system to assess
-population status and trajectory. Models might incorporate different data
-types, assume alternate population dynamics, or make contrasting assumptions
-about the starting state of a system. These models may suggest different
-conclusions about population status and trajectory and we often do not know
-which model is best. How can we reconcile multiple models to make robust
-management decisions about ecological resources?
+There are often multiple models we can apply to an ecological system
+[@hilborn1997; @burnham2002]. Models might incorporate different data types,
+assume alternate population dynamics, or make contrasting assumptions about the
+starting state of a system. These models may suggest different conclusions
+about population status and trajectory and we often do not know which model is
+best. How can we reconcile multiple models to make robust management decisions
+about ecological resources?
 
 Ensemble methods, which combine the output from multiple models as predictors
-in a new 'ensemble' model, are one solution to this problem. A simple ensemble
-model might take the average of multiple model outputs, perhaps weighted by
-some goodness of fit. A complex ensemble model might also take into account
-non-linear interactions between model outputs and additional covariates.
-Ensemble models are popular in other fields including public health
+in a new 'ensemble' model, are one possible solution to this problem. A simple
+ensemble model might take the average of multiple model predictions, perhaps
+weighted by some goodness of fit. A complex ensemble model might also take into
+account non-linear interactions between model predictions and additional
+covariates. Ensemble models are popular in other fields including public health
 [@thomson2006], agriculture [@cantelaube2005], climate science [@murphy2004;
-@tebaldi2007; @pierce2009], and machine learning [@dietterich2000]. We think
-they have tremendous opportunity in ecological resource management.
+@tebaldi2007; @pierce2009], and machine learning [@dietterich2000]. 
+<!--We think they have tremendous opportunity in ecological resource management.-->
 
@@
 
+<!--
 We tested the ability of ensemble models to improve estimates of population
@@
 to estimate population status.
+-->
 
+<!--
 ## Datasets
+-->
 
 We first developed and tested ensemble models with a fully factorial simulated
-dataset [@rosenberg2014]. In summary, the simulation models included three life
-histories: small pelagic, demersal, large pelagic; three levels of how depleted
-biomass was at the start of the dataset compared to carrying capacity: 100%,
-70%, and 40% of carrying capacity; and four exploitation dynamics: a constant
-exploitation rate, a exploitation rate that is coupled with biomass to mimic an
-open-access single-species fishery, a scenario where exploitation rate
-increases to a fixed level, and a 'roller-coaster' scenario where the
-exploitation rate increases and then decreases. Process noise was introduced to
-the models through two levels of multiplicative recruitment variability: $N(0,
-0.2)$ and $N(0, 0.6)$, and that process noise was either uncorrelated or had a
-first-order autoregressive correlation of 0.6. The simulation included a
-scenario without observation error and with multiplicative observation error
-around catch at $N(0, 0.2)$.
+dataset [@rosenberg2014]. These previously published simulation models included
+three life histories: small pelagic, demersal, large pelagic; three levels of
+how depleted biomass was at the start of the dataset compared to carrying
+capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation
+dynamics: a constant exploitation rate, a exploitation rate that is coupled
+with biomass to mimic an open-access single-species fishery, a scenario where
+exploitation rate increases to a fixed level, and a 'roller-coaster' scenario
+where the exploitation rate increases and then decreases. Process noise was
+introduced to the models through two levels of multiplicative recruitment
+variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either
+uncorrelated or had a first-order autoregressive correlation of 0.6. The
+simulation included a scenario without observation error and with
+multiplicative observation error around catch at $N(0, 0.2)$.
 
@@
 in the package `FLR` [@kell2007] for the statistical software \textsf{R}
-[@r2015]. The simulation model and the code to generate the simulations is
-available at <https://github.com/flr/StockSims> (TODO: not there).
+[@r2015]. The code to generate the simulations is available at
+<https://github.com/flr/StockSims> (TODO: not there).
 
@@
 version XX downloaded on XX. After removing stocks for which the models
-described below did not converge, this database included `r ram_stocks_n` from
-XX countries.
+described below did not converge, this database included `r ram_stocks_n`
+stocks.
 
@@
 growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$
-represents catch. @rosenberg2014 provide a full summary of these four methods
-and code to fit all the models is available in an accompanying package
-`datalimited` for the statistical software \textsf{R}. In summary:
+represents catch. Rosenberg et al. [@rosenberg2014] provide a full summary of
+these four methods and code to fit all the models is available in an
+accompanying package `datalimited` for the statistical software \textsf{R}. In
+summary:
 
@@
   Schaefer model and harvest dynamics are assumed to follow a logistic model.
-  The model is fit with a sample-importance-sampling algorithm developed in
-  @rosenberg2014.
+  The model is fit with a sample-importance-sampling algorithm [@rosenberg2014].
 
@@
   panel-regression model from Costello et al. [@costello2012]. Unlike the other
-  models considered, mPRM is empirical and not mechanistic---it uses the RAM
-  Legacy Stock Assessment database to fit a linear model to a series of
+  models, mPRM is empirical and not mechanistic---it uses the RAM
+  Legacy Stock Assessment database to fit a regression model to a series of
   characteristics of the catch time series and stock with stock-assessed
-  \bbmsy\\ as the response. The model used in this paper and @rosenberg2014 is
+  \bbmsy\\ as the response. The model used in this paper is
   modified from the original in that it condenses the life-history categories
@@
   al. [@costello2012] for deriving aggregate estimates of mean status across
-  multiple stocks. (*TODO: can we avoid listing most of these modifications or
-  relegate them to the supplement?*)
+  multiple stocks.
 
@@
 
-Ensemble methods allow us to incorporate additional covariates into our models
+Ensemble methods allow us to incorporate additional covariates 
 and potentially leverage interactions between these covariates and individual
@@
 properties of the data. For simplicity, and to allow us to apply models
-developed with the simulated dataset to the real-world dataset, we added one
+developed with the simulated dataset to the real-world dataset, we added only one
 set of additional covariates: spectral properties of the catch time series.
@@
 provides a means of describing the statistical properties of the catch series
-that is independent of time series length and magnitude of catch. We fit
+that is independent of time series length and absolute magnitude of catch. We fit
 spectral models to the scaled catch time series (catch divided by maximum
@@
 
-An ensemble model can combine models through anything from a simple average to
-a complex non-linear model. Here we chose four ensemble models to compare: an
-average, a linear model with two-way interactions, a random forest, and boosted
-regression tree. The general form of our models was
+Here we compare four ensemble models of varying complexity: an average, a
+linear model with two-way interactions, a random forest, and boosted regression
+tree. The general form of our models was
 
@@
 model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
-spectral density of the scaled catch series at frequencies of 0.2 and 0.05, $b$
+spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$
 is the mean or slope value the model is fitted to, and $\theta$ is the set of
 parameters for the regression function. For all ensemble models of mean
-\bbmsy\\, which is bounded at zero, we fit the models in log space and
-exponentiated the predictions to derive a median estimate of \bbmsy\\. For the
-estimates of \bbmsy\\ slope, we fit ensemble models on the natural
-untransformed scale.
+\bbmsy\\---a ratio bounded at zero---we fit the models in log space and
+exponentiated the predictions. For the estimates of \bbmsy\\ slope, we fit
+ensemble models on the natural untransformed scale.
 
@@
 splits the response data into two 'branches' based on a specified loss
-function. In random forests, a series of regression trees are built on a random
-subset of the data and random subset of the covariates of the model (REF). In
-generalized boosted models (GBMs), each subsequent model is fit to the
-residuals from the previous model; data points that are fit poorly in a given
-model are given more weight in the next model [@elith2008]. Random forests and
-GBMs can provide strong predictive performance and fit highly non-linear
-relationships [@elith2008; @hastie2009]. We fit random forest models with the
-`randomForest` package [@liaw2002] for \textsf{R} with the default argument
-values. We fit boosted regression tree models with the `gbm` package
-[@ridgeway2015] for \textsf{R}. Based on cross-validation with the **caret**
-\textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a
-learning rate (shrinkage parameter) of $0.01$, and all other arguments at their
-default values.
+function [@breiman1984]. In random forests, a series of regression trees are
+built on a random subset of the data and random subset of the covariates of the
+model [@breiman2001]. In generalized boosted models (GBMs), each subsequent
+model is fit to the residuals from the previous model; data points that are fit
+poorly in a given model are given more weight in the next model [@elith2008].
+Random forests and GBMs can provide strong predictive performance and fit
+highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest
+models with the `randomForest` package [@liaw2002] for \textsf{R} with the
+default argument values. We fit boosted regression tree models with the `gbm`
+package [@ridgeway2015] for \textsf{R}. Based on cross-validation with the
+**caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth
+of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other
+arguments at their default values.
 
@@
 bias, precision, accuracy (a combination of bias and precision), or the ability
-to correctly rank or correlate across populations (REFs). Here, we measure
-proportional error (PE; also called 'relative error') defined as $(\hat{b} -
-b)/b$. We summarize median proportional error to measure bias, median absolute
-proportional error to measure accuracy, and Spearman's rank-order correlation
-to measure the ability to correctly rank populations.
+to correctly rank or correlate across populations (e.g. [@walther2005]). Here,
+we measure proportional error, defined as
+$(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true'
+(or stock-assessed) mean or slope of \bbmsy\\. We calculated median proportional
+error to measure bias, median absolute proportional error to measure accuracy,
+and Spearman's rank-order correlation to measure the ability to correctly rank
+populations.
 
@@
 predicted trajectory for any of the individual models besides SSCOM (Figs
-\ref{scatter-sim-slope}a--d, \ref{performance-sim-slope}).
+\ref{scatter-sim-slope}a--d).
 
 Ensemble models, and in particular the machine learning ensemble models (random
-forest and GBM), generally improved estimates of status and trajectory (Fig.
+forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig.
 \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the
@@
 degraded by removing these covariates (Fig. \ref{hexagon} vs. Fig.
-\ref{hexagon-sim-basic}) TODO NEW FIG THAT QUANTIFIES THIS.
+\ref{hexagon-sim-basic}).
 
@@
 than the best individual models. The mean, random forest, and GBM ensembles
-even outperformed the mPRM method which is trained on the RAM Legacy Stock
+even outperformed the mPRM method which is trained directly on the RAM Legacy Stock
 Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
@@
 status, ensemble models were consistently the best or among the best across
-three performance dimensions, two response variables, two datasets, and
-multiple ensemble model types. Our results suggest choosing an ensemble model
-that allows for non-parametric relationships, such as machine learning
-methods---these methods provide added insight into individual model behaviour
-and generally performed the best---but even a simple average of predictions
-across multiple models may be useful.
+three performance dimensions (accuracy, bias, and rank-order performance), two
+response variables (status and trajectory), two datasets (simulated and global
+fisheries), and multiple ensemble model types (from a simple average to machine
+learning models). Our results suggest choosing an ensemble model that allows
+for non-parametric relationships, such as machine learning methods---these
+methods provide added insight into individual model behaviour and generally
+performed the best---but even a simple average of predictions across multiple
+models may be useful.
 
@@
 information theoretics such as the Akaike Information Criteron (AIC) is a
-common analytical approach in ecology [@burnham2002; @grueber2011]. The
-ensemble methods described in this paper share similarities with coefficient
-averaging but differ in other important ways. Ensemble models and coefficient
-averaging share the long-held notion that multiple working hypotheses can
-contribute useful information for inference [@chamberlin1890]. A fundamental
-difference is that coefficient averaging focuses on a averaging *coefficients*
-whereas ensembles instead average *predictions*. By focussing on predictions,
-ensembles do not suffer from the muddled inference of coefficient averaging
-when coefficients covary [@cade2015] and are a more general purpose tool: they
-do not require information theoretics, they can combine different types of
-models (e.g. parametric and non-parametric models or frequentist and
+common analytical approach in ecology [@burnham2002; @johnson2004;
+@grueber2011]. The ensemble methods described in this paper share similarities
+with coefficient averaging but differ in other important ways. Ensemble models
+and coefficient averaging share the long-held notion that multiple working
+hypotheses can contribute useful information for inference [@chamberlin1890]. A
+fundamental difference is that coefficient averaging focuses on a averaging
+*coefficients* whereas ensembles instead average *predictions*. By focussing on
+predictions, ensembles do not suffer from the muddled inference of coefficient
+averaging when coefficients covary [@cade2015] and are a more general purpose
+tool: they do not require information theoretics, they can combine different
+types of models (e.g. parametric and non-parametric models or frequentist and
 Bayesian predictions), and they can combine predictions in complex non-linear
@@
 levels of predicted \bbmsy\\, but contributed strongly to estimates of slope
-(TODO FIG). Formally, this fitting of ensemble models to specific quantities of
+(TODO FIG). Formally, fitting ensemble models to specific quantities of
 interest provides an additional calibration step [@rykiel1996]. Individual
@@
 single response variable; however, this model might then provide a biased or
-inaccurate estimate of quantities. For example, a data-limited fisheries model
-might calibrate parameters by maximizing likelihood with respect to observed
-and predicted catches but those parameters may then provide a biased or
-inaccurate estimate of the mean and slope of \bbmsy\\. Ensemble methods can
-provide a final calibration step to a quantity of interest.
+inaccurate estimate of other quantities of interest. For example, a
+data-limited fisheries model might calibrate parameters by maximizing
+likelihood with respect to observed and predicted catches but those parameters
+may then provide a biased or inaccurate estimate of the mean and slope of
+\bbmsy\\. Ensemble methods provide a final calibration step to a quantity of
+interest. This ensemble calibration could further include a loss function
+tailored to the goals of the model, say placing greater weight on accuracy at
+low status levels than higher values.
 
@@
 regression models to outcomes for individual species along with predictors of
-species risk (e.g. [@anderson2011a; @pinsky2011)], or population dynamics models
-to data for individual species (e.g. [@dfo2010]). Both types of models are prone
-to error caused by model-misspecification and therefore results are sensitive
-to decisions about model structure [@brooks2015]. Although there are options to
-account for potential model-misspecification in determination of species risk
-(e.g., coefficient averaging, [@burnham2002], or semi-parametric methods,
+species risk (e.g. [@anderson2011a; @pinsky2011)] (*TODO instead get references
+from official national/international risk assessments?*), or population
+dynamics models to data for individual species (e.g. [@dfo2010]). Both types of
+models are prone to error caused by model-misspecification and therefore
+results are sensitive to decisions about model structure [@brooks2015].
+Although there are options to account for potential model-misspecification in
+determination of species risk (e.g., coefficient averaging, [@burnham2002],
+generalized modeling [@yeakel2011], or semi-parametric methods,
 [@thorson2014]), ensemble methods are a relatively simple way to combine

### 2015-06-30 text/ms.Rmd
Expand bbmsy macros for RTF and Pandoc

@@
 Here we evaluate the potential for ensemble methods to improve estimates of fish population status and trajectory.
-We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees.
+We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees.
 We build our ensembles on a simulated dataset of 5760 fish stocks and test our models with cross-validation and a global database of `r ram_stocks_n` stock assessments.
@@
 
-We fit four individual data-limited models to estimate \bbmsy\\. Three of the
+We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the
 models are mechanistic and based generally on Schaefer biomass dynamics
@@
   characteristics of the catch time series and stock with stock-assessed
-  \bbmsy\\ as the response. The model used in this paper is
+  $B/B_\mathrm{MSY}$ as the response. The model used in this paper is
   modified from the original in that it condenses the life-history categories
@@
 The individual models we seek to combine with ensemble models provide time
-series of \bbmsy\\. Therefore, we can use ensemble models to estimate multiple
+series of $B/B_\mathrm{MSY}$. Therefore, we can use ensemble models to estimate multiple
 properties of these status time series. Here, we focus on two properties: the
-mean and slope of \bbmsy\\ in the last 5 years. Together, these quantities
+mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years. Together, these quantities
 address the recent state and trajectory of status, which are both of management
@@
 parameters for the regression function. For all ensemble models of mean
-\bbmsy\\---a ratio bounded at zero---we fit the models in log space and
-exponentiated the predictions. For the estimates of \bbmsy\\ slope, we fit
+$B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space and
+exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, we fit
 ensemble models on the natural untransformed scale.
@@
 We calculated the average ensemble models as the geometric mean of the
-individual model \bbmsy\\ estimates and the mean of the individual model
-\bbmsy\\ slopes. We fit the linear model ensemble with second-order
+individual model $B/B_\mathrm{MSY}$ estimates and the mean of the individual model
+$B/B_\mathrm{MSY}$ slopes. We fit the linear model ensemble with second-order
 interactions. We fit two machine learning ensemble models based on regression
@@
 $(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true'
-(or stock-assessed) mean or slope of \bbmsy\\. We calculated median proportional
+(or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$. We calculated median proportional
 error to measure bias, median absolute proportional error to measure accuracy,
@@
 variable success at recovering the mean (status) and slope (trajectory) of
-\bbmsy\\ in the last five years. All models exhibited a high degree of scatter
+$B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of scatter
 around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). CMSY
@@
 individual models, and interactions with other covariates. For example, SSCOM
-had strong predictive ability when it predicted low \bbmsy\\ (Fig.
+had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
 \ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
-approximately linearly related to \bbmsy\\ within the low and high prediction
+approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction
 clusters (Fig. \ref{partial-sim}). Ensembles also exploited the covariance
@@
 Some individual models performed nearly as well as the ensembles in one or two performance dimensions: e.g.
-SSCOM does nearly as well as the ensembles at predicting the slope of \bbmsy\\ (Fig.
+SSCOM does nearly as well as the ensembles at predicting the slope of $B/B_\mathrm{MSY}$ (Fig.
 \ref{performance-sim-slope}).
 But no individual method is consistently nearly as good as the ensembles.
-SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for both the simulated and stock-assessment datasets (Fig.
+SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean $B/B_\mathrm{MSY}$ for both the simulated and stock-assessment datasets (Fig.
 \ref{performance}).
@@
 A strength of ensemble models is that they can be tailored or specific response
-variables. For example, we built separate ensemble models of mean \bbmsy\\ and
-the slope of \bbmsy\\. The same set of model weights or non-linear
+variables. For example, we built separate ensemble models of mean $B/B_\mathrm{MSY}$ and
+the slope of $B/B_\mathrm{MSY}$. The same set of model weights or non-linear
 relationships need not hold across different response variables. For example,
 SSCOM contributed little to the GBM ensemble estimate of status at higher
-levels of predicted \bbmsy\\, but contributed strongly to estimates of slope
+levels of predicted $B/B_\mathrm{MSY}$, but contributed strongly to estimates of slope
 (TODO FIG). Formally, fitting ensemble models to specific quantities of
@@
 may then provide a biased or inaccurate estimate of the mean and slope of
-\bbmsy\\. Ensemble methods provide a final calibration step to a quantity of
+$B/B_\mathrm{MSY}$. Ensemble methods provide a final calibration step to a quantity of
 interest. This ensemble calibration could further include a loss function

### 2015-07-07 text/ms.Rmd
Major re-work of text; bring in 'superensembles'

@@
 ```
+
+\noindent
+Running head: Superensembles of population status
+
+\bigskip
+
+\begin{Large}
+\noindent
+Improving estimates of population status and trajectory with superensemble models
+\end{Large}
+
+\bigskip
 
@@
 
-<!--200 words-->
+<!--up to 350 words-->
 
-What can we do when multiple ecological models suggest different conclusions about population status?
-Any single model is unlikely to perform best under all scenarios.
-Ensemble methods, commonly used in fields such as climate science, provide a powerful tool to leverage information across multiple ecological models.
-Here we evaluate the potential for ensemble methods to improve estimates of fish population status and trajectory.
-We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees.
-We build our ensembles on a simulated dataset of 5760 fish stocks and test our models with cross-validation and a global database of `r ram_stocks_n` stock assessments.
-We find that ensemble models substantially improve estimates of population status and trajectory.
-Machine-learning ensembles performed the best: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
-We found similar improvements when predicting the slope of status.
-When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next-best individual model.
-Ensemble models have potential to improve estimates of population status and trajectory across many taxa beyond fish.
-However, ensemble methods must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations they are applied to.
+Ecological resource managers are often faced with reconciling multiple potentially
+conflicting estimates of population status and trajectory.  One solution is to
+average these predictions in an ensemble, but this approach ignores covariance
+between models and may not optimally reduce bias and improve accuracy.
+Superensemble models, commonly used in the fields of climate and weather
+science, may provide a superior solution. Superensembles use the predictions
+from multiple models as covariates in a new statistical model. Here we evaluate
+the potential for ensemble and superensemble methods to improve estimates of fish population
+status and trajectory. We fit four models of population and exploitation
+dynamics to data-limited populations and combine their estimates of the mean and
+slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an
+ensemble average and three superensembles: a linear model, a random forest, and
+a boosted regression tree. We build our models on a simulated dataset of 5760
+fish stocks and test them with cross-validation and a global database of 
+`r ram_stocks_n` stock assessments. 
+We find that ensemble and superensemble models
+substantially improve estimates of population status and trajectory.
+Machine-learning superensembles performed the best for population status: accuracy improved
+`r mean_sim$mach_vs_ind_mare_fold`%, correlation between predicted and true
+status improved from 
+`r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median
+proportional error) declined from 
+`r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. 
+We found similar improvements when predicting the trajectory of status. When the
+simulation-trained ensembles were applied to global fish stocks, the best
+ensembles still improved all dimensions of performance over the next-best
+individual model. Ensemble and superensemble models have the potential to
+improve estimates of population status and trajectory across many taxa beyond
+fish. However, they must be carefully tested, formed from a diverse set of
+accurate models, and built on a dataset sufficiently representative of the
+populations they are applied to.
 
-<!-- 3-6 keywords -->
-Keywords: data-limited fisheries, model averaging, population dynamics, ... *(up to 6)*
+<!-- up to 12 keywords -->
+\noindent 
+Keywords: data-limited fisheries, ensemble methods, model averaging, population dynamics, sustainable resource management
 
@@
 
-There are often multiple models we can apply to an ecological system
-[@hilborn1997; @burnham2002]. Models might incorporate different data types,
-assume alternate population dynamics, or make contrasting assumptions about the
-starting state of a system. These models may suggest different conclusions
-about population status and trajectory and we often do not know which model is
-best. How can we reconcile multiple models to make robust management decisions
-about ecological resources?
+Despite considerable advancements in the systematic collection of ecological
+population data, we still have little data with which to estimate population
+status for most species (IUCN, FAO, amphibian citations etc.). Yet, a large
+portion of these data-limited populations face human-caused pressures and may
+warrant conservation action (IUCN REF). This scenario calls for methods to
+assess the status of populations that rely on limited sources of data. While,
+such methods are increasingly available (e.g. REF, REF), their application faces
+two key challenges: (1) performance is relatively poor compared to more
+data-rich forms of population modelling (e.g. REF) and (2) resource managers are
+often faced with reconciling multiple potentially conflicting estimates of
+population status (e.g. REF).
 
-Ensemble methods, which combine the output from multiple models as predictors
-in a new 'ensemble' model, are one possible solution to this problem. A simple
-ensemble model might take the average of multiple model predictions, perhaps
-weighted by some goodness of fit. A complex ensemble model might also take into
-account non-linear interactions between model predictions and additional
-covariates. Ensemble models are popular in other fields including public health
-[@thomson2006], agriculture [@cantelaube2005], climate science [@murphy2004;
-@tebaldi2007; @pierce2009], and machine learning [@dietterich2000]. 
-<!--We think they have tremendous opportunity in ecological resource management.-->
+One solution is to take the average or weighted average of an ensemble of model
+predictions. Such ensembles are typically more accurate and less biased
+than individual model estimates and can integrate uncertainty in model
+structure, initial conditions, and parameter estimation [@araujo2007]. For
+instance, REF-international did this and REF-national did this.
+
+Whereas averages or weighted averages can improve predictions, they may not
+optimally leverage available data. Some models perform better than others in
+certain conditions and the covariance between model predictions may contain
+information that can both improve predictions and aid understanding of
+individual model performance. We can exploit these characteristics by using the
+predictions from a group of models as inputs into a separate statistical model.
+This technique, sometimes called 'superensemble' modelling, is common in climate
+and weather forecasting. The superensemble is fit to some training dataset where
+outcomes are well known and used to predict on a separate dataset. For example,
+@krishnamurti1999 combined predictions of wind and precipitation in Asian
+monsoons via a superensemble regression fit to observed data. Their
+superensemble considerably outperformed any individual prediction or weighted
+average of predictions in terms of accuracy.
+
+<!-- Ensemble methods, which combine the output from multiple models as
+predictors in a new 'ensemble' model, are one possible solution to this
+problem. A simple ensemble model might take the average of multiple model
+predictions, perhaps weighted by some goodness of fit. A complex ensemble model
+might also take into account non-linear interactions between model predictions
+and additional covariates. Ensemble models are popular in other fields
+including public health [@thomson2006], agriculture [@cantelaube2005], climate
+science [@murphy2004; @tebaldi2007; @pierce2009], and machine learning
+[@dietterich2000].-->
 
@@
 on the limited information that is broadly available: fisheries catch time
-series and basic knowledge about the productivity of the species (e.g.
-[@vasconcellos2005; @martell2013]). Recently, Rosenberg et al. [@rosenberg2014]
+series and basic knowledge about the productivity of the species 
+[e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014
 investigated the ability for four such models to estimate population status through
@@
 
-<!--
-Exploit interactions between individual models and relationships between other conditions of the data and the performance of the individual models.
-Fitted to populations of known or assumed status and this ensemble model is used to extrapolate to populations of interest.
--->
-
-Here, we develop ensemble models to estimate the population status and
-trajectory of exploited fish populations. We explore a variety of ensemble
-approaches applied to both simulated and real-world fish stocks and compare
-predictive performance. In developing these models we highlight the types of
-challenges one encounters when forming and evaluating ensemble models. We find
-that ensembles generally provide improved predictive performance over any
-single model---especially when considered across multiple dimensions of
-predictive ability and across multiple datasets. Furthermore, the output from
-ensemble models can provide better mechanistic understanding of how individual
-models of status and trajectory perform.
+Here, we estimate population status and trajectory of exploited fish populations
+using ensemble and superensemble models (collectively referred to 'ensemble
+methods'). We explore a variety of approaches applied to both simulated and
+real-world fish stocks and compare predictive performance. We find that
+ensembles, and in particular superensembles, generally improved predictive
+performance of status and trajectory of fish populations over any single
+model---especially when considered across multiple dimensions of predictive
+ability and across multiple datasets. Furthermore, the output from
+superensembles can aid mechanistic understanding of how individual models of
+status and trajectory perform.
 
@@
 
-We first developed and tested ensemble models with a fully factorial simulated
+We first developed and tested ensemble methods with a fully factorial simulated
 dataset [@rosenberg2014]. These previously published simulation models included
@@
 
-Rosenberg et al. [@rosenberg2014] ran ten iterations for each combination of
+@rosenberg2014 ran ten iterations for each combination of
 factors adding stochastic draws of recruitment and catch-recording variability
@@
 
-We also tested our ensemble models on the RAM Legacy Stock Assessment Database
+We also tested our ensemble methods on the RAM Legacy Stock Assessment Database
 [@ricard2012]. Our analysis of the stock-assessment database was based on
@@
 growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$
-represents catch. Rosenberg et al. [@rosenberg2014] provide a full summary of
+represents catch. @rosenberg2014 provide a full summary of
 these four methods and code to fit all the models is available in an
@@
 summary:
-
-* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
-  harvest-dynamics model @vasconcellos2005. Biomass is assumed to follow a
-  Schaefer model and harvest dynamics are assumed to follow a logistic model.
-  The model is fit with a sample-importance-sampling algorithm [@rosenberg2014].
 
@@
 
-* *SSCOM* (state-space catch-only model) TODO: Jim -- can you give a ~2 sentence summary?
+* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
+  harvest-dynamics model [@vasconcellos2005]. Biomass is assumed to follow a
+  Schaefer model and harvest dynamics are assumed to follow a logistic model.
+  The model is fit with a sample-importance-sampling algorithm [@rosenberg2014].
+
+* *SSCOM* (state-space catch-only model) is a hierarchical model that, similar
+  to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013].
+  SSCOM estimates unobserved dynamics in both fishing effort and the fished
+  population based on a catch time series and priors on $r$, the maximum rate
+  of increase of fishing effort, and the magnitude of various sources of
+  stochasticity. The model is fit in a Bayesian state-space framework to
+  integrate across three forms of stochasticity: variation in effort,
+  population dynamics, and fishing efficiency [@thorson2013].
 
 * *mPRM* (modified panel regression model) is a modified version of the
-  panel-regression model from Costello et al. [@costello2012]. Unlike the other
-  models, mPRM is empirical and not mechanistic---it uses the RAM
-  Legacy Stock Assessment database to fit a regression model to a series of
-  characteristics of the catch time series and stock with stock-assessed
-  $B/B_\mathrm{MSY}$ as the response. The model used in this paper is
-  modified from the original in that it condenses the life-history categories
-  into three categories to match the simulated dataset, removes the maximum
-  catch predictor since the absolute catch in the simulated dataset is
-  arbitrary, and does not implement the bias correction needed in Costello et
-  al. [@costello2012] for deriving aggregate estimates of mean status across
-  multiple stocks.
+  panel-regression model from @costello2012. Unlike the other
+  models, mPRM is empirical and not mechanistic---it uses the RAM Legacy Stock
+  Assessment database to fit a regression model to a series of characteristics
+  of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as
+  the response. The model used in this paper is modified from the original in
+  that it condenses the life-history categories into three categories to match
+  the simulated dataset, removes the maximum catch predictor since the absolute
+  catch in the simulated dataset is arbitrary, and does not implement the bias
+  correction needed in @costello2012 for deriving aggregate
+  estimates of mean status across multiple stocks.
 
@@
 
-Ensemble methods allow us to incorporate additional covariates 
+Superensemble models allow us to incorporate additional covariates 
 and potentially leverage interactions between these covariates and individual
@@
 
-## Ensemble models
+## Superensemble models
 
-The individual models we seek to combine with ensemble models provide time
-series of $B/B_\mathrm{MSY}$. Therefore, we can use ensemble models to estimate multiple
-properties of these status time series. Here, we focus on two properties: the
-mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years. Together, these quantities
-address the recent state and trajectory of status, which are both of management
-and conservation interest (e.g. [@mace2008]). To avoid undue influence of the
-end points of the time series on the calculated slope, we measured the slope as
-the Theil-Sen estimator of median slope [@theil1950].
+The individual models we seek to combine with superensembles provide time
+series of $B/B_\mathrm{MSY}$. Therefore, we can use superensembles to estimate
+multiple properties of these status time series. Here, we focus on two
+properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
+Together, these quantities address the recent state and trajectory of status,
+which are both of management and conservation interest [e.g. @mace2008]. To
+avoid undue influence of the end points of the time series on the calculated
+slope, we measured the slope as the Theil-Sen estimator of median slope
+[@theil1950].
 
-Here we compare four ensemble models of varying complexity: an average, a
-linear model with two-way interactions, a random forest, and boosted regression
-tree. The general form of our models was
+Here we compare an ensemble average and three superensembles of varying
+complexity: a linear model with two-way interactions, a random forest, and
+boosted regression tree. The general form of our models was
 
@@
 is the mean or slope value the model is fitted to, and $\theta$ is the set of
-parameters for the regression function. For all ensemble models of mean
-$B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space and
-exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, we fit
-ensemble models on the natural untransformed scale.
+parameters for the regression function. For all superensemble models of mean
+$B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space
+and exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$
+slope, which are not bounded at zero, we fit superensemble models on the natural
+untransformed scale.
 
-We calculated the average ensemble models as the geometric mean of the
-individual model $B/B_\mathrm{MSY}$ estimates and the mean of the individual model
-$B/B_\mathrm{MSY}$ slopes. We fit the linear model ensemble with second-order
-interactions. We fit two machine learning ensemble models based on regression
-trees. Regression trees sequentially determine what value of a predictor best
-splits the response data into two 'branches' based on a specified loss
-function [@breiman1984]. In random forests, a series of regression trees are
-built on a random subset of the data and random subset of the covariates of the
-model [@breiman2001]. In generalized boosted models (GBMs), each subsequent
-model is fit to the residuals from the previous model; data points that are fit
-poorly in a given model are given more weight in the next model [@elith2008].
-Random forests and GBMs can provide strong predictive performance and fit
-highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest
-models with the `randomForest` package [@liaw2002] for \textsf{R} with the
-default argument values. We fit boosted regression tree models with the `gbm`
-package [@ridgeway2015] for \textsf{R}. Based on cross-validation with the
-**caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth
-of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other
-arguments at their default values.
+We fit the linear model ensemble with second-order interactions. We fit two
+machine learning ensemble models based on regression trees. Regression trees
+sequentially determine what value of a predictor best splits the response data
+into two 'branches' based on a specified loss function [@breiman1984]. In random
+forests, a series of regression trees are built on a random subset of the data
+and random subset of the covariates of the model [@breiman2001]. In generalized
+boosted models (GBMs), each subsequent model is fit to the residuals from the
+previous model; data points that are fit poorly in a given model are given more
+weight in the next model [@elith2008]. Random forests and GBMs can provide
+strong predictive performance and fit highly non-linear relationships
+[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
+package [@liaw2002] for \textsf{R} with the default argument values. We fit
+boosted regression tree models with the `gbm` package [@ridgeway2015] for
+\textsf{R}. Based on cross-validation with the **caret** \textsf{R} package, we
+fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate
+(shrinkage parameter) of $0.01$, and all other arguments at their default
+values.
 
@@
 bias, precision, accuracy (a combination of bias and precision), or the ability
-to correctly rank or correlate across populations (e.g. [@walther2005]). Here,
+to correctly rank or correlate across populations [e.g. @walther2005]. Here,
 we measure proportional error, defined as
@@
 ensemble models that simply took a mean of the individual models ranked
-slightly behind the best individual model for estimating status (CMSY; Fig.
+slightly behind the best individual model for estimating fish stock status (CMSY; Fig.
 \ref{performance}a) and had slightly lower correlation but higher accuracy than
-the best individual model at predicting slope (SSCOM; Fig.
+the best individual model at predicting the trends of status (SSCOM; Fig.
 \ref{performance-sim-slope}).
 
+<!--TODO: check labels on Figure S4 – repeat of (d) and missing (f)-->
 The ensemble models were able to improve the predictions of status by
@@
 Ensemble methods provide a useful approach to situations where conservation and
-management decisions must be made on the basis of multiple, potentially
-contrasting estimates of status---a situation common to many ecological
-settings beyond fisheries. Compared to individual models of fish population
-status, ensemble models were consistently the best or among the best across
-three performance dimensions (accuracy, bias, and rank-order performance), two
-response variables (status and trajectory), two datasets (simulated and global
-fisheries), and multiple ensemble model types (from a simple average to machine
-learning models). Our results suggest choosing an ensemble model that allows
-for non-parametric relationships, such as machine learning methods---these
-methods provide added insight into individual model behaviour and generally
-performed the best---but even a simple average of predictions across multiple
-models may be useful.
+ecological resource management decisions must be made on the basis of multiple,
+potentially contrasting estimates of status---a situation common to many
+ecological settings beyond fisheries. Compared to individual models of fish
+population status, ensemble methods were consistently the best or among the best
+across three performance dimensions (accuracy, bias, and rank-order
+performance), two response variables (status and trajectory), two datasets
+(simulated and global fisheries), and multiple ensemble methods (from a
+simple average to machine learning models). Our results suggest choosing an
+superensemble model that allows for non-parametric relationships, such as machine
+learning methods. These methods provide added insight into individual model
+behaviour and generally performed the best, however even a simple average of
+predictions across multiple models may be useful.
 
@@
 representative of the dataset of interest [@knutti2009; @weigel2010].
-Cross-validation within a training dataset will provide a optimistically biased
+Cross-validation within a training dataset will provide an optimistically biased
 impression of predictive performance if the training dataset fundamentally
@@
 hypotheses can contribute useful information for inference [@chamberlin1890]. A
-fundamental difference is that coefficient averaging focuses on a averaging
+fundamental difference is that coefficient averaging focuses on averaging
 *coefficients* whereas ensembles instead average *predictions*. By focussing on
@@
 
-A strength of ensemble models is that they can be tailored or specific response
-variables. For example, we built separate ensemble models of mean $B/B_\mathrm{MSY}$ and
-the slope of $B/B_\mathrm{MSY}$. The same set of model weights or non-linear
-relationships need not hold across different response variables. For example,
-SSCOM contributed little to the GBM ensemble estimate of status at higher
-levels of predicted $B/B_\mathrm{MSY}$, but contributed strongly to estimates of slope
-(TODO FIG). Formally, fitting ensemble models to specific quantities of
-interest provides an additional calibration step [@rykiel1996]. Individual
-models are typically calibrated (i.e. parameters are estimated) based on a
-single response variable; however, this model might then provide a biased or
-inaccurate estimate of other quantities of interest. For example, a
-data-limited fisheries model might calibrate parameters by maximizing
-likelihood with respect to observed and predicted catches but those parameters
-may then provide a biased or inaccurate estimate of the mean and slope of
-$B/B_\mathrm{MSY}$. Ensemble methods provide a final calibration step to a quantity of
-interest. This ensemble calibration could further include a loss function
-tailored to the goals of the model, say placing greater weight on accuracy at
-low status levels than higher values.
+A strength of superensembles is that they can be tailored to specific response
+variables. For example, we built separate ensemble models of mean
+$B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$. The same set of model
+weights or non-linear relationships need not hold across different response
+variables. For example, SSCOM contributed little to the GBM ensemble estimate
+of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed
+strongly to estimates of slope (TODO FIG). Formally, fitting ensemble models to
+specific quantities of interest provides an additional calibration step
+[@rykiel1996]. Individual models are typically calibrated (i.e. parameters are
+estimated) based on a single response variable; however, this model might then
+provide a biased or inaccurate estimate of other quantities of interest. For
+example, a data-limited fisheries model might calibrate parameters by
+maximizing likelihood with respect to observed and predicted catches but those
+parameters may then provide a biased or inaccurate estimate of the mean and
+slope of $B/B_\mathrm{MSY}$. Ensemble methods provide a final calibration step
+to a quantity of interest. This ensemble calibration could further include a
+loss function tailored to the goals of the model, say placing greater weight on
+accuracy at low status levels than higher values.
 
@@
 
-Combining predictions from multiple models via ensemble methods is also broadly
-useful in other subfields of ecology and fisheries science. Predictions about
-extinction risk are widely used at the national (e.g. the US Endangered Species
-Act and the Canadian Species at Risk Act) and international (e.g. the IUCN Red
-List, [@iucn2015]) levels. These risk assessments generally involve fitting
-regression models to outcomes for individual species along with predictors of
-species risk (e.g. [@anderson2011a; @pinsky2011)] (*TODO instead get references
-from official national/international risk assessments?*), or population
-dynamics models to data for individual species (e.g. [@dfo2010]). Both types of
-models are prone to error caused by model-misspecification and therefore
-results are sensitive to decisions about model structure [@brooks2015].
-Although there are options to account for potential model-misspecification in
-determination of species risk (e.g., coefficient averaging, [@burnham2002],
-generalized modeling [@yeakel2011], or semi-parametric methods,
-[@thorson2014]), ensemble methods are a relatively simple way to combine
-predictions in a transparent manner. Beyond estimates of status and trajectory,
-ensemble methods could be used to increase the robustness of spatial
-predictions when designing networks of protected areas [@rassweiler2014], or to
-forecast potential spatial shifts in species distribution given climate impacts
-[@harsch2014].
+Combining predictions from multiple models via superensemble methods is also
+broadly useful in other subfields of ecology and fisheries science. Predictions
+about extinction risk are widely used at the national (e.g. the US Endangered
+Species Act and the Canadian Species at Risk Act) and international (e.g. the
+IUCN Red List, [@iucn2015]) levels. These risk assessments generally involve
+fitting regression models to outcomes for individual species along with
+predictors of species risk [e.g. @anderson2011a; @pinsky2011] (*TODO instead
+get references from official national/international risk assessments?*), or
+population dynamics models to data for individual species [e.g. @dfo2010].
+Both types of models are prone to error caused by model-misspecification and
+therefore results are sensitive to decisions about model structure
+[@brooks2015]. Although there are options to account for potential
+model-misspecification in determination of species risk (e.g., coefficient
+averaging, [@burnham2002], generalized modeling [@yeakel2011], or
+semi-parametric methods, [@thorson2014]), ensemble methods are a relatively
+simple way to combine predictions in a transparent manner. Beyond estimates of
+status and trajectory, ensemble methods could be used to increase the robustness
+of spatial predictions when designing networks of protected areas
+[@rassweiler2014], or to forecast potential spatial shifts in species
+distribution given climate impacts [@harsch2014].
 

### 2015-07-08 text/ms.Rmd
Rework intro yet again

@@
 from multiple models as covariates in a new statistical model. Here we evaluate
-the potential for ensemble and superensemble methods to improve estimates of fish population
+the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population
 status and trajectory. We fit four models of population and exploitation
@@
 `r ram_stocks_n` stock assessments. 
-We find that ensemble and superensemble models
+We find that ensemble methods
 substantially improve estimates of population status and trajectory.
@@
 We found similar improvements when predicting the trajectory of status. When the
-simulation-trained ensembles were applied to global fish stocks, the best
-ensembles still improved all dimensions of performance over the next-best
-individual model. Ensemble and superensemble models have the potential to
+simulation-trained superensembles were applied to global fish stocks, the best
+models still improved all dimensions of performance over the next-best
+individual model. Ensemble methods have the potential to
 improve estimates of population status and trajectory across many taxa beyond
@@
 
-Despite considerable advancements in the systematic collection of ecological
-population data, we still have little data with which to estimate population
-status for most species (IUCN, FAO, amphibian citations etc.). Yet, a large
-portion of these data-limited populations face human-caused pressures and may
-warrant conservation action (IUCN REF). This scenario calls for methods to
-assess the status of populations that rely on limited sources of data. While,
-such methods are increasingly available (e.g. REF, REF), their application faces
-two key challenges: (1) performance is relatively poor compared to more
-data-rich forms of population modelling (e.g. REF) and (2) resource managers are
-often faced with reconciling multiple potentially conflicting estimates of
-population status (e.g. REF).
+**Note if you're editing this: I recently started reworking the paper to
+incorporate 'superensembles' as a term and focus on management/conservation need
+up front instead of methods. As a result, some parts (mostly the introduction)
+are still in rough draft form.**
+
+\bigskip
+
+\noindent The estimation of ecological population status and trajectory are two
+of the most fundamental values to quantify in ecological resource management
+[e.g. @mace2008]. Status and trajectory are fundamental to everything from
+global assessments of extinction risk [e.g. @stuart2004; @schipper2008;
+@juan-jorda2011] to local management of hunting (REF) and fishing (REF). But
+ecological resource managers are often faced with reconciling multiple uncertain
+and potentially conflicting estimates of these quantities---some models may
+suggest a population is increasing in abundance while other may suggest it is
+declining (e.g. REF).
 
 One solution is to take the average or weighted average of an ensemble of model
-predictions. Such ensembles are typically more accurate and less biased
-than individual model estimates and can integrate uncertainty in model
-structure, initial conditions, and parameter estimation [@araujo2007]. For
-instance, REF-international did this and REF-national did this.
+predictions. Such ensembles are typically more accurate and less biased than
+individual model estimates and can integrate uncertainty in model structure,
+initial conditions, and parameter estimation [@araujo2007]. For instance,
+REF-international did this and REF-national did this.
 
 Whereas averages or weighted averages can improve predictions, they may not
-optimally leverage available data. Some models perform better than others in
+optimally leverage available data. The truth does not necessarily lie in the
+middle of multiple model predictions. Some models perform better than others in
 certain conditions and the covariance between model predictions may contain
-information that can both improve predictions and aid understanding of
-individual model performance. We can exploit these characteristics by using the
-predictions from a group of models as inputs into a separate statistical model.
-This technique, sometimes called 'superensemble' modelling, is common in climate
-and weather forecasting. The superensemble is fit to some training dataset where
-outcomes are well known and used to predict on a separate dataset. For example,
-@krishnamurti1999 combined predictions of wind and precipitation in Asian
-monsoons via a superensemble regression fit to observed data. Their
-superensemble considerably outperformed any individual prediction or weighted
-average of predictions in terms of accuracy.
+information that can improve predictions. We can exploit these characteristics
+by using the predictions from a group of models as inputs into a separate
+statistical model. This technique, sometimes called 'superensemble' modelling,
+is common in climate and weather forecasting [@krishnamurti1999]. The
+superensemble is fit to some training dataset where outcomes are well known and
+used to predict on a dataset of interest. For example, @krishnamurti1999
+combined predictions of wind and precipitation in Asian monsoons via a
+superensemble regression fit to observed data. Their superensemble considerably
+outperformed any individual prediction or weighted average of predictions in
+terms of accuracy.
 

### 2015-07-08 text/ms.Rmd
More edits to intro

@@
 
-\noindent The estimation of ecological population status and trajectory are two
-of the most fundamental values to quantify in ecological resource management
-[e.g. @mace2008]. Status and trajectory are fundamental to everything from
-global assessments of extinction risk [e.g. @stuart2004; @schipper2008;
+\noindent Ecological population status and trajectory are two of the most
+fundamental values to quantify in ecological resource management [e.g.
+@mace2008]. Population status and trajectory are key to everything from global
+assessments of extinction risk [e.g. @stuart2004; @schipper2008;
 @juan-jorda2011] to local management of hunting (REF) and fishing (REF). But
-ecological resource managers are often faced with reconciling multiple uncertain
-and potentially conflicting estimates of these quantities---some models may
-suggest a population is increasing in abundance while other may suggest it is
-declining (e.g. REF).
+ecological resource managers are often faced with reconciling multiple
+uncertain and potentially conflicting estimates of these quantities---some
+models may suggest a population is at risk and declining in abundance while
+others may suggest it is not at risk and stable (e.g. REF).
 
@@
 initial conditions, and parameter estimation [@araujo2007]. For instance,
-REF-international did this and REF-national did this.
+REF-international did this and REF-national did this...
 
@@
 
-In fisheries science, a common task is estimating the status and trajectory of
-an exploited fish population. For the majority of fish stocks, we have limited
-fisheries independent data to estimate stock status [@fao2014]. In recent
-years, a number of methods have been proposed to derive population status based
-on the limited information that is broadly available: fisheries catch time
-series and basic knowledge about the productivity of the species 
-[e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014
-investigated the ability for four such models to estimate population status through
-a large-scale simulation experiment. A key finding was that these models
-frequently disagree on population status and any one model has relatively poor
-predictive ability on average (e.g. Fig. \ref{motivate}).
+In fisheries science, a number of models have been proposed to assess
+population status (biomass compared to biomass at maximum sustainable yield,
+$B/B_\mathrm{MSY}$) and trajectory based on the limited data available for the
+majority of stocks: catches and a basic understanding of population
+productivity [e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014
+investigated the performance of four such models through a large-scale
+simulation experiment. The four models frequently disagreed about population
+status and any one model had relatively poor predictive ability on average
+(e.g. Fig. \ref{motivate}).
 
-Here, we estimate population status and trajectory of exploited fish populations
-using ensemble and superensemble models (collectively referred to 'ensemble
-methods'). We explore a variety of approaches applied to both simulated and
-real-world fish stocks and compare predictive performance. We find that
-ensembles, and in particular superensembles, generally improved predictive
-performance of status and trajectory of fish populations over any single
-model---especially when considered across multiple dimensions of predictive
-ability and across multiple datasets. Furthermore, the output from
-superensembles can aid mechanistic understanding of how individual models of
-status and trajectory perform.
+Here, we estimate population status and trajectory of exploited fish
+populations using ensembles and superensembles (collectively 'ensemble
+methods') of these four data-limited models. We explore a variety of approaches
+applied to both simulated and real-world fish stocks and compare predictive
+performance. We find that ensembles, and in particular superensembles,
+generally improved predictive performance of status and trajectory of fish
+populations over any single model---especially when considered across multiple
+dimensions of predictive ability and across multiple datasets. Furthermore, the
+output from superensembles can aid mechanistic understanding of how individual
+models of status and trajectory perform.
 

### 2015-07-08 text/ms.Rmd
Tweak intro

@@
 ecological resource managers are often faced with reconciling multiple
-uncertain and potentially conflicting estimates of these quantities---some
-models may suggest a population is at risk and declining in abundance while
-others may suggest it is not at risk and stable (e.g. REF).
+uncertain and potentially conflicting estimates of these quantities. For
+example, some models may suggest a population is at risk and declining in
+abundance while others may suggest it is not at risk and stable (e.g. REF).
 
@@
 initial conditions, and parameter estimation [@araujo2007]. For instance,
-REF-international did this and REF-national did this...
+REF-IUCN did this and REF-COSEWIC did this...
 
@@
 optimally leverage available data. The truth does not necessarily lie in the
-middle of multiple model predictions. Some models perform better than others in
-certain conditions and the covariance between model predictions may contain
+middle of multiple model predictions, some models perform better than others in
+certain conditions, and the covariance between model predictions may contain
 information that can improve predictions. We can exploit these characteristics
@@
 combined predictions of wind and precipitation in Asian monsoons via a
-superensemble regression fit to observed data. Their superensemble considerably
-outperformed any individual prediction or weighted average of predictions in
-terms of accuracy.
+superensemble regression fit to observed data. Their superensemble was
+considerably more accurate than any individual prediction or weighted average
+of predictions.
 

### 2015-07-09 text/ms.Rmd
Edits to paper
- includes Kristin's edits

@@
 \noindent
-^5^NOAA/NMFS/NEFSC, Ecosystem Assessment Program, 
+^5^NOAA/Northeast Fisheries Science Center, Ecosystem Assessment Program, 166 Water St., 
 Woods Hole, MA, 02543, USA
@@
 the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population
-status and trajectory. We fit four models of population and exploitation
-dynamics to data-limited populations and combine their estimates of the mean and
+status and trajectory. We fit four data-limited models of population and exploitation
+dynamics to marine fish stocks and combine their estimates of the mean and
 slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an
 ensemble average and three superensembles: a linear model, a random forest, and
-a boosted regression tree. We build our models on a simulated dataset of 5760
-fish stocks and test them with cross-validation and a global database of 
-`r ram_stocks_n` stock assessments. 
-We find that ensemble methods
-substantially improve estimates of population status and trajectory.
+a boosted regression tree. 
+We selected data-limited models for their applicability across the majority of stocks world wide.
+We build our models on a simulated dataset of 5760
+fish stocks and tested them with cross-validation and a global database of 
+`r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
+We find that ensemble methods substantially improve estimates of population status and trajectory.
 Machine-learning superensembles performed the best for population status: accuracy improved
@@
 We found similar improvements when predicting the trajectory of status. When the
-simulation-trained superensembles were applied to global fish stocks, the best
-models still improved all dimensions of performance over the next-best
+simulation-trained superensembles were applied to catch data for global fish stocks, the best
+superensembles improved all dimensions of performance over the next-best
 individual model. Ensemble methods have the potential to
@@
 accurate models, and built on a dataset sufficiently representative of the
-populations they are applied to.
+populations to which they are applied.
 
@@
 
-**Note if you're editing this: I recently started reworking the paper to
-incorporate 'superensembles' as a term and focus on management/conservation need
-up front instead of methods. As a result, some parts (mostly the introduction)
-are still in rough draft form.**
+Population status and trajectory are two of the most fundamental values to quantify in ecological resource management [e.g.
+@mace2008; @iucn2015].
+These properties are key to everything from global assessments of extinction risk [e.g.
+@stuart2004; @schipper2008; @juan-jorda2011] to setting hunting (REF) and fishing limits [e.g.
+@worm2009] (REFs).
+However, ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g.
+@branch2011] (REFs).
+Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
 
-\bigskip
+One solution is to take the average or weighted average of an ensemble of model predictions.
+Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
+This approach forms the basis of many machine learning methods [e.g.
+@dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.
+@murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006].
+In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g.
+@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trajectory [e.g.
+@brodziak2010] COSEWIC-IUCN REFS.
 
-\noindent Ecological population status and trajectory are two of the most
-fundamental values to quantify in ecological resource management [e.g.
-@mace2008]. Population status and trajectory are key to everything from global
-assessments of extinction risk [e.g. @stuart2004; @schipper2008;
-@juan-jorda2011] to local management of hunting (REF) and fishing (REF). But
-ecological resource managers are often faced with reconciling multiple
-uncertain and potentially conflicting estimates of these quantities. For
-example, some models may suggest a population is at risk and declining in
-abundance while others may suggest it is not at risk and stable (e.g. REF).
+<!--
+Biological ensemble modeling to evaluate potential futures of living marine resources
+Anna Gårdmark1,8, Martin Lindegren2, Stefan Neuenfeldt2, Thorsten Blenckner3, Outi Heikinheimo4, Bärbel Müller-Karulis3,5, Susa Niiranen3, Maciej T.
+Tomczak3, Eero Aro4, Anders Wikström6, and Christian Möllmann7
+Read More: http://www.esajournals.org/doi/abs/10.1890/12-0267.1
+-->
 
-One solution is to take the average or weighted average of an ensemble of model
-predictions. Such ensembles are typically more accurate and less biased than
-individual model estimates and can integrate uncertainty in model structure,
-initial conditions, and parameter estimation [@araujo2007]. For instance,
-REF-IUCN did this and REF-COSEWIC did this...
-
-Whereas averages or weighted averages can improve predictions, they may not
-optimally leverage available data. The truth does not necessarily lie in the
-middle of multiple model predictions, some models perform better than others in
-certain conditions, and the covariance between model predictions may contain
-information that can improve predictions. We can exploit these characteristics
-by using the predictions from a group of models as inputs into a separate
-statistical model. This technique, sometimes called 'superensemble' modelling,
-is common in climate and weather forecasting [@krishnamurti1999]. The
-superensemble is fit to some training dataset where outcomes are well known and
-used to predict on a dataset of interest. For example, @krishnamurti1999
-combined predictions of wind and precipitation in Asian monsoons via a
-superensemble regression fit to observed data. Their superensemble was
-considerably more accurate than any individual prediction or weighted average
-of predictions.
+Whereas averages or weighted averages may improve predictions over any single model, they may not optimally leverage available data.
+The truth does not necessarily lie in the middle of multiple model predictions, some models may perform better than others in certain conditions, and the covariance between models may contain information that can improve predictive accuracy.
+We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
+This technique, sometimes called 'superensemble' modelling, is common in climate and weather forecasting [e.g.
+@krishnamurti1999; @yun2005; @mote2015].
+The superensemble is fit to some training dataset where outcomes are well known and used to predict on a dataset of interest.
+For example, @krishnamurti1999 combined predictions of wind and precipitation in Asian monsoons via a superensemble regression fit to observed data.
+Their superensemble was considerably more accurate than any individual prediction or weighted average of predictions.
 
@@
 
-In fisheries science, a number of models have been proposed to assess
-population status (biomass compared to biomass at maximum sustainable yield,
-$B/B_\mathrm{MSY}$) and trajectory based on the limited data available for the
-majority of stocks: catches and a basic understanding of population
-productivity [e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014
-investigated the performance of four such models through a large-scale
-simulation experiment. The four models frequently disagreed about population
-status and any one model had relatively poor predictive ability on average
-(e.g. Fig. \ref{motivate}).
+In fisheries science, the gold standard for determining status and trajectory
+of exploited fish populations are stock assessments that
+incorporate all available data (e.g., catches, size and age distributions,
+surveys, and tagging information) to quantify values such as 
+the biomass that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
+However, these kinds of data are not available for the majority of stocks
+[@fao2014]. Therefore, a number of models have been proposed to assess
+$B/B_\mathrm{MSY}$ based on the limited data available for the majority of
+stocks: catches and a basic understanding of population productivity [e.g.
+@vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the
+performance of four data-limited models through a large-scale simulation
+experiment. The four models frequently disagreed about population status and
+any one model had relatively poor predictive ability on average (e.g. Fig.
+\ref{motivate}).
 
@@
 populations over any single model---especially when considered across multiple
-dimensions of predictive ability and across multiple datasets. Furthermore, the
-output from superensembles can aid mechanistic understanding of how individual
-models of status and trajectory perform.
+dimensions of predictive ability. Furthermore, the output from superensembles
+can aid mechanistic understanding of how individual models of status and
+trajectory perform.
 
@@
 We first developed and tested ensemble methods with a fully factorial simulated
-dataset [@rosenberg2014]. These previously published simulation models included
-three life histories: small pelagic, demersal, large pelagic; three levels of
-how depleted biomass was at the start of the dataset compared to carrying
-capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation
-dynamics: a constant exploitation rate, a exploitation rate that is coupled
-with biomass to mimic an open-access single-species fishery, a scenario where
-exploitation rate increases to a fixed level, and a 'roller-coaster' scenario
-where the exploitation rate increases and then decreases. Process noise was
+dataset [@rosenberg2014] with precisely known status. These published simulations included three life
+histories: small pelagic, demersal, and large pelagic; three levels of how
+depleted biomass was at the start of the dataset compared to carrying capacity:
+100%, 70%, and 40% of carrying capacity; and four exploitation dynamics: a
+constant exploitation rate, an exploitation rate coupled with biomass to mimic
+an open-access single-species fishery, a scenario where exploitation rate
+increases to a fixed level, and a 'roller-coaster' scenario where the
+exploitation rate increases and then decreases cyclically. Process noise was
 introduced to the models through two levels of multiplicative recruitment
-variability: $N(0, 0.2)$ and $N(0, 0.6)$, and that process noise was either
-uncorrelated or had a first-order autoregressive correlation of 0.6. The
+variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either
+uncorrelated or had a first-order autoregressive correlation of $0.6$. The
 simulation included a scenario without observation error and with
@@
 $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population
-growth rate, $B_0$ represents unfished biomass or carrying capacity, and $C$
+growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$
 represents catch. @rosenberg2014 provide a full summary of
@@
 
-* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
-  biomass dynamics [@martell2013]. It requires a prior distribution for $r$ and
-  assigns a prior to the relative proportion of biomass at the end compared to
-  unfished biomass (depletion) based on the percentage of maximum catch at the
-  end of the time series.
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
+  dynamics [@martell2013]. It requires a prior distribution for $r$ and assigns
+  a prior to the relative proportion of biomass at the end compared to unfished
+  biomass (depletion) based on the percentage of maximum catch at the end of the
+  time series. Our version of the model was modified from @martell2013 to
+  generate biomass trends from all viable $r$-$K$ pairs and produced an estimate of
+  $B/B_\mathrm{MSY}$ from the geometric mean trend.
 
@@
 \hat{b}_\mathrm{SSCOM}, \, \hat{b}_\mathrm{mPRM}, \,
-S(0.2), \, S(0.5), \, | \, \theta \right)
+S(0.2), \, S(0.5) \, | \, \theta \right)
 $$
@@
 
-We fit the linear model ensemble with second-order interactions. We fit two
-machine learning ensemble models based on regression trees. Regression trees
+We fit the linear model superensemble with second-order interactions. We fit two
+machine learning superensemble models based on regression trees. Regression trees
 sequentially determine what value of a predictor best splits the response data
@@
 cross validation to test predictive performance: we randomly divided the
-dataset into three sets, built ensemble models on two-thirds of the data, and
+dataset into three sets, built superensemble models on two-thirds of the data, and
 evaluated predictive performance on the remaining third. We repeated this
@@
 
-Ensemble models, and in particular the machine learning ensemble models (random
+Ensemble methods, and in particular the machine learning superensemble models (random
 forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig.
 \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the
-individual models, machine learning ensembles improved accuracy (median
+individual models, machine learning superensembles improved accuracy (median
 absolute proportional error) by 
@@
 `r mean_sim$mach_mre_range` (Fig. \ref{performance}a). 
-These ensembles also had better ability to distinguish if
+These superensembles also had better ability to distinguish if
 simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
 \ref{roc-sim}). Results were similar when predicting trajectory: compared to
-individual models, machine learning ensembles improved accuracy by 
+individual models, machine learning superensembles improved accuracy by 
 `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from 
@@
 <!--TODO: check labels on Figure S4 – repeat of (d) and missing (f)-->
-The ensemble models were able to improve the predictions of status by
+The superensemble models were able to improve the predictions of status by
 exploiting the best properties of individual models, the covariance between
@@
 approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction
-clusters (Fig. \ref{partial-sim}). Ensembles also exploited the covariance
+clusters (Fig. \ref{partial-sim}). Superensembles also exploited the covariance
 between individual model predictions. For instance, both the linear model and
@@
 status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l). The
-addition of spectral density covariates helped the ensemble models correctly
+addition of spectral density covariates helped the superensemble models correctly
 predict higher status values (Fig. \ref{hexagon}g, h) by exploiting
@@
 
-When applied to the stock assessment database, the ensemble models---trained
+When applied to the stock assessment database, the superensemble models---trained
 exclusively on the simulated dataset---generally performed as well or better
 than the best individual models. The mean, random forest, and GBM ensembles
-even outperformed the mPRM method which is trained directly on the RAM Legacy Stock
+outperformed the mPRM method which is trained directly on the RAM Legacy Stock
 Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
-Compared to the individual models, the machine learning ensembles increased
+Compared to the individual models, the machine learning superensembles increased
 accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, 
@@
 (simulated and global fisheries), and multiple ensemble methods (from a
-simple average to machine learning models). Our results suggest choosing an
+simple average to machine learning superensembles). Our results suggest choosing an
 superensemble model that allows for non-parametric relationships, such as machine
@@
 behaviour and generally performed the best, however even a simple average of
-predictions across multiple models may be useful.
+predictions across multiple models may be a robust solution compared to picking
+any one model.
 
@@
 @grueber2011]. The ensemble methods described in this paper share similarities
-with coefficient averaging but differ in other important ways. Ensemble models
+with coefficient averaging but differ in other important ways. Ensemble methods
 and coefficient averaging share the long-held notion that multiple working
@@
 
-A strength of superensembles is that they can be tailored to specific response
-variables. For example, we built separate ensemble models of mean
+A strength of superensembles is that they can be tailored to predict specific response
+variables. For example, we built separate superensemble models of mean
 $B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$. The same set of model
 weights or non-linear relationships need not hold across different response
-variables. For example, SSCOM contributed little to the GBM ensemble estimate
+variables. For example, SSCOM contributed little to the GBM superensemble estimate
 of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed
-strongly to estimates of slope (TODO FIG). Formally, fitting ensemble models to
+strongly to estimates of slope (TODO FIG). Formally, fitting superensemble models to
 specific quantities of interest provides an additional calibration step
@@
 
-* Coilin: How ensembles can help understand mechanistic underpinnings of individual 
-  models
-    * discover surprises: counterintuitive relationships and interactions
-    * can dig into surprising conditions and learn about why certain models fail; 
-      potentially learn how to improve models
-    * can learn under what conditions certain models perform well
-    * pick out an example
+*TODO: Coilin: How ensembles can help understand mechanistic underpinnings of individual models*
+Beyond predictive performance, superensembles can help understand mechanistic
+underpinnings of individual models. 
+First, at a most basic level, superensembles that allow for nonlinear
+relationships can describe at what ranges of a prediction a certain model is
+most informative. For example, we can see that SSCOM is most effective at
+distinguishing low values of $B/B_\mathrm{MSY}$ (Fig. S4).
+Second, the interaction between superensemble predictors can highlight
+under what conditions certain models perform best. 
+For instance, SAY SOMETHING ABOUT FIG S5 or FIG S6.
+*discover surprises: counterintuitive relationships and interactions*
+*can dig into surprising conditions and learn about why certain models fail; potentially learn how to improve models*
+*can learn under what conditions certain models perform well*
 
@@
 about extinction risk are widely used at the national (e.g. the US Endangered
-Species Act and the Canadian Species at Risk Act) and international (e.g. the
-IUCN Red List, [@iucn2015]) levels. These risk assessments generally involve
+Species Act and the Canadian Species at Risk Act) and international [e.g. the
+IUCN Red List, @iucn2015] levels. These risk assessments generally involve
 fitting regression models to outcomes for individual species along with
-predictors of species risk [e.g. @anderson2011a; @pinsky2011] (*TODO instead
-get references from official national/international risk assessments?*), or
+predictors of species risk [e.g. @anderson2011a; @pinsky2011], or
 population dynamics models to data for individual species [e.g. @dfo2010].
@@
 
--->
-
 # Authors' contributions
@@
 approval for publication.*
+-->
 

### 2015-07-09 text/ms.Rmd
More edits

@@
 We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
-This technique, sometimes called 'superensemble' modelling, is common in climate and weather forecasting [e.g.
-@krishnamurti1999; @yun2005; @mote2015].
+This technique, sometimes called 'superensemble' modelling [@krishnamurti1999], is common in climate and weather forecasting [e.g.
+@yun2005; @mote2015].
 The superensemble is fit to some training dataset where outcomes are well known and used to predict on a dataset of interest.
@@
   time series. Our version of the model was modified from @martell2013 to
-  generate biomass trends from all viable $r$-$K$ pairs and produced an estimate of
-  $B/B_\mathrm{MSY}$ from the geometric mean trend.
+  generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of
+  $B/B_\mathrm{MSY}$ from the median trend.
 
@@
 underpinnings of individual models. 
-First, at a most basic level, superensembles that allow for nonlinear
+First, superensembles that allow for nonlinear
 relationships can describe at what ranges of a prediction a certain model is

### 2015-07-09 text/ms.Rmd
More ms edits

@@
 Sean C. Anderson^1\*^, 
+Andrew B. Cooper^1^
+Olaf P. Jensen^4^, 
+C\'{o}il\'{i}n Minto^6^, 
+James T. Thorson^11^,
+Jessica C. Walsh^1^,
 Jamie Afflerbach^2^, 
 Mark Dickey-Collas^3^, 
-Olaf P. Jensen^4^, 
 Kristin M. Kleisner^5^, 
 Catherine Longo^2^, 
-C\'{o}il\'{i}n Minto^6^, 
 Giacomo Chato Osio^7^, 
@@
 Elizabeth R. Selig^10^, 
-James T. Thorson^11^,
-Jessica C. Walsh^1^,
-Andrew B. Cooper^1^
-*(order to be determined and others may be added)*
+*(order to be determined and superscripts out of order)*
 
@@
 between models and may not optimally reduce bias and improve accuracy.
-Superensemble models, commonly used in the fields of climate and weather
-science, may provide a superior solution. Superensembles use the predictions
+Superensemble models, commonly used for climate and weather
+forecasting, may provide a superior solution. Superensembles use the predictions
 from multiple models as covariates in a new statistical model. Here we evaluate
@@
 
+\bigskip
 <!-- up to 12 keywords -->
 \noindent 
-Keywords: data-limited fisheries, ensemble methods, model averaging, population dynamics, sustainable resource management
+Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
 

### 2015-07-09 analysis/8-main-figs.R
Switch to sans serif fonts

@@ -231,7 +231,7 @@ par(mfrow = c(1, 2))
 par(mgp = c(1.5, 0.5, 0), las = 1, tck = -0.015,
   oma = c(2.8, 0.5, 1.5, .5), cex = 0.8, mar = c(0, 3, 0, 0),
   xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
-par(family="serif")
+#par(family="serif")
 
 perf(d_sim_perf_wide, label = "(a) Simulation")
 
@@ -285,7 +285,7 @@ pdf("../figs/performance-slope-sim.pdf", width = 5, height = 4)
 par(mgp = c(1.5, 0.5, 0), las = 1, tck = -0.015,
   oma = c(2.8, 0.5, 1.5, .5), cex = 0.8, mar = c(0, 3, 0, 0),
   xaxs = "i", yaxs = "i", col.axis = "grey50", col.lab = "grey50")
-par(family="serif")
+#par(family="serif")
 
 pal_df <- data_frame(mre = seq(-max(abs(d_slope_error$mre)+0.02),
   max(abs(d_slope_error$mre)+0.02),

### 2015-07-09 analysis/8-motivate-fig.R

@@ -74,7 +74,7 @@ ylim <- range(c(dl_dat$b_bmsyiq25, dl_dat$b_bmsyiq75))
 pdf("../figs/motivate.pdf", width = 4.6, height = 2.9)
 par(mar = c(3, 3.4, .5, 4.5), cex = 0.8, oma = c(0, 0, 0, 0), tck = -0.015,
   mgp = c(2, 0.5, 0), col.axis = "grey40", las = 1)
-par(family="serif")
+#par(family="serif")
 plot(1, 1, type = "n", xlim = xlim, ylim = ylim, xlab = "", ylab = "", axes = FALSE)
 abline(h = 1, lty = 2, col = "grey40", lwd = 2)
 plyr::d_ply(dl_dat, "method", plot_method)

### 2015-07-09 analysis/0-ensemble-functions.R
Add 'superensemble' to figures

@@ -129,7 +129,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   par(mfrow = c(rows, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
     oma = oma, cex = 0.8, mar = c(0, 0, 0, 0),
     xaxs = "i", yaxs = "i", col.axis = "grey60", col.lab = "grey50")
-  par(family="serif")
+  #par(family="serif")
 
   hexcol1 <- RColorBrewer::brewer.pal(9, "Blues")
   #hexcol1 <- c("#FFFFFF", "#6A4A3C")
@@ -250,7 +250,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
     box(col = "grey60")
     add_label(-0.01, 0.08, paste0("(", letters[m], ") ", unique(filter(dat, order == m)$clean_method)),
       #col = "grey20", add_bg = ifelse(m %in% 1:4, FALSE, TRUE))
-      col = "grey20", add_bg = FALSE)
+      col = "grey20", add_bg = FALSE, cex = 0.95)
     if (m %in% c(1, 5, 9)) axis(2, at = axis_ticks[-length(axis_ticks)], col = "grey60")
     if (m %in% (rows * 4 - 3):(rows * 4)) axis(1, at = axis_ticks, col = "grey60")
   })

### 2015-07-09 analysis/8-main-figs.R

@@ -10,7 +10,7 @@ clean_names <- dplyr::data_frame(
   method = c("CMSY", "COMSIR", "mPRM", "SSCOM",
     "gbm_ensemble", "rf_ensemble", "lm_ensemble", "mean_ensemble"),
   clean_method = c("CMSY", "COM-SIR", "mPRM", "SSCOM",
-    "GBM Ensemble", "RF Ensemble", "LM Ensemble", "Mean Ensemble"),
+    "GBM Superensemble", "RF Superensemble", "LM Superensemble", "Mean Ensemble"),
   order = c(1, 2, 4, 3, 8, 7, 6, 5),
   label_fudge_x = c(
     0, -0.03,0,-0.04,
@@ -43,7 +43,7 @@ d_sim_perf_summ <- d_sim_perf_long %>%
     l = quantile(value, 0.25),
     u = quantile(value, 0.75)) %>%
   as.data.frame
-d_sim_perf_summ$text_col <- ifelse(grepl("Ensemble", d_sim_perf_summ$clean_method),
+d_sim_perf_summ$text_col <- ifelse(grepl("ensemble", d_sim_perf_summ$clean_method, ignore.case = TRUE),
   "grey20", "grey50")
 d_sim_perf_wide <- reshape2::dcast(d_sim_perf_summ, clean_method + order ~ variable,
   value.var = "m")
@@ -80,7 +80,7 @@ re_ram_sum_long <- reshape2::melt(re_ram_sum, id.vars = "clean_method") %>%
   inner_join(clean_names) %>%
   rename(m = value) %>%
   mutate(l = m, u = m) %>%
-  mutate(text_col = ifelse(grepl("Ensemble", clean_method), "grey20", "grey50"))
+  mutate(text_col = ifelse(grepl("ensemble", clean_method, ignore.case = TRUE), "grey20", "grey50"))
 re_ram_sum_long$clean_method <- reorder(re_ram_sum_long$clean_method, re_ram_sum_long$order)
 re_ram_sum_long$label_fudge_x <- 0
 re_ram_sum_long$label_fudge_y <- 0
@@ -98,7 +98,7 @@ d_sim_mean <- d_sim %>% filter(type == "mean") %>%
 rocs_sim_mean <- d_sim_mean %>% group_by(clean_method, order) %>%
   do({get_roc(true = .$above1, est = .$bbmsy_est)}) %>%
   as.data.frame
[…]

### 2015-07-09 analysis/5-simulation-ensembles.R
Add partial dependence plot of slope model

@@ -136,6 +136,10 @@ m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
   data = d_mean, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
 d_mean$gbm_pred <- exp(gbm::predict.gbm(m, n.trees = 2000L))
 
+m_slope <- gbm::gbm(bbmsy_true_slope ~ CMSY + COMSIR + mPRM + SSCOM +
+    spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
+  data = d_slope, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
+
 p1 <- ggplot(d_mean, aes(CMSY, gbm_pred)) + geom_point()
 p2 <- ggplot(d_mean, aes(COMSIR, gbm_pred)) + geom_point()
 p3 <- ggplot(d_mean, aes(mPRM, gbm_pred)) + geom_point()
@@ -200,18 +204,32 @@ partial <- plyr::ldply(seq_len(nvar), function(i) {
   dd
 })
 
+partial_slope <- plyr::ldply(seq_len(nvar), function(i) {
+  dd <- gbm::plot.gbm(m_slope, i.var = i, return.grid = TRUE)
+  dd$predictor <- names(dd)[1]
+  names(dd)[1] <- "predictor_value"
+  dd
+})
+
 partial_names <- data_frame(
   predictor = c("CMSY", "COMSIR", "mPRM", "SSCOM",
     "spec_freq_0.05", "spec_freq_0.2"),
   predictor_label = c("(a) CMSY", "(b) COM-SIR", "(d) mPRM", "(c) SSCOM",
-    "(e) Long-term spectral density", "(d) Short-term spectral density"))
+    "(f) Long-term spectral density", "(e) Short-term spectral density"))
 partial <- inner_join(partial, partial_names)
+partial_slope <- inner_join(partial_slope, partial_names)
 # partial dependence plot:
 p <- ggplot(partial, aes(predictor_value, y)) + geom_line() +
   facet_wrap(~predictor_label, scales = "free_x") + ylim(0.3, 1.6) + theme_bw() +
   ylab("Expected B/B_MSY integrating out other predictors") +
   xlab("Predictor value")
 ggsave("../figs/partial-sim.pdf", width = 7, height = 5)
+
[…]

### 2015-07-09 text/ms.Rmd
More edits to paper

@@
 Sean C. Anderson^1\*^, 
-Andrew B. Cooper^1^
+Andrew B. Cooper^1^,
 Olaf P. Jensen^4^, 
@@
 Andrew A. Rosenberg^9^, 
-Elizabeth R. Selig^10^, 
-*(order to be determined and superscripts out of order)*
+Elizabeth R. Selig^10^
+
+\noindent
+\textit{order to be determined and superscripts out of order}
 
@@
 a boosted regression tree. 
-We selected data-limited models for their applicability across the majority of stocks world wide.
+We selected data-limited models for their applicability across the majority of fish stocks world wide.
 We build our models on a simulated dataset of 5760
-fish stocks and tested them with cross-validation and a global database of 
+stocks and tested them with cross-validation and a global database of 
 `r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
@@
 These properties are key to everything from global assessments of extinction risk [e.g.
-@stuart2004; @schipper2008; @juan-jorda2011] to setting hunting (REF) and fishing limits [e.g.
-@worm2009] (REFs).
+@stuart2004; @schipper2008] to assessing hunting and fishing exploitation levels [e.g.
+@worm2009; @juan-jorda2011].
 However, ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g.
-@branch2011] (REFs).
+@brodziak2010; @branch2011].
 Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
@@
 of exploited fish populations are stock assessments that
-incorporate all available data (e.g., catches, size and age distributions,
+incorporate all available data (e.g. catches, size and age distributions,
 surveys, and tagging information) to quantify values such as 
 the biomass that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
-However, these kinds of data are not available for the majority of stocks
+However, these kinds of data are not available for the majority of fish
+populations
 [@fao2014]. Therefore, a number of models have been proposed to assess
 $B/B_\mathrm{MSY}$ based on the limited data available for the majority of
-stocks: catches and a basic understanding of population productivity [e.g.
+fish stocks: catches and a basic understanding of population productivity [e.g.
 @vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the
@@
   correction needed in @costello2012 for deriving aggregate
-  estimates of mean status across multiple stocks.
+  estimates of median status across multiple stocks.
 
@@
 and Spearman's rank-order correlation to measure the ability to correctly rank
-populations.
+populations. Code to reproduce our analysis is available at
+<https://github.com/datalimited/ensembles>.
 
@@
 
-<!--TODO: check labels on Figure S4 – repeat of (d) and missing (f)-->
 The superensemble models were able to improve the predictions of status by
@@
 of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed
-strongly to estimates of slope (TODO FIG). Formally, fitting superensemble models to
+strongly to estimates of slope (Fig. \ref{fig:partial-sim-slope}). Formally,
+fitting superensemble models to
 specific quantities of interest provides an additional calibration step
@@
 
-TODO
-
-# Funding
-
-TODO
+We thank members of Phase I of the working group 'INSERT NAME HERE' who
+contributed to developing the data-limited methods and simulations used in our
+analysis.
+We thank J.A. Hutchings for helpful discussions on ensemble methods for
+assessing population status.
+We thank the Gordon and Betty Moore Foundation for funding of the working group
+'INSERT NAME HERE'.
 

### 2015-07-14 analysis/9-output-values.R
Fix percent change in MARE
- bug noted by Coilin
- was 'folds' original, but now quoted as percent change in paper

@@ -16,8 +16,8 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   ind <- dat %>% filter(ensemble == FALSE)
 
   x$mach_vs_ind_mare_fold <- c(
-    round(min(ind$mare) / max(mach$mare), digits_fold),
-    round(max(ind$mare) / min(mach$mare), digits_fold)
+    round((min(ind$mare) - max(mach$mare)) / max(mach$mare), digits_fold),
+    round((max(ind$mare) - min(mach$mare)) / min(mach$mare), digits_fold)
   ) * 100
 
 #   x$mach_vs_ind_bias <- c(

### 2015-07-14 analysis/9-output-values.R
Lowercase the 'ensemble' grep

@@ -6,7 +6,7 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   collapse = "--", mre_digits = 2) {
 
   dat <- dat %>%
-    mutate(ensemble = ifelse(grepl("Ensemble", clean_method), TRUE, FALSE)) %>%
+    mutate(ensemble = ifelse(grepl("ensemble", clean_method), TRUE, FALSE)) %>%
     mutate(machine = ifelse(grepl("GBM|RF", clean_method), TRUE, FALSE))
 
   x <- list()

### 2015-07-14 text/ms.Rmd
Rework paper after co-author comments

@@
 ```
+
+<!--notes:
+http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1523-1739/homepage/ForAuthors.html
+
+con bio: American spelling
+
+Contributed Papers: 6000 words
+
+Submit your cover page as a separate document. It should not be part of the
+manuscript itself. The cover page must include the title of the paper; a running
+head (short title of 40 or fewer characters); a list of 5-8 keywords; word count
+(all text from the first word of the Abstract through the last word of the
+Literature Cited but not including table or figure legends or the body of
+tables); authors' complete mailing addresses (including postal code) at the time
+the work was conducted and present addresses if different; name and complete
+address (including email) of the person to whom correspondence should be sent;
+and text of your acknowledgments section.
+
+Contributed Papers, Research Notes, and Conservation Practice and Policy papers
+should contain the following sections in the following order: Abstract,
+Introduction, Methods, Results, Discussion, Supporting Information paragraph (if
+there are online appendices), Literature Cited, tables, figure legends, and
+figures. Do not combine sections (e.g., Results and Discussion). The
+Acknowledgments section will be added to the body of the paper after the
+manuscript has been accepted. Do not number section headings or subheadings. Do
+not include a Conclusion section (conclusions are part of the Discussion).
+
+You may have color figures in the online version and gray-scale figures in
+print. However, reference to color cannot be made in the figure legend or in the
+text, and elements in the gray-scale version must be distinguishable. Supply
+separate files for color and gray-scale figures.
+
+-->
 
@@
 Jamie Afflerbach^2^, 
-Mark Dickey-Collas^3^, 
+Mark Dickey-Collas^3,12^, 
 Kristin M. Kleisner^5^, 
@@
 \noindent
-^3^International Council for the Exploration of the Sea
+^3^International Council for the Exploration of the Sea, 
+H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark
 
 \noindent
-^5^NOAA/Northeast Fisheries Science Center, Ecosystem Assessment Program, 166 Water St., 
-Woods Hole, MA, 02543, USA
+^4^Institute of Marine & Coastal Sciences, Rutgers University, 
+71 Dudley Road, New Brunswick, NJ, 08901-8525, USA
+
+\noindent
+^5^NOAA/Northeast Fisheries Science Center, Ecosystem Assessment Program, 
+166 Water St., Woods Hole, MA, 02543, USA
 
@@
 \noindent
+^12^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark
+
+\noindent
 ^\*^Corresponding author: Sean C. Anderson,
@@
 
-<!--up to 350 words-->
+<!--up to 300 words-->
 
-Ecological resource managers are often faced with reconciling multiple potentially
-conflicting estimates of population status and trajectory. One solution is to
-average these predictions in an ensemble, but this approach ignores covariance
-between models and may not optimally reduce bias and improve accuracy.
-Superensemble models, commonly used for climate and weather
+Ecological resource managers are often faced with reconciling multiple
+potentially conflicting estimates of population status and trajectory. One
+solution is to average these predictions in an ensemble, but this approach
+ignores covariance between models and may not optimally reduce bias and improve
+accuracy. Superensemble models, commonly used for climate and weather
 forecasting, may provide a superior solution. Superensembles use the predictions
 from multiple models as covariates in a new statistical model. Here we evaluate
-the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population
-status and trajectory. We fit four data-limited models of population and exploitation
-dynamics to marine fish stocks and combine their estimates of the mean and
-slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an
-ensemble average and three superensembles: a linear model, a random forest, and
-a boosted regression tree. 
-We selected data-limited models for their applicability across the majority of fish stocks world wide.
-We build our models on a simulated dataset of 5760
+the potential for ensemble and superensemble models ('ensemble methods') to
+improve estimates of fish population status and trajectory. We fit four
+data-limited models of population and exploitation dynamics to marine fish
+stocks and combine their estimates of the mean and slope of biomass at maximum
+sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three
+superensembles: a linear model, a random forest, and a boosted regression tree. 
+We selected data-limited models for their applicability across the majority of
+fish stocks world wide. We build our models on a simulated dataset of 5760
 stocks and tested them with cross-validation and a global database of 
-`r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
-We find that ensemble methods substantially improve estimates of population status and trajectory.
-Machine-learning superensembles performed the best for population status: accuracy improved
-`r mean_sim$mach_vs_ind_mare_fold`%, correlation between predicted and true
-status improved from 
+`r ram_stocks_n` stock assessments that have also been assessed with more
+data-intensive stock assessment models. We find that ensemble methods
+substantially improve estimates of population status and trajectory.
+Machine-learning superensembles performed the best for population status:
+accuracy improved
+`r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted
+and true status improved from 
 `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median
@@
 `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. 
-We found similar improvements when predicting the trajectory of status. When the
-simulation-trained superensembles were applied to catch data for global fish stocks, the best
-superensembles improved all dimensions of performance over the next-best
-individual model. Ensemble methods have the potential to
-improve estimates of population status and trajectory across many taxa beyond
-fish. However, they must be carefully tested, formed from a diverse set of
-accurate models, and built on a dataset sufficiently representative of the
-populations to which they are applied.
+We found similar improvements when predicting the trajectory of status and when
+applying the simulation-trained superensembles to catch data for global fish
+stocks. Ensemble methods have the potential to improve estimates of population
+status and trajectory across many taxa beyond fish. However, they must be
+carefully tested, formed from a diverse set of accurate models, and built on a
+dataset sufficiently representative of the populations to which they are
+applied.
 
 \bigskip
-<!-- up to 12 keywords -->
 \noindent 
 Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
+<!--Include on the cover page five to eight words or phrases that will be useful for indexing and literature searches.
+Do not use words in the title as keywords, and avoid general terms such as conservation.-->
 
@@
 
-Population status and trajectory are two of the most fundamental values to quantify in ecological resource management [e.g.
-@mace2008; @iucn2015].
-These properties are key to everything from global assessments of extinction risk [e.g.
-@stuart2004; @schipper2008] to assessing hunting and fishing exploitation levels [e.g.
-@worm2009; @juan-jorda2011].
-However, ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g.
-@brodziak2010; @branch2011].
-Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
+Population status and trajectory are two of the most fundamental values to
+quantify in conservation biology [e.g. @mace2008; @iucn2015]. These properties
+are key to everything from global assessments of extinction risk [e.g.
+@stuart2004; @schipper2008] to assessing the sustainability of hunting and
+fishing exploitation levels [e.g. @worm2009; @juan-jorda2011]. However,
+conservation and ecological resource managers are often faced with reconciling
+multiple uncertain and potentially conflicting estimates of status and
+trajectory [e.g. @brodziak2010; @branch2011]. Some models may suggest a
+population is at risk and declining in abundance while others may suggest it is
+not at risk and stable.
 
-One solution is to take the average or weighted average of an ensemble of model predictions.
-Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
+One solution is to take the average or weighted average of an ensemble of model
+predictions. Such ensembles are typically more accurate and less biased than
+individual model estimates and can integrate uncertainty in model structure,
+initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
 This approach forms the basis of many machine learning methods [e.g.
-@dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.
-@murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006].
-In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g.
-@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trajectory [e.g.
-@brodziak2010] COSEWIC-IUCN REFS.
+@dietterich2000], has helped reconcile climate forecasts from dozens of models
+[e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning
+signs of malaria outbreaks [@thomson2006]. In ecology, ensemble methods are
+frequently used to improve species distribution modelling [e.g. @araujo2007;
+@breiner2015] and indeed have been used to combine estimates of population
+status and trajectory [e.g. @brodziak2010].
 
-<!--
-Biological ensemble modeling to evaluate potential futures of living marine resources
-Anna Gårdmark1,8, Martin Lindegren2, Stefan Neuenfeldt2, Thorsten Blenckner3, Outi Heikinheimo4, Bärbel Müller-Karulis3,5, Susa Niiranen3, Maciej T.
-Tomczak3, Eero Aro4, Anders Wikström6, and Christian Möllmann7
-Read More: http://www.esajournals.org/doi/abs/10.1890/12-0267.1
--->
+Whereas averages or weighted averages may improve predictions over any single
+model, they may not optimally leverage available data. The truth does not
+necessarily lie in the middle of multiple model predictions, some models may
+perform better than others in certain conditions, and the covariance between
+models may contain information that can improve predictive accuracy. We can
+exploit these characteristics by using the predictions from a group of models as
+inputs into a separate statistical model. This technique, sometimes called
+superensemble modelling [@krishnamurti1999], is common in climate and weather
+forecasting [e.g. @yun2005; @mote2015]. The superensemble is fit to some
+training dataset where outcomes are well known and used to predict on a dataset
+of interest. For example, @krishnamurti1999 combined predictions of wind and
+precipitation in Asian monsoons via a superensemble regression fit to observed
+data. Their superensemble was considerably more accurate than any individual
+prediction or weighted average of predictions.
 
-Whereas averages or weighted averages may improve predictions over any single model, they may not optimally leverage available data.
-The truth does not necessarily lie in the middle of multiple model predictions, some models may perform better than others in certain conditions, and the covariance between models may contain information that can improve predictive accuracy.
-We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
-This technique, sometimes called 'superensemble' modelling [@krishnamurti1999], is common in climate and weather forecasting [e.g.
-@yun2005; @mote2015].
-The superensemble is fit to some training dataset where outcomes are well known and used to predict on a dataset of interest.
-For example, @krishnamurti1999 combined predictions of wind and precipitation in Asian monsoons via a superensemble regression fit to observed data.
-Their superensemble was considerably more accurate than any individual prediction or weighted average of predictions.
+In fisheries science, the commonly used operational model for determining status
+and trajectory of exploited fish populations are stock assessments that
+incorporate all appropriate data (e.g. catches, size and age distributions,
+surveys, and tagging information) to quantify values such as the biomass that
+can produce maximum sustainable yield ($B/B_\mathrm{MSY}$). However, a broad
+range of data are not available for the majority of fish populations [@fao2014].
+Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$
+based on the limited data available for the majority of fish stocks: catches and
+a basic understanding of population productivity [e.g. @vasconcellos2005;
+@martell2013]. Recently, @rosenberg2014 investigated the performance of four
+data-limited models through a large-scale simulation experiment. The four models
+frequently disagreed about population status and any one model had relatively
+poor predictive ability on average (e.g. Fig. \ref{motivate}).
 
-<!-- Ensemble methods, which combine the output from multiple models as
-predictors in a new 'ensemble' model, are one possible solution to this
-problem. A simple ensemble model might take the average of multiple model
-predictions, perhaps weighted by some goodness of fit. A complex ensemble model
-might also take into account non-linear interactions between model predictions
-and additional covariates. Ensemble models are popular in other fields
-including public health [@thomson2006], agriculture [@cantelaube2005], climate
-science [@murphy2004; @tebaldi2007; @pierce2009], and machine learning
-[@dietterich2000].-->
-
-In fisheries science, the gold standard for determining status and trajectory
-of exploited fish populations are stock assessments that
-incorporate all available data (e.g. catches, size and age distributions,
-surveys, and tagging information) to quantify values such as 
-the biomass that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
-However, these kinds of data are not available for the majority of fish
-populations
-[@fao2014]. Therefore, a number of models have been proposed to assess
-$B/B_\mathrm{MSY}$ based on the limited data available for the majority of
-fish stocks: catches and a basic understanding of population productivity [e.g.
-@vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the
-performance of four data-limited models through a large-scale simulation
-experiment. The four models frequently disagreed about population status and
-any one model had relatively poor predictive ability on average (e.g. Fig.
-\ref{motivate}).
-
-Here, we estimate population status and trajectory of exploited fish
-populations using ensembles and superensembles (collectively 'ensemble
-methods') of these four data-limited models. We explore a variety of approaches
-applied to both simulated and real-world fish stocks and compare predictive
-performance. We find that ensembles, and in particular superensembles,
-generally improved predictive performance of status and trajectory of fish
-populations over any single model---especially when considered across multiple
-dimensions of predictive ability. Furthermore, the output from superensembles
-can aid mechanistic understanding of how individual models of status and
-trajectory perform.
+Here, we estimate population status and trajectory of exploited fish populations
+using ensembles and superensembles (collectively 'ensemble methods') of these
+four data-limited models. We explore a variety of approaches applied to both
+simulated and real-world fish stocks and compare predictive performance. We find
+that ensembles, and in particular superensembles, generally improved predictive
+performance of status and trajectory of fish populations over any single
+model---especially when considered across multiple dimensions of predictive
+ability. Furthermore, the output from superensembles can aid mechanistic
+understanding of how individual models of status and trajectory perform.
 
@@
 We first developed and tested ensemble methods with a fully factorial simulated
-dataset [@rosenberg2014] with precisely known status. These published simulations included three life
-histories: small pelagic, demersal, and large pelagic; three levels of how
-depleted biomass was at the start of the dataset compared to carrying capacity:
-100%, 70%, and 40% of carrying capacity; and four exploitation dynamics: a
-constant exploitation rate, an exploitation rate coupled with biomass to mimic
-an open-access single-species fishery, a scenario where exploitation rate
-increases to a fixed level, and a 'roller-coaster' scenario where the
-exploitation rate increases and then decreases cyclically. Process noise was
-introduced to the models through two levels of multiplicative recruitment
-variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either
+dataset [@rosenberg2014] with precisely known status. These published
+simulations included three fish life histories: small pelagic, demersal, and large
+pelagic; three levels of how depleted biomass was at the start of the dataset
+compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four
+harvest dynamics: a constant harvest rate, a harvest rate
+coupled with biomass to mimic an open-access single-species fishery, a scenario
+where harvest rate increases continuously, and a 'roller-coaster'
+scenario where the harvest rate increases and then decreases.
+Process noise was introduced to the models through two levels of multiplicative
+recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either
 uncorrelated or had a first-order autoregressive correlation of $0.6$. The
-simulation included a scenario without observation error and with
-multiplicative observation error around catch at $N(0, 0.2)$.
+simulation included a scenario without observation error and with multiplicative
+observation error around catch at $N(0, 0.2)$.
 
@@
 
-  <!--TODO: build in fig:didactic -->
-
 ## Additional covariates
 
+<!--TODO Kristin and Mark: It might be a good idea to add a figure that
+illustrates the spectrum of ensemble/superensembles that are applied here and
+what is included (e.g., non-linear relationships, etc.). This will help the
+reader understand the progression in complexity of the ensemble approaches I
+think.-->
 Superensemble models allow us to incorporate additional covariates 
@@
 error to measure bias, median absolute proportional error to measure accuracy,
-and Spearman's rank-order correlation to measure the ability to correctly rank
-populations. Code to reproduce our analysis is available at
+and Spearman's rank-order correlation between predicted and 'true' values to
+measure the ability to correctly rank populations. Code to reproduce our
+analysis is available at
 <https://github.com/datalimited/ensembles>.
@@
 absolute proportional error) by 
+<!--TODO: check these percentage values- right?-->
 `r mean_sim$mach_vs_ind_mare_fold`%, 
@@
 across three performance dimensions (accuracy, bias, and rank-order
-performance), two response variables (status and trajectory), two datasets
+correlation), two response variables (status and trajectory), two datasets
 (simulated and global fisheries), and multiple ensemble methods (from a
@@
 learning methods. These methods provide added insight into individual model
-behaviour and generally performed the best, however even a simple average of
-predictions across multiple models may be a robust solution compared to picking
-any one model.
+behaviour and generally performed the best; however, even a simple average of
+predictions across multiple models may provide a more useful measure of the
+desired parameter than a single model in many cases.
 
@@
 contrasting but plausible ranges of parameter values, or make errors in
-uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007]. Such
-individual models would be expected to perform well in different conditions and
-an ensemble model can exploit the best predictive performance of each. Second,
-ensemble models will be most effective when they are not overfit to the
-training dataset. Cross-validation testing [@caruana2004; @hastie2009] and
-methods that are robust to over-fitting such as random forests [@breiman2001],
-may help avoid overfitting ensemble models. We note that our simplest ensemble
-model, an average of individual model predictions, performed approximately as
-well as complex machine learning models when we applied our ensembles to an
-entirely new dataset (Fig. \ref{performance}b) and averages are a common
-ensemble approach in climate science [@ipcc2013]. Third, ensemble
+uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007]. Such individual
+models would be expected to perform well in different conditions and an ensemble
+model can exploit the best predictive performance of each. Second, ensemble
+models will be most effective when they are not overfit to the training dataset.
+Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust
+to over-fitting such as random forests [@breiman2001], may help avoid
+overfitting ensemble models. We note that our simplest ensemble model, an
+average of individual model predictions, performed approximately as well as
+complex machine learning models when we trained our superensembles on the
+simulation dataset and tested them on a separate 'real' dataset (i.e. the RAM
+Legacy Stock Assessment database, Fig. \ref{performance}b). Third, ensemble
 models will be most effective when they are trained on data that are
@@
 Multi-model inference in the form of coefficient averaging weighted by
-information theoretics such as the Akaike Information Criteron (AIC) is a
+information theoretics such as the Akaike Information Criterion (AIC) is a
 common analytical approach in ecology [@burnham2002; @johnson2004;
@@
 fundamental difference is that coefficient averaging focuses on averaging
-*coefficients* whereas ensembles instead average *predictions*. By focussing on
-predictions, ensembles do not suffer from the muddled inference of coefficient
-averaging when coefficients covary [@cade2015] and are a more general purpose
-tool: they do not require information theoretics, they can combine different
-types of models (e.g. parametric and non-parametric models or frequentist and
-Bayesian predictions), and they can combine predictions in complex non-linear
-ways that may depend on additional covariates.
+*coefficients* whereas ensembles instead average *predictions*. While the
+outcome will be the same in some cases, ensembles provide a more general
+purpose tool: they do not require information theoretics and they can combine
+different types of models (e.g. parametric and non-parametric models or
+frequentist and Bayesian predictions). Furthermore, superensembles extend these
+benefits by allowing model predictions be combined in via non-linear functions
+that are tuned to known data.
 
-A strength of superensembles is that they can be tailored to predict specific response
-variables. For example, we built separate superensemble models of mean
+A strength of superensembles is that they can be tailored to predict specific
+response variables. For example, we built separate superensemble models of mean
 $B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$. The same set of model
 weights or non-linear relationships need not hold across different response
-variables. For example, SSCOM contributed little to the GBM superensemble estimate
-of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed
-strongly to estimates of slope (Fig. \ref{fig:partial-sim-slope}). Formally,
-fitting superensemble models to
-specific quantities of interest provides an additional calibration step
-[@rykiel1996]. Individual models are typically calibrated (i.e. parameters are
-estimated) based on a single response variable; however, this model might then
-provide a biased or inaccurate estimate of other quantities of interest. For
-example, a data-limited fisheries model might calibrate parameters by
-maximizing likelihood with respect to observed and predicted catches but those
-parameters may then provide a biased or inaccurate estimate of the mean and
-slope of $B/B_\mathrm{MSY}$. Ensemble methods provide a final calibration step
-to a quantity of interest. This ensemble calibration could further include a
-loss function tailored to the goals of the model, say placing greater weight on
-accuracy at low status levels than higher values.
+variables. For instance, SSCOM contributed little to the GBM superensemble
+estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$, but
+contributed strongly to estimates of trajectory (Fig. \ref{partial-sim-slope}).
+Formally, fitting superensemble models to specific quantities of interest
+provides an additional calibration step [@rykiel1996]. Individual models are
+typically calibrated (i.e. parameters are estimated) based on a single response
+variable; however, this model might then provide a biased or inaccurate estimate
+of other quantities of interest. Ensemble methods provide a final calibration
+step to a quantity of interest. This ensemble calibration could further include
+a loss function tailored to the goals of the model, say placing greater weight
+on accuracy at low status levels than higher values. Conversely, because
+superensembles are tailored to a specific response and loss function,
+superensembles force a modeller to choose an operational purpose for their model
+up front [*sensu* @dickey-collas2014].
 
-*TODO: Coilin: How ensembles can help understand mechanistic underpinnings of individual models*
-Beyond predictive performance, superensembles can help understand mechanistic
-underpinnings of individual models. 
-First, superensembles that allow for nonlinear
-relationships can describe at what ranges of a prediction a certain model is
-most informative. For example, we can see that SSCOM is most effective at
-distinguishing low values of $B/B_\mathrm{MSY}$ (Fig. S4).
-Second, the interaction between superensemble predictors can highlight
-under what conditions certain models perform best. 
-For instance, SAY SOMETHING ABOUT FIG S5 or FIG S6.
-*discover surprises: counterintuitive relationships and interactions*
-*can dig into surprising conditions and learn about why certain models fail; potentially learn how to improve models*
-*can learn under what conditions certain models perform well*
+As @box1987 noted, all models are wrong, but they may still be useful. The
+ensemble methods we investigated attempt to piece together the useful parts of
+candidate models to build a model with improved performance. Instead of viewing
+the superensemble as a black box, we think considerable mechanistic
+understanding can be gained by studying the structure of the superensemble. For
+example, when SSCOM estimates very low status this is likely the case,
+conversely when COMSIR estimates low status, the truth is likely that the status
+is high (Fig. \ref{partial-sim}). These models have two main differences: (1)
+the form of effort dynamics and (2) the allowance for both measurement and
+process error in SSCOM, whereas the implemented COMSIR admits measurement error
+only. Were the methods to differ only in effort dynamics, the results point
+towards a more suitable representation of effort dynamics at low biomasses in
+SSCOM. We think that such investigation of the parameters of a superensemble may
+lead to improvement in the mechanisms imparted to the individual models.
 
@@
 fitting regression models to outcomes for individual species along with
-predictors of species risk [e.g. @anderson2011a; @pinsky2011], or
-population dynamics models to data for individual species [e.g. @dfo2010].
-Both types of models are prone to error caused by model-misspecification and
-therefore results are sensitive to decisions about model structure
-[@brooks2015]. Although there are options to account for potential
-model-misspecification in determination of species risk (e.g., coefficient
-averaging, [@burnham2002], generalized modeling [@yeakel2011], or
-semi-parametric methods, [@thorson2014]), ensemble methods are a relatively
-simple way to combine predictions in a transparent manner. Beyond estimates of
-status and trajectory, ensemble methods could be used to increase the robustness
-of spatial predictions when designing networks of protected areas
+predictors of species risk [e.g. @anderson2011a; @pinsky2011], or population
+dynamics models to data for individual species [e.g. @dfo2010]. Both types of
+models are prone to error caused by model-misspecification and therefore results
+are sensitive to decisions about model structure [@brooks2015]. Although there
+are options to account for potential model-misspecification in determination of
+species risk [e.g. coefficient averaging, @burnham2002; generalized modeling,
+@yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are
+a relatively simple way to combine predictions in a transparent manner. Beyond
+estimates of status and trajectory, ensemble methods could be used to increase
+the robustness of spatial predictions when designing networks of protected areas
 [@rassweiler2014], or to forecast potential spatial shifts in species
@@
 <!--
+
+and different models may result in different scalings of population status
+[Deroba JJ, Butterworth DS, Methot RD Jr., De Oliveira JAA, Fernandez C, Nielsen
+A, Cadrin SX, Dickey-Collas M, Legault CM, Ianelli J, et al. 2015. Simulation
+testing the robustness of stock assessment models to error: some results from
+the ICES Strategic Initiative on Stock Assessment Methods. ICES Journal of
+Marine Science 72: 19–30 doi:10.1093/icesjms/fst237]
+CM: It would be good to keep this in mind from a management perspective as
+depending on the assumptions in the assessment, biomass could be high and
+fishing mortality low to predict the same catches as biomass low and fishing
+mortality high. Combining for a single scaled variable B/Bmsy removes some of
+the scaling but when there is a strong covariation in B and F this would need to
+be (and likely can be) accounted for in a bivariate ensemble. Not our focus here
+but could enter the Discussion.
+
+Biological ensemble modeling to evaluate potential futures of living marine
+resources
+Anna Gårdmark1,8, Martin Lindegren2, Stefan Neuenfeldt2, Thorsten Blenckner3,
+Outi Heikinheimo4, Bärbel Müller-Karulis3,5, Susa Niiranen3, Maciej T. Tomczak3,
+Eero Aro4, Anders Wikström6, and Christian Möllmann7
+Read More: http://www.esajournals.org/doi/abs/10.1890/12-0267.1
+
 * Other possible topics:
@@
   ensembles have the property of rarely being far from the best
-
-# Authors' contributions
-
-TODO *Example: AB carried out the molecular lab work, participated in data
-analysis, carried out sequence alignments, participated in the design of the
-study and drafted the manuscript; CD carried out the statistical analyses; EF
-collected field data; GH conceived of the study, designed the study,
-coordinated the study and helped draft the manuscript. All authors gave final
-approval for publication.*
 -->
@@
 
+<!--Place the acknowledgment paragraph on the cover page of your manuscript.
+(Reviewers are not provided with a cover page.) Do not spell out first (given)
+names. Provide the first initial of the first name, even if the initial starts a
+sentence. Do not use titles (e.g., Dr. or Professor). Refer to authors of the
+manuscript by their initials only (e.g., “S.T.W. was supported by a grant from
+the Torry Foundation.”).-->
 We thank members of Phase I of the working group 'INSERT NAME HERE' who

### 2015-07-14 text/ms.Rmd
Add reference to intro
Simplify paragraph

@@
 multiple uncertain and potentially conflicting estimates of status and
-trajectory [e.g. @brodziak2010; @branch2011]. Some models may suggest a
-population is at risk and declining in abundance while others may suggest it is
-not at risk and stable.
+trajectory [e.g. @brodziak2010; @branch2011; @deroba2015]. Some models may
+suggest a population is at risk and declining in abundance while others may
+suggest it is not at risk and stable.
 
@@
 the superensemble as a black box, we think considerable mechanistic
-understanding can be gained by studying the structure of the superensemble. For
-example, when SSCOM estimates very low status this is likely the case,
-conversely when COMSIR estimates low status, the truth is likely that the status
-is high (Fig. \ref{partial-sim}). These models have two main differences: (1)
-the form of effort dynamics and (2) the allowance for both measurement and
-process error in SSCOM, whereas the implemented COMSIR admits measurement error
-only. Were the methods to differ only in effort dynamics, the results point
-towards a more suitable representation of effort dynamics at low biomasses in
-SSCOM. We think that such investigation of the parameters of a superensemble may
-lead to improvement in the mechanisms imparted to the individual models.
+understanding can be gained by studying its structure. For example, when SSCOM
+estimates very low status this is likely the case, conversely when COMSIR
+estimates low status, the truth is likely that the status is high (Fig.
+\ref{partial-sim}). These models have two main differences: (1) the form of
+effort dynamics and (2) the allowance for both measurement and process error in
+SSCOM, whereas the implemented COMSIR admits measurement error only. Were the
+methods to differ only in effort dynamics, the results point towards a more
+suitable representation of effort dynamics at low biomasses in SSCOM. We think
+that such investigation of the parameters of a superensemble may lead to
+improvement in the mechanisms imparted to the individual models.
 

### 2015-07-20 text/ms.Rmd
Edits before sending out draft

@@
 
-Submit your cover page as a separate document. It should not be part of the
-manuscript itself. The cover page must include the title of the paper; a running
-head (short title of 40 or fewer characters); a list of 5-8 keywords; word count
-(all text from the first word of the Abstract through the last word of the
-Literature Cited but not including table or figure legends or the body of
-tables); authors' complete mailing addresses (including postal code) at the time
-the work was conducted and present addresses if different; name and complete
-address (including email) of the person to whom correspondence should be sent;
-and text of your acknowledgments section.
+Submit your cover page as a separate document.
+It should not be part of the manuscript itself.
+The cover page must include the title of the paper; a running head (short title of 40 or fewer characters); a list of 5-8 keywords; word count (all text from the first word of the Abstract through the last word of the Literature Cited but not including table or figure legends or the body of tables); authors' complete mailing addresses (including postal code) at the time the work was conducted and present addresses if different; name and complete address (including email) of the person to whom correspondence should be sent; and text of your acknowledgments section.
 
-Contributed Papers, Research Notes, and Conservation Practice and Policy papers
-should contain the following sections in the following order: Abstract,
-Introduction, Methods, Results, Discussion, Supporting Information paragraph (if
-there are online appendices), Literature Cited, tables, figure legends, and
-figures. Do not combine sections (e.g., Results and Discussion). The
-Acknowledgments section will be added to the body of the paper after the
-manuscript has been accepted. Do not number section headings or subheadings. Do
-not include a Conclusion section (conclusions are part of the Discussion).
+Contributed Papers, Research Notes, and Conservation Practice and Policy papers should contain the following sections in the following order: Abstract, Introduction, Methods, Results, Discussion, Supporting Information paragraph (if there are online appendices), Literature Cited, tables, figure legends, and figures.
+Do not combine sections (e.g., Results and Discussion).
+The Acknowledgments section will be added to the body of the paper after the manuscript has been accepted.
+Do not number section headings or subheadings.
+Do not include a Conclusion section (conclusions are part of the Discussion).
 
-You may have color figures in the online version and gray-scale figures in
-print. However, reference to color cannot be made in the figure legend or in the
-text, and elements in the gray-scale version must be distinguishable. Supply
-separate files for color and gray-scale figures.
+You may have color figures in the online version and gray-scale figures in print.
+However, reference to color cannot be made in the figure legend or in the text, and elements in the gray-scale version must be distinguishable.
+Supply separate files for color and gray-scale figures.
 
@@
 \noindent
-^11^National Marine Fisheries Service, 
-National Oceanic and Atmospheric Administration, 
-Fisheries Resource Assessment and Monitoring Division, 
-Northwest Fisheries Science Center, Seattle, WA, 98112, USA
+^11^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112
 
@@
 
-Ecological resource managers are often faced with reconciling multiple
-potentially conflicting estimates of population status and trajectory. One
-solution is to average these predictions in an ensemble, but this approach
-ignores covariance between models and may not optimally reduce bias and improve
-accuracy. Superensemble models, commonly used for climate and weather
-forecasting, may provide a superior solution. Superensembles use the predictions
-from multiple models as covariates in a new statistical model. Here we evaluate
-the potential for ensemble and superensemble models ('ensemble methods') to
-improve estimates of fish population status and trajectory. We fit four
-data-limited models of population and exploitation dynamics to marine fish
-stocks and combine their estimates of the mean and slope of biomass at maximum
-sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three
-superensembles: a linear model, a random forest, and a boosted regression tree. 
-We selected data-limited models for their applicability across the majority of
-fish stocks world wide. We build our models on a simulated dataset of 5760
-stocks and tested them with cross-validation and a global database of 
-`r ram_stocks_n` stock assessments that have also been assessed with more
-data-intensive stock assessment models. We find that ensemble methods
-substantially improve estimates of population status and trajectory.
-Machine-learning superensembles performed the best for population status:
-accuracy improved
-`r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted
-and true status improved from 
-`r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median
-proportional error) declined from 
-`r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. 
-We found similar improvements when predicting the trajectory of status and when
-applying the simulation-trained superensembles to catch data for global fish
-stocks. Ensemble methods have the potential to improve estimates of population
-status and trajectory across many taxa beyond fish. However, they must be
-carefully tested, formed from a diverse set of accurate models, and built on a
-dataset sufficiently representative of the populations to which they are
-applied.
+Ecological resource managers are often faced with reconciling multiple potentially conflicting estimates of population status and trajectory.
+One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
+Superensemble models, commonly used for climate and weather forecasting, may provide a superior solution.
+Superensembles use the predictions from multiple models as covariates in a new statistical model.
+Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trajectory.
+We fit four data-limited models of population and exploitation dynamics to marine fish stocks and combine their estimates of the mean and slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
+We selected data-limited models for their applicability across the majority of fish stocks world wide.
+We build our models on a simulated dataset of 5760 stocks and tested them with cross-validation and a global database of `r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
+We find that ensemble methods substantially improve estimates of population status and trajectory.
+Machine-learning superensembles performed the best for population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements when predicting the trajectory of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
+Ensemble methods have the potential to improve estimates of population status and trajectory across many taxa beyond fish.
+However, they must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations to which they are applied.
 
@@
 
-Population status and trajectory are two of the most fundamental values to
-quantify in conservation biology [e.g. @mace2008; @iucn2015]. These properties
-are key to everything from global assessments of extinction risk [e.g.
-@stuart2004; @schipper2008] to assessing the sustainability of hunting and
-fishing exploitation levels [e.g. @worm2009; @juan-jorda2011]. However,
-conservation and ecological resource managers are often faced with reconciling
-multiple uncertain and potentially conflicting estimates of status and
-trajectory [e.g. @brodziak2010; @branch2011; @deroba2015]. Some models may
-suggest a population is at risk and declining in abundance while others may
-suggest it is not at risk and stable.
+Population status and trajectory are two of the most fundamental values to quantify in conservation biology [e.g.
+@mace2008; @iucn2015].
+These properties are key to everything from global assessments of extinction risk [e.g.
+@stuart2004; @schipper2008] to assessing the sustainability of hunting and fishing exploitation levels [e.g.
+@worm2009; @juan-jorda2011].
+However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g.
+@brodziak2010; @branch2011; @deroba2015].
+Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
 
-One solution is to take the average or weighted average of an ensemble of model
-predictions. Such ensembles are typically more accurate and less biased than
-individual model estimates and can integrate uncertainty in model structure,
-initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
+One solution is to take the average or weighted average of an ensemble of model predictions.
+Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
 This approach forms the basis of many machine learning methods [e.g.
-@dietterich2000], has helped reconcile climate forecasts from dozens of models
-[e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning
-signs of malaria outbreaks [@thomson2006]. In ecology, ensemble methods are
-frequently used to improve species distribution modelling [e.g. @araujo2007;
-@breiner2015] and indeed have been used to combine estimates of population
-status and trajectory [e.g. @brodziak2010].
+@dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.
+@murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006].
+In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g.
+@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trajectory [e.g.
+@brodziak2010].
 
-Whereas averages or weighted averages may improve predictions over any single
-model, they may not optimally leverage available data. The truth does not
-necessarily lie in the middle of multiple model predictions, some models may
-perform better than others in certain conditions, and the covariance between
-models may contain information that can improve predictive accuracy. We can
-exploit these characteristics by using the predictions from a group of models as
-inputs into a separate statistical model. This technique, sometimes called
-superensemble modelling [@krishnamurti1999], is common in climate and weather
-forecasting [e.g. @yun2005; @mote2015]. The superensemble is fit to some
-training dataset where outcomes are well known and used to predict on a dataset
-of interest. For example, @krishnamurti1999 combined predictions of wind and
-precipitation in Asian monsoons via a superensemble regression fit to observed
-data. Their superensemble was considerably more accurate than any individual
-prediction or weighted average of predictions.
+Whereas averages or weighted averages may improve predictions over any single model, they may not optimally leverage available data.
+The best prediction does not necessarily lie in the middle of multiple model predictions, some models may perform better than others in certain conditions, and the covariance between models may contain information that can improve predictive accuracy.
+We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
+This technique, sometimes called superensemble modelling [@krishnamurti1999], is common in climate and weather forecasting [e.g.
+@yun2005; @mote2015].
+The superensemble is fit to some training dataset where outcomes are well known and used to predict on a dataset of interest.
+For example, @krishnamurti1999 combined predictions of wind and precipitation in Asian monsoons via a superensemble regression fit to observed data.
+Their superensemble was considerably more accurate than any individual prediction or weighted average of predictions.
 
-In fisheries science, the commonly used operational model for determining status
-and trajectory of exploited fish populations are stock assessments that
-incorporate all appropriate data (e.g. catches, size and age distributions,
-surveys, and tagging information) to quantify values such as the biomass that
-can produce maximum sustainable yield ($B/B_\mathrm{MSY}$). However, a broad
-range of data are not available for the majority of fish populations [@fao2014].
-Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$
-based on the limited data available for the majority of fish stocks: catches and
-a basic understanding of population productivity [e.g. @vasconcellos2005;
-@martell2013]. Recently, @rosenberg2014 investigated the performance of four
-data-limited models through a large-scale simulation experiment. The four models
-frequently disagreed about population status and any one model had relatively
-poor predictive ability on average (e.g. Fig. \ref{motivate}).
+In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments that incorporate all appropriate data (e.g.
+catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
+However, a broad range of data are not available for the majority of fish populations [@fao2014].
+Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: catches and a basic understanding of population productivity [e.g.
+@vasconcellos2005; @martell2013].
+Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment.
+The four models frequently disagreed about population status and any one model had relatively poor predictive ability on average (e.g.
+Fig.
+\ref{motivate}).
 
-Here, we estimate population status and trajectory of exploited fish populations
-using ensembles and superensembles (collectively 'ensemble methods') of these
-four data-limited models. We explore a variety of approaches applied to both
-simulated and real-world fish stocks and compare predictive performance. We find
-that ensembles, and in particular superensembles, generally improved predictive
-performance of status and trajectory of fish populations over any single
-model---especially when considered across multiple dimensions of predictive
-ability. Furthermore, the output from superensembles can aid mechanistic
-understanding of how individual models of status and trajectory perform.
+Here, we estimate population status and trajectory of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
+We explore a variety of approaches applied to both simulated and real-world fish stocks and compare predictive performance.
+We find that ensembles, and in particular superensembles, generally improved predictive performance of status and trajectory of fish populations over any single model---especially when considered across multiple dimensions of predictive ability.
+Furthermore, the output from superensembles can aid mechanistic understanding of how individual models of status and trajectory perform.
 
@@
 
-We first developed and tested ensemble methods with a fully factorial simulated
-dataset [@rosenberg2014] with precisely known status. These published
-simulations included three fish life histories: small pelagic, demersal, and large
-pelagic; three levels of how depleted biomass was at the start of the dataset
-compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four
-harvest dynamics: a constant harvest rate, a harvest rate
-coupled with biomass to mimic an open-access single-species fishery, a scenario
-where harvest rate increases continuously, and a 'roller-coaster'
-scenario where the harvest rate increases and then decreases.
-Process noise was introduced to the models through two levels of multiplicative
-recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either
-uncorrelated or had a first-order autoregressive correlation of $0.6$. The
-simulation included a scenario without observation error and with multiplicative
-observation error around catch at $N(0, 0.2)$.
+We first developed and tested ensemble methods with a fully factorial simulated dataset [@rosenberg2014] with precisely known status.
+These published simulations included three fish life histories: small pelagic, demersal, and large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamics: a constant harvest rate, a harvest rate coupled with biomass to mimic an open-access single-species fishery, a scenario where harvest rate increases continuously, and a 'roller-coaster' scenario where the harvest rate increases and then decreases.
+Process noise was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
+The simulation included a scenario without observation error and with multiplicative observation error around catch at $N(0, 0.2)$.
 
-@rosenberg2014 ran ten iterations for each combination of
-factors adding stochastic draws of recruitment and catch-recording variability
-each time to generate a total of 5760 stocks. The simulation models were built
-in the package `FLR` [@kell2007] for the statistical software \textsf{R}
-[@r2015]. The code to generate the simulations is available at
-<https://github.com/flr/StockSims> (TODO: not there).
+@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
+The simulation models were built in the package `FLR` [@kell2007] for the statistical software \textsf{R} [@r2015].
+The code to generate the simulations is available at <https://github.com/flr/StockSims> (TODO: not there).
 
-We also tested our ensemble methods on the RAM Legacy Stock Assessment Database
-[@ricard2012]. Our analysis of the stock-assessment database was based on
-version XX downloaded on XX. After removing stocks for which the models
-described below did not converge, this database included `r ram_stocks_n`
-stocks.
+We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012].
+Our analysis of the stock-assessment database was based on version XX downloaded on XX.
+After removing stocks for which the models described below did not converge, this database included `r ram_stocks_n` stocks.
 
@@
 
-We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the
-models are mechanistic and based generally on Schaefer biomass dynamics
-[@schaefer1954] of the form
+We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
 
@@
 \noindent
-where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year,
-$B_t$ represents biomass at time $t$, $r$ represents the intrinsic population
-growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$
-represents catch. @rosenberg2014 provide a full summary of
-these four methods and code to fit all the models is available in an
-accompanying package `datalimited` for the statistical software \textsf{R}. In
-summary:
+where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch.
+@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` for the statistical software \textsf{R}.
+In summary:
 
-* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
-  dynamics [@martell2013]. It requires a prior distribution for $r$ and assigns
-  a prior to the relative proportion of biomass at the end compared to unfished
-  biomass (depletion) based on the percentage of maximum catch at the end of the
-  time series. Our version of the model was modified from @martell2013 to
-  generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of
-  $B/B_\mathrm{MSY}$ from the median trend.
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass dynamics [@martell2013].
+It requires a prior distribution for $r$ and assigns a prior to the relative proportion of biomass at the end compared to unfished biomass (depletion) based on the percentage of maximum catch at the end of the time series.
+Our version of the model was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
 
-* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled
-  harvest-dynamics model [@vasconcellos2005]. Biomass is assumed to follow a
-  Schaefer model and harvest dynamics are assumed to follow a logistic model.
+* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled harvest-dynamics model [@vasconcellos2005].
+Biomass is assumed to follow a Schaefer model and harvest dynamics are assumed to follow a logistic model.
 The model is fit with a sample-importance-sampling algorithm [@rosenberg2014].
 
-* *SSCOM* (state-space catch-only model) is a hierarchical model that, similar
-  to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013].
-  SSCOM estimates unobserved dynamics in both fishing effort and the fished
-  population based on a catch time series and priors on $r$, the maximum rate
-  of increase of fishing effort, and the magnitude of various sources of
-  stochasticity. The model is fit in a Bayesian state-space framework to
-  integrate across three forms of stochasticity: variation in effort,
-  population dynamics, and fishing efficiency [@thorson2013].
+* *SSCOM* (state-space catch-only model) is a hierarchical model that, similar to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013].
+SSCOM estimates unobserved dynamics in both fishing effort and the fished population based on a catch time series and priors on $r$, the maximum rate of increase of fishing effort, and the magnitude of various forms of stochasticity.
+The model is fit in a Bayesian state-space framework to integrate across three forms of stochasticity: variation in effort, population dynamics, and fishing efficiency [@thorson2013].
 
-* *mPRM* (modified panel regression model) is a modified version of the
-  panel-regression model from @costello2012. Unlike the other
-  models, mPRM is empirical and not mechanistic---it uses the RAM Legacy Stock
-  Assessment database to fit a regression model to a series of characteristics
-  of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as
-  the response. The model used in this paper is modified from the original in
-  that it condenses the life-history categories into three categories to match
-  the simulated dataset, removes the maximum catch predictor since the absolute
-  catch in the simulated dataset is arbitrary, and does not implement the bias
-  correction needed in @costello2012 for deriving aggregate
-  estimates of median status across multiple stocks.
+* *mPRM* (modified panel regression model) is a modified version of the panel-regression model from @costello2012.
+Unlike the other models, mPRM is empirical and not mechanistic---it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
+The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 for deriving aggregate estimates of median status across multiple stocks.
 
@@
 
-<!--TODO Kristin and Mark: It might be a good idea to add a figure that
-illustrates the spectrum of ensemble/superensembles that are applied here and
-what is included (e.g., non-linear relationships, etc.). This will help the
-reader understand the progression in complexity of the ensemble approaches I
-think.-->
-Superensemble models allow us to incorporate additional covariates 
-and potentially leverage interactions between these covariates and individual
-model predictions. Additional covariates could be, for example, life-history
-characteristics, information on exploitation patterns, or statistical
-properties of the data. For simplicity, and to allow us to apply models
-developed with the simulated dataset to the real-world dataset, we added only one
-set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and
-provides a means of describing the statistical properties of the catch series
-that is independent of time series length and absolute magnitude of catch. We fit
-spectral models to the scaled catch time series (catch divided by maximum
-catch) with the `spec.ar` function in \textsf{R} and recorded representative
-short- and long-term spectral densities at frequencies of 0.20 and 0.05, which
-correspond to 5- and 20-year cycles.
+<!--TODO Kristin and Mark: It might be a good idea to add a figure that illustrates the spectrum of ensemble/superensembles that are applied here and what is included (e.g., non-linear relationships, etc.).
+This will help the reader understand the progression in complexity of the ensemble approaches I think.-->
+Superensemble models allow us to incorporate additional covariates and potentially leverage interactions between these covariates and individual model predictions.
+Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data.
+For simplicity, and to allow us to apply models developed with the simulated dataset to the real-world dataset, we added only one set of additional covariates: spectral properties of the catch time series.
+Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length and absolute magnitude of catch.
+We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
 
@@
 
-The individual models we seek to combine with superensembles provide time
-series of $B/B_\mathrm{MSY}$. Therefore, we can use superensembles to estimate
-multiple properties of these status time series. Here, we focus on two
-properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
-Together, these quantities address the recent state and trajectory of status,
-which are both of management and conservation interest [e.g. @mace2008]. To
-avoid undue influence of the end points of the time series on the calculated
-slope, we measured the slope as the Theil-Sen estimator of median slope
-[@theil1950].
+The individual models we seek to combine with superensembles provide time series of $B/B_\mathrm{MSY}$.
+Therefore, we can use superensembles to estimate multiple properties of these status time series.
+Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
+Together, these quantities address the recent state and trajectory of status, which are both of management and conservation interest [e.g.
+@mace2008].
+To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
 
-Here we compare an ensemble average and three superensembles of varying
-complexity: a linear model with two-way interactions, a random forest, and
-boosted regression tree. The general form of our models was
+Here we compare an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and boosted regression tree.
+The general form of our models was
 
@@
 \noindent 
-where $L$ is a generalized loss function, $f$ is some generalized regression
-function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY
-model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
-spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$
-is the mean or slope value the model is fitted to, and $\theta$ is the set of
-parameters for the regression function. For all superensemble models of mean
-$B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space
-and exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$
-slope, which are not bounded at zero, we fit superensemble models on the natural
-untransformed scale.
+where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$ is the mean or slope value the model is fitted to, and $\theta$ is the set of parameters for the regression function.
+For all superensemble models of mean $B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space and exponentiated the predictions.
+For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
 
-We fit the linear model superensemble with second-order interactions. We fit two
-machine learning superensemble models based on regression trees. Regression trees
-sequentially determine what value of a predictor best splits the response data
-into two 'branches' based on a specified loss function [@breiman1984]. In random
-forests, a series of regression trees are built on a random subset of the data
-and random subset of the covariates of the model [@breiman2001]. In generalized
-boosted models (GBMs), each subsequent model is fit to the residuals from the
-previous model; data points that are fit poorly in a given model are given more
-weight in the next model [@elith2008]. Random forests and GBMs can provide
-strong predictive performance and fit highly non-linear relationships
-[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
-package [@liaw2002] for \textsf{R} with the default argument values. We fit
-boosted regression tree models with the `gbm` package [@ridgeway2015] for
-\textsf{R}. Based on cross-validation with the **caret** \textsf{R} package, we
-fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate
-(shrinkage parameter) of $0.01$, and all other arguments at their default
-values.
+We fit the linear model superensemble with second-order interactions.
+We fit two machine learning superensemble models based on regression trees.
+Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984].
+In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model [@breiman2001].
+In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008].
+Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [@elith2008; @hastie2009].
+We fit random forest models with the `randomForest` package [@liaw2002] for \textsf{R} with the default argument values.
+We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for \textsf{R}.
+Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
 
@@
 
-A critical component to any predictive modelling exercise is to evaluate the
-performance of a model on new data [@hastie2009]. We used repeated three-fold
-cross validation to test predictive performance: we randomly divided the
-dataset into three sets, built superensemble models on two-thirds of the data, and
-evaluated predictive performance on the remaining third. We repeated this
-across each of the three splits and then repeated the whole procedure 50 times
-to account for bias that may result from any one set of validation splits.
-Since the dynamics of simulated populations differing only in the stochastic
-draw of random values are likely similar, we grouped these stocks (10 per
-factorial cell) in the cross-validation process into either the training or
-testing split. Since mPRM is built on the RAM Legacy Stock Assessment database,
-when testing model performance on the stock assessment database, we refit mPRM
-on each training split.
+A critical component to any predictive modelling exercise is to evaluate the performance of a model on new data [@hastie2009].
+We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built superensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third.
+We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits.
+Since the dynamics of simulated populations differing only in the stochastic draw of random values are likely similar, we grouped these stocks (10 per factorial cell) in the cross-validation process into either the training or testing split.
+When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset.
+However, since mPRM is built on the same stock assessment database, we refit mPRM on each training split.
 
-Predictive performance can be evaluated with metrics that represent a variety
-of modelling goals. For continuous response variables such as the mean and
-slope of population status, performance metrics often measure some form of
-bias, precision, accuracy (a combination of bias and precision), or the ability
-to correctly rank or correlate across populations [e.g. @walther2005]. Here,
-we measure proportional error, defined as
-$(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true'
-(or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$. We calculated median proportional
-error to measure bias, median absolute proportional error to measure accuracy,
-and Spearman's rank-order correlation between predicted and 'true' values to
-measure the ability to correctly rank populations. Code to reproduce our
-analysis is available at
-<https://github.com/datalimited/ensembles>.
+Predictive performance can be evaluated with metrics that represent a variety of modelling goals.
+For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of bias, precision, accuracy (a combination of bias and precision), or the ability to correctly rank or correlate across populations [e.g.
+@walther2005].
+Here, we measure proportional error, defined as $(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$.
+We calculated median proportional error to measure bias, median absolute proportional error to measure accuracy, and Spearman's rank-order correlation between predicted and 'true' values to measure the ability to correctly rank populations.
+We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 1$ at any decision threshold by plotting receiver operating characteristic curves.
+Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles>.
 
@@
 
-Applied to the simulated dataset of known status, the individual models had
-variable success at recovering the mean (status) and slope (trajectory) of
-$B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of scatter
-around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). CMSY
-exhibited bimodal predictions of status but had the best rank-order correlation
-and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly
-identified a number of low status stocks, but frequently predicted a high
-status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had
-relatively poor ability to predict status for the simulated dataset (Fig.
-\ref{hexagon}d). There was generally little correlation between true and
-predicted trajectory for any of the individual models besides SSCOM (Figs
-\ref{scatter-sim-slope}a--d).
+Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of $B/B_\mathrm{MSY}$ in the last five years.
+All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig.
+\ref{hexagon}).
+CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig.
+\ref{performance}a).
+COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig.
+\ref{hexagon}b, c).
+mPRM had relatively poor ability to predict status for the simulated dataset (Fig.
+\ref{hexagon}d).
+There was generally little correlation between true and predicted trajectory for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) besides SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
 
-Ensemble methods, and in particular the machine learning superensemble models (random
-forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig.
-\ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the
-individual models, machine learning superensembles improved accuracy (median
-absolute proportional error) by 
-<!--TODO: check these percentage values- right?-->
-`r mean_sim$mach_vs_ind_mare_fold`%, 
-increased rank-order correlation from `r mean_sim$ind_corr_range` 
-to `r mean_sim$mach_corr_range`, 
-and reduced bias (median proportional error) from 
-`r mean_sim$ind_mre_range` to 
-`r mean_sim$mach_mre_range` (Fig. \ref{performance}a). 
-These superensembles also had better ability to distinguish if
-simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
-\ref{roc-sim}). Results were similar when predicting trajectory: compared to
-individual models, machine learning superensembles improved accuracy by 
-`r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from 
-`r slope_sim$ind_corr_range` to 
-`r slope_sim$mach_corr_range`, and reduced bias
-from `r slope_sim$ind_mre_range` 
-to `r slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The
-ensemble models that simply took a mean of the individual models ranked
-slightly behind the best individual model for estimating fish stock status (CMSY; Fig.
-\ref{performance}a) and had slightly lower correlation but higher accuracy than
-the best individual model at predicting the trends of status (SSCOM; Fig.
+Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig.
+\ref{hexagon}e--h, Fig.
+\ref{scatter-sim-slope}e--h).
+Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig.
+\ref{performance}a).
+These superensembles also had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
+\ref{roc-sim}).
+Results were similar when predicting trajectory: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
+\ref{performance-sim-slope}).
+The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating fish stock status (CMSY; Fig.
+\ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting the trends of status (SSCOM; Fig.
 \ref{performance-sim-slope}).
 
-The superensemble models were able to improve the predictions of status by
-exploiting the best properties of individual models, the covariance between
-individual models, and interactions with other covariates. For example, SSCOM
-had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
-\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
-approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction
-clusters (Fig. \ref{partial-sim}). Superensembles also exploited the covariance
-between individual model predictions. For instance, both the linear model and
-GBM ensemble suggest that if mPRM and SSCOM predict high status, the true
-status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l). The
-addition of spectral density covariates helped the superensemble models correctly
-predict higher status values (Fig. \ref{hexagon}g, h) by exploiting
-interactions between the spectral density predictors and the individual model
-predictions (Fig. \ref{partial-2d-sim}). Although the additional covariates
-improved ensemble fit, performance of the ensembles was only marginally
-degraded by removing these covariates (Fig. \ref{hexagon} vs. Fig.
+The superensemble models were able to improve the predictions of status by exploiting the best properties of individual models, the covariance between individual models, and interactions with other covariates.
+For example, SSCOM had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
+\ref{hexagon}c, Fig.
+\ref{partial-sim}c) and CMSY predictions were approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction clusters (Fig.
+\ref{partial-sim}).
+Superensembles also exploited the covariance between individual model predictions.
+For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
+The addition of spectral density covariates helped the superensemble models correctly predict higher status values (Fig.
+\ref{hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
+\ref{partial-2d-sim}).
+Although the additional covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
+\ref{hexagon} vs.
+Fig.
 \ref{hexagon-sim-basic}).
 
-When applied to the stock assessment database, the superensemble models---trained
-exclusively on the simulated dataset---generally performed as well or better
-than the best individual models. The mean, random forest, and GBM ensembles
-outperformed the mPRM method which is trained directly on the RAM Legacy Stock
-Assessment database itself (Fig. \ref{performance}b, Fig. \ref{hexagon-ram}).
-Compared to the individual models, the machine learning superensembles increased
-accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, 
-improved correlation from `r mean_ram$ind_corr_range` to 
-`r mean_ram$mach_corr_range`, and reduced bias from
-`r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
+When applied to the stock assessment database, the superensemble models---trained exclusively on the simulated dataset---generally performed as well or better than the best individual models.
+The mean, random forest, and GBM ensembles outperformed the mPRM method which is trained directly on the RAM Legacy Stock Assessment database itself (Fig.
+\ref{performance}b, Fig.
+\ref{hexagon-ram}).
+Compared to the individual models, the machine learning superensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
 
@@
 
-Ensemble methods provide a useful approach to situations where conservation and
-ecological resource management decisions must be made on the basis of multiple,
-potentially contrasting estimates of status---a situation common to many
-ecological settings beyond fisheries. Compared to individual models of fish
-population status, ensemble methods were consistently the best or among the best
-across three performance dimensions (accuracy, bias, and rank-order
-correlation), two response variables (status and trajectory), two datasets
-(simulated and global fisheries), and multiple ensemble methods (from a
-simple average to machine learning superensembles). Our results suggest choosing an
-superensemble model that allows for non-parametric relationships, such as machine
-learning methods. These methods provide added insight into individual model
-behaviour and generally performed the best; however, even a simple average of
-predictions across multiple models may provide a more useful measure of the
-desired parameter than a single model in many cases.
+Ensemble methods provide a useful approach to situations where conservation and ecological resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status---a situation common to many settings beyond fisheries.
+Compared to individual models of fish population status, ensemble methods were consistently the best or among the best across three performance dimensions (accuracy, bias, and rank-order correlation), two response variables (status and trajectory), two datasets (simulated and global fisheries), and multiple ensemble methods (from a simple average to machine learning superensembles).
+Our results suggest choosing an superensemble model that allows for non-parametric relationships, such as machine learning methods.
+These methods provide added insight into individual model behaviour and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
 
 Certain conditions will make some ensemble models more effective than others.
-First, ensembles will be most effective when they are comprised of diverse
-individual models that choose different structural model forms, explore
-contrasting but plausible ranges of parameter values, or make errors in
-uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007]. Such individual
-models would be expected to perform well in different conditions and an ensemble
-model can exploit the best predictive performance of each. Second, ensemble
-models will be most effective when they are not overfit to the training dataset.
-Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust
-to over-fitting such as random forests [@breiman2001], may help avoid
-overfitting ensemble models. We note that our simplest ensemble model, an
-average of individual model predictions, performed approximately as well as
-complex machine learning models when we trained our superensembles on the
-simulation dataset and tested them on a separate 'real' dataset (i.e. the RAM
-Legacy Stock Assessment database, Fig. \ref{performance}b). Third, ensemble
-models will be most effective when they are trained on data that are
-representative of the dataset of interest [@knutti2009; @weigel2010].
-Cross-validation within a training dataset will provide an optimistically biased
-impression of predictive performance if the training dataset fundamentally
-differs from the dataset of interest [@hastie2009].
+First, ensembles will be most effective when they are comprised of diverse individual models that choose different structural model forms, explore contrasting but plausible ranges of parameter values, or make errors in uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007].
+Such individual models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance of each.
+Second, ensemble models will be most effective when they are not overfit to the training dataset.
+Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust to over-fitting such as random forests [@breiman2001], may help avoid overfitting ensemble models.
+We note that our simplest ensemble model, an average of individual model predictions, performed approximately as well as complex machine learning models when we trained our superensembles on the simulation dataset and tested them on a separate 'real' dataset (i.e.
+the RAM Legacy Stock Assessment database, Fig.
+\ref{performance}b).
+Third, ensemble models will be most effective when they are trained on data that are representative of the dataset of interest [@knutti2009; @weigel2010].
+Cross-validation within a training dataset will provide an optimistically biased impression of predictive performance if the training dataset fundamentally differs from the dataset of interest [@hastie2009].
 
-Multi-model inference in the form of coefficient averaging weighted by
-information theoretics such as the Akaike Information Criterion (AIC) is a
-common analytical approach in ecology [@burnham2002; @johnson2004;
-@grueber2011]. The ensemble methods described in this paper share similarities
-with coefficient averaging but differ in other important ways. Ensemble methods
-and coefficient averaging share the long-held notion that multiple working
-hypotheses can contribute useful information for inference [@chamberlin1890]. A
-fundamental difference is that coefficient averaging focuses on averaging
-*coefficients* whereas ensembles instead average *predictions*. While the
-outcome will be the same in some cases, ensembles provide a more general
-purpose tool: they do not require information theoretics and they can combine
-different types of models (e.g. parametric and non-parametric models or
-frequentist and Bayesian predictions). Furthermore, superensembles extend these
-benefits by allowing model predictions be combined in via non-linear functions
-that are tuned to known data.
+Multi-model inference in the form of coefficient averaging weighted by information theoretics such as the Akaike Information Criterion (AIC) is a common analytical approach in ecology [@burnham2002; @johnson2004; @grueber2011].
+The ensemble methods described in this paper share similarities with coefficient averaging but differ in other important ways.
+Ensemble methods and coefficient averaging share the long-held notion that multiple working hypotheses can contribute useful information for inference [@chamberlin1890].
+A fundamental difference is that coefficient averaging focuses on averaging *coefficients* whereas ensembles instead average *predictions*.
+While the outcome will be the same in some cases, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
+parametric and non-parametric models or frequentist and Bayesian predictions).
+Furthermore, superensembles extend these benefits by allowing model predictions to be combined via non-linear functions that are tuned to known data.
 
-A strength of superensembles is that they can be tailored to predict specific
-response variables. For example, we built separate superensemble models of mean
-$B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$. The same set of model
-weights or non-linear relationships need not hold across different response
-variables. For instance, SSCOM contributed little to the GBM superensemble
-estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$, but
-contributed strongly to estimates of trajectory (Fig. \ref{partial-sim-slope}).
-Formally, fitting superensemble models to specific quantities of interest
-provides an additional calibration step [@rykiel1996]. Individual models are
-typically calibrated (i.e. parameters are estimated) based on a single response
-variable; however, this model might then provide a biased or inaccurate estimate
-of other quantities of interest. Ensemble methods provide a final calibration
-step to a quantity of interest. This ensemble calibration could further include
-a loss function tailored to the goals of the model, say placing greater weight
-on accuracy at low status levels than higher values. Conversely, because
-superensembles are tailored to a specific response and loss function,
-superensembles force a modeller to choose an operational purpose for their model
-up front [*sensu* @dickey-collas2014].
+A strength of superensembles is that they can be tailored to predict specific response variables.
+For example, we built separate superensemble models of mean $B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$.
+The same set of model weights or non-linear relationships need not hold across different response variables.
+For instance, SSCOM contributed little to the GBM superensemble estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed strongly to estimates of trajectory (Fig.
+\ref{partial-sim-slope}).
+Formally, fitting superensemble models to specific quantities of interest provides an additional calibration step [@rykiel1996].
+Individual models are typically calibrated (i.e.
+parameters are estimated) based on a single response variable; however, this model might then provide a biased or inaccurate estimate of other quantities of interest.
+Ensemble methods provide a final calibration step to a quantity of interest.
+This ensemble calibration could further include a loss function tailored to the goals of the model, say placing greater weight on accuracy at low status levels than higher values.
+Conversely, because superensembles are tailored to a specific response and loss function, superensembles force a modeller to choose an operational purpose for their model up front [*sensu* @dickey-collas2014].
 
-As @box1987 noted, all models are wrong, but they may still be useful. The
-ensemble methods we investigated attempt to piece together the useful parts of
-candidate models to build a model with improved performance. Instead of viewing
-the superensemble as a black box, we think considerable mechanistic
-understanding can be gained by studying its structure. For example, when SSCOM
-estimates very low status this is likely the case, conversely when COMSIR
-estimates low status, the truth is likely that the status is high (Fig.
-\ref{partial-sim}). These models have two main differences: (1) the form of
-effort dynamics and (2) the allowance for both measurement and process error in
-SSCOM, whereas the implemented COMSIR admits measurement error only. Were the
-methods to differ only in effort dynamics, the results point towards a more
-suitable representation of effort dynamics at low biomasses in SSCOM. We think
-that such investigation of the parameters of a superensemble may lead to
-improvement in the mechanisms imparted to the individual models.
+As @box1987 noted, all models are wrong, but some may still be useful.
+The ensemble methods we investigated attempt to piece together the useful parts of candidate models to build a model with improved performance.
+Instead of viewing the superensemble as a black box, we think considerable mechanistic understanding can be gained by studying its structure.
+For example, when SSCOM estimates very low status this is likely the case, conversely when COMSIR estimates low status, the truth is likely that the status is high (Fig.
+\ref{partial-sim}).
+These models have two main differences: (1) the form of effort dynamics and (2) the allowance for both measurement and process error in SSCOM, whereas the implemented COMSIR admits measurement error only.
+Were the methods to differ only in effort dynamics, the results point towards a more suitable representation of effort dynamics at low biomasses in SSCOM.
+We think that such investigation of the parameters of a superensemble may lead to improvement in the mechanisms imparted to the individual models.
 
-Combining predictions from multiple models via superensemble methods is also
-broadly useful in other subfields of ecology and fisheries science. Predictions
-about extinction risk are widely used at the national (e.g. the US Endangered
-Species Act and the Canadian Species at Risk Act) and international [e.g. the
-IUCN Red List, @iucn2015] levels. These risk assessments generally involve
-fitting regression models to outcomes for individual species along with
-predictors of species risk [e.g. @anderson2011a; @pinsky2011], or population
-dynamics models to data for individual species [e.g. @dfo2010]. Both types of
-models are prone to error caused by model-misspecification and therefore results
-are sensitive to decisions about model structure [@brooks2015]. Although there
-are options to account for potential model-misspecification in determination of
-species risk [e.g. coefficient averaging, @burnham2002; generalized modeling,
-@yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are
-a relatively simple way to combine predictions in a transparent manner. Beyond
-estimates of status and trajectory, ensemble methods could be used to increase
-the robustness of spatial predictions when designing networks of protected areas
-[@rassweiler2014], or to forecast potential spatial shifts in species
-distribution given climate impacts [@harsch2014].
+Combining predictions from multiple models via superensemble methods is also broadly useful in other subfields of ecology and fisheries science.
+Predictions about extinction risk are widely used at the national (e.g.
+the US Endangered Species Act and the Canadian Species at Risk Act) and international [e.g.
+the IUCN Red List, @iucn2015] levels.
+These risk assessments generally involve fitting regression models to outcomes for individual species along with predictors of species risk [e.g.
+@anderson2011a; @pinsky2011], or population dynamics models to data for individual species [e.g.
+@dfo2010].
+Both types of models are prone to error caused by model-misspecification and therefore results are sensitive to decisions about model structure [@brooks2015].
+Although there are options to account for potential model-misspecification in determination of species risk [e.g.
+coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
+Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], or to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014].
 
@@
 'INSERT NAME HERE'.
+'INSERT OTHER REQUIRED FUNDING ACKNOWLEDGEMENTS HERE'
 

### 2015-07-20 analysis/9-output-values.R

@@ -2,11 +2,13 @@
 # for example, ranges of MARE or correlation or AUC values. They get saved to
 # `../text/values.rda` and are input through knitr in `../text/ms.Rmd`.
 
+library(dplyr)
+
 get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   collapse = "--", mre_digits = 2) {
 
   dat <- dat %>%
-    mutate(ensemble = ifelse(grepl("ensemble", clean_method), TRUE, FALSE)) %>%
+    mutate(ensemble = ifelse(grepl("ensemble", clean_method, ignore.case = TRUE), TRUE, FALSE)) %>%
     mutate(machine = ifelse(grepl("GBM|RF", clean_method), TRUE, FALSE))
 
   x <- list()
@@ -47,6 +49,10 @@ mean_ram <- get_performance_stats(re_ram_sum)
 d_slope_error <- readRDS("generated-data/d_slope_error.rds")
 slope_sim <- get_performance_stats(d_slope_error, mre_digits = 3L)
 
+range_slope_corr_non_sscom <- paste(round(range(filter(d_slope_error, !grepl("ensemble", clean_method, ignore.case = TRUE), clean_method != "SSCOM")$corr), 2), collapse = "--")
+
+slope_corr_sscom <- round(filter(d_slope_error, clean_method == "SSCOM")$corr, 2)
+
 auc_sim_mean <- readRDS("generated-data/auc_sim_mean.rds")
 auc_sim <- list()
 
@@ -59,4 +65,4 @@ auc_sim$ind_range <- auc_sim_mean %>% filter(grepl("Ensemble", clean_method) == 
 d_ram <- readRDS("generated-data/ram-ensemble-predicted.rds")
 ram_stocks_n <- length(unique(d_ram$stockid))
 
-save(mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n, file = "../text/values.rda")
+save(mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n, slope_corr_sscom, range_slope_corr_non_sscom, file = "../text/values.rda")

### 2015-07-20 text/ms.Rmd
edits from JCW - a few minor typos, and lots of comments and suggestions added at the top.

@@
 We selected data-limited models for their applicability across the majority of fish stocks world wide.
-We build our models on a simulated dataset of 5760 stocks and tested them with cross-validation and a global database of `r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
+We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and a global database of `r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
 We find that ensemble methods substantially improve estimates of population status and trajectory.

### 2015-07-20 text/ms.Rmd
Jess edits - minor typos and many comments at top of doc

@@
 ```
+
+<!--Jessicas Comments: 
+1.
+Massive improvement in abstract and introduction.
+I really like the strucutre and how you explain the models.
+makes sense to people who aren't familiar with the problems around datalimited fisheries, and the ideas and need for ensemble models.
+This is an amazing paper - quite complicated, but well explained to show people why it matters and how they can use it.
+2. If I remember correctly, Cons Biol doesn't like having the text wrapped around between two lines with hyphens as the PDF shows. Not sure how to fix that, but just something to note. 
+3.
+is it normal to describe the results and findings at the end of the introduction?
+i see it sometimes, but not sure if Cons BIol like it.
+If not, you could put that final paragraph of introduction at start of discussion.
+4.
+First paragraph in Methods describing simulations line 113 in pdf - you may want to describe the recrutiment variability a bit more.
+I know its not an important detail, but because its in the first paragraph of methods, it might turn people off reading further.
+Most people wouldn't understand what 'multiplcative' (including me) is or may not know what 'process error/noise' refers to.
+5. First paragraph in Methods last sentence - needed a bit of rewording because as was, i got quite confused. hopefully this new sentence makes sense and doesnt change the meaning. 
+6. Fix up Simulation FLR package
+7. Update Version xx downloaded on xx RAM legacy database
+8. line 125 in pdf - Ram legacy database - how many stocks didn't converge? readers might want to know how many stocks you threw away.
+9. make datalimited package and ensembles code public when published
+10.
+Sorrry - dumb questions, but probably should ask to properly understand equations...
+line 184 pdf - what is theta-hat?
+and what do you use it for?
+should you describe this, or is it obvious to more technically minded folk than me?
+I don't fully understand what b is.
+11.
+Also, why do you use 'arg min' and 'a loss function'?
+Am i right in thinking that the first equation is to find out which superensemble models are best predictors (different theta-hats) and the different 'loss functions' are those used to determine bias and accuracy?
+Or am I completely wrong?
+Would it help to explain what arg min and the loss functions are used for (as it took me a long time to figure out!)?
+you could just say 'for performance metrics, see below'
+12. line 202 - is this cross validation of GBM the same as the cross-validation described in the section below about performance metrics?
+13. FIgure 1 - I like it as is. 
+14. Figure 4 - tweak labels when all else is completed.
+15.
+Line 255 from pdf - CMSY predictions were approximately linear  (Fig S4) - Do you mean Fig S4a?
+Maybe I don't know how to interpret these graphs - I don't see how you can extract that info from the CMSY graph.
+I don't really know wht you mean by 'high and low prediction clusters'.
+16.
+line 258-59 - if mPRM and SSCOM are high, then true status tends to be high.
+Is there another example that shows how covariates are exploited.
+The example shown is convincing in Fig S5, but not in Fig S6l, because the predicted B/Bmsy is only just above (light pink shading).
+though ths may not matter.
+you can ignore this comment if you want.
+17. check for consistent US spelling - e.g. behaviour - line 279-280. there might be more that I didn't notice. also - line 290 and 291 - overfitting vs over-fitting. I'm not sure which is correct. 
+18. line 314-15 need a reference showing that SSCOM contributed little to status from GBM. e.g. which figure shows this?
+19. line 334 - should this be "effort dynamics at low biomasses in COM-SIR", not SSCOM? 
+20.
+I said that i'd help with the last paragraph about other uses for ensemble methos.
+If there is space within the word limit, could we add just one other potential use?
+Are the references listed in 350-51 using ensemble models or do they simply provide examples of where ensembles could be used.
+Just check my addition - and delete it if you aren't happy with it.
+I don't know how to insert references - i'll leave that up to you.
+21. fix acknowledgements section
+22. I didn't check references. 
+23. other than that! Awesome! Hope my comments were helpful.
+
+-->
 
@@
 In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments that incorporate all appropriate data (e.g.
-catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
-However, a broad range of data are not available for the majority of fish populations [@fao2014].
+catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
+However, the broad range of data required to conduct stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
 Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: catches and a basic understanding of population productivity [e.g.
@@
 Here, we estimate population status and trajectory of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
-We explore a variety of approaches applied to both simulated and real-world fish stocks and compare predictive performance.
+We apply a variety of approaches to both simulated and real-world fish stocks and compare their predictive performance.
 We find that ensembles, and in particular superensembles, generally improved predictive performance of status and trajectory of fish populations over any single model---especially when considered across multiple dimensions of predictive ability.
@@
 We first developed and tested ensemble methods with a fully factorial simulated dataset [@rosenberg2014] with precisely known status.
-These published simulations included three fish life histories: small pelagic, demersal, and large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamics: a constant harvest rate, a harvest rate coupled with biomass to mimic an open-access single-species fishery, a scenario where harvest rate increases continuously, and a 'roller-coaster' scenario where the harvest rate increases and then decreases.
-Process noise was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
-The simulation included a scenario without observation error and with multiplicative observation error around catch at $N(0, 0.2)$.
+These published simulations included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamics: a constant harvest rate, a harvest rate coupled with biomass to mimic an open-access single-species fishery, a scenario where harvest rate increases continuously, and a 'roller-coaster' scenario where the harvest rate increases and then decreases.
+Process noise (i.e.
+variability in population dynamics) was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
+The simulation also included a scenario with multiplicative observation error around catch at $N(0, 0.2)$ and one without observation error.
 
@@
 * *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass dynamics [@martell2013].
-It requires a prior distribution for $r$ and assigns a prior to the relative proportion of biomass at the end compared to unfished biomass (depletion) based on the percentage of maximum catch at the end of the time series.
+It requires a prior distribution for $r$ and assigns a prior to the relative proportion of biomass at the end of the time series compared to unfished biomass (depletion) based on the percentage of maximum catch at the end of the time series.
 Our version of the model was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
@@
 * *mPRM* (modified panel regression model) is a modified version of the panel-regression model from @costello2012.
-Unlike the other models, mPRM is empirical and not mechanistic---it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
+Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
 The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 for deriving aggregate estimates of median status across multiple stocks.
@@
 where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$ is the mean or slope value the model is fitted to, and $\theta$ is the set of parameters for the regression function.
-For all superensemble models of mean $B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space and exponentiated the predictions.
+For all superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions.
 For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
@@
 We fit the linear model superensemble with second-order interactions.
-We fit two machine learning superensemble models based on regression trees.
+We fit the two machine learning superensemble models based on regression trees.
 Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984].
@@
 \ref{hexagon}).
-CMSY exhibited bimodal predictions of status but had the best rank-order correlation and accuracy scores (Fig.
+CMSY exhibited bimodal predictions of status (Fig.
+\ref{hexagon}a), but had the best rank-order correlation and accuracy scores (Fig.
 \ref{performance}a).
@@
 
-When applied to the stock assessment database, the superensemble models---trained exclusively on the simulated dataset---generally performed as well or better than the best individual models.
+When applied to the stock assessment database, the superensemble models --- trained exclusively on the simulated dataset --- generally performed as well or better than the best individual models.
 The mean, random forest, and GBM ensembles outperformed the mPRM method which is trained directly on the RAM Legacy Stock Assessment database itself (Fig.
@@
 
-Ensemble methods provide a useful approach to situations where conservation and ecological resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status---a situation common to many settings beyond fisheries.
+Ensemble methods provide a useful approach to situations where conservation and environmental resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status --- a situation common to many settings beyond fisheries.
 Compared to individual models of fish population status, ensemble methods were consistently the best or among the best across three performance dimensions (accuracy, bias, and rank-order correlation), two response variables (status and trajectory), two datasets (simulated and global fisheries), and multiple ensemble methods (from a simple average to machine learning superensembles).
-Our results suggest choosing an superensemble model that allows for non-parametric relationships, such as machine learning methods.
+Our results suggest choosing a superensemble model that allows for non-parametric relationships, such as machine learning methods.
 These methods provide added insight into individual model behaviour and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
@@
 coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
-Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], or to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014].
+Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation or environmental management actions when experimental or observational data are limited *(Walsh et al.
+2012)*.
+<!-- insert reference here: Walsh, J.C., Wilson, K.A., Benshemesh, J., Possingham, H.P., 2012.
+Unexpected outcomes of invasive predator control: the importance of evaluating conservation management actions.
+Animal Conservation 15, 319–328.
+-->
 

### 2015-07-21 text/ms.Rmd

@@
 catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
-However, the broad range of data required to conduct stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
+However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
 Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: catches and a basic understanding of population productivity [e.g.
@@
 The simulation also included a scenario with multiplicative observation error around catch at $N(0, 0.2)$ and one without observation error.
+<!--TODO: make last sentence clearer-->
 
@@
 coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
-Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation or environmental management actions when experimental or observational data are limited *(Walsh et al.
-2012)*.
-<!-- insert reference here: Walsh, J.C., Wilson, K.A., Benshemesh, J., Possingham, H.P., 2012.
-Unexpected outcomes of invasive predator control: the importance of evaluating conservation management actions.
-Animal Conservation 15, 319–328.
--->
-
+Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].
 <!--

### 2015-07-21 text/ms.Rmd
Incorporate more of Jessica's comments

@@
 
-<!--Jessicas Comments: 
-1.
-Massive improvement in abstract and introduction.
-I really like the strucutre and how you explain the models.
-makes sense to people who aren't familiar with the problems around datalimited fisheries, and the ideas and need for ensemble models.
-This is an amazing paper - quite complicated, but well explained to show people why it matters and how they can use it.
-2. If I remember correctly, Cons Biol doesn't like having the text wrapped around between two lines with hyphens as the PDF shows. Not sure how to fix that, but just something to note. 
-3.
-is it normal to describe the results and findings at the end of the introduction?
-i see it sometimes, but not sure if Cons BIol like it.
-If not, you could put that final paragraph of introduction at start of discussion.
-4.
-First paragraph in Methods describing simulations line 113 in pdf - you may want to describe the recrutiment variability a bit more.
-I know its not an important detail, but because its in the first paragraph of methods, it might turn people off reading further.
-Most people wouldn't understand what 'multiplcative' (including me) is or may not know what 'process error/noise' refers to.
-5. First paragraph in Methods last sentence - needed a bit of rewording because as was, i got quite confused. hopefully this new sentence makes sense and doesnt change the meaning. 
-6. Fix up Simulation FLR package
-7. Update Version xx downloaded on xx RAM legacy database
-8. line 125 in pdf - Ram legacy database - how many stocks didn't converge? readers might want to know how many stocks you threw away.
-9. make datalimited package and ensembles code public when published
-10.
-Sorrry - dumb questions, but probably should ask to properly understand equations...
-line 184 pdf - what is theta-hat?
-and what do you use it for?
-should you describe this, or is it obvious to more technically minded folk than me?
-I don't fully understand what b is.
-11.
-Also, why do you use 'arg min' and 'a loss function'?
-Am i right in thinking that the first equation is to find out which superensemble models are best predictors (different theta-hats) and the different 'loss functions' are those used to determine bias and accuracy?
-Or am I completely wrong?
-Would it help to explain what arg min and the loss functions are used for (as it took me a long time to figure out!)?
-you could just say 'for performance metrics, see below'
-12. line 202 - is this cross validation of GBM the same as the cross-validation described in the section below about performance metrics?
-13. FIgure 1 - I like it as is. 
-14. Figure 4 - tweak labels when all else is completed.
+<!--TODO: Jessicas Comments: 
 15.
@@
 you can ignore this comment if you want.
-17. check for consistent US spelling - e.g. behaviour - line 279-280. there might be more that I didn't notice. also - line 290 and 291 - overfitting vs over-fitting. I'm not sure which is correct. 
 18. line 314-15 need a reference showing that SSCOM contributed little to status from GBM. e.g. which figure shows this?
@@
 I don't know how to insert references - i'll leave that up to you.
-21. fix acknowledgements section
-22. I didn't check references. 
-23. other than that! Awesome! Hope my comments were helpful.
+-->
 
+<!-- 
+TODO
+- justified or ragged?
+- spell check American
 -->
@@
 The simulation also included a scenario with multiplicative observation error around catch at $N(0, 0.2)$ and one without observation error.
-<!--TODO: make last sentence clearer-->
+<!--TODO: make last sentence clearer
+4.
+First paragraph in Methods describing simulations line 113 in pdf - you may want to describe the recrutiment variability a bit more.
+I know its not an important detail, but because its in the first paragraph of methods, it might turn people off reading further.
+Most people wouldn't understand what 'multiplcative' (including me) is or may not know what 'process error/noise' refers to.
+5. First paragraph in Methods last sentence - needed a bit of rewording because as was, i got quite confused. hopefully this new sentence makes sense and doesnt change the meaning. 
+-->
 
@@
 We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012].
-Our analysis of the stock-assessment database was based on version XX downloaded on XX.
-After removing stocks for which the models described below did not converge, this database included `r ram_stocks_n` stocks.
+Our analysis of the stock-assessment database was based on version XX downloaded on XX (TODO).
+After removing stocks for which the models described below did not converge (TODO: add number), this database included `r ram_stocks_n` stocks.
 
@@
 where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch.
-@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` for the statistical software \textsf{R}.
+@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}.
 In summary:
@@
 For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
+<!--add theta-hat description, b hat, make sure these are right, maybe simplify, describe arg min-->
 
@@
 Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
+<!--TODO give more details on tuning or gloss over more-->
 
@@
 We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 1$ at any decision threshold by plotting receiver operating characteristic curves.
-Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles>.
+Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles> (TODO: make public).
 
@@
 Our results suggest choosing a superensemble model that allows for non-parametric relationships, such as machine learning methods.
-These methods provide added insight into individual model behaviour and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
+These methods provide added insight into individual model behavior and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
 
@@
 Second, ensemble models will be most effective when they are not overfit to the training dataset.
-Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust to over-fitting such as random forests [@breiman2001], may help avoid overfitting ensemble models.
+Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust to overfitting such as random forests [@breiman2001], may help avoid overfitting ensemble models.
 We note that our simplest ensemble model, an average of individual model predictions, performed approximately as well as complex machine learning models when we trained our superensembles on the simulation dataset and tested them on a separate 'real' dataset (i.e.

### 2015-07-27 text/ms.Rmd
Trivial tweaks
Incorporate a round of co-author comments

@@
 <!--TODO: Jessicas Comments: 
+TODO - should the table go in the main text, new fig, grid image?
 15.
@@
 Giacomo Chato Osio^7^, 
-Dan Ovando^8^, 
+Daniel Ovando^8^, 
 Andrew A. Rosenberg^9^, 
@@
 \noindent
-^5^NOAA/Northeast Fisheries Science Center, Ecosystem Assessment Program, 
-166 Water St., Woods Hole, MA, 02543, USA
+^5^National Marine Fisheries Service, National Oceanographic and Atmospheric
+Administration, Northeast Fisheries Science Center, Ecosystem Assessment
+Program, 166 Water St., Woods Hole, MA, 02543, USA
 
@@
 Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trajectory.
-We fit four data-limited models of population and exploitation dynamics to marine fish stocks and combine their estimates of the mean and slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
-We selected data-limited models for their applicability across the majority of fish stocks world wide.
-We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and a global database of `r ram_stocks_n` stock assessments that have also been assessed with more data-intensive stock assessment models.
-We find that ensemble methods substantially improve estimates of population status and trajectory.
-Machine-learning superensembles performed the best for population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We fit four data-limited models of population and exploitation dynamics to marine fish stocks and combine their estimates of the mean and recent trends in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
+We selected data-limited models for minimal data requirements and thus wide applicability across the majority of the world's fish stocks.
+We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments that incorporated data-intensive stock assessment models.
+We found that ensemble methods substantially improved estimates of population status and trajectory.
+Machine-learning superensembles performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
 We found similar improvements when predicting the trajectory of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
-Ensemble methods have the potential to improve estimates of population status and trajectory across many taxa beyond fish.
+Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species.
 However, they must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations to which they are applied.
@@
 These properties are key to everything from global assessments of extinction risk [e.g.
-@stuart2004; @schipper2008] to assessing the sustainability of hunting and fishing exploitation levels [e.g.
+@stuart2004; @schipper2008] to assessing the sustainability of hunting [e.g.
+@fa2002] and fishing exploitation levels [e.g.
 @worm2009; @juan-jorda2011].
@@
 Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
+
+<!-- One solution is to...
+TODO KL: As part of setting the stage, right off the bat I'd delimit the set of conditions that make ensemble models an appropriate and useful choice, e.g.: When there are several models (how many to make it useful?) and there is a way to validate them, one solution is..."-->
 
@@
 
-In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments that incorporate all appropriate data (e.g.
-catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$).
+In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
+catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992].
 However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
-Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: catches and a basic understanding of population productivity [e.g.
+Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: (1) a time series of the total biomass of catch and (2) a basic understanding of population productivity [e.g.
 @vasconcellos2005; @martell2013].
 Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment.
-The four models frequently disagreed about population status and any one model had relatively poor predictive ability on average (e.g.
+The four models frequently disagreed about population status and any one model had relatively poor predictive ability on average, but there was evidence suggesting that some models performed better than others depending on circumstances (e.g.
 Fig.
@@
 We apply a variety of approaches to both simulated and real-world fish stocks and compare their predictive performance.
-We find that ensembles, and in particular superensembles, generally improved predictive performance of status and trajectory of fish populations over any single model---especially when considered across multiple dimensions of predictive ability.
-Furthermore, the output from superensembles can aid mechanistic understanding of how individual models of status and trajectory perform.
+*TODO Add a few sentences on the rationale for how the approaches were identified and the testing was designed.*
+
+<!--We find that ensembles, and in particular superensembles, generally improved predictive performance of status and trajectory of fish populations over any single model---especially when considered across multiple dimensions of predictive ability.
+Furthermore, the output from superensembles can aid mechanistic understanding of how individual models of status and trajectory perform.-->
 
@@
 
-We first developed and tested ensemble methods with a fully factorial simulated dataset [@rosenberg2014] with precisely known status.
-These published simulations included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of how depleted biomass was at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamics: a constant harvest rate, a harvest rate coupled with biomass to mimic an open-access single-species fishery, a scenario where harvest rate increases continuously, and a 'roller-coaster' scenario where the harvest rate increases and then decreases.
+We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with precisely known status.
+These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of biomass depletion at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamic settings: (1) a constant harvest rate, (2) a harvest rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where harvest rate increases continuously, and (4) a 'roller-coaster' scenario where the harvest rate increases and then decreases.
 Process noise (i.e.
@@
 After removing stocks for which the models described below did not converge (TODO: add number), this database included `r ram_stocks_n` stocks.
+<!-- Dan: Do we address the implications of only running our method on stocks that converged later on?
+i.e.
+is there something about those omitted stocks that would make them less suited for this type of analysis?-->
 
@@
 
-We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
+We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer (logistic) biomass dynamics [@schaefer1954] of the form
 
@@
 where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch.
+The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database.
 @rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}.
@@
 It requires a prior distribution for $r$ and assigns a prior to the relative proportion of biomass at the end of the time series compared to unfished biomass (depletion) based on the percentage of maximum catch at the end of the time series.
-Our version of the model was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
+The version of the model used in @rosenberg2014 was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
 
-* *COM-SIR* (catch-only-model with sample-importance-resampling) is a coupled harvest-dynamics model [@vasconcellos2005].
+* *COM-SIR* (catch-only-model with sampling-importance-resampling) is a coupled harvest-dynamics model [@vasconcellos2005].
 Biomass is assumed to follow a Schaefer model and harvest dynamics are assumed to follow a logistic model.
-The model is fit with a sample-importance-sampling algorithm [@rosenberg2014].
+The model is fit with a sampling-importance-sampling algorithm [@rosenberg2014].
 
@@
 Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
-The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 for deriving aggregate estimates of median status across multiple stocks.
+The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.
 
@@
 Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data.
-For simplicity, and to allow us to apply models developed with the simulated dataset to the real-world dataset, we added only one set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length and absolute magnitude of catch.
+For simplicity, and since many possible covariates, such as fishing effort dynamics, would not be known for many real-world data-limited scenarios, we added only one set of additional covariates: spectral properties of the catch time series.
+Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch.
 We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
+*TODO add new supp.
+fig.?*
+
+<!-- TODO KL: I think it's worth explaining that we were interested in how the predictions changed wrt the shape of the real time-series, i.e.
+how their cyclic patterns of peaks and troughs matched, and that's why other summary statistics, dependent on abundance or length but insensitive to the shape of the trajectory, weren't chosen.-->
 
@@
 
-The individual models we seek to combine with superensembles provide time series of $B/B_\mathrm{MSY}$.
-Therefore, we can use superensembles to estimate multiple properties of these status time series.
+The individual models we seek to combine with superensembles provide time series of stock status ($B/B_\mathrm{MSY}$).
+Therefore, we can use superensembles to estimate multiple properties of these time series.
 Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
-Together, these quantities address the recent state and trajectory of status, which are both of management and conservation interest [e.g.
+Together, these quantities address the recent state and trajectory of stock status, which are both of management and conservation interest [e.g.
 @mace2008].
@@
 We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits.
-Since the dynamics of simulated populations differing only in the stochastic draw of random values are likely similar, we grouped these stocks (10 per factorial cell) in the cross-validation process into either the training or testing split.
+In the simulated dataset, there were 10 replicates of each unique combination of simulation parameters that differed only in stochastic variability.
+Since the dynamics of these populations were often similar, we grouped these stocks in the cross-validation process into either the training or testing split.
 When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset.
@@
 Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles> (TODO: make public).
+When cross-validating against the RAM Legacy Stock Assessment database, we treat the estimates from these data-rich stock assessments as known without error.
+Thus, any error in the stock assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
+
+*TODO Olaf: B/Bmsy = 0.5 may be a better threshold as this corresponds to the overfished definition in the US and a few other places.
+As Ray always points out, if Bmsy is the target, then you’d expect to be below it about half of the time in a well managed fishery.*
+
+<!-- TODO PE ....
+mean or slope of b/bmsy: Dan: logged or normal space?
+For the status regression where you are estimating log(B/Bmsy), wouldn't you need to correct for the retransformation bias in the estimating median proportional error later down?
+I could be missing something though in whether that applied once you've calculated proportional error-->
+
+<!-- TODO Dan: Should we discuss the issue with MAPE that we ran into before, in that something can have bimodal predictions that are really positively or really negatively biased, and the net result its a MAPE of 0?-->
 
@@
 \ref{hexagon}).
-CMSY exhibited bimodal predictions of status (Fig.
+In contrast to the known unimodal distribution of status, CMSY exhibited bimodal predictions (Fig.
 \ref{hexagon}a), but had the best rank-order correlation and accuracy scores (Fig.
 \ref{performance}a).
-COM-SIR and SSCOM both correctly identified a number of low status stocks, but frequently predicted a high status when status was in fact low (Fig.
+COM-SIR and SSCOM both correctly identified a number of stocks with low status, but frequently predicted a high status when status was in fact low (Fig.
 \ref{hexagon}b, c).
@@
 \ref{hexagon}d).
-There was generally little correlation between true and predicted trajectory for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) besides SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
+There was generally little correlation between true and predicted trajectory for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
 
@@
 
+<!-- TODO KMK: "by exploiting interactions between the spectral density predictors and the individual model predictions" I'm not exactly sure I follow this and I suspect that the subtleties would be lost on others.
+Can you explain this a little more?
+-->
+
 When applied to the stock assessment database, the superensemble models --- trained exclusively on the simulated dataset --- generally performed as well or better than the best individual models.
@@
 Compared to the individual models, the machine learning superensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
+
+<!-- TODO Dan: worth noting in the discussion that this is still pretty low: These methods clearly improve the predictive ability of the catch-based DLAS, but we should point out that it hardly moves it to a perfect estimator; there's still a lot of error in any individual estimate-->
 
@@
 
+<!-- Dan:
+A few general comments.
+Great job, this is well written and clear, especially given the esoteric nature of the content.
+In the discussion, can we address actual limitations of ensemble methods, if any?
+We say when it works, when might it not?
+Are there times when it would actually be a bad idea to use ensembles, or would it always be as good or better than indivduals or simple means?
+
+Second, it might be worth mentioning that the ensembles here are helping make sense of conflicting results from the same data.
+Do these methods work if you were trying to compare methods each individually estimated using different data?
+This same method could presumably work for any metric or data, but it's worth noting that we've only tested catch-based methods here.
+
+Lastly, it's worth noting that this approach, as laid out here is for helping improve potentially conflicting estimates of the same metric, B/Bmsy.
+In many cases though a fishery is going to be basing its decisions on streams of multiple indicators, e.g trends in SPR and B/Bmsy.
+While the ensemble methods here can help refine estimates of individual metrics, it's still worth looking at what alternative data streams are telling you if possible.
+E.g.
+if a purely catch based B/Bmsy estimate is high and has a positive slope, but mean lengths and SPR are decreasing, managers will still need to think about how to deal with these conflicting signals.
+So, this ensemble method, as I interpret it, is helpful for resolving conflicts within metric, but doesn't necesarily resolve conflicts across metrics (if I'm understanding these methods right)
+-->
+
 Ensemble methods provide a useful approach to situations where conservation and environmental resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status --- a situation common to many settings beyond fisheries.
@@
 
+<!-- TODO KMK: Does it make a difference that 3 of our models are based on Schaefer? Definitely they have different structural forms, but not sure if this warrants any discussion?-->
+
 Certain conditions will make some ensemble models more effective than others.
-First, ensembles will be most effective when they are comprised of diverse individual models that choose different structural model forms, explore contrasting but plausible ranges of parameter values, or make errors in uncorrelated ways [@ali1996; @dietterich2000; @tebaldi2007].
+First, ensembles will be most effective when they are comprised of diverse individual models that choose different structural model forms, explore contrasting but plausible ranges of parameter values, or make uncorrelated errors [@ali1996; @dietterich2000; @tebaldi2007].
 Such individual models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance of each.
@@
 A fundamental difference is that coefficient averaging focuses on averaging *coefficients* whereas ensembles instead average *predictions*.
-While the outcome will be the same in some cases, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
+Thus, while the outcome will be the same in some cases, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
 parametric and non-parametric models or frequentist and Bayesian predictions).
@@
 
+<!--TODO: Olaf: pick trajectory or slope throughout -->
 A strength of superensembles is that they can be tailored to predict specific response variables.
@@
 Instead of viewing the superensemble as a black box, we think considerable mechanistic understanding can be gained by studying its structure.
-For example, when SSCOM estimates very low status this is likely the case, conversely when COMSIR estimates low status, the truth is likely that the status is high (Fig.
+For example, when SSCOM estimates very low status this is likely the case, conversely when COMSIR estimates low status, the true status is more likely to be high (Fig.
 \ref{partial-sim}).
@@
 Were the methods to differ only in effort dynamics, the results point towards a more suitable representation of effort dynamics at low biomasses in SSCOM.
-We think that such investigation of the parameters of a superensemble may lead to improvement in the mechanisms imparted to the individual models.
+We think that such investigation of the parameters of a superensemble may lead to improvement in the mechanisms assumed in the individual models.
 
@@
 the IUCN Red List, @iucn2015] levels.
-These risk assessments generally involve fitting regression models to outcomes for individual species along with predictors of species risk [e.g.
+These risk assessments generally involve fitting regression models to outcomes for individual species along with predictors of extinction risk [e.g.
 @anderson2011a; @pinsky2011], or population dynamics models to data for individual species [e.g.

### 2015-07-27 analysis/8-main-figs.R
Switch ROC to > 0.5 b/bmsy

@@ -93,8 +93,9 @@ get_roc <- function(true, est) {
 }
 
 # get ROC and AUC for models applied to simulation data with b/bmsy as the response:
+# we'll use 0.5 B/Bmsy as the treshold as in US overfished definition
 d_sim_mean <- d_sim %>% filter(type == "mean") %>%
-  mutate(above1 = ifelse(bbmsy_true > 1, 1, 0))
+  mutate(above1 = ifelse(bbmsy_true > 0.5, 1, 0))
 rocs_sim_mean <- d_sim_mean %>% group_by(clean_method, order) %>%
   do({get_roc(true = .$above1, est = .$bbmsy_est)}) %>%
   as.data.frame
@@ -109,7 +110,7 @@ saveRDS(auc_sim_mean, file = "generated-data/auc_sim_mean.rds")
 
 # and do the same for the RAM data:
 d_ram_mean <- d_ram %>%
-  mutate(above1 = ifelse(bbmsy_true > 1, 1, 0))
+  mutate(above1 = ifelse(bbmsy_true > 0.5, 1, 0))
 rocs_ram_mean <- d_ram_mean %>% group_by(clean_method, order) %>%
   do({get_roc(true = .$above1, est = .$bbmsy_est)}) %>%
   as.data.frame
@@ -126,8 +127,8 @@ make_roc_curves <- function(dat) {
     geom_line() + coord_equal() +
     geom_abline(intercept = 1, slope = 1, lty = 2, col = "darkgrey") +
     xlim(1, 0) +
-    xlab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as < 1)) +
-    ylab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as > 1)) +
+    xlab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as <= 0.5)) +
+    ylab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as > 0.5)) +
     theme_bw() +
     theme(plot.background = element_blank(),
       panel.grid.major = element_blank(),

### 2015-07-27 analysis/5-simulation-ensembles.R
abs() PE denominator
- I believe this is necessary for the slope bias calculations

@@ -378,7 +378,7 @@ saveRDS(cv_sim_long, "generated-data/cv_sim_long.rds")
 
 re <- cv_sim_long %>% mutate(
   sq_er = (bbmsy_est - bbmsy_true)^2,
-  re    = (bbmsy_est - bbmsy_true) / bbmsy_true)
+  re    = (bbmsy_est - bbmsy_true) / abs(bbmsy_true))
 
 re %>% saveRDS(file = "generated-data/cv_sim_long.rds")
 

### 2015-07-27 analysis/8-main-figs.R

@@ -65,7 +65,7 @@ d_ram <- filter(d_ram, bbmsy_true < 4, bbmsy_est < 4)
 
 re_ram <- d_ram %>% mutate(
   sq_er = (bbmsy_est - bbmsy_true)^2,
-  re    = (bbmsy_est - bbmsy_true) / bbmsy_true)
+  re    = (bbmsy_est - bbmsy_true) / abs(bbmsy_true))
 
 re_ram_sum <- re_ram %>% group_by(clean_method) %>%
   summarise(mare = median(abs(re)),

### 2015-07-27 analysis/5-simulation-ensembles.R
Delete extra line

@@ -374,7 +374,6 @@ cv_sim_slope_long <- cv_sim_slope %>%
 
 cv_sim_long <- suppressWarnings(
   dplyr::bind_rows(cv_sim_mean_long, cv_sim_slope_long))
-saveRDS(cv_sim_long, "generated-data/cv_sim_long.rds")
 
 re <- cv_sim_long %>% mutate(
   sq_er = (bbmsy_est - bbmsy_true)^2,

### 2015-07-27 text/ms.Rmd
Incorporate more comments

@@
 \noindent
-^2^National Center for Ecological Analysis and Synthesis
+^2^National Center for Ecological Analysis and Synthesis, University of
+California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
+USA
 
@@
 5. First paragraph in Methods last sentence - needed a bit of rewording because as was, i got quite confused. hopefully this new sentence makes sense and doesnt change the meaning. 
+TODO: Jamie: add table instead / in addition to this paragraph?
 -->
@@
 
-We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012].
+We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world.
 Our analysis of the stock-assessment database was based on version XX downloaded on XX (TODO).
@@
 When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset.
-However, since mPRM is built on the same stock assessment database, we refit mPRM on each training split.
+However, since mPRM is built on the same stock-assessment database, we refit mPRM on each training split.
 
@@
 @walther2005].
-Here, we measure proportional error, defined as $(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$.
+Here, we measure proportional error, defined as $(\hat{b} - b)/\abs{b}$, where $\hat{b}$ and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$.
 We calculated median proportional error to measure bias, median absolute proportional error to measure accuracy, and Spearman's rank-order correlation between predicted and 'true' values to measure the ability to correctly rank populations.
-We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 1$ at any decision threshold by plotting receiver operating characteristic curves.
+When cross-validating against the RAM Legacy Stock Assessment database, we treated the estimates from these data-rich stock assessments as known without error.
+Thus, any error in the stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
+We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
 Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles> (TODO: make public).
-When cross-validating against the RAM Legacy Stock Assessment database, we treat the estimates from these data-rich stock assessments as known without error.
-Thus, any error in the stock assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
-
-*TODO Olaf: B/Bmsy = 0.5 may be a better threshold as this corresponds to the overfished definition in the US and a few other places.
-As Ray always points out, if Bmsy is the target, then you’d expect to be below it about half of the time in a well managed fishery.*
 
@@
 
-Applied to the simulated dataset of known status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of $B/B_\mathrm{MSY}$ in the last five years.
+Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of $B/B_\mathrm{MSY}$ in the last five years.
 All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig.
@@
 
-When applied to the stock assessment database, the superensemble models --- trained exclusively on the simulated dataset --- generally performed as well or better than the best individual models.
+When applied to the stock-assessment database, the superensemble models --- trained exclusively on the simulated dataset --- generally performed as well or better than the best individual models.
 The mean, random forest, and GBM ensembles outperformed the mPRM method which is trained directly on the RAM Legacy Stock Assessment database itself (Fig.

### 2015-08-15 text/ms.Rmd
Edits to ms

@@
 
-<!-- One solution is to...
-TODO KL: As part of setting the stage, right off the bat I'd delimit the set of conditions that make ensemble models an appropriate and useful choice, e.g.: When there are several models (how many to make it useful?) and there is a way to validate them, one solution is..."-->
-
 One solution is to take the average or weighted average of an ensemble of model predictions.
@@
 Here, we estimate population status and trajectory of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
-We apply a variety of approaches to both simulated and real-world fish stocks and compare their predictive performance.
-*TODO Add a few sentences on the rationale for how the approaches were identified and the testing was designed.*
-
-<!--We find that ensembles, and in particular superensembles, generally improved predictive performance of status and trajectory of fish populations over any single model---especially when considered across multiple dimensions of predictive ability.
-Furthermore, the output from superensembles can aid mechanistic understanding of how individual models of status and trajectory perform.-->
+We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and against individual model productions.
 
@@
 <!--
-We tested the ability of ensemble models to improve estimates of population
-status and trajectory when applied to both a large simulated dataset and a
-global database of assessed stock status. Here we describe the datasets,
-individual models of population status, and ensemble methods to combine those
-estimates. We then describe how we evaluated the ability of the various models
-to estimate population status.
--->
+We tested the ability of ensemble methods to improve estimates of population status and trajectory when applied to both a large simulated dataset and a global database of assessed stock status.
+Here we describe the datasets, individual models of population status, and ensemble methods to combine those estimates.
+We then describe how we evaluated the ability of the various models to estimate population status.
 
-<!--
 ## Datasets
@@
 
-We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with precisely known status.
-These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of biomass depletion at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four harvest dynamic settings: (1) a constant harvest rate, (2) a harvest rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where harvest rate increases continuously, and (4) a 'roller-coaster' scenario where the harvest rate increases and then decreases.
-Process noise (i.e.
-variability in population dynamics) was introduced to the models through two levels of multiplicative recruitment variability: $N(0, 0.2)$ and $N(0, 0.6)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
-The simulation also included a scenario with multiplicative observation error around catch at $N(0, 0.2)$ and one without observation error.
-<!--TODO: make last sentence clearer
-4.
-First paragraph in Methods describing simulations line 113 in pdf - you may want to describe the recrutiment variability a bit more.
-I know its not an important detail, but because its in the first paragraph of methods, it might turn people off reading further.
-Most people wouldn't understand what 'multiplcative' (including me) is or may not know what 'process error/noise' refers to.
-5. First paragraph in Methods last sentence - needed a bit of rewording because as was, i got quite confused. hopefully this new sentence makes sense and doesnt change the meaning. 
+We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with known status.
+These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of biomass depletion at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) a exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increases continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increases and then decreases.
+Process noise (recruitment variability; i.e.
+unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
+The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.
+
+<!--
 TODO: Jamie: add table instead / in addition to this paragraph?
@@
 The simulation models were built in the package `FLR` [@kell2007] for the statistical software \textsf{R} [@r2015].
-The code to generate the simulations is available at <https://github.com/flr/StockSims> (TODO: not there).
+The code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.
 
@@
 After removing stocks for which the models described below did not converge (TODO: add number), this database included `r ram_stocks_n` stocks.
-<!-- Dan: Do we address the implications of only running our method on stocks that converged later on?
+<!-- TODO Dan: Do we address the implications of only running our method on stocks that converged later on?
 i.e.
@@
 \noindent 
-where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$ is the mean or slope value the model is fitted to, and $\theta$ is the set of parameters for the regression function.
+where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $\hat{b}$ are the mean or slope values the model is fitted to, and $\hat{\theta}$ is the set of estimated parameters for the superensemble regression function.
+The $\argmin_\theta$ refers to finding the set of $\theta$ that minimize the loss function.
 For all superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions.
@@
 Ensemble methods provide a final calibration step to a quantity of interest.
-This ensemble calibration could further include a loss function tailored to the goals of the model, say placing greater weight on accuracy at low status levels than higher values.
+This ensemble calibration could further include a loss function tailored to
+ the goals of the model, say placing greater weight on accuracy at low status levels than higher values.
 Conversely, because superensembles are tailored to a specific response and loss function, superensembles force a modeller to choose an operational purpose for their model up front [*sensu* @dickey-collas2014].

### 2015-09-14 analysis/1-read-old-ram-fits.R
fix path

@@ -14,7 +14,7 @@ if (Sys.info()[["user"]] == "seananderson") {
   all.stocks<-all.stocks[order(all.stocks)]
 
   ## old prior
-  setwd("../ensembles-old/RAM_Legacy_fits/All_fits/")
+  setwd("../../ensembles-old/RAM_Legacy_fits/All_fits/")
   load("../CMSY/cmsy_rlegacy_results_table_v0.RData")
   cmsy.oldprior.df<-cmsy.rlegacy.df0
   ## new prior
@@ -72,6 +72,6 @@ if (Sys.info()[["user"]] == "seananderson") {
     sscom.df[,keep.names]
     )
   all.fits.df$stock <- as.character(all.fits.df$stock)
-  setwd("../../../ensembles/")
+  setwd("../../../ensembles/analysis/")
   saveRDS(all.fits.df, "generated-data/ram-orig-fits.rds")
 }

### 2015-09-14 text/ms.Rmd
incorporate comments and edit manuscript

@@
 ```
-
-<!--TODO: Jessicas Comments: 
-TODO - should the table go in the main text, new fig, grid image?
-15.
-Line 255 from pdf - CMSY predictions were approximately linear  (Fig S4) - Do you mean Fig S4a?
-Maybe I don't know how to interpret these graphs - I don't see how you can extract that info from the CMSY graph.
-I don't really know wht you mean by 'high and low prediction clusters'.
-16.
-line 258-59 - if mPRM and SSCOM are high, then true status tends to be high.
-Is there another example that shows how covariates are exploited.
-The example shown is convincing in Fig S5, but not in Fig S6l, because the predicted B/Bmsy is only just above (light pink shading).
-though ths may not matter.
-you can ignore this comment if you want.
-18. line 314-15 need a reference showing that SSCOM contributed little to status from GBM. e.g. which figure shows this?
-19. line 334 - should this be "effort dynamics at low biomasses in COM-SIR", not SSCOM? 
-20.
-I said that i'd help with the last paragraph about other uses for ensemble methos.
-If there is space within the word limit, could we add just one other potential use?
-Are the references listed in 350-51 using ensemble models or do they simply provide examples of where ensembles could be used.
-Just check my addition - and delete it if you aren't happy with it.
-I don't know how to insert references - i'll leave that up to you.
--->
 
@@
 \noindent
-Improving estimates of population status and trajectory with superensemble models
+Improving estimates of population status and trend with superensemble models
 \end{Large}
@@
 Andrew B. Cooper^1^,
-Olaf P. Jensen^4^, 
-C\'{o}il\'{i}n Minto^6^, 
-James T. Thorson^11^,
+Olaf P. Jensen^2^, 
+C\'{o}il\'{i}n Minto^3^, 
+James T. Thorson^4^,
 Jessica C. Walsh^1^,
-Jamie Afflerbach^2^, 
-Mark Dickey-Collas^3,12^, 
-Kristin M. Kleisner^5^, 
-Catherine Longo^2^, 
-Giacomo Chato Osio^7^, 
-Daniel Ovando^8^, 
-Andrew A. Rosenberg^9^, 
-Elizabeth R. Selig^10^
-
-\noindent
-\textit{order to be determined and superscripts out of order}
+Jamie Afflerbach^5^, 
+Mark Dickey-Collas^6,7^, 
+Kristin M. Kleisner^8^, 
+Catherine Longo^5^, 
+Giacomo Chato Osio^9^, 
+Daniel Ovando^10^, 
+Iago Mosqueira^9^,
+Andrew A. Rosenberg^11^, 
+Elizabeth R. Selig^12^
 
@@
 \noindent
-^2^National Center for Ecological Analysis and Synthesis, University of
-California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
-USA
+^2^Institute of Marine & Coastal Sciences, Rutgers University, 
+71 Dudley Road, New Brunswick, NJ, 08901-8525, USA
 
 \noindent
-^3^International Council for the Exploration of the Sea, 
-H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark
+^3^Galway-Mayo Institute of Technology, 
+Marine and Freshwater Research Centre, Galway, 00000, Ireland
 
 \noindent
-^4^Institute of Marine & Coastal Sciences, Rutgers University, 
-71 Dudley Road, New Brunswick, NJ, 08901-8525, USA
+^4^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112
 
@@
 \noindent
-^6^Galway-Mayo Institute of Technology, 
-Marine and Freshwater Research Centre, Galway, 00000, Ireland
+^6^International Council for the Exploration of the Sea, 
+H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark
 
 \noindent
-^7^EC JRC, IPSC, MAU, Italy
+^7^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark
 
 \noindent
-^8^University of California Santa Barbara, 
+^8^National Center for Ecological Analysis and Synthesis, University of
+California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
+USA
+
+\noindent
+^9^European Commission, Joint Research Centre, Institute for the Protection and
+Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi 2749, 21027
+Ispra (VA), Italy
+
+\noindent
+^10^University of California Santa Barbara, 
 Bren School of Environmental Science and Management, 
@@
 \noindent
-^9^Union of Concerned Scientists, Cambridge, MA, USA
+^11^Union of Concerned Scientists, Cambridge, MA, USA
 
 \noindent
-^10^Conservation International, Arlington, VA, USA
+^12^Conservation International, Arlington, VA, USA
 
 \noindent
-^11^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112
-
-\noindent
-^12^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark
-
-\noindent
-^\*^Corresponding author: Sean C. Anderson,
-School of Resource and Environmental Management,
-Simon Fraser University,
-Burnaby BC, V5A 1S6;
-E-mail: sean_anderson@sfu.ca
-
-\clearpage
-
-# Abstract
-
-<!--up to 300 words-->
-
-Ecological resource managers are often faced with reconciling multiple potentially conflicting estimates of population status and trajectory.
-One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
-Superensemble models, commonly used for climate and weather forecasting, may provide a superior solution.
-Superensembles use the predictions from multiple models as covariates in a new statistical model.
-Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trajectory.
-We fit four data-limited models of population and exploitation dynamics to marine fish stocks and combine their estimates of the mean and recent trends in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
-We selected data-limited models for minimal data requirements and thus wide applicability across the majority of the world's fish stocks.
-We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments that incorporated data-intensive stock assessment models.
-We found that ensemble methods substantially improved estimates of population status and trajectory.
-Machine-learning superensembles performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
-We found similar improvements when predicting the trajectory of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
-Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species.
-However, they must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations to which they are applied.
+^\*^Corresponding author, present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA  98195, USA
+E-mail: sandrsn@uw.edu
 
@@
 Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
+
+\bigskip
+\noindent 
+Word count: TODO
+
+\bigskip
+<!--Place the acknowledgment paragraph on the cover page of your manuscript.
+(Reviewers are not provided with a cover page.) Do not spell out first (given)
+names. Provide the first initial of the first name, even if the initial starts a
+sentence. Do not use titles (e.g., Dr. or Professor). Refer to authors of the
+manuscript by their initials only (e.g., “S.T.W. was supported by a grant from
+the Torry Foundation.”).-->
+\noindent
+Acknowledgements: We thank members of Phase I of the working group 'INSERT NAME HERE' who
+contributed to developing the data-limited methods and simulations used in our
+analysis.
+We thank J.A. Hutchings for helpful discussions on ensemble methods for
+assessing population status.
+We thank the Gordon and Betty Moore Foundation for funding of the working group
+'INSERT NAME HERE'.
+'INSERT OTHER REQUIRED FUNDING ACKNOWLEDGEMENTS HERE'
+
+\clearpage
+
+\begin{LARGE}
+\noindent
+Improving estimates of population status and trend with superensemble models
+\end{LARGE}
+
+<!--up to 300 words-->
+
+\bigskip
+\noindent
+\textbf{Abstract:} Ecological resource managers must often reconcile multiple conflicting estimates of population status and trend.
+One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
+Superensemble models, commonly used in climate and weather forecasting, may provide a superior solution.
+Superensembles use the predictions from multiple models as covariates in an additional statistical model fit to known data.
+Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trend.
+We fit four data-limited and widely applicable models of population and exploitation dynamics and combine the models' estimates of the mean and recent trend in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
+We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments.
+We found that ensemble methods substantially improved estimates of population status and trend.
+Machine-learning superensembles (random forest and boosted regression trees) performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements when predicting the trend of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
+Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species.
+However, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
+
+\bigskip
 <!--Include on the cover page five to eight words or phrases that will be useful for indexing and literature searches.
@@
 
-Population status and trajectory are two of the most fundamental values to quantify in conservation biology [e.g.
+Population status and trend are two of the most fundamental values to quantify in conservation biology [e.g.
 @mace2008; @iucn2015].
@@
 @worm2009; @juan-jorda2011].
-However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g.
+However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trend [e.g.
 @brodziak2010; @branch2011; @deroba2015].
@@
 In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g.
-@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trajectory [e.g.
+@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trend [e.g.
 @brodziak2010].
@@
 
-In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
+In fisheries science, the commonly used operational model for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
 catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992].
@@
 Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment.
-The four models frequently disagreed about population status and any one model had relatively poor predictive ability on average, but there was evidence suggesting that some models performed better than others depending on circumstances (e.g.
+The four models frequently disagreed about population status, no one model had strong performance across all fish stocks, and some models performed better than others depending on circumstances (e.g.
 Fig.
@@
 
-Here, we estimate population status and trajectory of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
+Here, we estimate population status and trend of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
 We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and against individual model productions.
@@
 
-<!--
-We tested the ability of ensemble methods to improve estimates of population status and trajectory when applied to both a large simulated dataset and a global database of assessed stock status.
-Here we describe the datasets, individual models of population status, and ensemble methods to combine those estimates.
-We then describe how we evaluated the ability of the various models to estimate population status.
-
-## Datasets
--->
-
 We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with known status.
-These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of biomass depletion at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) a exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increases continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increases and then decreases.
+These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of initial biomass depletion compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) an exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increased continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increased and then decreased.
 Process noise (recruitment variability; i.e.
@@
 
-<!--
-TODO: Jamie: add table instead / in addition to this paragraph?
--->
-
 @rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-The simulation models were built in the package `FLR` [@kell2007] for the statistical software \textsf{R} [@r2015].
-The code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.
+The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software \textsf{R} [@r2015].
+Code to generate the simulations is available at <https://github.com/datalimited/stocksims>.
 
 We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world.
-Our analysis of the stock-assessment database was based on version XX downloaded on XX (TODO).
-After removing stocks for which the models described below did not converge (TODO: add number), this database included `r ram_stocks_n` stocks.
-<!-- TODO Dan: Do we address the implications of only running our method on stocks that converged later on?
-i.e.
-is there something about those omitted stocks that would make them less suited for this type of analysis?-->
+Our analysis of the stock-assessment database was based on version TODO; after removing stocks for which the models described below did not converge (121), this database included `r ram_stocks_n` stocks.
 
@@
 The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database.
-@rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}.
+@rosenberg2014 provide a full background on these four methods and code to fit the models.
 In summary:
+
+<!--and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}. In summary:-->
 
@@
 Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
-The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.
+The model used in this paper is modified from the original --- it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.
 
@@
 For simplicity, and since many possible covariates, such as fishing effort dynamics, would not be known for many real-world data-limited scenarios, we added only one set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch.
+Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the cyclical shape of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch.
 We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
-*TODO add new supp.
-fig.?*
 
-<!-- TODO KL: I think it's worth explaining that we were interested in how the predictions changed wrt the shape of the real time-series, i.e.
-how their cyclic patterns of peaks and troughs matched, and that's why other summary statistics, dependent on abundance or length but insensitive to the shape of the trajectory, weren't chosen.-->
+<!--*TODO add new supp. fig.?*-->
 
@@
 Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
-Together, these quantities address the recent state and trajectory of stock status, which are both of management and conservation interest [e.g.
+Together, these quantities address the recent state and trend of stock status, which are both of management and conservation interest [e.g.
 @mace2008].
@@
 
-Here we compare an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and boosted regression tree.
-The general form of our models was
+For all superensembles, we used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models, and the spectral values, as the predictors.
+For superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions.
+For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
 
-$$
-\hat{\theta} = 
-\argmin_\theta \left( L(b \, | \, \hat{b}_{\mathrm{ensemble}}) \right)
-$$
-$$
-\hat{b}_\mathrm{ensemble} = 
-f \left( \hat{b}_\mathrm{CMSY}, \, \hat{b}_\mathrm{COM-SIR}, \,
-\hat{b}_\mathrm{SSCOM}, \, \hat{b}_\mathrm{mPRM}, \,
-S(0.2), \, S(0.5) \, | \, \theta \right)
-$$
-
-\noindent 
-where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $\hat{b}$ are the mean or slope values the model is fitted to, and $\hat{\theta}$ is the set of estimated parameters for the superensemble regression function.
-The $\argmin_\theta$ refers to finding the set of $\theta$ that minimize the loss function.
-For all superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions.
-For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
-<!--add theta-hat description, b hat, make sure these are right, maybe simplify, describe arg min-->
-
-We fit the linear model superensemble with second-order interactions.
-We fit the two machine learning superensemble models based on regression trees.
+We compared an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and a boosted regression tree.
+We fit the linear model superensemble with all second-order interactions.
+Our two machine learning superensemble models, a random forest and a boosted regression tree, were based on regression trees.
 Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984].
@@
 We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for \textsf{R}.
-Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
-<!--TODO give more details on tuning or gloss over more-->
+Based on initial tuning with the **caret** \textsf{R} package [@caret2008], we chose to fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
 
@@
 
-A critical component to any predictive modelling exercise is to evaluate the performance of a model on new data [@hastie2009].
+A critical component to predictive modelling is to evaluate the performance of a model on new data [@hastie2009].
 We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built superensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third.
@@
 
-<!-- TODO Dan: Should we discuss the issue with MAPE that we ran into before, in that something can have bimodal predictions that are really positively or really negatively biased, and the net result its a MAPE of 0?-->
-
 # Results
 
-Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of $B/B_\mathrm{MSY}$ in the last five years.
+Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trend) of $B/B_\mathrm{MSY}$ in the last five years.
 All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig.
@@
 \ref{hexagon}d).
-There was generally little correlation between true and predicted trajectory for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
+There was generally little correlation between true and predicted trend for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
 
-Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig.
+Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trend over any individual model (Fig.
 \ref{hexagon}e--h, Fig.
@@
 \ref{performance}a).
-These superensembles also had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig.
+These superensembles also generally had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 0.5$ (Fig.
 \ref{roc-sim}).
-Results were similar when predicting trajectory: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
+Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
 \ref{performance-sim-slope}).
@@
 
-The superensemble models were able to improve the predictions of status by exploiting the best properties of individual models, the covariance between individual models, and interactions with other covariates.
+The superensemble models were able to improve the predictive performance by harnessing the best properties of individual models, the covariance between individual models, and interactions with other covariates.
 For example, SSCOM had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
 \ref{hexagon}c, Fig.
-\ref{partial-sim}c) and CMSY predictions were approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction clusters (Fig.
+\ref{partial-sim}c) and CMSY predictions were approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high clusters of predictions (Fig.
 \ref{partial-sim}).
@@
 The addition of spectral density covariates helped the superensemble models correctly predict higher status values (Fig.
-\ref{hexagon}g, h) by exploiting interactions between the spectral density predictors and the individual model predictions (Fig.
+\ref{hexagon}g, h) due to the interactions between the spectral density predictors and the individual model predictors (Fig.
 \ref{partial-2d-sim}).
@@
 \ref{hexagon-sim-basic}).
-
-<!-- TODO KMK: "by exploiting interactions between the spectral density predictors and the individual model predictions" I'm not exactly sure I follow this and I suspect that the subtleties would be lost on others.
-Can you explain this a little more?
--->
 
@@
 
-<!-- TODO Dan: worth noting in the discussion that this is still pretty low: These methods clearly improve the predictive ability of the catch-based DLAS, but we should point out that it hardly moves it to a perfect estimator; there's still a lot of error in any individual estimate-->
-
-<!--
-Some individual models performed nearly as well as the ensembles in one or two performance dimensions: e.g.
-SSCOM does nearly as well as the ensembles at predicting the slope of $B/B_\mathrm{MSY}$ (Fig.
-\ref{performance-sim-slope}).
-But no individual method is consistently nearly as good as the ensembles.
-SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean $B/B_\mathrm{MSY}$ for both the simulated and stock-assessment datasets (Fig.
-\ref{performance}).
--->
-
 # Discussion
@@
 <!-- Dan:
-A few general comments.
-Great job, this is well written and clear, especially given the esoteric nature of the content.
-In the discussion, can we address actual limitations of ensemble methods, if any?
-We say when it works, when might it not?
-Are there times when it would actually be a bad idea to use ensembles, or would it always be as good or better than indivduals or simple means?
-
-Second, it might be worth mentioning that the ensembles here are helping make sense of conflicting results from the same data.
-Do these methods work if you were trying to compare methods each individually estimated using different data?
-This same method could presumably work for any metric or data, but it's worth noting that we've only tested catch-based methods here.
-
 Lastly, it's worth noting that this approach, as laid out here is for helping improve potentially conflicting estimates of the same metric, B/Bmsy.
@@
 Ensemble methods provide a useful approach to situations where conservation and environmental resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status --- a situation common to many settings beyond fisheries.
-Compared to individual models of fish population status, ensemble methods were consistently the best or among the best across three performance dimensions (accuracy, bias, and rank-order correlation), two response variables (status and trajectory), two datasets (simulated and global fisheries), and multiple ensemble methods (from a simple average to machine learning superensembles).
+Compared to individual models of fish population status, ensemble methods were consistently the best or among the best across three performance dimensions (accuracy, bias, and rank-order correlation), two response variables (status and trend), two datasets (simulated and global fisheries), and multiple ensemble methods (from a simple average to machine learning superensembles).
 Our results suggest choosing a superensemble model that allows for non-parametric relationships, such as machine learning methods.
 These methods provide added insight into individual model behavior and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
-
-<!-- TODO KMK: Does it make a difference that 3 of our models are based on Schaefer? Definitely they have different structural forms, but not sure if this warrants any discussion?-->
 
@@
 
-<!--TODO: Olaf: pick trajectory or slope throughout -->
 A strength of superensembles is that they can be tailored to predict specific response variables.
@@
 The same set of model weights or non-linear relationships need not hold across different response variables.
-For instance, SSCOM contributed little to the GBM superensemble estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$, but contributed strongly to estimates of trajectory (Fig.
+For instance, SSCOM contributed little to the GBM superensemble estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$ (Fig.
+\ref{partial-sim}), but contributed strongly to estimates of trend (Fig.
 \ref{partial-sim-slope}).
@@
 coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
-Beyond estimates of status and trajectory, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].
-<!--
-
-and different models may result in different scalings of population status
-[Deroba JJ, Butterworth DS, Methot RD Jr., De Oliveira JAA, Fernandez C, Nielsen
-A, Cadrin SX, Dickey-Collas M, Legault CM, Ianelli J, et al. 2015. Simulation
-testing the robustness of stock assessment models to error: some results from
-the ICES Strategic Initiative on Stock Assessment Methods. ICES Journal of
-Marine Science 72: 19–30 doi:10.1093/icesjms/fst237]
-CM: It would be good to keep this in mind from a management perspective as
-depending on the assumptions in the assessment, biomass could be high and
-fishing mortality low to predict the same catches as biomass low and fishing
-mortality high. Combining for a single scaled variable B/Bmsy removes some of
-the scaling but when there is a strong covariation in B and F this would need to
-be (and likely can be) accounted for in a bivariate ensemble. Not our focus here
-but could enter the Discussion.
-
-Biological ensemble modeling to evaluate potential futures of living marine
-resources
-Anna Gårdmark1,8, Martin Lindegren2, Stefan Neuenfeldt2, Thorsten Blenckner3,
-Outi Heikinheimo4, Bärbel Müller-Karulis3,5, Susa Niiranen3, Maciej T. Tomczak3,
-Eero Aro4, Anders Wikström6, and Christian Möllmann7
-Read More: http://www.esajournals.org/doi/abs/10.1890/12-0267.1
-
-* Other possible topics:
-    * how much better are the ensembles really? translate the results into
-      some concrete examples in an absolute not relative sense
-    * other types of ensembles: assigning weights, incorporating uncertainty
-      around estimates, GAMs, ...
-    * when wouldn't you want to use ensembles? perhaps if models represent 
-      dichotomous assumptions where either one or the other is right and imply
-      different management actions? ...
-    * your ideas here...
-
-Why do ensembles work?
-
-- 'representational' no one model can contain all 'hypothetical functional
-  forms'... many models can cover many hypotheses @dietterich2000
-- error cancellation --- especially if multiple models make errors in
-  uncorrelated ways [@ali1996]
-- they provide an additional step of 'calibration' [@rykiel1996] where the
-  predictions are calibrated to data of known or assumed truth about a
-  particular aspect of interest in the data (e.g. mean status, slope of status)
-  whereas individual model may be calibrated (parameters estimated) with a
-  slightly different response in mind (e.g. maximizing likelihood with respect
-  to observed and predicted catches)
-- we usually do not know which individual model is 'best' [@hagedorn2005];
-  ensembles have the property of rarely being far from the best
--->
-
-# Acknowledgements
-
-<!--Place the acknowledgment paragraph on the cover page of your manuscript.
-(Reviewers are not provided with a cover page.) Do not spell out first (given)
-names. Provide the first initial of the first name, even if the initial starts a
-sentence. Do not use titles (e.g., Dr. or Professor). Refer to authors of the
-manuscript by their initials only (e.g., “S.T.W. was supported by a grant from
-the Torry Foundation.”).-->
-We thank members of Phase I of the working group 'INSERT NAME HERE' who
-contributed to developing the data-limited methods and simulations used in our
-analysis.
-We thank J.A. Hutchings for helpful discussions on ensemble methods for
-assessing population status.
-We thank the Gordon and Betty Moore Foundation for funding of the working group
-'INSERT NAME HERE'.
-'INSERT OTHER REQUIRED FUNDING ACKNOWLEDGEMENTS HERE'
+Beyond estimates of status and trend, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].
 

### 2015-12-29 text/ms.Rmd
Previous final edits + format switch

@@
 <!--
-TODO
-- justified or ragged?
-- spell check American
+ The full title of the paper, with two alternatives. Keep titles short and simple.
+• The full names of all authors.
+• The name(s) and address(es) of the institution(s) at which the work was carried out. If the present address of any author is different from the above, give it as a footnote.
+• The name, address, telephone, fax and e-mail contacts of the author to whom all correspondence and proofs should be sent.
+• A suggested running title of not more than 40 characters, including spaces.
 -->
-
-<!--notes:
-http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1523-1739/homepage/ForAuthors.html
-
-con bio: American spelling
-
-Contributed Papers: 6000 words
-
-Submit your cover page as a separate document.
-It should not be part of the manuscript itself.
-The cover page must include the title of the paper; a running head (short title of 40 or fewer characters); a list of 5-8 keywords; word count (all text from the first word of the Abstract through the last word of the Literature Cited but not including table or figure legends or the body of tables); authors' complete mailing addresses (including postal code) at the time the work was conducted and present addresses if different; name and complete address (including email) of the person to whom correspondence should be sent; and text of your acknowledgments section.
-
-Contributed Papers, Research Notes, and Conservation Practice and Policy papers should contain the following sections in the following order: Abstract, Introduction, Methods, Results, Discussion, Supporting Information paragraph (if there are online appendices), Literature Cited, tables, figure legends, and figures.
-Do not combine sections (e.g., Results and Discussion).
-The Acknowledgments section will be added to the body of the paper after the manuscript has been accepted.
-Do not number section headings or subheadings.
-Do not include a Conclusion section (conclusions are part of the Discussion).
-
-You may have color figures in the online version and gray-scale figures in print.
-However, reference to color cannot be made in the figure legend or in the text, and elements in the gray-scale version must be distinguishable.
-Supply separate files for color and gray-scale figures.
-
--->
-
-\noindent
-Running head: Superensembles of population status
-
-\bigskip
 
@@
 Kristin M. Kleisner^8^, 
-Catherine Longo^5^, 
-Giacomo Chato Osio^9^, 
-Daniel Ovando^10^, 
-Iago Mosqueira^9^,
-Andrew A. Rosenberg^11^, 
-Elizabeth R. Selig^12^
+Catherine Longo^9^, 
+Giacomo Chato Osio^10^, 
+Daniel Ovando^11^, 
+Iago Mosqueira^10^,
+Andrew A. Rosenberg^12^, 
+Elizabeth R. Selig^13^
 
@@
 \noindent
-^3^Galway-Mayo Institute of Technology, 
-Marine and Freshwater Research Centre, Galway, 00000, Ireland
+^3^Marine and Freshwater Research Centre, Galway-Mayo Institute of Technology, 
+Galway, 00000, Ireland
 
 \noindent
-^4^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112
+^4^Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 2725 Montlake Boulevard E., Seattle, WA, 98112, USA
 
 \noindent
-^5^National Marine Fisheries Service, National Oceanographic and Atmospheric
-Administration, Northeast Fisheries Science Center, Ecosystem Assessment
-Program, 166 Water St., Woods Hole, MA, 02543, USA
+^5^National Center for Ecological Analysis and Synthesis, University of
+California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA, 93103,
+USA
 
@@
 \noindent
-^8^National Center for Ecological Analysis and Synthesis, University of
-California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
-USA
+^8^Ecosystem Assessment Program, Northeast Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 166 Water St., Woods Hole, MA, 02543, USA
 
 \noindent
-^9^European Commission, Joint Research Centre, Institute for the Protection and
+^9^Marine Stewarship Council, Marine House, 1 Snow Hill, London, EC1A 2DH, United Kingdom
+
+\noindent
+^10^European Commission, Joint Research Centre, Institute for the Protection and
 Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi 2749, 21027
@@
 \noindent
-^10^University of California Santa Barbara, 
-Bren School of Environmental Science and Management, 
+^11^Bren School of Environmental Science and Management, 
+University of California Santa Barbara, 
 Santa Barbara, CA, 93106-5131, USA
@@
 \noindent
-^11^Union of Concerned Scientists, Cambridge, MA, USA
+^12^Union of Concerned Scientists, Cambridge, MA, USA
 
 \noindent
-^12^Conservation International, Arlington, VA, USA
+^13^Conservation International, Arlington, VA, USA
 
@@
 
-\bigskip
 \noindent
-Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
-
-\bigskip
-\noindent 
-Word count: TODO
-
-\bigskip
-<!--Place the acknowledgment paragraph on the cover page of your manuscript.
-(Reviewers are not provided with a cover page.) Do not spell out first (given)
-names. Provide the first initial of the first name, even if the initial starts a
-sentence. Do not use titles (e.g., Dr. or Professor). Refer to authors of the
-manuscript by their initials only (e.g., “S.T.W. was supported by a grant from
-the Torry Foundation.”).-->
-\noindent
-Acknowledgements: We thank members of Phase I of the working group 'INSERT NAME HERE' who
-contributed to developing the data-limited methods and simulations used in our
-analysis.
-We thank J.A. Hutchings for helpful discussions on ensemble methods for
-assessing population status.
-We thank the Gordon and Betty Moore Foundation for funding of the working group
-'INSERT NAME HERE'.
-'INSERT OTHER REQUIRED FUNDING ACKNOWLEDGEMENTS HERE'
+Running head: Superensembles of population status
 
 \clearpage
-
-\begin{LARGE}
-\noindent
-Improving estimates of population status and trend with superensemble models
-\end{LARGE}
-
-<!--up to 300 words-->
+<!--up to 250 words-->
 
@@
 \bigskip
-<!--Include on the cover page five to eight words or phrases that will be useful for indexing and literature searches.
-Do not use words in the title as keywords, and avoid general terms such as conservation.-->
+
+\noindent 
+Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
+
+\tableofcontents
 
@@
 @yun2005; @mote2015].
-The superensemble is fit to some training dataset where outcomes are well known and used to predict on a dataset of interest.
+The superensemble is fit to a training dataset where outcomes are well known and used to predict on a dataset of interest (Fig.
+\ref{didactic}).
 For example, @krishnamurti1999 combined predictions of wind and precipitation in Asian monsoons via a superensemble regression fit to observed data.
@@
 
-In fisheries science, the commonly used operational model for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
+In fisheries science, the commonly used operational models for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
 catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992].
@@
 Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment.
+Three of these models are based on Schaefer (logistic) biomass dynamics and one is an empirical model fitted to more data-rich stock assessment output.
 The four models frequently disagreed about population status, no one model had strong performance across all fish stocks, and some models performed better than others depending on circumstances (e.g.
@@
 Here, we estimate population status and trend of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
-We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and against individual model productions.
+We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and the individual models.
+
+<!--
+ TODO: It might be good to throw in a single sentence here describing the 4 models, or at least listing them and maybe a couple key differences among them.-->
 
@@
 Process noise (recruitment variability; i.e.
-unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$.
+unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through time or had a first-order autoregressive correlation of $0.6$.
 The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.
+@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
 
-@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software \textsf{R} [@r2015].
-Code to generate the simulations is available at <https://github.com/datalimited/stocksims>.
+<!--TODO add back when identity public: The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software \textsf{R} [@r2015]. -->
+<!--Code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.-->
 
 We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world.
-Our analysis of the stock-assessment database was based on version TODO; after removing stocks for which the models described below did not converge (121), this database included `r ram_stocks_n` stocks.
+Our analysis of the stock-assessment database was based on version 2.5; after removing stocks for which at least one of the individual models described below did not converge (121), this database included `r ram_stocks_n` stocks.
 
@@
 
-We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer (logistic) biomass dynamics [@schaefer1954] of the form
+We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
 
@@
 
-For all superensembles, we used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models, and the spectral values, as the predictors.
-For superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions.
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models, and the spectral values, as predictors in our superensemble models.
+When modelling mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the superensemble models in log space and exponentiated the predictions.
 For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
@@
 Our two machine learning superensemble models, a random forest and a boosted regression tree, were based on regression trees.
-Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984].
+Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a loss function [@breiman1984].
 In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model [@breiman2001].
@@
 Since the dynamics of these populations were often similar, we grouped these stocks in the cross-validation process into either the training or testing split.
-When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset.
-However, since mPRM is built on the same stock-assessment database, we refit mPRM on each training split.
+We also validated the models using the RAM Legacy Stock Assessment database.
+In this case, we used superensembles trained on the entire simulation dataset.
+However, since mPRM is built on the same stock-assessment database, we applied three-fold cross-validation to the RAM Legacy Stock Assessment data underlying the mPRM model so that the dataset that mPRM was trained with (for the individual model and superensemble) was separate from the dataset it was tested on.
 
@@
 We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
-Code to reproduce our analysis is available at <https://github.com/datalimited/ensembles> (TODO: make public).
+Code to reproduce our analysis and an \textsf{R} package to fit the data-limited assessment models are available at <https://github.com/...> (*Private while in review*).
 
@@
 \ref{hexagon}d).
-There was generally little correlation between true and predicted trend for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
+There was generally little correlation between true and predicted recent trend in status for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
 
@@
 coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
-Beyond estimates of status and trend, ensemble methods could be used to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014], to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014], or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].
+Beyond estimates of status and trend, ensemble methods could be used, for example, to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014] or to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014].
+
+<!-- TODO put back when space: or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].-->
+
+\noindent
+**Acknowledgements**: We thank members of Phase I of the working group 'Developing new approaches to global stock status assessment and fishery production potential of the seas' who
+contributed to developing the data-limited methods and simulations used in our
+analysis.
+We thank E. Jardim, F. Scott, and J.A. Hutchings for helpful comments during the development of this project, and R.D. Methot for comments on an earlier version of the manuscript.
+We thank the Gordon and Betty Moore Foundation for funding the working group
+'Applying data-limited stock status models and developing management guidance for unassessed fish stocks'.
 

### 2015-12-29 analysis/0-ensemble-functions.R

@@ -127,7 +127,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
 
   rows <- max(dat$order) / 4
   par(mfrow = c(rows, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
-    oma = oma, cex = 0.8, mar = c(0, 0, 0, 0),
+    oma = oma, cex = 0.8, mar = c(0, 0, 0.27, 0),
     xaxs = "i", yaxs = "i", col.axis = "grey60", col.lab = "grey50")
   #par(family="serif")
 
@@ -220,7 +220,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
     if (m %in% 9:12) {
       pal_function <- colorRampPalette(hexcol_third_row, space = space, bias = bias[9])
       #add_hex <- FALSE
-      if (m %in% 9) par(mar = par("mar") + c(0, 0, 0.27, 0))
+      #if (m %in% 9) par(mar = par("mar") + c(0, 0, 0.27, 0))
     }
     if (add_hex) {
       pal <- pal_function(max(counts))

### 2016-02-22 text/ms.Rmd
Edits to text for revision

@@
+<!-- TODO
+- [ ] trash slope?
+- [ ] trash covariates?
+- [ ] add discussion of uncertainty (CIs) and ML (and incorporating uncertainty into pred.)
+- [ ] add table of fish methods and ensemble models
+- [ ] add flow chart of cross validation / methods
+-->
+
 ```{r, echo=FALSE}
@@
 \end{Large}
+
+\noindent
+Alternative 1: Combining stock-assessment output with ensemble modelling
+
+\noindent
+Alternative 2: Ensembles of data-limited stock-assessments can improve accuracy and reduce bias of $B/B_\mathrm{MSY}$ estimates
 
@@
 \noindent
-Running head: Superensembles of population status
+Running title: Superensembles of population status
 
@@
 \noindent
-\textbf{Abstract:} Ecological resource managers must often reconcile multiple conflicting estimates of population status and trend.
-One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
+\textbf{Abstract:} Fishery managers must often reconcile conflicting estimates of population status and trend.
+One solution is to average predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
 Superensemble models, commonly used in climate and weather forecasting, may provide a superior solution.
-Superensembles use the predictions from multiple models as covariates in an additional statistical model fit to known data.
-Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trend.
-We fit four data-limited and widely applicable models of population and exploitation dynamics and combine the models' estimates of the mean and recent trend in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
-We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments.
+Superensembles use predictions from multiple models as covariates in an additional model fit to known data.
+We evaluated the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of population status and trend.
+We fit four data-limited and widely applicable models of population and exploitation dynamics and combined the models' estimates of the mean and recent trend in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
+We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a database of `r ram_stocks_n` stock assessments.
 We found that ensemble methods substantially improved estimates of population status and trend.
-Machine-learning superensembles (random forest and boosted regression trees) performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+Random forest and boosted regression trees performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
 We found similar improvements when predicting the trend of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
-Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species.
-However, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
+Ensemble methods can improve estimates of status and trends; however, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
+*TODO: Still need to cut ~30 words*
 
@@
 
-Population status and trend are two of the most fundamental values to quantify in conservation biology [e.g.
-@mace2008; @iucn2015].
-These properties are key to everything from global assessments of extinction risk [e.g.
-@stuart2004; @schipper2008] to assessing the sustainability of hunting [e.g.
-@fa2002] and fishing exploitation levels [e.g.
-@worm2009; @juan-jorda2011].
-However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trend [e.g.
+Population status and trend are two of the most fundamental values to quantify in fisheries management [e.g.
+TODO NEW REFS].
+However, fisheries managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trend [e.g.
 @brodziak2010; @branch2011; @deroba2015].
@@
 @murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006].
-In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g.
+In ecology, ensemble methods are sometimes used to improve species distribution modelling [e.g.
 @araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trend [e.g.
@@
 Here, we estimate population status and trend of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
-We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and the individual models.
+We apply four ensemble-method approaches of varying complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and the individual models.
 
@@
 Process noise (recruitment variability; i.e.
-unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through time or had a first-order autoregressive correlation of $0.6$.
+unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through time or had first-order autoregressive correlation of $0.6$.
 The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.
 @rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-
-<!--TODO add back when identity public: The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software \textsf{R} [@r2015]. -->
-<!--Code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.-->
+The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software R [@r2015].
+Code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.
 
 We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world.
-Our analysis of the stock-assessment database was based on version 2.5; after removing stocks for which at least one of the individual models described below did not converge (121), this database included `r ram_stocks_n` stocks.
+Our analysis of the stock-assessment database was based on version 2.5.
+After removing stocks for which at least one of the individual models described below did not converge (121), this database included `r ram_stocks_n` stocks.
 
@@
 The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database.
-@rosenberg2014 provide a full background on these four methods and code to fit the models.
-In summary:
-
-<!--and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}. In summary:-->
+@rosenberg2014 provide a full background on these four methods
+and code to fit all the models is available in an accompanying package `datalimited` for the statistical software R TODO ADD URL. In summary:
 
 * *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass dynamics [@martell2013].
-It requires a prior distribution for $r$ and assigns a prior to the relative proportion of biomass at the end of the time series compared to unfished biomass (depletion) based on the percentage of maximum catch at the end of the time series.
+It requires a prior distribution for $r$ and a prior on the relative proportion of biomass at the end of the time series compared to unfished biomass (depletion), which is derived from the percentage of maximum catch at the end of the time series.
 The version of the model used in @rosenberg2014 was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
@@
 Biomass is assumed to follow a Schaefer model and harvest dynamics are assumed to follow a logistic model.
-The model is fit with a sampling-importance-sampling algorithm [@rosenberg2014].
+The model is fit with a sampling-importance-resampling algorithm [@rosenberg2014].
 
@@
 This will help the reader understand the progression in complexity of the ensemble approaches I think.-->
+<!-- TODO deal with removed spectral -->
+<!-- 
 Superensemble models allow us to incorporate additional covariates and potentially leverage interactions between these covariates and individual model predictions.
@@
 Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the cyclical shape of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch.
-We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
+We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in R and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
+-->
 
@@
 The individual models we seek to combine with superensembles provide time series of stock status ($B/B_\mathrm{MSY}$).
-Therefore, we can use superensembles to estimate multiple properties of these time series.
-Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years.
+Therefore, we can use superensembles to estimate any property of these time series.
+Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five years.
 Together, these quantities address the recent state and trend of stock status, which are both of management and conservation interest [e.g.
-@mace2008].
+@mace2008] TODO add fishery REF.
 To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
 
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models, and the spectral values, as predictors in our superensemble models.
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models
+<!--, and the spectral values, -->
+as predictors in our superensemble models.
 When modelling mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the superensemble models in log space and exponentiated the predictions.
@@
 We compared an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and a boosted regression tree.
-We fit the linear model superensemble with all second-order interactions.
+We describe these models as estimating $\hat{\theta}$, which represents either
+the ensemble estimated $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$.
+The individual model estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are represented as $b$ for models $i$ $1$ through $n$ ($4$).
+The ensemble average was calculated as:
+
+\begin{equation} 
+\hat{\theta} = \frac{1}{n} \sum_{i=1}^n \hat{b_i}.
+\end{equation}
+
+$$ \hat{\theta} = \argmin_\theta \left( L(b \, | \, \hat{b}{\mathrm{ensemble}}) \right) $$ $$ \hat{b}\mathrm{ensemble} = f \left( \hat{b}\mathrm{CMSY}, \, \hat{b}\mathrm{COM-SIR}, \, \hat{b}\mathrm{SSCOM}, \, \hat{b}\mathrm{mPRM}, \, S(0.2), \, S(0.5), \, | \, \theta \right) $$
+
+We fit the linear model superensemble with all second-order interactions:
+
+\begin{equation} \theta_{\mathrm{ensemble}(i)} = \beta_0 + \sum_{j=1}^m \beta_j \theta_{j(i)} + \sum_{j=1}^m \sum_{\substack{k =1 \ j \neq k}}^m \beta_{kj} \theta_{k(i)} \theta_{j(i)} + \epsilon_i, \quad \epsilon_i \sim \mathrm{Normal}(0, \sigma), \end{equation}
+
+TODO ADD EQN
+
 Our two machine learning superensemble models, a random forest and a boosted regression tree, were based on regression trees.
@@
 Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [@elith2008; @hastie2009].
-We fit random forest models with the `randomForest` package [@liaw2002] for \textsf{R} with the default argument values.
-We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for \textsf{R}.
-Based on initial tuning with the **caret** \textsf{R} package [@caret2008], we chose to fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.
+We fit random forest models with the `randomForest` package [@liaw2002] for R with the default argument values.
+We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for R.
+Based on initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values:
+
+TODO ADD EQN
 
@@
 We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
-Code to reproduce our analysis and an \textsf{R} package to fit the data-limited assessment models are available at <https://github.com/...> (*Private while in review*).
+Code to reproduce our analysis and an R package to fit the data-limited assessment models are available at <https://github.com/datalimited/...> (*Private while in review*).
 
@@
 \ref{hexagon-sim-basic}).
+<!--TODO trash this?-->
 
@@
 A fundamental difference is that coefficient averaging focuses on averaging *coefficients* whereas ensembles instead average *predictions*.
-Thus, while the outcome will be the same in some cases, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
+Thus, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
 parametric and non-parametric models or frequentist and Bayesian predictions).

### 2016-02-22 text/ms.Rmd
jessica edits - minor rewording and comments

@@
 \textbf{Abstract:} Fishery managers must often reconcile conflicting estimates of population status and trend.
-One solution is to average predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
+One solution is to average predictions using an ensemble method, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
 Superensemble models, commonly used in climate and weather forecasting, may provide a superior solution.
-Superensembles use predictions from multiple models as covariates in an additional model fit to known data.
-We evaluated the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of population status and trend.
-We fit four data-limited and widely applicable models of population and exploitation dynamics and combined the models' estimates of the mean and recent trend in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree.
-We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a database of `r ram_stocks_n` stock assessments.
+Superensembles use predictions from multiple models as covariates in an additional model.
+By training this superensemble model using known data, we can then use it to predict unknown situations.
+We evaluated the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of population status and trend for fisheries.
+We fit four widely applicable data-limited models that estimate stock biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$).
+We combined the models' estimates to build four ensemble methods: an ensemble average and three superensembles (a linear model, a random forest, and a boosted regression tree).
+We used the ensemble methods to predict both (i) fishery status and (ii) recent trends of $B/B_\mathrm{MSY}$.
+We trained our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` real stock assessments.
 We found that ensemble methods substantially improved estimates of population status and trend.
-Random forest and boosted regression trees performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+Random forest and boosted regression trees performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range` across models?, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
 We found similar improvements when predicting the trend of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
@@
 
-One solution is to take the average or weighted average of an ensemble of model predictions.
+One solution is to take the average or weighted average of several model predictions, i.e.
+an ensemble.
 Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
-This approach forms the basis of many machine learning methods [e.g.
+<!--explain how this is possible.--> This approach forms the basis of many machine learning methods [e.g.
 @dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.
@@
 
-Whereas averages or weighted averages may improve predictions over any single model, they may not optimally leverage available data.
+Whereas averages or weighted averages of model estimates may improve predictions compared to a single model, they may not optimally leverage available data.
 The best prediction does not necessarily lie in the middle of multiple model predictions, some models may perform better than others in certain conditions, and the covariance between models may contain information that can improve predictive accuracy.
-We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
+<!--can you give an example here?
+e.g.
+a model may behave poorly at estimating low biomass.
+By training the superensbmeles models with real data, it can identify the pros and cons of each model and incorporate them into the estimates.
+.--> We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
 This technique, sometimes called superensemble modelling [@krishnamurti1999], is common in climate and weather forecasting [e.g.
 @yun2005; @mote2015].
-The superensemble is fit to a training dataset where outcomes are well known and used to predict on a dataset of interest (Fig.
+The superensemble is fit to a training dataset where outcomes are well known and then used to predict on a dataset of interest (Fig.
 \ref{didactic}).
@@
 In fisheries science, the commonly used operational models for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
-catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992].
+catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B_\mathrm{MSY}$) [@hilborn1992].
 However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
@@
 
-We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with known status.
-These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of initial biomass depletion compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) an exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increased continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increased and then decreased.
+<!--
+ Structure of methods: summary of methods
+ 1. indiviudla models
+ 2. ensemble models
+ 3. additional covariates
+ 4. data used to test and cross validation,
+ 5. testing model perforance. -->
+
+We first developed and tested ensemble methods on a fully factorial simulated dataset of fisheries [@rosenberg2014] with known status.
+These simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of initial biomass depletion compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) an exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increased continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increased and then decreased.
 Process noise (recruitment variability; i.e.
@@
 @rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software R [@r2015].
+The simulation models were built in the `FLR` packages [@kell2007, <http://flr-project.org>] for the statistical software R [@r2015].
 Code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.
@@
 
-We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
+We fit four individual data-limited models that use data on catch and basic life history parameters to estimate $B/B_\mathrm{MSY}$.
+<!-- explain why we choose these models not others?
+--> Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
 
@@
 @rosenberg2014 provide a full background on these four methods
-and code to fit all the models is available in an accompanying package `datalimited` for the statistical software R TODO ADD URL. In summary:
+and code to fit all the models is available in an accompanying package `datalimited` for the statistical software R TODO ADD URL.
+As described in the Introduction, the performance of these models at estimating $B/B_\mathrm{MSY}$ is variable.
+In summary:
 
@@
 The model used in this paper is modified from the original --- it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.
+<!--TODO is this last sentence necessary?
+-->
 
@@
 <!--TODO Kristin and Mark: It might be a good idea to add a figure that illustrates the spectrum of ensemble/superensembles that are applied here and what is included (e.g., non-linear relationships, etc.).
-This will help the reader understand the progression in complexity of the ensemble approaches I think.-->
+This will help the reader understand the progression in complexity of the ensemble approaches I think.
+Jess: probably add this below the section explaining superensembles-->
 <!-- TODO deal with removed spectral -->
@@
 
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models
-<!--, and the spectral values, -->
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and the predictions from the individual models
+<!--, and the spectral values, fix this explanation -->
 as predictors in our superensemble models.
@@
 Thus, any error in the stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
-We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
+We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold <!-- Jess: not sure what this means - at any decision threshold--> by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
 Code to reproduce our analysis and an R package to fit the data-limited assessment models are available at <https://github.com/datalimited/...> (*Private while in review*).
@@
 First, ensembles will be most effective when they are comprised of diverse individual models that choose different structural model forms, explore contrasting but plausible ranges of parameter values, or make uncorrelated errors [@ali1996; @dietterich2000; @tebaldi2007].
-Such individual models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance of each.
+Such individual models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance of each <!-- is this only true for superensembles?-->.
 Second, ensemble models will be most effective when they are not overfit to the training dataset.

### 2016-04-01 text/ms.Rmd
Edit the introduction

@@
 @brodziak2010; @branch2011; @deroba2015].
-Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
+For example, some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
 
@@
 Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
-<!--explain how this is possible.--> This approach forms the basis of many machine learning methods [e.g.
+Ensembles are superior to individual models for at least three reasons including statistical, computational, and representational reasons [@dietterich2000].
+Statistically by averaging across models representing different hypotheses about reality, computationally by reducing the risk of getting stuck in a local optima, and representationally he by expanding representational form<!--explain how this is possible.--> This approach forms the basis of many machine learning methods [e.g.
 @dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.

### 2016-04-01 text/ms.Rmd
Remove the ROC section from the main text

@@
 Thus, any error in the stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
-We measured the ability for each model to correctly place stocks into categories of above or below $B/B_\mathrm{MSY} = 0.5$ at any decision threshold <!-- Jess: not sure what this means - at any decision threshold--> by plotting receiver operating characteristic curves; a threshold of $B/B_\mathrm{MSY} = 0.5$ is used to classify a stock as overfished in Australia and the United States [@daff2007; @hilborn2010].
-Code to reproduce our analysis and an R package to fit the data-limited assessment models are available at <https://github.com/datalimited/...> (*Private while in review*).
+Code to reproduce our analysis and an R package to fit the data-limited assessment models are available at <https://github.com/datalimited/ensembles> (*Private while in review*).
 
@@
 \ref{performance}a).
-These superensembles also generally had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 0.5$ (Fig.
-\ref{roc-sim}).
 Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.

### 2016-04-01 text/ms.Rmd
At a missing line

@@
 \noindent
+
 **Acknowledgements**: We thank members of Phase I of the working group 'Developing new approaches to global stock status assessment and fishery production potential of the seas' who

### 2016-04-04 text/ms.Rmd
Wrap text to work in vim by voice

@@
 • The full names of all authors.
-• The name(s) and address(es) of the institution(s) at which the work was carried out. If the present address of any author is different from the above, give it as a footnote.
-• The name, address, telephone, fax and e-mail contacts of the author to whom all correspondence and proofs should be sent.
+• The name(s) and address(es) of the institution(s) at which the work was
+carried out. If the present address of any author is different from the above,
+give it as a footnote.
+• The name, address, telephone, fax and e-mail contacts of the author to whom
+all correspondence and proofs should be sent.
 • A suggested running title of not more than 40 characters, including spaces.
@@
 \noindent
-Alternative 2: Ensembles of data-limited stock-assessments can improve accuracy and reduce bias of $B/B_\mathrm{MSY}$ estimates
+Alternative 2: Ensembles of data-limited stock-assessments can improve accuracy
+and reduce bias of $B/B_\mathrm{MSY}$ estimates
 
@@
 
-\noindent
-^2^Institute of Marine & Coastal Sciences, Rutgers University, 
+\noindent ^2^Institute of Marine & Coastal Sciences, Rutgers University,
 71 Dudley Road, New Brunswick, NJ, 08901-8525, USA
 
-\noindent
-^3^Marine and Freshwater Research Centre, Galway-Mayo Institute of Technology, 
-Galway, 00000, Ireland
+\noindent ^3^Marine and Freshwater Research Centre, Galway-Mayo Institute of
+Technology, Galway, 00000, Ireland
 
-\noindent
-^4^Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 2725 Montlake Boulevard E., Seattle, WA, 98112, USA
+\noindent ^4^Fisheries Resource Analysis and Monitoring Division, Northwest
+Fisheries Science Center, National Marine Fisheries Service, National
+Oceanographic and Atmospheric Administration, 2725 Montlake Boulevard E.,
+Seattle, WA, 98112, USA
 
-\noindent
-^5^National Center for Ecological Analysis and Synthesis, University of
-California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA, 93103,
-USA
+\noindent ^5^National Center for Ecological Analysis and Synthesis, University
+of California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA,
+93103, USA
 
-\noindent
-^6^International Council for the Exploration of the Sea, 
-H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark
+\noindent ^6^International Council for the Exploration of the Sea, H.C.
+Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark
 
-\noindent
-^7^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark
+\noindent ^7^DTU Aqua National Institute of Aquatic Resources, Technical
+University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark
 
-\noindent
-^8^Ecosystem Assessment Program, Northeast Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 166 Water St., Woods Hole, MA, 02543, USA
+\noindent ^8^Ecosystem Assessment Program, Northeast Fisheries Science Center,
+National Marine Fisheries Service, National Oceanographic and Atmospheric
+Administration,
+166 Water St., Woods Hole, MA, 02543, USA
 
-\noindent
-^9^Marine Stewarship Council, Marine House, 1 Snow Hill, London, EC1A 2DH, United Kingdom
+\noindent ^9^Marine Stewarship Council, Marine House, 1 Snow Hill, London, EC1A
+2DH, United Kingdom
 
-\noindent
-^10^European Commission, Joint Research Centre, Institute for the Protection and
-Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi 2749, 21027
-Ispra (VA), Italy
+\noindent ^10^European Commission, Joint Research Centre, Institute for the
+Protection and Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi
+2749, 21027 Ispra (VA), Italy
 
-\noindent
-^11^Bren School of Environmental Science and Management, 
-University of California Santa Barbara, 
-Santa Barbara, CA, 93106-5131, USA
+\noindent ^11^Bren School of Environmental Science and Management, University of
+California Santa Barbara, Santa Barbara, CA, 93106-5131, USA
 
-\noindent
-^12^Union of Concerned Scientists, Cambridge, MA, USA
+\noindent ^12^Union of Concerned Scientists, Cambridge, MA, USA
 
-\noindent
-^13^Conservation International, Arlington, VA, USA
+\noindent ^13^Conservation International, Arlington, VA, USA
 
-\noindent
-^\*^Corresponding author, present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA  98195, USA
+\noindent ^\*^Corresponding author, present address: School of Aquatic and
+Fishery Sciences, Box 355020, University of Washington, Seattle, WA  98195, USA
 E-mail: sandrsn@uw.edu
 
-\noindent
-Running title: Superensembles of population status
+\noindent Running title: Superensembles of population status
 
@@
 \bigskip
-\noindent
-\textbf{Abstract:} Fishery managers must often reconcile conflicting estimates of population status and trend.
-One solution is to average predictions using an ensemble method, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy.
-Superensemble models, commonly used in climate and weather forecasting, may provide a superior solution.
-Superensembles use predictions from multiple models as covariates in an additional model.
-By training this superensemble model using known data, we can then use it to predict unknown situations.
-We evaluated the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of population status and trend for fisheries.
-We fit four widely applicable data-limited models that estimate stock biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$).
-We combined the models' estimates to build four ensemble methods: an ensemble average and three superensembles (a linear model, a random forest, and a boosted regression tree).
-We used the ensemble methods to predict both (i) fishery status and (ii) recent trends of $B/B_\mathrm{MSY}$.
-We trained our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` real stock assessments.
-We found that ensemble methods substantially improved estimates of population status and trend.
-Random forest and boosted regression trees performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range` across models?, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
-We found similar improvements when predicting the trend of status and when applying the simulation-trained superensembles to catch data for global fish stocks.
-Ensemble methods can improve estimates of status and trends; however, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
-*TODO: Still need to cut ~30 words*
+
+\noindent \textbf{Abstract:} Fishery managers must often reconcile conflicting
+estimates of population status and trend. One solution is to average
+predictions using an ensemble method, but this approach ignores covariance
+between models and may not optimally reduce bias and improve accuracy.
+Superensemble models, commonly used in climate and weather forecasting, may
+provide a superior solution. Superensembles use predictions from multiple
+models as covariates in an additional model. By training this superensemble
+model using known data, we can then use it to predict unknown situations. We
+evaluated the potential for ensemble and superensemble models ('ensemble
+methods') to improve estimates of population status and trend for fisheries. We
+fit four widely applicable data-limited models that estimate stock biomass
+relative to the equilibrium biomass at maximum sustainable yield
+($B/B_\mathrm{MSY}$). We combined the models' estimates to build four ensemble
+methods: an ensemble average and three superensembles (a linear model, a random
+forest, and a boosted regression tree). We used the ensemble methods to predict
+both (i) fishery status and (ii) recent trends of $B/B_\mathrm{MSY}$. We
+trained our models on a simulated dataset of 5760 stocks and tested them with
+cross-validation and against a global database of `r ram_stocks_n` real stock
+assessments. We found that ensemble methods substantially improved estimates of
+population status and trend. Random forest and boosted regression trees
+performed the best for estimating population status: accuracy improved `r
+mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and
+true status improved from `r mean_sim$ind_corr_range` to `r
+mean_sim$mach_corr_range` across models?, and bias (median proportional error)
+declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We
+found similar improvements when predicting the trend of status and when
+applying the simulation-trained superensembles to catch data for global fish
+stocks. Ensemble methods can improve estimates of status and trends; however,
+they must be tested, formed from a diverse set of accurate models, and built on
+a dataset representative of the populations to which they are applied. *TODO:
+Still need to cut ~30 words*
 
@@
 
-\noindent 
-Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
+\noindent Keywords: data-limited fisheries, ensemble methods, multi-model
+averaging, population dynamics, sustainable resource management
 
@@
 
-Population status and trend are two of the most fundamental values to quantify in fisheries management [e.g.
-TODO NEW REFS].
-However, fisheries managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trend [e.g.
-@brodziak2010; @branch2011; @deroba2015].
-For example, some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.
+Population status and trend are two of the most fundamental values to quantify
+in fisheries management [e.g. TODO NEW REFS]. However, fisheries managers are
+often faced with reconciling multiple uncertain and potentially conflicting
+estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015].
+For example, some models may suggest a population is at risk and declining in
+abundance while others may suggest it is not at risk and stable.
 
-One solution is to take the average or weighted average of several model predictions, i.e.
-an ensemble.
-Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007].
-Ensembles are superior to individual models for at least three reasons including statistical, computational, and representational reasons [@dietterich2000].
-Statistically by averaging across models representing different hypotheses about reality, computationally by reducing the risk of getting stuck in a local optima, and representationally he by expanding representational form<!--explain how this is possible.--> This approach forms the basis of many machine learning methods [e.g.
-@dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g.
-@murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006].
-In ecology, ensemble methods are sometimes used to improve species distribution modelling [e.g.
-@araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trend [e.g.
+One solution is to take the average or weighted average of several model
+predictions, i.e. an ensemble. Such ensembles are typically more accurate and
+less biased than individual model estimates and can integrate uncertainty in
+model structure, initial conditions, and parameter estimation [@dietterich2000;
+@araujo2007]. Ensembles are superior to individual models for at least three
+reasons including statistical, computational, and representational reasons
+[@dietterich2000]. Statistically by averaging across models representing
+different hypotheses about reality, computationally by reducing the risk of
+getting stuck in a local optima, and representationally he by expanding
+representational form<!--explain how this is possible.--> This approach forms
+the basis of many machine learning methods [e.g. @dietterich2000], has helped
+reconcile climate forecasts from dozens of models [e.g. @murphy2004;
+@tebaldi2007; @ipcc2013], and even improved early warning signs of malaria
+outbreaks [@thomson2006]. In ecology, ensemble methods are sometimes used to
+improve species distribution modelling [e.g. @araujo2007; @breiner2015] and
+indeed have been used to combine estimates of population status and trend [e.g.
 @brodziak2010].
 
-Whereas averages or weighted averages of model estimates may improve predictions compared to a single model, they may not optimally leverage available data.
-The best prediction does not necessarily lie in the middle of multiple model predictions, some models may perform better than others in certain conditions, and the covariance between models may contain information that can improve predictive accuracy.
-<!--can you give an example here?
-e.g.
-a model may behave poorly at estimating low biomass.
-By training the superensbmeles models with real data, it can identify the pros and cons of each model and incorporate them into the estimates.
-.--> We can exploit these characteristics by using the predictions from a group of models as inputs into a separate statistical model.
-This technique, sometimes called superensemble modelling [@krishnamurti1999], is common in climate and weather forecasting [e.g.
-@yun2005; @mote2015].
-The superensemble is fit to a training dataset where outcomes are well known and then used to predict on a dataset of interest (Fig.
-\ref{didactic}).
-For example, @krishnamurti1999 combined predictions of wind and precipitation in Asian monsoons via a superensemble regression fit to observed data.
-Their superensemble was considerably more accurate than any individual prediction or weighted average of predictions.
+Whereas averages or weighted averages of model estimates may improve
+predictions compared to a single model, they may not optimally leverage
+available data. The best prediction does not necessarily lie in the middle of
+multiple model predictions, some models may perform better than others in
+certain conditions, and the covariance between models may contain information
+that can improve predictive accuracy. <!--can you give an example here? e.g.
+a model may behave poorly at estimating low biomass. By training the
+superensbmeles models with real data, it can identify the pros and cons of each
+model and incorporate them into the estimates. .--> We can exploit these
+characteristics by using the predictions from a group of models as inputs into
+a separate statistical model. This technique, sometimes called superensemble
+modelling [@krishnamurti1999], is common in climate and weather forecasting
+[e.g. @yun2005; @mote2015]. The superensemble is fit to a training dataset
+where outcomes are well known and then used to predict on a dataset of interest
+(Fig. \ref{didactic}). For example, @krishnamurti1999 combined predictions of
+wind and precipitation in Asian monsoons via a superensemble regression fit to
+observed data. Their superensemble was considerably more accurate than any
+individual prediction or weighted average of predictions.
 
-In fisheries science, the commonly used operational models for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g.
-catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B_\mathrm{MSY}$) [@hilborn1992].
-However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014].
-Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: (1) a time series of the total biomass of catch and (2) a basic understanding of population productivity [e.g.
-@vasconcellos2005; @martell2013].
-Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment.
-Three of these models are based on Schaefer (logistic) biomass dynamics and one is an empirical model fitted to more data-rich stock assessment output.
-The four models frequently disagreed about population status, no one model had strong performance across all fish stocks, and some models performed better than others depending on circumstances (e.g.
-Fig.
+In fisheries science, the commonly used operational models for determining
+status and trend of exploited fish populations are stock assessments, i.e.,
+population models coupled to an observation model, that incorporate all
+appropriate data (e.g. catches, size and age distributions, surveys, and
+tagging information) to quantify values such as the biomass of a stock that can
+produce maximum sustainable yield ($B_\mathrm{MSY}$) [@hilborn1992]. However,
+the broad range of data required to conduct these stock assessments are not
+available for the majority of fish populations, including those of conservation
+concern and of economic interest to fisheries [@fao2014]. Therefore, a number
+of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited
+data available for the majority of fish stocks: (1) a time series of the total
+biomass of catch and (2) a basic understanding of population productivity [e.g.
+@vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the
+performance of four data-limited models through a large-scale simulation
+experiment. Three of these models are based on Schaefer (logistic) biomass
+dynamics and one is an empirical model fitted to more data-rich stock
+assessment output. The four models frequently disagreed about population
+status, no one model had strong performance across all fish stocks, and some
+models performed better than others depending on circumstances (e.g. Fig.
 \ref{motivate}).
 
-Here, we estimate population status and trend of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models.
-We apply four ensemble-method approaches of varying complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and the individual models.
+Here, we estimate population status and trend of exploited fish populations
+using ensembles and superensembles (collectively 'ensemble methods') of these
+four data-limited models. We apply four ensemble-method approaches of varying
+complexity to both simulated and real-world fish stocks and compare their
+predictive performance against each other and the individual models.
 
-<!--
- TODO: It might be good to throw in a single sentence here describing the 4 models, or at least listing them and maybe a couple key differences among them.-->
+<!-- TODO: It might be good to throw in a single sentence here describing the
+4 models, or at least listing them and maybe a couple key differences among
+them.-->
 
@@
 
-We first developed and tested ensemble methods on a fully factorial simulated dataset of fisheries [@rosenberg2014] with known status.
-These simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of initial biomass depletion compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) an exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increased continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increased and then decreased.
-Process noise (recruitment variability; i.e.
-unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through time or had first-order autoregressive correlation of $0.6$.
-The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.
-@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks.
-The simulation models were built in the `FLR` packages [@kell2007, <http://flr-project.org>] for the statistical software R [@r2015].
-Code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.
+We first developed and tested ensemble methods on a fully factorial simulated
+dataset of fisheries [@rosenberg2014] with known status. These simulated data
+included the following scenarios: three fish life histories: small pelagic,
+demersal, and large pelagic; three levels of initial biomass depletion compared
+to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and
+four exploitation settings: (1) a constant exploitation rate, (2) an
+exploitation rate coupled with biomass to mimic an open-access single-species
+fishery, (3) a scenario where exploitation rate increased continuously, and (4)
+a 'roller-coaster' scenario where the exploitation rate increased and then
+decreased. Process noise (recruitment variability; i.e. unexplained variability
+in population dynamics) was introduced to the models at two magnitudes in log
+space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through
+time or had first-order autoregressive correlation of $0.6$. The simulation
+also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch)
+and one scenario without measurement error. @rosenberg2014 ran ten iterations
+for each combination of factors adding stochastic draws of recruitment and
+catch-recording variability each time to generate a total of 5760 stocks. The
+simulation models were built in the `FLR` packages [@kell2007,
+<http://flr-project.org>] for the statistical software R [@r2015]. Code to
+generate the simulations is available at
+<https://github.com/iagomosqueira/stockStatusFAO>.
 
-We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world.
-Our analysis of the stock-assessment database was based on version 2.5.
-After removing stocks for which at least one of the individual models described below did not converge (121), this database included `r ram_stocks_n` stocks.
+We also tested our ensemble methods on the RAM Legacy Stock Assessment Database
+[@ricard2012] --- a compilation of stock-assessment output from hundreds of
+exploited marine populations around the world. Our analysis of the
+stock-assessment database was based on version 2.5. After removing stocks for
+which at least one of the individual models described below did not converge
+(121), this database included `r ram_stocks_n` stocks.
 
@@
 
-We fit four individual data-limited models that use data on catch and basic life history parameters to estimate $B/B_\mathrm{MSY}$.
-<!-- explain why we choose these models not others?
---> Three of the models are mechanistic and based generally on Schaefer biomass dynamics [@schaefer1954] of the form
+We fit four individual data-limited models that use data on catch and basic life
+history parameters to estimate $B/B_\mathrm{MSY}$. <!-- explain why we choose
+these models not others? --> Three of the models are mechanistic and based
+generally on Schaefer biomass dynamics [@schaefer1954] of the form
 
@@
 
-\noindent
-where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch.
-The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database.
-@rosenberg2014 provide a full background on these four methods
-and code to fit all the models is available in an accompanying package `datalimited` for the statistical software R TODO ADD URL.
-As described in the Introduction, the performance of these models at estimating $B/B_\mathrm{MSY}$ is variable.
-In summary:
+\noindent where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus
+one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic
+population growth rate, $B_0$ represents unfished biomass or carrying capacity
+$K$, and $C$ represents catch. The fourth model is an empirically derived model
+based on the RAM Legacy Stock Assessment Database. @rosenberg2014 provide a full
+background on these four methods and code to fit all the models is available in
+an accompanying package `datalimited` for the statistical software R TODO ADD
+URL. As described in the Introduction, the performance of these models at
+estimating $B/B_\mathrm{MSY}$ is variable. In summary:
 
-* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass dynamics [@martell2013].
-It requires a prior distribution for $r$ and a prior on the relative proportion of biomass at the end of the time series compared to unfished biomass (depletion), which is derived from the percentage of maximum catch at the end of the time series.
-The version of the model used in @rosenberg2014 was modified from @martell2013 to generate biomass trends from all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the median trend.
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
+dynamics [@martell2013]. It requires a prior distribution for $r$ and a prior on
+the relative proportion of biomass at the end of the time series compared to
+unfished biomass (depletion), which is derived from the percentage of maximum
+catch at the end of the time series. The version of the model used in
+@rosenberg2014 was modified from @martell2013 to generate biomass trends from
+all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the
+median trend.
 
-* *COM-SIR* (catch-only-model with sampling-importance-resampling) is a coupled harvest-dynamics model [@vasconcellos2005].
-Biomass is assumed to follow a Schaefer model and harvest dynamics are assumed to follow a logistic model.
-The model is fit with a sampling-importance-resampling algorithm [@rosenberg2014].
+* *COM-SIR* (catch-only-model with sampling-importance-resampling) is a coupled
+harvest-dynamics model [@vasconcellos2005]. Biomass is assumed to follow
+a Schaefer model and harvest dynamics are assumed to follow a logistic model.
+The model is fit with a sampling-importance-resampling algorithm
+[@rosenberg2014].
 
-* *SSCOM* (state-space catch-only model) is a hierarchical model that, similar to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013].
-SSCOM estimates unobserved dynamics in both fishing effort and the fished population based on a catch time series and priors on $r$, the maximum rate of increase of fishing effort, and the magnitude of various forms of stochasticity.
-The model is fit in a Bayesian state-space framework to integrate across three forms of stochasticity: variation in effort, population dynamics, and fishing efficiency [@thorson2013].
+* *SSCOM* (state-space catch-only model) is a hierarchical model that, similar
+to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013]. SSCOM
+estimates unobserved dynamics in both fishing effort and the fished population
+based on a catch time series and priors on $r$, the maximum rate of increase of
+fishing effort, and the magnitude of various forms of stochasticity. The model
+is fit in a Bayesian state-space framework to integrate across three forms of
+stochasticity: variation in effort, population dynamics, and fishing efficiency
+[@thorson2013].
 
-* *mPRM* (modified panel regression model) is a modified version of the panel-regression model from @costello2012.
-Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
-The model used in this paper is modified from the original --- it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.
-<!--TODO is this last sentence necessary?
--->
+* *mPRM* (modified panel regression model) is a modified version of the
+panel-regression model from @costello2012. Unlike the other models, mPRM is
+empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment
+database to fit a regression model to a series of characteristics of the catch
+time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
+The model used in this paper is modified from the original --- it condenses the
+life-history categories into three categories to match the simulated dataset,
+removes the maximum catch predictor since the absolute catch in the simulated
+dataset is arbitrary, and does not implement the bias correction needed in
+@costello2012 since we do not derive estimates of median status across multiple
+stocks. <!--TODO is this last sentence necessary? -->
 
@@
 
-<!--TODO Kristin and Mark: It might be a good idea to add a figure that illustrates the spectrum of ensemble/superensembles that are applied here and what is included (e.g., non-linear relationships, etc.).
-This will help the reader understand the progression in complexity of the ensemble approaches I think.
-Jess: probably add this below the section explaining superensembles-->
+<!--TODO Kristin and Mark: It might be a good idea to add a figure that
+illustrates the spectrum of ensemble/superensembles that are applied here and
+what is included (e.g., non-linear relationships, etc.). This will help the
+reader understand the progression in complexity of the ensemble approaches
+I think. Jess: probably add this below the section explaining superensembles-->
 <!-- TODO deal with removed spectral -->
-<!-- 
-Superensemble models allow us to incorporate additional covariates and potentially leverage interactions between these covariates and individual model predictions.
-Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data.
-For simplicity, and since many possible covariates, such as fishing effort dynamics, would not be known for many real-world data-limited scenarios, we added only one set of additional covariates: spectral properties of the catch time series.
-Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the cyclical shape of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch.
-We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in R and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.
--->
+
+<!-- Superensemble models allow us to incorporate additional covariates and
+potentially leverage interactions between these covariates and individual model
+predictions. Additional covariates could be, for example, life-history
+characteristics, information on exploitation patterns, or statistical properties
+of the data. For simplicity, and since many possible covariates, such as fishing
+effort dynamics, would not be known for many real-world data-limited scenarios,
+we added only one set of additional covariates: spectral properties of the catch
+time series. Spectral analysis decomposes a time series into the frequency
+domain and provides a means of describing the cyclical shape of the catch series
+that is independent of time series length (except in affecting precision) and
+absolute magnitude of catch. We fit spectral models to the scaled catch time
+series (catch divided by maximum catch) with the `spec.ar` function in R and
+recorded representative short- and long-term spectral densities at frequencies
+of 0.20 and 0.05, which correspond to 5- and 20-year cycles. -->
 
@@
 
-The individual models we seek to combine with superensembles provide time series of stock status ($B/B_\mathrm{MSY}$).
-Therefore, we can use superensembles to estimate any property of these time series.
-Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five years.
-Together, these quantities address the recent state and trend of stock status, which are both of management and conservation interest [e.g.
-@mace2008] TODO add fishery REF.
-To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
+The individual models we seek to combine with superensembles provide time series
+of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use superensembles to
+estimate any property of these time series. Here, we focus on two properties:
+the mean and slope of $B/B_\mathrm{MSY}$ in the last five years. Together, these
+quantities address the recent state and trend of stock status, which are both of
+management and conservation interest [e.g. @mace2008] TODO add fishery REF. To
+avoid undue influence of the end points of the time series on the calculated
+slope, we measured the slope as the Theil-Sen estimator of median slope
+[@theil1950].
 
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and the predictions from the individual models
-<!--, and the spectral values, fix this explanation -->
-as predictors in our superensemble models.
-When modelling mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the superensemble models in log space and exponentiated the predictions.
-For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and the
+predictions from the individual models <!--, and the spectral values, fix this
+explanation --> as predictors in our superensemble models. When modelling mean
+$B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the superensemble
+models in log space and exponentiated the predictions. For the estimates of
+$B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble
+models on the natural untransformed scale.
 
-We compared an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and a boosted regression tree.
-We describe these models as estimating $\hat{\theta}$, which represents either
-the ensemble estimated $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$.
-The individual model estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are represented as $b$ for models $i$ $1$ through $n$ ($4$).
-The ensemble average was calculated as:
+We compared an ensemble average and three superensembles of varying complexity:
+a linear model with two-way interactions, a random forest, and a boosted
+regression tree. We describe these models as estimating $\hat{\theta}$, which
+represents either the ensemble estimated $\log B/B_\mathrm{MSY}$ or slope of
+$B/B_\mathrm{MSY}$. The individual model estimates of $\log B/B_\mathrm{MSY}$ or
+slope of $B/B_\mathrm{MSY}$ are represented as $b$ for models $i$ $1$ through
+$n$ ($4$). The ensemble average was calculated as:
 
@@
 
-$$ \hat{\theta} = \argmin_\theta \left( L(b \, | \, \hat{b}{\mathrm{ensemble}}) \right) $$ $$ \hat{b}\mathrm{ensemble} = f \left( \hat{b}\mathrm{CMSY}, \, \hat{b}\mathrm{COM-SIR}, \, \hat{b}\mathrm{SSCOM}, \, \hat{b}\mathrm{mPRM}, \, S(0.2), \, S(0.5), \, | \, \theta \right) $$
+$$ \hat{\theta} = \argmin_\theta \left( L(b \, | \, \hat{b}{\mathrm{ensemble}})
+\right) $$ $$ \hat{b}\mathrm{ensemble} = f \left( \hat{b}\mathrm{CMSY}, \,
+\hat{b}\mathrm{COM-SIR}, \, \hat{b}\mathrm{SSCOM}, \, \hat{b}\mathrm{mPRM}, \,
+S(0.2), \, S(0.5), \, | \, \theta \right) $$
 
@@
 
-\begin{equation} \theta_{\mathrm{ensemble}(i)} = \beta_0 + \sum_{j=1}^m \beta_j \theta_{j(i)} + \sum_{j=1}^m \sum_{\substack{k =1 \ j \neq k}}^m \beta_{kj} \theta_{k(i)} \theta_{j(i)} + \epsilon_i, \quad \epsilon_i \sim \mathrm{Normal}(0, \sigma), \end{equation}
+\begin{equation} \theta_{\mathrm{ensemble}(i)} = \beta_0 + \sum_{j=1}^m \beta_j
+\theta_{j(i)} + \sum_{j=1}^m \sum_{\substack{k =1 \ j \neq k}}^m \beta_{kj}
+\theta_{k(i)} \theta_{j(i)} + \epsilon_i, \quad \epsilon_i \sim
+\mathrm{Normal}(0, \sigma), \end{equation}
 
@@
 
-Our two machine learning superensemble models, a random forest and a boosted regression tree, were based on regression trees.
-Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a loss function [@breiman1984].
-In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model [@breiman2001].
-In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008].
-Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [@elith2008; @hastie2009].
-We fit random forest models with the `randomForest` package [@liaw2002] for R with the default argument values.
-We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for R.
-Based on initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values:
+Our two machine learning superensemble models, a random forest and a boosted
+regression tree, were based on regression trees. Regression trees sequentially
+determine what value of a predictor best splits the response data into two
+'branches' based on a loss function [@breiman1984]. In random forests, a series
+of regression trees are built on a random subset of the data and random subset
+of the covariates of the model [@breiman2001]. In generalized boosted models
+(GBMs), each subsequent model is fit to the residuals from the previous model;
+data points that are fit poorly in a given model are given more weight in the
+next model [@elith2008]. Random forests and GBMs can provide strong predictive
+performance and fit highly non-linear relationships [@elith2008; @hastie2009].
+We fit random forest models with the `randomForest` package [@liaw2002] for
+R with the default argument values. We fit boosted regression tree models with
+the `gbm` package [@ridgeway2015] for R. Based on initial tuning with the
+**caret** R package [@caret2008], we chose to fit GBMs with 2000 trees, an
+interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and
+all other arguments at their default values:
 
@@
 
-A critical component to predictive modelling is to evaluate the performance of a model on new data [@hastie2009].
-We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built superensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third.
-We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits.
-In the simulated dataset, there were 10 replicates of each unique combination of simulation parameters that differed only in stochastic variability.
-Since the dynamics of these populations were often similar, we grouped these stocks in the cross-validation process into either the training or testing split.
-We also validated the models using the RAM Legacy Stock Assessment database.
-In this case, we used superensembles trained on the entire simulation dataset.
-However, since mPRM is built on the same stock-assessment database, we applied three-fold cross-validation to the RAM Legacy Stock Assessment data underlying the mPRM model so that the dataset that mPRM was trained with (for the individual model and superensemble) was separate from the dataset it was tested on.
+A critical component to predictive modelling is to evaluate the performance of
+a model on new data [@hastie2009]. We used repeated three-fold cross validation
+to test predictive performance: we randomly divided the dataset into three sets,
+built superensemble models on two-thirds of the data, and evaluated predictive
+performance on the remaining third. We repeated this across each of the three
+splits and then repeated the whole procedure 50 times to account for bias that
+may result from any one set of validation splits. In the simulated dataset,
+there were 10 replicates of each unique combination of simulation parameters
+that differed only in stochastic variability. Since the dynamics of these
+populations were often similar, we grouped these stocks in the cross-validation
+process into either the training or testing split. We also validated the models
+using the RAM Legacy Stock Assessment database. In this case, we used
+superensembles trained on the entire simulation dataset. However, since mPRM is
+built on the same stock-assessment database, we applied three-fold
+cross-validation to the RAM Legacy Stock Assessment data underlying the mPRM
+model so that the dataset that mPRM was trained with (for the individual model
+and superensemble) was separate from the dataset it was tested on.
 
-Predictive performance can be evaluated with metrics that represent a variety of modelling goals.
-For continuous response variables such as the mean and slope of population status, performance metrics often measure some form of bias, precision, accuracy (a combination of bias and precision), or the ability to correctly rank or correlate across populations [e.g.
-@walther2005].
-Here, we measure proportional error, defined as $(\hat{b} - b)/\abs{b}$, where $\hat{b}$ and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$.
-We calculated median proportional error to measure bias, median absolute proportional error to measure accuracy, and Spearman's rank-order correlation between predicted and 'true' values to measure the ability to correctly rank populations.
-When cross-validating against the RAM Legacy Stock Assessment database, we treated the estimates from these data-rich stock assessments as known without error.
-Thus, any error in the stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for each of the four data-limited models and the ensembles.
-Code to reproduce our analysis and an R package to fit the data-limited assessment models are available at <https://github.com/datalimited/ensembles> (*Private while in review*).
+Predictive performance can be evaluated with metrics that represent a variety of
+modelling goals. For continuous response variables such as the mean and slope of
+population status, performance metrics often measure some form of bias,
+precision, accuracy (a combination of bias and precision), or the ability to
+correctly rank or correlate across populations [e.g. @walther2005]. Here, we
+measure proportional error, defined as $(\hat{b} - b)/\abs{b}$, where $\hat{b}$
+and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of
+$B/B_\mathrm{MSY}$. We calculated median proportional error to measure bias,
+median absolute proportional error to measure accuracy, and Spearman's
+rank-order correlation between predicted and 'true' values to measure the
+ability to correctly rank populations. When cross-validating against the RAM
+Legacy Stock Assessment database, we treated the estimates from these data-rich
+stock assessments as known without error. Thus, any error in the
+stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also
+contributes to our estimates of prediction error for each of the four
+data-limited models and the ensembles. Code to reproduce our analysis and an
+R package to fit the data-limited assessment models are available at
+<https://github.com/datalimited/ensembles> (*Private while in review*).
 
-<!-- TODO PE ....
-mean or slope of b/bmsy: Dan: logged or normal space?
-For the status regression where you are estimating log(B/Bmsy), wouldn't you need to correct for the retransformation bias in the estimating median proportional error later down?
-I could be missing something though in whether that applied once you've calculated proportional error-->
+<!-- TODO PE .... mean or slope of b/bmsy: Dan: logged or normal space? For the
+status regression where you are estimating log(B/Bmsy), wouldn't you need to
+correct for the retransformation bias in the estimating median proportional
+error later down? I could be missing something though in whether that applied
+once you've calculated proportional error-->
 
@@
 
-Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trend) of $B/B_\mathrm{MSY}$ in the last five years.
-All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig.
-\ref{hexagon}).
-In contrast to the known unimodal distribution of status, CMSY exhibited bimodal predictions (Fig.
-\ref{hexagon}a), but had the best rank-order correlation and accuracy scores (Fig.
-\ref{performance}a).
-COM-SIR and SSCOM both correctly identified a number of stocks with low status, but frequently predicted a high status when status was in fact low (Fig.
-\ref{hexagon}b, c).
-mPRM had relatively poor ability to predict status for the simulated dataset (Fig.
-\ref{hexagon}d).
-There was generally little correlation between true and predicted recent trend in status for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
+Applied to the simulated dataset of known stock status, the individual models
+had variable success at recovering the mean (status) and slope (trend) of
+$B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of
+scatter around the one-to-one line of perfect status prediction (Fig.
+\ref{hexagon}). In contrast to the known unimodal distribution of status, CMSY
+exhibited bimodal predictions (Fig. \ref{hexagon}a), but had the best rank-order
+correlation and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM
+both correctly identified a number of stocks with low status, but frequently
+predicted a high status when status was in fact low (Fig. \ref{hexagon}b, c).
+mPRM had relatively poor ability to predict status for the simulated dataset
+(Fig. \ref{hexagon}d). There was generally little correlation between true and
+predicted recent trend in status for any of the individual models (rank-order
+correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM
+(correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).
 
-Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trend over any individual model (Fig.
-\ref{hexagon}e--h, Fig.
-\ref{scatter-sim-slope}e--h).
-Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig.
-\ref{performance}a).
-Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
-\ref{performance-sim-slope}).
-The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating fish stock status (CMSY; Fig.
-\ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting the trends of status (SSCOM; Fig.
+Ensemble methods, and in particular the machine learning superensemble models
+(random forest and GBM), generally improved estimates of status and trend over
+any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h).
+Compared to the individual models, machine learning superensembles improved
+accuracy (median absolute proportional error) by `r
+mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r
+mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias
+(median proportional error) from `r mean_sim$ind_mre_range` to `r
+mean_sim$mach_mre_range` (Fig. \ref{performance}a). Results were similar when
+predicting trend: compared to individual models, machine learning superensembles
+improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order
+correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`,
+and reduced bias from `r slope_sim$ind_mre_range` to `r
+slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The ensemble
+models that simply took a mean of the individual models ranked slightly behind
+the best individual model for estimating fish stock status (CMSY; Fig.
+\ref{performance}a) and had slightly lower correlation but higher accuracy than
+the best individual model at predicting the trends of status (SSCOM; Fig.
 \ref{performance-sim-slope}).
 
-The superensemble models were able to improve the predictive performance by harnessing the best properties of individual models, the covariance between individual models, and interactions with other covariates.
-For example, SSCOM had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
-\ref{hexagon}c, Fig.
-\ref{partial-sim}c) and CMSY predictions were approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high clusters of predictions (Fig.
-\ref{partial-sim}).
-Superensembles also exploited the covariance between individual model predictions.
-For instance, both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
-The addition of spectral density covariates helped the superensemble models correctly predict higher status values (Fig.
-\ref{hexagon}g, h) due to the interactions between the spectral density predictors and the individual model predictors (Fig.
-\ref{partial-2d-sim}).
-Although the additional covariates improved ensemble fit, performance of the ensembles was only marginally degraded by removing these covariates (Fig.
-\ref{hexagon} vs.
-Fig.
+The superensemble models were able to improve the predictive performance by
+harnessing the best properties of individual models, the covariance between
+individual models, and interactions with other covariates. For example, SSCOM
+had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
+\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were approximately
+linearly related to $B/B_\mathrm{MSY}$ within the low and high clusters of
+predictions (Fig. \ref{partial-sim}). Superensembles also exploited the
+covariance between individual model predictions. For instance, both the linear
+model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the
+true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
+The addition of spectral density covariates helped the superensemble models
+correctly predict higher status values (Fig. \ref{hexagon}g, h) due to the
+interactions between the spectral density predictors and the individual model
+predictors (Fig. \ref{partial-2d-sim}). Although the additional covariates
+improved ensemble fit, performance of the ensembles was only marginally degraded
+by removing these covariates (Fig. \ref{hexagon} vs. Fig.
 \ref{hexagon-sim-basic}).
+
 <!--TODO trash this?-->
 
-When applied to the stock-assessment database, the superensemble models --- trained exclusively on the simulated dataset --- generally performed as well or better than the best individual models.
-The mean, random forest, and GBM ensembles outperformed the mPRM method which is trained directly on the RAM Legacy Stock Assessment database itself (Fig.
-\ref{performance}b, Fig.
-\ref{hexagon-ram}).
-Compared to the individual models, the machine learning superensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from `r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
+When applied to the stock-assessment database, the superensemble models ---
+trained exclusively on the simulated dataset --- generally performed as well or
+better than the best individual models. The mean, random forest, and GBM
+ensembles outperformed the mPRM method which is trained directly on the RAM
+Legacy Stock Assessment database itself (Fig. \ref{performance}b, Fig.
+\ref{hexagon-ram}). Compared to the individual models, the machine learning
+superensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%,
+improved correlation from `r mean_ram$ind_corr_range` to `r
+mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to
+`r mean_ram$mach_mre_range`.
 
@@
 
-<!-- Dan:
-Lastly, it's worth noting that this approach, as laid out here is for helping improve potentially conflicting estimates of the same metric, B/Bmsy.
-In many cases though a fishery is going to be basing its decisions on streams of multiple indicators, e.g trends in SPR and B/Bmsy.
-While the ensemble methods here can help refine estimates of individual metrics, it's still worth looking at what alternative data streams are telling you if possible.
-E.g.
-if a purely catch based B/Bmsy estimate is high and has a positive slope, but mean lengths and SPR are decreasing, managers will still need to think about how to deal with these conflicting signals.
-So, this ensemble method, as I interpret it, is helpful for resolving conflicts within metric, but doesn't necesarily resolve conflicts across metrics (if I'm understanding these methods right)
--->
+<!-- Dan: Lastly, it's worth noting that this approach, as laid out here is for
+helping improve potentially conflicting estimates of the same metric, B/Bmsy. In
+many cases though a fishery is going to be basing its decisions on streams of
+multiple indicators, e.g trends in SPR and B/Bmsy. While the ensemble methods
+here can help refine estimates of individual metrics, it's still worth looking
+at what alternative data streams are telling you if possible. E.g. if a purely
+catch based B/Bmsy estimate is high and has a positive slope, but mean lengths
+and SPR are decreasing, managers will still need to think about how to deal with
+these conflicting signals. So, this ensemble method, as I interpret it, is
+helpful for resolving conflicts within metric, but doesn't necesarily resolve
+conflicts across metrics (if I'm understanding these methods right) -->
 
-Ensemble methods provide a useful approach to situations where conservation and environmental resource management decisions must be made on the basis of multiple, potentially contrasting estimates of status --- a situation common to many settings beyond fisheries.
-Compared to individual models of fish population status, ensemble methods were consistently the best or among the best across three performance dimensions (accuracy, bias, and rank-order correlation), two response variables (status and trend), two datasets (simulated and global fisheries), and multiple ensemble methods (from a simple average to machine learning superensembles).
-Our results suggest choosing a superensemble model that allows for non-parametric relationships, such as machine learning methods.
-These methods provide added insight into individual model behavior and generally performed the best; however, even a simple average of predictions across multiple models may provide a more useful measure of the desired parameter than a single model in many cases.
+Ensemble methods provide a useful approach to situations where conservation and
+environmental resource management decisions must be made on the basis of
+multiple, potentially contrasting estimates of status --- a situation common to
+many settings beyond fisheries. Compared to individual models of fish population
+status, ensemble methods were consistently the best or among the best across
+three performance dimensions (accuracy, bias, and rank-order correlation), two
+response variables (status and trend), two datasets (simulated and global
+fisheries), and multiple ensemble methods (from a simple average to machine
+learning superensembles). Our results suggest choosing a superensemble model
+that allows for non-parametric relationships, such as machine learning methods.
+These methods provide added insight into individual model behavior and generally
+performed the best; however, even a simple average of predictions across
+multiple models may provide a more useful measure of the desired parameter than
+a single model in many cases.
 
 Certain conditions will make some ensemble models more effective than others.
-First, ensembles will be most effective when they are comprised of diverse individual models that choose different structural model forms, explore contrasting but plausible ranges of parameter values, or make uncorrelated errors [@ali1996; @dietterich2000; @tebaldi2007].
-Such individual models would be expected to perform well in different conditions and an ensemble model can exploit the best predictive performance of each <!-- is this only true for superensembles?-->.
-Second, ensemble models will be most effective when they are not overfit to the training dataset.
-Cross-validation testing [@caruana2004; @hastie2009] and methods that are robust to overfitting such as random forests [@breiman2001], may help avoid overfitting ensemble models.
-We note that our simplest ensemble model, an average of individual model predictions, performed approximately as well as complex machine learning models when we trained our superensembles on the simulation dataset and tested them on a separate 'real' dataset (i.e.
-the RAM Legacy Stock Assessment database, Fig.
-\ref{performance}b).
-Third, ensemble models will be most effective when they are trained on data that are representative of the dataset of interest [@knutti2009; @weigel2010].
-Cross-validation within a training dataset will provide an optimistically biased impression of predictive performance if the training dataset fundamentally differs from the dataset of interest [@hastie2009].
+First, ensembles will be most effective when they are comprised of diverse
+individual models that choose different structural model forms, explore
+contrasting but plausible ranges of parameter values, or make uncorrelated
+errors [@ali1996; @dietterich2000; @tebaldi2007]. Such individual models would
+be expected to perform well in different conditions and an ensemble model can
+exploit the best predictive performance of each <!-- is this only true for
+superensembles?-->. Second, ensemble models will be most effective when they are
+not overfit to the training dataset. Cross-validation testing [@caruana2004;
+@hastie2009] and methods that are robust to overfitting such as random forests
+[@breiman2001], may help avoid overfitting ensemble models. We note that our
+simplest ensemble model, an average of individual model predictions, performed
+approximately as well as complex machine learning models when we trained our
+superensembles on the simulation dataset and tested them on a separate 'real'
+dataset (i.e. the RAM Legacy Stock Assessment database, Fig.
+\ref{performance}b). Third, ensemble models will be most effective when they are
+trained on data that are representative of the dataset of interest [@knutti2009;
+@weigel2010]. Cross-validation within a training dataset will provide an
+optimistically biased impression of predictive performance if the training
+dataset fundamentally differs from the dataset of interest [@hastie2009].
 
-Multi-model inference in the form of coefficient averaging weighted by information theoretics such as the Akaike Information Criterion (AIC) is a common analytical approach in ecology [@burnham2002; @johnson2004; @grueber2011].
-The ensemble methods described in this paper share similarities with coefficient averaging but differ in other important ways.
-Ensemble methods and coefficient averaging share the long-held notion that multiple working hypotheses can contribute useful information for inference [@chamberlin1890].
-A fundamental difference is that coefficient averaging focuses on averaging *coefficients* whereas ensembles instead average *predictions*.
-Thus, ensembles provide a more general purpose tool: they do not require information theoretics and they can combine different types of models (e.g.
-parametric and non-parametric models or frequentist and Bayesian predictions).
-Furthermore, superensembles extend these benefits by allowing model predictions to be combined via non-linear functions that are tuned to known data.
+Multi-model inference in the form of coefficient averaging weighted by
+information theoretics such as the Akaike Information Criterion (AIC) is
+a common analytical approach in ecology [@burnham2002; @johnson2004;
+@grueber2011]. The ensemble methods described in this paper share similarities
+with coefficient averaging but differ in other important ways. Ensemble methods
+and coefficient averaging share the long-held notion that multiple working
+hypotheses can contribute useful information for inference [@chamberlin1890].
+A fundamental difference is that coefficient averaging focuses on averaging
+*coefficients* whereas ensembles instead average *predictions*. Thus, ensembles
+provide a more general purpose tool: they do not require information theoretics
+and they can combine different types of models (e.g. parametric and
+non-parametric models or frequentist and Bayesian predictions). Furthermore,
+superensembles extend these benefits by allowing model predictions to be
+combined via non-linear functions that are tuned to known data.
 
-A strength of superensembles is that they can be tailored to predict specific response variables.
-For example, we built separate superensemble models of mean $B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$.
-The same set of model weights or non-linear relationships need not hold across different response variables.
-For instance, SSCOM contributed little to the GBM superensemble estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$ (Fig.
+A strength of superensembles is that they can be tailored to predict specific
+response variables. For example, we built separate superensemble models of mean
+$B/B_\mathrm{MSY}$ and the slope of $B/B_\mathrm{MSY}$. The same set of model
+weights or non-linear relationships need not hold across different response
+variables. For instance, SSCOM contributed little to the GBM superensemble
+estimate of status at higher levels of predicted $B/B_\mathrm{MSY}$ (Fig.
 \ref{partial-sim}), but contributed strongly to estimates of trend (Fig.
-\ref{partial-sim-slope}).
-Formally, fitting superensemble models to specific quantities of interest provides an additional calibration step [@rykiel1996].
-Individual models are typically calibrated (i.e.
-parameters are estimated) based on a single response variable; however, this model might then provide a biased or inaccurate estimate of other quantities of interest.
-Ensemble methods provide a final calibration step to a quantity of interest.
-This ensemble calibration could further include a loss function tailored to
- the goals of the model, say placing greater weight on accuracy at low status levels than higher values.
-Conversely, because superensembles are tailored to a specific response and loss function, superensembles force a modeller to choose an operational purpose for their model up front [*sensu* @dickey-collas2014].
+\ref{partial-sim-slope}). Formally, fitting superensemble models to specific
+quantities of interest provides an additional calibration step [@rykiel1996].
+Individual models are typically calibrated (i.e. parameters are estimated) based
+on a single response variable; however, this model might then provide a biased
+or inaccurate estimate of other quantities of interest. Ensemble methods provide
+a final calibration step to a quantity of interest. This ensemble calibration
+could further include a loss function tailored to the goals of the model, say
+placing greater weight on accuracy at low status levels than higher values.
+Conversely, because superensembles are tailored to a specific response and loss
+function, superensembles force a modeller to choose an operational purpose for
+their model up front [*sensu* @dickey-collas2014].
 
-As @box1987 noted, all models are wrong, but some may still be useful.
-The ensemble methods we investigated attempt to piece together the useful parts of candidate models to build a model with improved performance.
-Instead of viewing the superensemble as a black box, we think considerable mechanistic understanding can be gained by studying its structure.
-For example, when SSCOM estimates very low status this is likely the case, conversely when COMSIR estimates low status, the true status is more likely to be high (Fig.
-\ref{partial-sim}).
-These models have two main differences: (1) the form of effort dynamics and (2) the allowance for both measurement and process error in SSCOM, whereas the implemented COMSIR admits measurement error only.
-Were the methods to differ only in effort dynamics, the results point towards a more suitable representation of effort dynamics at low biomasses in SSCOM.
-We think that such investigation of the parameters of a superensemble may lead to improvement in the mechanisms assumed in the individual models.
+As @box1987 noted, all models are wrong, but some may still be useful. The
+ensemble methods we investigated attempt to piece together the useful parts of
+candidate models to build a model with improved performance. Instead of viewing
+the superensemble as a black box, we think considerable mechanistic
+understanding can be gained by studying its structure. For example, when SSCOM
+estimates very low status this is likely the case, conversely when COMSIR
+estimates low status, the true status is more likely to be high (Fig.
+\ref{partial-sim}). These models have two main differences: (1) the form of
+effort dynamics and (2) the allowance for both measurement and process error in
+SSCOM, whereas the implemented COMSIR admits measurement error only. Were the
+methods to differ only in effort dynamics, the results point towards a more
+suitable representation of effort dynamics at low biomasses in SSCOM. We think
+that such investigation of the parameters of a superensemble may lead to
+improvement in the mechanisms assumed in the individual models.
 
-Combining predictions from multiple models via superensemble methods is also broadly useful in other subfields of ecology and fisheries science.
-Predictions about extinction risk are widely used at the national (e.g.
-the US Endangered Species Act and the Canadian Species at Risk Act) and international [e.g.
-the IUCN Red List, @iucn2015] levels.
-These risk assessments generally involve fitting regression models to outcomes for individual species along with predictors of extinction risk [e.g.
-@anderson2011a; @pinsky2011], or population dynamics models to data for individual species [e.g.
-@dfo2010].
-Both types of models are prone to error caused by model-misspecification and therefore results are sensitive to decisions about model structure [@brooks2015].
-Although there are options to account for potential model-misspecification in determination of species risk [e.g.
-coefficient averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are a relatively simple way to combine predictions in a transparent manner.
-Beyond estimates of status and trend, ensemble methods could be used, for example, to increase the robustness of spatial predictions when designing networks of protected areas [@rassweiler2014] or to forecast potential spatial shifts in species distribution given climate impacts [@harsch2014].
+Combining predictions from multiple models via superensemble methods is also
+broadly useful in other subfields of ecology and fisheries science. Predictions
+about extinction risk are widely used at the national (e.g. the US Endangered
+Species Act and the Canadian Species at Risk Act) and international [e.g. the
+IUCN Red List, @iucn2015] levels. These risk assessments generally involve
+fitting regression models to outcomes for individual species along with
+predictors of extinction risk [e.g. @anderson2011a; @pinsky2011], or population
+dynamics models to data for individual species [e.g. @dfo2010]. Both types of
+models are prone to error caused by model-misspecification and therefore results
+are sensitive to decisions about model structure [@brooks2015]. Although there
+are options to account for potential model-misspecification in determination of
+species risk [e.g. coefficient averaging, @burnham2002; generalized modeling,
+@yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are
+a relatively simple way to combine predictions in a transparent manner. Beyond
+estimates of status and trend, ensemble methods could be used, for example, to
+increase the robustness of spatial predictions when designing networks of
+protected areas [@rassweiler2014] or to forecast potential spatial shifts in
+species distribution given climate impacts [@harsch2014].
 
-<!-- TODO put back when space: or to evaluate the effect of conservation management actions when experimental or observational data are limited [@walsh2012].-->
+<!-- TODO put back when space: or to evaluate the effect of conservation
+management actions when experimental or observational data are limited
+[@walsh2012].-->
 
@@
 
-**Acknowledgements**: We thank members of Phase I of the working group 'Developing new approaches to global stock status assessment and fishery production potential of the seas' who
-contributed to developing the data-limited methods and simulations used in our
-analysis.
-We thank E. Jardim, F. Scott, and J.A. Hutchings for helpful comments during the development of this project, and R.D. Methot for comments on an earlier version of the manuscript.
-We thank the Gordon and Betty Moore Foundation for funding the working group
-'Applying data-limited stock status models and developing management guidance for unassessed fish stocks'.
+**Acknowledgements**: We thank members of Phase I of the working group
+'Developing new approaches to global stock status assessment and fishery
+production potential of the seas' who contributed to developing the data-limited
+methods and simulations used in our analysis. We thank E. Jardim, F. Scott, and
+J.A. Hutchings for helpful comments during the development of this project, and
+R.D. Methot for comments on an earlier version of the manuscript. We thank the
+Gordon and Betty Moore Foundation for funding the working group 'Applying
+data-limited stock status models and developing management guidance for
+unassessed fish stocks'.
 

### 2016-04-06 text/ms.Rmd
Wrap the manuscript for editing by voice
Editing of the main manuscript file

@@
 evaluated the potential for ensemble and superensemble models ('ensemble
-methods') to improve estimates of population status and trend for fisheries. We
-fit four widely applicable data-limited models that estimate stock biomass
+methods') to improve estimates of population status and trend for fisheries.
+We fit four widely applicable data-limited models that estimate stock biomass
 relative to the equilibrium biomass at maximum sustainable yield
 ($B/B_\mathrm{MSY}$). We combined the models' estimates to build four ensemble
-methods: an ensemble average and three superensembles (a linear model, a random
-forest, and a boosted regression tree). We used the ensemble methods to predict
-both (i) fishery status and (ii) recent trends of $B/B_\mathrm{MSY}$. We
-trained our models on a simulated dataset of 5760 stocks and tested them with
-cross-validation and against a global database of `r ram_stocks_n` real stock
-assessments. We found that ensemble methods substantially improved estimates of
-population status and trend. Random forest and boosted regression trees
-performed the best for estimating population status: accuracy improved `r
-mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and
-true status improved from `r mean_sim$ind_corr_range` to `r
-mean_sim$mach_corr_range` across models?, and bias (median proportional error)
-declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We
-found similar improvements when predicting the trend of status and when
-applying the simulation-trained superensembles to catch data for global fish
-stocks. Ensemble methods can improve estimates of status and trends; however,
-they must be tested, formed from a diverse set of accurate models, and built on
-a dataset representative of the populations to which they are applied. *TODO:
-Still need to cut ~30 words*
+methods: an ensemble average and three superensembles (a linear model,
+a random forest, and a boosted regression tree). We used the ensemble methods
+to predict both (i) fishery status and (ii) recent trends of
+$B/B_\mathrm{MSY}$. We trained our models on a simulated dataset of 5760
+stocks and tested them with cross-validation and against a global database of
+`r ram_stocks_n` real stock assessments. We found that ensemble methods
+substantially improved estimates of population status and trend. Random forest
+and boosted regression trees performed the best for estimating population
+status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order
+correlation between predicted and true status improved from `r
+mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range` across models?, and
+bias (median proportional error) declined from `r mean_sim$ind_mre_range` to
+`r mean_sim$mach_mre_range`. We found similar improvements when predicting the
+trend of status and when applying the simulation-trained superensembles to
+catch data for global fish stocks. Ensemble methods can improve estimates of
+status and trends; however, they must be tested, formed from a diverse set of
+accurate models, and built on a dataset representative of the populations to
+which they are applied. *TODO: Still need to cut ~30 words*
 
@@
 less biased than individual model estimates and can integrate uncertainty in
-model structure, initial conditions, and parameter estimation [@dietterich2000;
-@araujo2007]. Ensembles are superior to individual models for at least three
-reasons including statistical, computational, and representational reasons
-[@dietterich2000]. Statistically by averaging across models representing
-different hypotheses about reality, computationally by reducing the risk of
-getting stuck in a local optima, and representationally he by expanding
-representational form<!--explain how this is possible.--> This approach forms
-the basis of many machine learning methods [e.g. @dietterich2000], has helped
-reconcile climate forecasts from dozens of models [e.g. @murphy2004;
+model structure, initial conditions, and parameter estimation
+[@dietterich2000; @araujo2007]. Ensembles are superior to individual models
+for at least three reasons: statistically by averaging across models and
+therefore being less likely to pick the "wrong" model, computationally by
+reducing the risk of getting stuck in a local optima, and representationally
+by expanding the range of hypotheses explored [@dietterich2000]. This approach
+forms the basis of many machine learning methods [e.g. @dietterich2000], has
+helped reconcile climate forecasts from dozens of models [e.g. @murphy2004;
 @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria
@@
 improve species distribution modelling [e.g. @araujo2007; @breiner2015] and
-indeed have been used to combine estimates of population status and trend [e.g.
-@brodziak2010].
+indeed have been used to combine estimates of population status and trend
+[e.g. @brodziak2010].
 
@@
 certain conditions, and the covariance between models may contain information
-that can improve predictive accuracy. <!--can you give an example here? e.g.
-a model may behave poorly at estimating low biomass. By training the
-superensbmeles models with real data, it can identify the pros and cons of each
-model and incorporate them into the estimates. .--> We can exploit these
-characteristics by using the predictions from a group of models as inputs into
-a separate statistical model. This technique, sometimes called superensemble
-modelling [@krishnamurti1999], is common in climate and weather forecasting
-[e.g. @yun2005; @mote2015]. The superensemble is fit to a training dataset
-where outcomes are well known and then used to predict on a dataset of interest
-(Fig. \ref{didactic}). For example, @krishnamurti1999 combined predictions of
-wind and precipitation in Asian monsoons via a superensemble regression fit to
-observed data. Their superensemble was considerably more accurate than any
-individual prediction or weighted average of predictions.
+that can improve predictive accuracy. For example, one model might perform
+well at estimating high levels of abundance but be biased at low levels of
+abundance, while another model might have the opposite properties. An optimal
+combination of these models is not simply an average of the two.
+
+We can exploit these characteristics by using the predictions from a group of
+models as inputs into a separate statistical model. This technique, sometimes
+called superensemble modelling [@krishnamurti1999], is common in climate and
+weather forecasting [e.g. @yun2005; @mote2015]. The superensemble is fit to a
+training dataset where outcomes are well known and then used to predict on a
+dataset of interest (Fig. \ref{didactic}). For example, @krishnamurti1999
+combined predictions of wind and precipitation in Asian monsoons via a
+superensemble regression fit to observed data. Their superensemble was
+considerably more accurate than any individual prediction or weighted average
+of predictions.
 
@@
 appropriate data (e.g. catches, size and age distributions, surveys, and
-tagging information) to quantify values such as the biomass of a stock that can
-produce maximum sustainable yield ($B_\mathrm{MSY}$) [@hilborn1992]. However,
-the broad range of data required to conduct these stock assessments are not
-available for the majority of fish populations, including those of conservation
-concern and of economic interest to fisheries [@fao2014]. Therefore, a number
-of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited
-data available for the majority of fish stocks: (1) a time series of the total
-biomass of catch and (2) a basic understanding of population productivity [e.g.
-@vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the
-performance of four data-limited models through a large-scale simulation
-experiment. Three of these models are based on Schaefer (logistic) biomass
-dynamics and one is an empirical model fitted to more data-rich stock
-assessment output. The four models frequently disagreed about population
-status, no one model had strong performance across all fish stocks, and some
-models performed better than others depending on circumstances (e.g. Fig.
-\ref{motivate}).
+tagging information) to quantify values such as the biomass of a stock that
+can produce maximum sustainable yield ($B_\mathrm{MSY}$) [@hilborn1992].
+However, the broad range of data required to conduct these stock assessments
+are not available for the majority of fish populations, including those of
+conservation concern and of economic interest to fisheries [@fao2014].
+
+Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$
+based on the limited data available for the majority of fish stocks: (1)
+a time series of the total biomass of catch and (2) a basic understanding of
+population productivity [e.g. @vasconcellos2005; @martell2013]. Recently,
+@rosenberg2014 investigated the performance of four data-limited models
+through a large-scale simulation experiment. Three of these models are based
+on Schaefer (logistic) biomass dynamics and one is an empirical model fitted
+to more data-rich stock assessment output. The four models frequently
+disagreed about population status, no one model had strong performance across
+all fish stocks, and some models performed better than others depending on
+circumstances (e.g. Fig. \ref{motivate}).
 
@@
 predictive performance against each other and the individual models.
-
-<!-- TODO: It might be good to throw in a single sentence here describing the
-4 models, or at least listing them and maybe a couple key differences among
-them.-->
 
@@
 
-We first developed and tested ensemble methods on a fully factorial simulated
-dataset of fisheries [@rosenberg2014] with known status. These simulated data
-included the following scenarios: three fish life histories: small pelagic,
-demersal, and large pelagic; three levels of initial biomass depletion compared
-to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and
-four exploitation settings: (1) a constant exploitation rate, (2) an
-exploitation rate coupled with biomass to mimic an open-access single-species
-fishery, (3) a scenario where exploitation rate increased continuously, and (4)
-a 'roller-coaster' scenario where the exploitation rate increased and then
-decreased. Process noise (recruitment variability; i.e. unexplained variability
-in population dynamics) was introduced to the models at two magnitudes in log
-space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated through
-time or had first-order autoregressive correlation of $0.6$. The simulation
-also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch)
-and one scenario without measurement error. @rosenberg2014 ran ten iterations
-for each combination of factors adding stochastic draws of recruitment and
-catch-recording variability each time to generate a total of 5760 stocks. The
-simulation models were built in the `FLR` packages [@kell2007,
-<http://flr-project.org>] for the statistical software R [@r2015]. Code to
-generate the simulations is available at
-<https://github.com/iagomosqueira/stockStatusFAO>.
-
-We also tested our ensemble methods on the RAM Legacy Stock Assessment Database
-[@ricard2012] --- a compilation of stock-assessment output from hundreds of
-exploited marine populations around the world. Our analysis of the
-stock-assessment database was based on version 2.5. After removing stocks for
-which at least one of the individual models described below did not converge
-(121), this database included `r ram_stocks_n` stocks.
+<!-- TODO Add a summary of the methods here  -->
 
@@
 
-We fit four individual data-limited models that use data on catch and basic life
-history parameters to estimate $B/B_\mathrm{MSY}$. <!-- explain why we choose
-these models not others? --> Three of the models are mechanistic and based
-generally on Schaefer biomass dynamics [@schaefer1954] of the form
+We fit four individual data-limited models that use data on catch and basic
+life history parameters to estimate $B/B_\mathrm{MSY}$. We chose these models
+because TODO. Three of the models are mechanistic and based generally on
+Schaefer biomass dynamics [@schaefer1954] of the form
 
@@
 population growth rate, $B_0$ represents unfished biomass or carrying capacity
-$K$, and $C$ represents catch. The fourth model is an empirically derived model
-based on the RAM Legacy Stock Assessment Database. @rosenberg2014 provide a full
-background on these four methods and code to fit all the models is available in
-an accompanying package `datalimited` for the statistical software R TODO ADD
-URL. As described in the Introduction, the performance of these models at
-estimating $B/B_\mathrm{MSY}$ is variable. In summary:
+$K$, and $C$ represents catch. The fourth model is an empirically derived
+model based on the RAM Legacy Stock Assessment Database. @rosenberg2014
+provide a full background on these four methods and code to fit all the models
+is available in an accompanying package `datalimited` for the statistical
+software R TODO ADD URL. As described in the Introduction, the performance of
+these models at estimating $B/B_\mathrm{MSY}$ is variable. In summary:
 
-* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
-dynamics [@martell2013]. It requires a prior distribution for $r$ and a prior on
-the relative proportion of biomass at the end of the time series compared to
-unfished biomass (depletion), which is derived from the percentage of maximum
-catch at the end of the time series. The version of the model used in
-@rosenberg2014 was modified from @martell2013 to generate biomass trends from
-all viable $r$-$K$ pairs and produce an estimate of $B/B_\mathrm{MSY}$ from the
-median trend.
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
+  biomass dynamics [@martell2013]. It requires a prior distribution for $r$
+  and a prior on the relative proportion of biomass at the end of the time
+  series compared to unfished biomass (depletion), which is derived from the
+  percentage of maximum catch at the end of the time series. The version of
+  the model used in @rosenberg2014 was modified from @martell2013 to generate
+  biomass trends from all viable $r$-$K$ pairs and produce an estimate of
+  $B/B_\mathrm{MSY}$ from the median trend.
 
-* *COM-SIR* (catch-only-model with sampling-importance-resampling) is a coupled
-harvest-dynamics model [@vasconcellos2005]. Biomass is assumed to follow
-a Schaefer model and harvest dynamics are assumed to follow a logistic model.
-The model is fit with a sampling-importance-resampling algorithm
-[@rosenberg2014].
+* *COM-SIR* (catch-only-model with sampling-importance-resampling) is
+  a coupled harvest-dynamics model [@vasconcellos2005]. Biomass is assumed to
+  follow a Schaefer model and harvest dynamics are assumed to follow
+  a logistic model. The model is fit with a sampling-importance-resampling
+  algorithm [@rosenberg2014].
 
 * *SSCOM* (state-space catch-only model) is a hierarchical model that, similar
-to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013]. SSCOM
-estimates unobserved dynamics in both fishing effort and the fished population
-based on a catch time series and priors on $r$, the maximum rate of increase of
-fishing effort, and the magnitude of various forms of stochasticity. The model
-is fit in a Bayesian state-space framework to integrate across three forms of
-stochasticity: variation in effort, population dynamics, and fishing efficiency
-[@thorson2013].
+  to COM-SIR, is based on a coupled harvest-dynamics model [@thorson2013].
+  SSCOM estimates unobserved dynamics in both fishing effort and the fished
+  population based on a catch time series and priors on $r$, the maximum rate
+  of increase of fishing effort, and the magnitude of various forms of
+  stochasticity. The model is fit in a Bayesian state-space framework to
+  integrate across three forms of stochasticity: variation in effort,
+  population dynamics, and fishing efficiency [@thorson2013].
 
@@
   empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment
-database to fit a regression model to a series of characteristics of the catch
-time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response.
-The model used in this paper is modified from the original --- it condenses the
-life-history categories into three categories to match the simulated dataset,
-removes the maximum catch predictor since the absolute catch in the simulated
-dataset is arbitrary, and does not implement the bias correction needed in
-@costello2012 since we do not derive estimates of median status across multiple
-stocks. <!--TODO is this last sentence necessary? -->
+  database to fit a regression model to a series of characteristics of the
+  catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the
+  response. The model used in this paper is modified from the original --- it
+  condenses the life-history categories into three categories to match the
+  simulated dataset, removes the maximum catch predictor since the absolute
+  catch in the simulated dataset is arbitrary, and does not implement the bias
+  correction needed in @costello2012 since we do not derive estimates of
+  median status across multiple stocks. <!--TODO is this last sentence
+  necessary? -->
+
+## Simulated data set to build the ensemble TODO: finish this
+
+We first developed and tested ensemble methods on a fully factorial simulated
+dataset of fisheries [@rosenberg2014] with known status. These simulated data
+included the following scenarios: three fish life histories: small pelagic,
+demersal, and large pelagic; three levels of initial biomass depletion
+compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying
+capacity; and four exploitation settings: (1) a constant exploitation rate,
+(2) an exploitation rate coupled with biomass to mimic an open-access
+single-species fishery, (3) a scenario where exploitation rate increased
+continuously, and (4) a 'roller-coaster' scenario where the exploitation
+rate increased and then decreased. Process noise (recruitment variability;
+i.e. unexplained variability in population dynamics) was introduced to the
+models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$,
+and was either uncorrelated through time or had first-order autoregressive
+correlation of $0.6$. The simulation also included a scenario with $N(0,
+0.2^2)$ measurement error around log(catch) and one scenario without
+measurement error. @rosenberg2014 ran ten iterations for each combination
+of factors adding stochastic draws of recruitment and catch-recording
+variability each time to generate a total of 5760 stocks. The simulation
+models were built in the `FLR` packages [@kell2007,
+<http://flr-project.org>] for the statistical software R [@r2015]. Code to
+generate the simulations is available at
+<https://github.com/iagomosqueira/stockStatusFAO>.
+
+We also tested our ensemble methods on the RAM Legacy Stock Assessment
+Database [@ricard2012] --- a compilation of stock-assessment output from
+hundreds of exploited marine populations around the world. Our analysis of the
+stock-assessment database was based on version 2.5. After removing stocks for
+which at least one of the individual models described below did not converge
+(121), this database included `r ram_stocks_n` stocks.
 
@@
 
-The individual models we seek to combine with superensembles provide time series
-of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use superensembles to
-estimate any property of these time series. Here, we focus on two properties:
-the mean and slope of $B/B_\mathrm{MSY}$ in the last five years. Together, these
-quantities address the recent state and trend of stock status, which are both of
-management and conservation interest [e.g. @mace2008] TODO add fishery REF. To
-avoid undue influence of the end points of the time series on the calculated
-slope, we measured the slope as the Theil-Sen estimator of median slope
-[@theil1950].
+The individual models we seek to combine with superensembles provide time
+series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use
+superensembles to estimate any property of these time series. Here, we focus
+on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five
+years. Together, these quantities address the recent state and trend of stock
+status, which are both of management and conservation interest [e.g.
+@mace2008] TODO add fishery REF. To avoid undue influence of the end points of
+the time series on the calculated slope, we measured the slope as the
+Theil-Sen estimator of median slope [@theil1950].
 
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and the
-predictions from the individual models <!--, and the spectral values, fix this
-explanation --> as predictors in our superensemble models. When modelling mean
-$B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the superensemble
-models in log space and exponentiated the predictions. For the estimates of
-$B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble
-models on the natural untransformed scale.
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and
+the predictions from the individual models <!--, and the spectral values, fix
+this explanation --> as predictors in our superensemble models. When modelling
+mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the
+superensemble models in log space and exponentiated the predictions. For the
+estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit
+superensemble models on the natural untransformed scale.
 
-We compared an ensemble average and three superensembles of varying complexity:
-a linear model with two-way interactions, a random forest, and a boosted
-regression tree. We describe these models as estimating $\hat{\theta}$, which
-represents either the ensemble estimated $\log B/B_\mathrm{MSY}$ or slope of
-$B/B_\mathrm{MSY}$. The individual model estimates of $\log B/B_\mathrm{MSY}$ or
-slope of $B/B_\mathrm{MSY}$ are represented as $b$ for models $i$ $1$ through
-$n$ ($4$). The ensemble average was calculated as:
+We compared an ensemble average and three superensembles of varying
+complexity: a linear model with two-way interactions, a random forest, and
+a boosted regression tree. We describe these models as estimating
+$\hat{\theta}$, which represents either the ensemble estimated $\log
+B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$. The individual model
+estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
+represented as $b$ for models $i$ $1$ through $n$ ($4$). The ensemble average
+was calculated as:
 
@@
 determine what value of a predictor best splits the response data into two
-'branches' based on a loss function [@breiman1984]. In random forests, a series
-of regression trees are built on a random subset of the data and random subset
-of the covariates of the model [@breiman2001]. In generalized boosted models
-(GBMs), each subsequent model is fit to the residuals from the previous model;
-data points that are fit poorly in a given model are given more weight in the
-next model [@elith2008]. Random forests and GBMs can provide strong predictive
-performance and fit highly non-linear relationships [@elith2008; @hastie2009].
-We fit random forest models with the `randomForest` package [@liaw2002] for
-R with the default argument values. We fit boosted regression tree models with
-the `gbm` package [@ridgeway2015] for R. Based on initial tuning with the
-**caret** R package [@caret2008], we chose to fit GBMs with 2000 trees, an
-interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and
-all other arguments at their default values:
+'branches' based on a loss function [@breiman1984]. In random forests,
+a series of regression trees are built on a random subset of the data and
+random subset of the covariates of the model [@breiman2001]. In generalized
+boosted models (GBMs), each subsequent model is fit to the residuals from the
+previous model; data points that are fit poorly in a given model are given
+more weight in the next model [@elith2008]. Random forests and GBMs can
+provide strong predictive performance and fit highly non-linear relationships
+[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
+package [@liaw2002] for R with the default argument values. We fit boosted
+regression tree models with the `gbm` package [@ridgeway2015] for R. Based on
+initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs
+with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage
+parameter) of $0.01$, and all other arguments at their default values:
 

### 2016-04-06 text/ms.Rmd
Mainly edits to the paper

@@
 We fit four individual data-limited models that use data on catch and basic
-life history parameters to estimate $B/B_\mathrm{MSY}$. We chose these models
-because TODO. Three of the models are mechanistic and based generally on
-Schaefer biomass dynamics [@schaefer1954] of the form
+life-history parameters to estimate $B/B_\mathrm{MSY}$. We chose these models
+because because they can be fit to the vast majority of fisheries around the
+world, are established in the literature, and have been extensively simulation
+tested [@rosenberg2014].
+
+Three of the models are mechanistic and based generally on Schaefer biomass
+dynamics [@schaefer1954] of the form
 
@@
   correction needed in @costello2012 since we do not derive estimates of
-  median status across multiple stocks. <!--TODO is this last sentence
-  necessary? -->
+  median status across multiple stocks.
 
-## Simulated data set to build the ensemble TODO: finish this
+## Simulated data set to build the ensemble
 
@@
 <https://github.com/iagomosqueira/stockStatusFAO>.
+
+## Superensemble models
+
+The individual models we seek to combine with superensembles provide time
+series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use
+superensembles to estimate any property of these time series. Here, we focus
+on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five
+years. Together, these quantities address the recent state and trend of stock
+status, which are both of management and conservation interest [e.g.
+@mace2008] TODO add fishery REF. To avoid undue influence of the end points of
+the time series on the calculated slope, we measured the slope as the
+Theil-Sen estimator of median slope [@theil1950].
+
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and
+the predictions from the individual models <!--, and the spectral values, fix
+this explanation --> as predictors in our superensemble models. When modelling
+mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the
+superensemble models in log space and exponentiated the predictions. For the
+estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit
+superensemble models on the natural untransformed scale.
+
+We compared an ensemble average and three superensembles of varying
+complexity: a linear model with two-way interactions, a random forest, and
+a boosted regression tree. We describe these models as estimating
+$\hat{\theta}$, which represents either the ensemble estimated $\log
+B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$. The individual model
+estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
+represented as $b$ for models $i$ $1$ through $n$. The subscript $j$ indexes
+the fisheries from $1$ through $m$. The ensemble average was calculated as:
+
+\begin{equation}
+\hat{\theta_j} = \frac{1}{n} \sum_{i=1}^n \hat{b}_{i,j} , \quad
+\mathrm{for}\ j  =  1,...,m.
+\end{equation}
+
+\noindent We fit the linear model superensemble with all second-order
+interactions:
+
+\begin{equation}
+\hat{\theta_j} = \beta_0 + \beta_1 \hat{b}_{j,0} + ... +
+\beta_{1,2} \hat{b}_{j,1} \hat{b}_{j,2} + ... +
+\epsilon_j,
+\quad
+\mathrm{Normal}(0, \sigma^2)
+\quad \mathrm{for}\ j  =  1,...,m
+\end{equation}
+
+Our two machine learning superensemble models, a random forest and a boosted
+regression tree, were based on regression trees. Regression trees sequentially
+determine what value of a predictor best splits the response data into two
+'branches' based on a loss function [@breiman1984]. In random forests,
+a series of regression trees are built on a random subset of the data and
+random subset of the covariates of the model [@breiman2001]. In generalized
+boosted models (GBMs), each subsequent model is fit to the residuals from the
+previous model; data points that are fit poorly in a given model are given
+more weight in the next model [@elith2008]. Random forests and GBMs can
+provide strong predictive performance and fit highly non-linear relationships
+[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
+package [@liaw2002] for R with the default argument values. We fit boosted
+regression tree models with the `gbm` package [@ridgeway2015] for R. Based on
+initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs
+with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage
+parameter) of $0.01$, and all other arguments at their default values:
+
+TODO ADD EQN
 
@@
 <!--*TODO add new supp. fig.?*-->
-
-## Superensemble models
-
-The individual models we seek to combine with superensembles provide time
-series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use
-superensembles to estimate any property of these time series. Here, we focus
-on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five
-years. Together, these quantities address the recent state and trend of stock
-status, which are both of management and conservation interest [e.g.
-@mace2008] TODO add fishery REF. To avoid undue influence of the end points of
-the time series on the calculated slope, we measured the slope as the
-Theil-Sen estimator of median slope [@theil1950].
-
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and
-the predictions from the individual models <!--, and the spectral values, fix
-this explanation --> as predictors in our superensemble models. When modelling
-mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the
-superensemble models in log space and exponentiated the predictions. For the
-estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit
-superensemble models on the natural untransformed scale.
-
-We compared an ensemble average and three superensembles of varying
-complexity: a linear model with two-way interactions, a random forest, and
-a boosted regression tree. We describe these models as estimating
-$\hat{\theta}$, which represents either the ensemble estimated $\log
-B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$. The individual model
-estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
-represented as $b$ for models $i$ $1$ through $n$ ($4$). The ensemble average
-was calculated as:
-
-\begin{equation}
-\hat{\theta} = \frac{1}{n} \sum_{i=1}^n \hat{b_i}.
-\end{equation}
-
-$$ \hat{\theta} = \argmin_\theta \left( L(b \, | \, \hat{b}{\mathrm{ensemble}})
-\right) $$ $$ \hat{b}\mathrm{ensemble} = f \left( \hat{b}\mathrm{CMSY}, \,
-\hat{b}\mathrm{COM-SIR}, \, \hat{b}\mathrm{SSCOM}, \, \hat{b}\mathrm{mPRM}, \,
-S(0.2), \, S(0.5), \, | \, \theta \right) $$
-
-We fit the linear model superensemble with all second-order interactions:
-
-\begin{equation} \theta_{\mathrm{ensemble}(i)} = \beta_0 + \sum_{j=1}^m \beta_j
-\theta_{j(i)} + \sum_{j=1}^m \sum_{\substack{k =1 \ j \neq k}}^m \beta_{kj}
-\theta_{k(i)} \theta_{j(i)} + \epsilon_i, \quad \epsilon_i \sim
-\mathrm{Normal}(0, \sigma), \end{equation}
-
-TODO ADD EQN
-
-Our two machine learning superensemble models, a random forest and a boosted
-regression tree, were based on regression trees. Regression trees sequentially
-determine what value of a predictor best splits the response data into two
-'branches' based on a loss function [@breiman1984]. In random forests,
-a series of regression trees are built on a random subset of the data and
-random subset of the covariates of the model [@breiman2001]. In generalized
-boosted models (GBMs), each subsequent model is fit to the residuals from the
-previous model; data points that are fit poorly in a given model are given
-more weight in the next model [@elith2008]. Random forests and GBMs can
-provide strong predictive performance and fit highly non-linear relationships
-[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
-package [@liaw2002] for R with the default argument values. We fit boosted
-regression tree models with the `gbm` package [@ridgeway2015] for R. Based on
-initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs
-with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage
-parameter) of $0.01$, and all other arguments at their default values:
-
-TODO ADD EQN
 

### 2016-04-06 analysis/1-read-old-ram-fits.R

@@ -2,6 +2,7 @@
 # analysis). Instead of running this script, you can skip it and load the
 # cached .rds file at the end in subsequent scripts.
 # I.e., you can safely ignore this code.
+#
 
 if (Sys.info()[["user"]] == "seananderson") {
   cdat <- read.csv("raw-data/RAM_bmsy_Ctousev4.csv", stringsAsFactors=FALSE)

### 2016-04-06 text/pseudocode.R

@@ -1,0 +1,23 @@
+# Fit models to known data:
+cmsy_status <- cmsy(...)
+comsir_status <- comsir(...)
+# Combine and reshape output into wide format (not shown)
+head(data_known)
+# > cmsy_status comsir_status known_status
+# > 1.4         1.3           1.2
+# > ...         ...           ...
+# Build super ensemble model:
+ens_model <- lm(known_status ~ cmsy_status + comsir_status,
+  data = data_known)
+
+# Fit models to data of interest:
+cmsy_status <- cmsy(...)
+comsir_status <- comsir(...)
+# Combine and reshape output into wide format (not shown)
+head(data_unknown)
+# > cmsy_status comsir_status ...
+# > 0.9         1.1
+# > ...         ...
+# Predict status using super ensemble model:
+predict(model, newdata = data_unknown)
+# > 1.06  0.98  0.85 ...

### 2016-04-13 text/ms.Rmd
Many edits to the text

@@
 Population status and trend are two of the most fundamental values to quantify
-in fisheries management [e.g. TODO NEW REFS]. However, fisheries managers are
-often faced with reconciling multiple uncertain and potentially conflicting
-estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015].
-For example, some models may suggest a population is at risk and declining in
-abundance while others may suggest it is not at risk and stable.
+in fisheries management. However, fisheries managers are often faced with
+reconciling multiple uncertain and potentially conflicting estimates of status
+and trend [e.g. @brodziak2010; @branch2011; @deroba2015]. For example, some
+models may suggest a population is at risk and declining in abundance while
+others may suggest it is not at risk and stable.
 
@@
 [@dietterich2000; @araujo2007]. Ensembles are superior to individual models
-for at least three reasons: statistically by averaging across models and
-therefore being less likely to pick the "wrong" model, computationally by
-reducing the risk of getting stuck in a local optima, and representationally
-by expanding the range of hypotheses explored [@dietterich2000]. This approach
-forms the basis of many machine learning methods [e.g. @dietterich2000], has
-helped reconcile climate forecasts from dozens of models [e.g. @murphy2004;
-@tebaldi2007; @ipcc2013], and even improved early warning signs of malaria
-outbreaks [@thomson2006]. In ecology, ensemble methods are sometimes used to
-improve species distribution modelling [e.g. @araujo2007; @breiner2015] and
-indeed have been used to combine estimates of population status and trend
-[e.g. @brodziak2010].
+for at least three reasons: (1) statistically by averaging across models and
+therefore being less likely to pick the "wrong" model, (2) computationally by
+reducing the risk of getting stuck in a local optima, and (3)
+representationally by expanding the range of hypotheses explored
+[@dietterich2000]. This approach forms the basis of many machine learning
+methods [e.g. @dietterich2000], has helped reconcile climate forecasts from
+dozens of models [e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even
+improved early warning signs of malaria outbreaks [@thomson2006]. In ecology,
+ensemble methods are sometimes used to improve species distribution modelling
+[e.g. @araujo2007; @breiner2015] and indeed have been used to combine
+estimates of population status and trend [e.g. @brodziak2010].
 
@@
 @rosenberg2014 investigated the performance of four data-limited models
-through a large-scale simulation experiment. Three of these models are based
-on Schaefer (logistic) biomass dynamics and one is an empirical model fitted
+through a large-scale simulation experiment. Three of these models were based
+on Schaefer (logistic) biomass dynamics and one was an empirical model fitted
 to more data-rich stock assessment output. The four models frequently
@@
 
-<!-- TODO Add a summary of the methods here  -->
-
 ## Individual models of population status
@@
 \noindent where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus
-one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic
+one year, $B_t$ represents biomass at time $t$, $r$ represents intrinsic
 population growth rate, $B_0$ represents unfished biomass or carrying capacity
-$K$, and $C$ represents catch. The fourth model is an empirically derived
-model based on the RAM Legacy Stock Assessment Database. @rosenberg2014
+$K$, and $C$ represents catch. The fourth model is an empirically derived model
+based on the RAM Legacy Stock Assessment Database [@ricard2012]. @rosenberg2014
 provide a full background on these four methods and code to fit all the models
 is available in an accompanying package `datalimited` for the statistical
-software R TODO ADD URL. As described in the Introduction, the performance of
-these models at estimating $B/B_\mathrm{MSY}$ is variable. In summary:
+software R <https://github.com/datalimited/datalimited> (*Private while in
+review*). In summary:
 
@@
 
-## Simulated data set to build the ensemble
+## Simulated dataset to build the superensemble
 
@@
 
-## Superensemble models
+## Building the superensemble models
 
@@
 We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and
-the predictions from the individual models <!--, and the spectral values, fix
-this explanation --> as predictors in our superensemble models. When modelling
-mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the
-superensemble models in log space and exponentiated the predictions. For the
-estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit
-superensemble models on the natural untransformed scale.
+the predictions from the individual models as predictors in our superensemble
+models. When modelling mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero ---
+we fit the superensemble models in log space and exponentiated the
+predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, which are not
+bounded at zero, we fit superensemble models on the natural untransformed
+scale.
 
 We compared an ensemble average and three superensembles of varying
-complexity: a linear model with two-way interactions, a random forest, and
-a boosted regression tree. We describe these models as estimating
+complexity: a linear model with two-way interactions, a random forest, and a
+boosted regression tree. We describe these models as estimating
 $\hat{\theta}$, which represents either the ensemble estimated $\log
@@
 estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
-represented as $b$ for models $i$ $1$ through $n$. The subscript $j$ indexes
-the fisheries from $1$ through $m$. The ensemble average was calculated as:
+represented as $\hat{b}$ for models $i$ $1$ through $n$. The subscript $j$
+indexes the simulated fisheries from $1$ through $m$. The ensemble average was
+calculated as:
 
@@
 \quad
-\mathrm{Normal}(0, \sigma^2)
-\quad \mathrm{for}\ j  =  1,...,m
+\epsilon \sim \mathrm{Normal}(0, \sigma^2),
+\quad \mathrm{for}\ j  =  1,...,m.
 \end{equation}
 
-Our two machine learning superensemble models, a random forest and a boosted
-regression tree, were based on regression trees. Regression trees sequentially
-determine what value of a predictor best splits the response data into two
-'branches' based on a loss function [@breiman1984]. In random forests,
-a series of regression trees are built on a random subset of the data and
-random subset of the covariates of the model [@breiman2001]. In generalized
-boosted models (GBMs), each subsequent model is fit to the residuals from the
-previous model; data points that are fit poorly in a given model are given
-more weight in the next model [@elith2008]. Random forests and GBMs can
-provide strong predictive performance and fit highly non-linear relationships
-[@elith2008; @hastie2009]. We fit random forest models with the `randomForest`
-package [@liaw2002] for R with the default argument values. We fit boosted
-regression tree models with the `gbm` package [@ridgeway2015] for R. Based on
-initial tuning with the **caret** R package [@caret2008], we chose to fit GBMs
-with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage
-parameter) of $0.01$, and all other arguments at their default values:
-
-TODO ADD EQN
-
-We also tested our ensemble methods on the RAM Legacy Stock Assessment
-Database [@ricard2012] --- a compilation of stock-assessment output from
-hundreds of exploited marine populations around the world. Our analysis of the
-stock-assessment database was based on version 2.5. After removing stocks for
-which at least one of the individual models described below did not converge
-(121), this database included `r ram_stocks_n` stocks.
+Our two machine learning superensemble models, a random forest and a
+generalized boosted model (GBM), were based on regression trees. Regression
+trees sequentially determine what value of a predictor best splits the
+response data into two 'branches' based on a loss function [@breiman1984]. In
+random forests, a series of regression trees are built on a random subset of
+the data and random subset of the covariates of the model [@breiman2001]. In
+GBMs, each subsequent model is fit to the residuals from the previous model;
+data points that are fit poorly in a given model are given more weight in the
+next model [@elith2008]. Random forests and GBMs can provide strong predictive
+performance and fit highly non-linear relationships [@elith2008; @hastie2009].
+We fit random forest models with the `randomForest` package [@liaw2002] for R
+with the default argument values. We fit boosted regression tree models with
+the `gbm` package [@ridgeway2015] for R. We fit GBMs with 2000 trees, an
+interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and
+all other arguments at their default values.
 
@@
 
-<!-- Superensemble models allow us to incorporate additional covariates and
+Superensemble models allow us to incorporate additional covariates and
 potentially leverage interactions between these covariates and individual model
@@
 that is independent of time series length (except in affecting precision) and
-absolute magnitude of catch. We fit spectral models to the scaled catch time
-series (catch divided by maximum catch) with the `spec.ar` function in R and
-recorded representative short- and long-term spectral densities at frequencies
-of 0.20 and 0.05, which correspond to 5- and 20-year cycles. -->
+independent of absolute magnitude of catch. We fit spectral models to the
+scaled catch time series (catch divided by maximum catch) with the `spec.ar`
+function in R and recorded representative short- and long-term spectral
+densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year
+cycles. We include the results of adding these additional covariates in the
+supplementary materials.
 
-<!--*TODO add new supp. fig.?*-->
+## Applying the superensemble models and testing performance
 
-## Testing model performance
+Once the superensemble models are built, we can apply them to new stocks that
+we'd like to know the status of. To do this we applied the individual models
+to our stocks of interest and then applied these data to our already built
+superensemble models. In this paper we applied the superensemble models to
+subsets of the simulated data as a cross validation test and to the RAM Legacy
+Stock Assessment database to test on real stocks.
 
-A critical component to predictive modelling is to evaluate the performance of
-a model on new data [@hastie2009]. We used repeated three-fold cross validation
-to test predictive performance: we randomly divided the dataset into three sets,
-built superensemble models on two-thirds of the data, and evaluated predictive
-performance on the remaining third. We repeated this across each of the three
-splits and then repeated the whole procedure 50 times to account for bias that
-may result from any one set of validation splits. In the simulated dataset,
-there were 10 replicates of each unique combination of simulation parameters
-that differed only in stochastic variability. Since the dynamics of these
-populations were often similar, we grouped these stocks in the cross-validation
-process into either the training or testing split. We also validated the models
-using the RAM Legacy Stock Assessment database. In this case, we used
-superensembles trained on the entire simulation dataset. However, since mPRM is
-built on the same stock-assessment database, we applied three-fold
-cross-validation to the RAM Legacy Stock Assessment data underlying the mPRM
-model so that the dataset that mPRM was trained with (for the individual model
-and superensemble) was separate from the dataset it was tested on.
+We used repeated three-fold cross validation to test predictive performance: we
+randomly divided the dataset into three sets, built superensemble models on
+two-thirds of the data, and evaluated predictive performance on the remaining
+third. We repeated this across each of the three splits and then repeated the
+whole procedure 50 times to account for bias that may result from any one set
+of validation splits. In the simulated dataset, there were 10 replicates of
+each unique combination of simulation parameters that differed only in
+stochastic variability. Since the dynamics of these populations were often
+similar, we grouped these stocks in the cross-validation process into either
+the training or testing split.
 
-Predictive performance can be evaluated with metrics that represent a variety of
-modelling goals. For continuous response variables such as the mean and slope of
-population status, performance metrics often measure some form of bias,
-precision, accuracy (a combination of bias and precision), or the ability to
-correctly rank or correlate across populations [e.g. @walther2005]. Here, we
-measure proportional error, defined as $(\hat{b} - b)/\abs{b}$, where $\hat{b}$
-and $b$ represent estimated and 'true' (or stock-assessed) mean or slope of
-$B/B_\mathrm{MSY}$. We calculated median proportional error to measure bias,
-median absolute proportional error to measure accuracy, and Spearman's
-rank-order correlation between predicted and 'true' values to measure the
-ability to correctly rank populations. When cross-validating against the RAM
-Legacy Stock Assessment database, we treated the estimates from these data-rich
-stock assessments as known without error. Thus, any error in the
-stock-assessment estimates of the mean or slope of $B/B_\mathrm{MSY}$ also
-contributes to our estimates of prediction error for each of the four
-data-limited models and the ensembles. Code to reproduce our analysis and an
-R package to fit the data-limited assessment models are available at
-<https://github.com/datalimited/ensembles> (*Private while in review*).
+We also tested our ensemble methods on the RAM Legacy Stock Assessment Database
+--- a compilation of stock-assessment output from hundreds of exploited marine
+populations around the world. Our analysis of the stock-assessment database was
+based on version 2.5. After removing stocks for which at least one of the
+individual models described below did not converge (121), this database
+included `r ram_stocks_n` stocks. We removed these stocks for all methods ---
+both for the individual and superensemble models. An alternative would be to
+fit superensemble models to subsets of the individual models that did converge,
+but for the purposes of this paper we will only use superensemble models that
+use all four individual models.
+
+In the case of the RAM Legacy Stock Assessment Database, we used superensembles
+trained on the entire simulation dataset. However, since mPRM is built on the
+same stock assessment database, we applied three-fold cross-validation to the
+data underlying the mPRM model so that the dataset with which mPRM was trained
+(for the individual model and superensemble) was separate from the dataset with
+which it was tested.
+
+Predictive performance can be evaluated with metrics that represent a variety
+of modelling goals. For continuous response variables such as the mean and
+slope of population status, performance metrics often measure some form of
+bias, precision, accuracy (a combination of bias and precision), or the ability
+to correctly rank or correlate across populations [e.g. @walther2005]. Here, we
+measure proportional error, defined as $(\hat{\theta} - \theta)/\abs{\theta}$,
+where $\hat{\theta}$ and $\theta$ represent estimated and 'true' (or
+stock-assessed) mean or slope of $B/B_\mathrm{MSY}$. We calculated median
+proportional error to measure bias, median absolute proportional error to
+measure accuracy, and Spearman's rank-order correlation between predicted and
+'true' values to measure the ability to correctly rank populations. When
+testing with the RAM Legacy Stock Assessment database, we treated the
+estimates from these data-rich stock assessments as known without error. Thus,
+any error in the stock assessment estimates of the mean or slope of
+$B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for
+each of the four data-limited models and the ensembles. Code to reproduce our
+analysis and an R package to fit the data-limited assessment models are
+available at <https://github.com/datalimited/ensembles> (*Private while in
+review*).
 
@@
 any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h).
-Compared to the individual models, machine learning superensembles improved
-accuracy (median absolute proportional error) by `r
-mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r
-mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias
-(median proportional error) from `r mean_sim$ind_mre_range` to `r
-mean_sim$mach_mre_range` (Fig. \ref{performance}a). Results were similar when
-predicting trend: compared to individual models, machine learning superensembles
-improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order
-correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`,
-and reduced bias from `r slope_sim$ind_mre_range` to `r
-slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The ensemble
-models that simply took a mean of the individual models ranked slightly behind
-the best individual model for estimating fish stock status (CMSY; Fig.
-\ref{performance}a) and had slightly lower correlation but higher accuracy than
-the best individual model at predicting the trends of status (SSCOM; Fig.
+Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig.
+\ref{performance}a).
+Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
 \ref{performance-sim-slope}).
+The ensemble models that simply took a mean of the individual models ranked
+slightly behind the best individual model for estimating fish stock status
+(CMSY; Fig. \ref{performance}a) and had slightly lower correlation but higher
+accuracy than the best individual model at predicting the trends of status
+(SSCOM; Fig. \ref{performance-sim-slope}).
 
@@
 
-<!--TODO trash this?-->
+<!--> TODO update the result numbers to report the models that don't have the covariates -->
 
@@
 
-**Acknowledgements**: We thank members of Phase I of the working group
-'Developing new approaches to global stock status assessment and fishery
-production potential of the seas' who contributed to developing the data-limited
-methods and simulations used in our analysis. We thank E. Jardim, F. Scott, and
-J.A. Hutchings for helpful comments during the development of this project, and
-R.D. Methot for comments on an earlier version of the manuscript. We thank the
-Gordon and Betty Moore Foundation for funding the working group 'Applying
-data-limited stock status models and developing management guidance for
-unassessed fish stocks'.
+# Acknowledgments
+
+We thank members of Phase I of the working group 'Developing new approaches to
+global stock status assessment and fishery production potential of the seas'
+who contributed to developing the data-limited methods and simulations used in
+our analysis. We thank E. Jardim, F. Scott, and J.A. Hutchings for helpful
+comments during the development of this project, and R.D. Methot for comments
+on an earlier version of the manuscript. We thank the Gordon and Betty Moore
+Foundation for funding the working group 'Applying data-limited stock status
+models and developing management guidance for unassessed fish stocks'.
 
@@
 -->
+
+<!-- vim: set formatoptions=nroq: -->

### 2016-04-19 text/ms.Rmd
Edit main text

@@
 <!-- TODO
-- [ ] trash slope?
-- [ ] trash covariates?
 - [ ] add discussion of uncertainty (CIs) and ML (and incorporating uncertainty into pred.)
-- [ ] add table of fish methods and ensemble models
-- [ ] add flow chart of cross validation / methods
+- [ ] add back in the roc plot and sentence in the main text
+- [ ] rerun the main simulation analysis because the equations of change for the slope
 -->
@@
 
-\tableofcontents
+<!-- \tableofcontents -->
 
@@
 
-Population status and trend are two of the most fundamental values to quantify
-in fisheries management. However, fisheries managers are often faced with
-reconciling multiple uncertain and potentially conflicting estimates of status
-and trend [e.g. @brodziak2010; @branch2011; @deroba2015]. For example, some
-models may suggest a population is at risk and declining in abundance while
-others may suggest it is not at risk and stable.
+Status and trend are two of the most fundamental values to quantify in the
+management of ecological populations [e.g. @iucn2015]. However, managers are
+often faced with reconciling multiple uncertain and potentially conflicting
+estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015].
+For example, one models may suggest a population is at risk and declining in
+abundance while others may suggest it is not at risk and stable.
 
@@
 training dataset where outcomes are well known and then used to predict on a
-dataset of interest (Fig. \ref{didactic}). For example, @krishnamurti1999
+dataset of interest. For example, @krishnamurti1999
 combined predictions of wind and precipitation in Asian monsoons via a
@@
  5. testing model perforance. -->
+
+<!--TODO Add in the reference to (Fig. \ref{didactic}) and reorder the figures and add
+in the figure with the partial dependence -->
 
@@
 had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
-\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were approximately
-linearly related to $B/B_\mathrm{MSY}$ within the low and high clusters of
-predictions (Fig. \ref{partial-sim}). Superensembles also exploited the
-covariance between individual model predictions. For instance, both the linear
-model and GBM ensemble suggest that if mPRM and SSCOM predict high status, the
-true status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l).
-The addition of spectral density covariates helped the superensemble models
-correctly predict higher status values (Fig. \ref{hexagon}g, h) due to the
-interactions between the spectral density predictors and the individual model
-predictors (Fig. \ref{partial-2d-sim}). Although the additional covariates
-improved ensemble fit, performance of the ensembles was only marginally degraded
-by removing these covariates (Fig. \ref{hexagon} vs. Fig.
-\ref{hexagon-sim-basic}).
+\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
+approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high
+clusters of predictions (Fig. \ref{partial-sim}). Superensembles also exploited
+the covariance between individual model predictions. For instance, both the
+linear model and GBM ensemble suggest that if mPRM and SSCOM predict high
+status, the true status also tends to be high (Figs \ref{lm-coefs},
+\ref{partial-2d-sim}l). The addition of spectral density covariates helped the
+superensemble models correctly predict higher status values (Fig.
+\ref{hexagon-sim-covariance}g, h). The performance of the ensembles was only
+marginally improved by including these covariates (Fig.
+\ref{performance-with-covariance} vs. Fig. \ref{performance}.
 

### 2016-04-19 analysis/6-sim-to-ram.R
To rerun the analysis on the RAM data with the basic model

@@ -53,29 +53,19 @@ d_slope <- reshape2::dcast(ram_sum, stockid + scientificname ~ method,
 # bring in the simulation formatted data to build the simulation-trained models:
 d_mean_sim <- readRDS("generated-data/sim-mean-dat.rds")
 m_rf <- randomForest::randomForest(
-  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
-    spec_freq_0.05 + spec_freq_0.2,
+  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM,
+    # spec_freq_0.05 + spec_freq_0.2,
   data = d_mean_sim, ntree = 1000L)
 
 m_gbm <- gbm::gbm(
-  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
-    spec_freq_0.05 + spec_freq_0.2,
+  log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM,
+    # spec_freq_0.05 + spec_freq_0.2,
   data = d_mean_sim, distribution = "gaussian",
   n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
 
-m_lm0 <- lm(
-  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
-      spec_freq_0.05 + spec_freq_0.2),
-  data = d_mean_sim)
-
 m_lm <- lm(
-  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
-      spec_freq_0.05 + spec_freq_0.2)^2,
-  data = d_mean_sim)
-
-m_lm3 <- lm(
-  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
-      spec_freq_0.05 + spec_freq_0.2)^3,
+  log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM)^2,
+      # spec_freq_0.05 + spec_freq_0.2)^2,
   data = d_mean_sim)
 
 pdf("../figs/lm-coefs.pdf", width = 5, height = 6)
@@ -86,23 +76,6 @@ arm::coefplot(arm::standardize(m_lm), main = "")
 par(xpd = NA)
 mtext("Standardized regression coefficient", side = 3, line = 3)
 dev.off()
-
[…]

### 2016-04-19 analysis/5-simulation-ensembles.R
Integrate the basic model into the main simulation analysis

@@ -140,6 +140,14 @@ m_slope <- gbm::gbm(bbmsy_true_slope ~ CMSY + COMSIR + mPRM + SSCOM +
     spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
   data = d_slope, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
 
+m_basic <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM,
+  distribution = "gaussian",
+  data = d_mean, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
+
+m_slope_basic <- gbm::gbm(bbmsy_true_slope ~ CMSY + COMSIR + mPRM + SSCOM,
+  distribution = "gaussian",
+  data = d_slope, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)
+
 p1 <- ggplot(d_mean, aes(CMSY, gbm_pred)) + geom_point()
 p2 <- ggplot(d_mean, aes(COMSIR, gbm_pred)) + geom_point()
 p3 <- ggplot(d_mean, aes(mPRM, gbm_pred)) + geom_point()
@@ -160,52 +168,52 @@ pdf("../figs/predictor-vs-truth-gbm.pdf", width = 10, height = 5)
 gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 2)
 dev.off()
 
-make_partial_resid_gbm <- function(var = "SSCOM") {
-  d_temp <- d_mean
-  d_temp$x <- d_mean[,var]
-  d_temp[,var] <- mean(d_temp[,var])
-  d_temp$p <- predict(m, n.trees = m$n.trees, type = "response",
-    newdata = d_temp)
-  d_temp$res <- log(d_temp$bbmsy_true_mean) - d_temp$p
-  # ggplot(d_temp, aes(x, exp(res))) + geom_point() + stat_smooth(se = FALSE, col = "red", method = "loess")
+## make_partial_resid_gbm <- function(var = "SSCOM") {
+##   d_temp <- d_mean
+##   d_temp$x <- d_mean[,var]
+##   d_temp[,var] <- mean(d_temp[,var])
+##   d_temp$p <- predict(m, n.trees = m$n.trees, type = "response",
+##     newdata = d_temp)
+##   d_temp$res <- log(d_temp$bbmsy_true_mean) - d_temp$p
[…]

### 2016-04-19 analysis/8-main-figs.R
Revise the main figures to deal with the basic superensemble

@@ -4,8 +4,6 @@ library(ggplot2)
 source("0-ensemble-functions.R")
 
 # bring in the simulated data and shape it for plotting:
-d_sim <- readRDS("generated-data/cv_sim_long.rds")
-
 clean_names <- dplyr::data_frame(
   method = c("CMSY", "COMSIR", "mPRM", "SSCOM",
     "gbm_ensemble", "rf_ensemble", "lm_ensemble", "mean_ensemble"),
@@ -19,42 +17,47 @@ clean_names <- dplyr::data_frame(
     0, -0.02,0.03,0.04,
     0, 0.02,0,-0.02))
 
+d_sim <- readRDS("generated-data/cv_sim_long.rds")
 d_sim <- suppressWarnings(inner_join(d_sim, clean_names))
 
-d_sim_perf <- d_sim %>% group_by(test_iter, type, clean_method, order, label_fudge_x, label_fudge_y) %>%
-  summarise(
-    mare = median(abs(re)),
-    mre = median(re),
-    corr = cor(bbmsy_true_trans, bbmsy_est_trans, method = "spearman",
-      use = "pairwise.complete.obs"))
+get_performance <- function(validation_data) {
+  d_sim_perf <- validation_data %>% group_by(test_iter, type, clean_method, order, label_fudge_x, label_fudge_y) %>%
+    summarise(
+      mare = median(abs(re)),
+      mre = median(re),
+      corr = cor(bbmsy_true_trans, bbmsy_est_trans, method = "spearman",
+        use = "pairwise.complete.obs"))
 
-d_sim_perf$clean_method <- as.factor(d_sim_perf$clean_method)
-d_sim_perf$clean_method <- reorder(d_sim_perf$clean_method, d_sim_perf$order)
+  d_sim_perf$clean_method <- as.factor(d_sim_perf$clean_method)
+  d_sim_perf$clean_method <- reorder(d_sim_perf$clean_method, d_sim_perf$order)
 
-d_sim_perf_slope <- filter(d_sim_perf, type == "slope")
-d_sim_perf <- filter(d_sim_perf, type == "mean")
+  d_sim_perf_slope <- filter(d_sim_perf, type == "slope")
+  d_sim_perf <- filter(d_sim_perf, type == "mean")
 
[…]

### 2016-04-19 text/ms.Rmd
Rewrap a couple paragraphs and add a to do

@@
 - [ ] rerun the main simulation analysis because the equations of change for the slope
+- [ ] update the magic number get inserted to use the equation that only has the basic model
 -->
@@
 any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h).
-Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig.
-\ref{performance}a).
-Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
+Compared to the individual models, machine learning superensembles improved
+accuracy (median absolute proportional error) by
+`r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from
+`r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias
+(median proportional error) from `r mean_sim$ind_mre_range` to
+`r mean_sim$mach_mre_range` (Fig. \ref{performance}a).
+Results were similar when predicting trend: compared to individual models,
+machine learning superensembles improved accuracy by
+`r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from
+`r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced
+bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig.
 \ref{performance-sim-slope}).
@@
 \ref{hexagon-ram}). Compared to the individual models, the machine learning
-superensembles increased accuracy by `r mean_ram$mach_vs_ind_mare_fold`%,
-improved correlation from `r mean_ram$ind_corr_range` to `r
-mean_ram$mach_corr_range`, and reduced bias from `r mean_ram$ind_mre_range` to
-`r mean_ram$mach_mre_range`.
+superensembles increased accuracy by
+`r mean_ram$mach_vs_ind_mare_fold`%, improved correlation from
+`r mean_ram$ind_corr_range` to `r mean_ram$mach_corr_range`, and reduced bias
+from `r mean_ram$ind_mre_range` to `r mean_ram$mach_mre_range`.
 

### 2016-04-19 text/ms.Rmd
Update to do items

--- 2016-04-19 text/ms.Rmd
@@
 - [ ] add discussion of uncertainty (CIs) and ML (and incorporating uncertainty into pred.)
-- [ ] add back in the roc plot and sentence in the main text
+- [x] add back in the roc plot and sentence in the main text
 - [ ] rerun the main simulation analysis because the equations of change for the slope
+- [ ] manually adjusted the labels in figure 4
+- [ ] update the magic number get inserted to use the equation that only has the basic model
 -->

### 2016-04-19 analysis/5-simulation-ensembles.R
Add a hat to a figure

@@ -300,7 +300,7 @@ plyr::d_ply(partial_2d, c("var1", "var2"), function(x) {
         text(0.9, blocks[1], round(min(zlim), 1), pos = 2)
         text(0.9, 0.5, 1, pos = 2)
         text(0.9, blocks[length(blocks)], round(max(zlim), 1), pos = 2)
-        mtext(expression(B/B[MSY]), side = 2, las = 0, cex = 0.8)
+        mtext(expression(widehat(B/B[MSY])), side = 2, las = 0, cex = 0.8)
       }
     }
   }

### 2016-04-19 analysis/8-main-figs.R
Capture the output from the basic analysis

@@ -56,6 +56,7 @@ saveRDS(d_sim_perf_wide, "generated-data/d_sim_perf_wide.rds")
 d_sim_basic <- readRDS("generated-data/cv_sim_mean_basic_long.rds")
 d_sim_basic <- suppressWarnings(inner_join(d_sim_basic, clean_names))
 d_sim_perf_wide_basic <- get_performance(d_sim_basic)
+saveRDS(d_sim_perf_wide_basic, "generated-data/d_sim_perf_wide_basic.rds")
 d_sim_basic$bbmsy_est[d_sim_basic$bbmsy_est > 10] <- NA
 d_sim_basic <- na.omit(d_sim_basic)
 

### 2016-04-19 analysis/9-output-values.R

@@ -43,6 +43,9 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
 d_sim_perf_wide <- readRDS("generated-data/d_sim_perf_wide.rds")
 mean_sim <- get_performance_stats(d_sim_perf_wide)
 
+d_sim_perf_wide_basic <- readRDS("generated-data/d_sim_perf_wide_basic.rds")
+mean_sim_basic <- get_performance_stats(d_sim_perf_wide_basic)
+
 re_ram_sum <- readRDS("generated-data/re_ram_sum.rds")
 mean_ram <- get_performance_stats(re_ram_sum)
 

### 2016-04-19 analysis/9-output-values.R
Update the output values

@@ -68,4 +68,5 @@ auc_sim$ind_range <- auc_sim_mean %>% filter(grepl("Ensemble", clean_method) == 
 d_ram <- readRDS("generated-data/ram-ensemble-predicted.rds")
 ram_stocks_n <- length(unique(d_ram$stockid))
 
-save(mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n, slope_corr_sscom, range_slope_corr_non_sscom, file = "../text/values.rda")
+save(mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n, slope_corr_sscom,
+  range_slope_corr_non_sscom, mean_sim_basic, file = "../text/values.rda")

### 2016-04-19 text/ms.Rmd
And back ROC section in the results
Fix 2 figure labels
Insert basic performance measures in the results

@@
 accuracy (median absolute proportional error) by
-`r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from
-`r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias
-(median proportional error) from `r mean_sim$ind_mre_range` to
-`r mean_sim$mach_mre_range` (Fig. \ref{performance}a).
+`r mean_sim_basic$mach_vs_ind_mare_fold`%, increased rank-order correlation from
+`r mean_sim_basic$ind_corr_range` to `r mean_sim_basic$mach_corr_range`, and reduced bias
+(median proportional error) from `r mean_sim_basic$ind_mre_range` to
+`r mean_sim_basic$mach_mre_range` (Fig. \ref{performance}a).
+These superensembles also generally had better ability to distinguish if
+simulated stocks were above or below $B/B_\mathrm{MSY} = 0.5$ (Fig. \ref{roc-sim}).
 Results were similar when predicting trend: compared to individual models,
@@
 superensemble models correctly predict higher status values (Fig.
-\ref{hexagon-sim-covariance}g, h). The performance of the ensembles was only
+\ref{hexagon-sim-covariates}g, h). The performance of the ensembles was only
 marginally improved by including these covariates (Fig.
-\ref{performance-with-covariance} vs. Fig. \ref{performance}.
+\ref{performance-with-covariates} vs. Fig. \ref{performance}).
 

### 2016-04-19 text/ms.Rmd
Start editing discussion and removing comments

@@
 - [ ] manually adjusted the labels in figure 4
-- [ ] update the magic number get inserted to use the equation that only has the basic model
+- [x] update the magic number get inserted to use the equation that only has the basic model
 -->
@@
 
-<!-- TODO PE .... mean or slope of b/bmsy: Dan: logged or normal space? For the
-status regression where you are estimating log(B/Bmsy), wouldn't you need to
-correct for the retransformation bias in the estimating median proportional
-error later down? I could be missing something though in whether that applied
-once you've calculated proportional error-->
-
 # Results
@@
 
-<!--> TODO update the result numbers to report the models that don't have the covariates -->
-
 When applied to the stock-assessment database, the superensemble models ---
@@
 
-Ensemble methods provide a useful approach to situations where conservation and
-environmental resource management decisions must be made on the basis of
-multiple, potentially contrasting estimates of status --- a situation common to
-many settings beyond fisheries. Compared to individual models of fish population
-status, ensemble methods were consistently the best or among the best across
-three performance dimensions (accuracy, bias, and rank-order correlation), two
-response variables (status and trend), two datasets (simulated and global
-fisheries), and multiple ensemble methods (from a simple average to machine
-learning superensembles). Our results suggest choosing a superensemble model
-that allows for non-parametric relationships, such as machine learning methods.
-These methods provide added insight into individual model behavior and generally
-performed the best; however, even a simple average of predictions across
-multiple models may provide a more useful measure of the desired parameter than
-a single model in many cases.
+Ensemble methods provide a useful approach to situations where environmental
+resource management decisions must be made on the basis of multiple,
+potentially contrasting estimates of status. Compared to individual models of
+fish population status, ensemble methods were consistently the best or among
+the best across three performance dimensions (accuracy, bias, and rank-order
+correlation), two response variables (status and trend), two datasets
+(simulated and global fisheries), and multiple ensemble methods (from a simple
+average to machine learning superensembles). Our results suggest choosing a
+superensemble model that allows for non-parametric relationships, such as
+machine learning methods. These methods provide added insight into individual
+model behavior and generally performed the best; however, even a simple average
+of predictions across multiple models may provide a more useful measure of the
+desired parameter than a single model in many cases.
 
@@
 be expected to perform well in different conditions and an ensemble model can
-exploit the best predictive performance of each <!-- is this only true for
-superensembles?-->. Second, ensemble models will be most effective when they are
-not overfit to the training dataset. Cross-validation testing [@caruana2004;
-@hastie2009] and methods that are robust to overfitting such as random forests
-[@breiman2001], may help avoid overfitting ensemble models. We note that our
-simplest ensemble model, an average of individual model predictions, performed
-approximately as well as complex machine learning models when we trained our
-superensembles on the simulation dataset and tested them on a separate 'real'
-dataset (i.e. the RAM Legacy Stock Assessment database, Fig.
-\ref{performance}b). Third, ensemble models will be most effective when they are
-trained on data that are representative of the dataset of interest [@knutti2009;
-@weigel2010]. Cross-validation within a training dataset will provide an
-optimistically biased impression of predictive performance if the training
-dataset fundamentally differs from the dataset of interest [@hastie2009].
+exploit the best predictive performance of each superensembles? Second,
+ensemble models will be most effective when they are not overfit to the
+training dataset. Cross-validation testing [@caruana2004; @hastie2009] and
+methods that are robust to overfitting such as random forests [@breiman2001],
+may help avoid overfitting ensemble models. We note that our simplest ensemble
+model, an average of individual model predictions, performed approximately as
+well as complex machine learning models when we trained our superensembles on
+the simulation dataset and tested them on a separate 'real' dataset (i.e. the
+RAM Legacy Stock Assessment database, Fig. \ref{performance}b). Third, ensemble
+models will be most effective when they are trained on data that are
+representative of the dataset of interest [@knutti2009; @weigel2010].
+Cross-validation within a training dataset will provide an optimistically
+biased impression of predictive performance if the training dataset
+fundamentally differs from the dataset of interest [@hastie2009].
 

### 2016-04-27 text/ms.Rmd
Edit main text

@@
 - [x] add back in the roc plot and sentence in the main text
-- [ ] rerun the main simulation analysis because the equations of change for the slope
+- [x] rerun the main simulation analysis because the equations of change for the slope
 - [ ] manually adjusted the labels in figure 4
 - [x] update the magic number get inserted to use the equation that only has the basic model
+- [ ] check figure order
+- [ ] check all remaining in-text todo items
 -->
@@
 \noindent
-Alternative 2: Ensembles of data-limited stock-assessments can improve accuracy
+Alternative 2: Ensembles of data-limited stock assessments improve accuracy
 and reduce bias of $B/B_\mathrm{MSY}$ estimates
+
+\noindent Running title: Superensembles of population status
 
@@
 
-\noindent Running title: Superensembles of population status
-
 \clearpage
@@
 and boosted regression trees performed the best for estimating population
-status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order
-correlation between predicted and true status improved from `r
-mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range` across models?, and
-bias (median proportional error) declined from `r mean_sim$ind_mre_range` to
+status: accuracy improved
+`r mean_sim$mach_vs_ind_mare_fold`%, rank-order
+correlation between predicted and true status improved from
+`r mean_sim$ind_corr_range` to
+`r mean_sim$mach_corr_range` across models, and
+bias (median proportional error) declined from
+`r mean_sim$ind_mre_range` to
 `r mean_sim$mach_mre_range`. We found similar improvements when predicting the
@@
 estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015].
-For example, one models may suggest a population is at risk and declining in
+For example, one model may suggest a population is at risk and declining in
 abundance while others may suggest it is not at risk and stable.
@@
 [@dietterich2000; @araujo2007]. Ensembles are superior to individual models
-for at least three reasons: (1) statistically by averaging across models and
+in at least three ways: (1) statistically by averaging across models and
 therefore being less likely to pick the "wrong" model, (2) computationally by
@@
 Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$
-based on the limited data available for the majority of fish stocks: (1)
-a time series of the total biomass of catch and (2) a basic understanding of
+based on the limited data available for the majority of fish stocks: (1) a time
+series of the total biomass of catch and (2) a basic understanding of
 population productivity [e.g. @vasconcellos2005; @martell2013]. Recently,
-@rosenberg2014 investigated the performance of four data-limited models
-through a large-scale simulation experiment. Three of these models were based
-on Schaefer (logistic) biomass dynamics and one was an empirical model fitted
-to more data-rich stock assessment output. The four models frequently
-disagreed about population status, no one model had strong performance across
-all fish stocks, and some models performed better than others depending on
-circumstances (e.g. Fig. \ref{motivate}).
+@rosenberg2014 investigated the performance of four data-limited models through
+a large-scale simulation experiment. Three of these models were based on
+Schaefer (logistic) biomass dynamics and one was an empirical model fitted to
+more data-rich stock-assessment output. The four models frequently disagreed
+about population status (e.g. Fig. \ref{motivate}), no one model had strong
+performance across all fish stocks, and some models performed better than
+others depending on circumstances.
 
@@
 is available in an accompanying package `datalimited` for the statistical
-software R <https://github.com/datalimited/datalimited> (*Private while in
+software R [@r2015] (<https://github.com/datalimited/datalimited>) (*Private while in
 review*). In summary:
@@
 models were built in the `FLR` packages [@kell2007,
-<http://flr-project.org>] for the statistical software R [@r2015]. Code to
+<http://flr-project.org>] for the statistical software R. Code to
 generate the simulations is available at
@@
 scale.
+
+TODO: standardize notation of the following with figure 2
 
@@
 
+TODO add short description of spectral densities
+
 ## Applying the superensemble models and testing performance
@@
 trained on the entire simulation dataset. However, since mPRM is built on the
-same stock assessment database, we applied three-fold cross-validation to the
+same stock-assessment database, we applied three-fold cross-validation to the
 data underlying the mPRM model so that the dataset with which mPRM was trained
@@
 estimates from these data-rich stock assessments as known without error. Thus,
-any error in the stock assessment estimates of the mean or slope of
+any error in the stock-assessment estimates of the mean or slope of
 $B/B_\mathrm{MSY}$ also contributes to our estimates of prediction error for
 each of the four data-limited models and the ensembles. Code to reproduce our
-analysis and an R package to fit the data-limited assessment models are
-available at <https://github.com/datalimited/ensembles> (*Private while in
+analysis is available at <https://github.com/datalimited/ensembles> (*Private while in
 review*).

### 2016-05-02 text/ms.Rmd
Minor edits to the manuscript

@@
 
-We fit four individual data-limited models that use data on catch and basic
+We fit four individual data-limited models that use catch data and basic
 life-history parameters to estimate $B/B_\mathrm{MSY}$. We chose these models
-because because they can be fit to the vast majority of fisheries around the
+because they can be fit to the vast majority of fisheries around the
 world, are established in the literature, and have been extensively simulation
@@
 predictions. Additional covariates could be, for example, life-history
-characteristics, information on exploitation patterns, or statistical properties
-of the data. For simplicity, and since many possible covariates, such as fishing
-effort dynamics, would not be known for many real-world data-limited scenarios,
-we added only one set of additional covariates: spectral properties of the catch
-time series. Spectral analysis decomposes a time series into the frequency
-domain and provides a means of describing the cyclical shape of the catch series
-that is independent of time series length (except in affecting precision) and
+characteristics, information on exploitation patterns, or statistical
+properties of the data. We tested the performance benefits of including one set
+of additional covariates: spectral properties of the catch time series.
+Spectral analysis decomposes a time series into the frequency domain and
+provides a means of describing the cyclical shape of the catch series that is
+independent of time series length (except in affecting precision) and
 independent of absolute magnitude of catch. We fit spectral models to the
@@
 superensemble models. In this paper we applied the superensemble models to
-subsets of the simulated data as a cross validation test and to the RAM Legacy
+subsets of the simulated data as a cross-validation test and to the RAM Legacy
 Stock Assessment database to test on real stocks.
@@
 fit superensemble models to subsets of the individual models that did converge,
-but for the purposes of this paper we will only use superensemble models that
-use all four individual models.
+but for simplicity we will only use superensemble models fitted to all four
+individual models.
 
@@
 be expected to perform well in different conditions and an ensemble model can
-exploit the best predictive performance of each superensembles? Second,
+exploit the best predictive performance of each superensembles. Second,
 ensemble models will be most effective when they are not overfit to the
@@
 information theoretics such as the Akaike Information Criterion (AIC) is
-a common analytical approach in ecology [@burnham2002; @johnson2004;
+a common analytical approach in fisheries and ecology [@burnham2002; @johnson2004;
 @grueber2011]. The ensemble methods described in this paper share similarities
@@
 *coefficients* whereas ensembles instead average *predictions*. Thus, ensembles
-provide a more general purpose tool: they do not require information theoretics
+provide a general purpose tool: they do not require information theoretics
 and they can combine different types of models (e.g. parametric and

### 2016-05-16 text/ms.Rmd
Copy editing

@@
 and reduce bias of $B/B_\mathrm{MSY}$ estimates
-
-\noindent Running title: Superensembles of population status
 
@@
 
+\bigskip
+
+\noindent Running title: Superensembles of population status
+
 \clearpage
@@
 
-<!-- \tableofcontents -->
+\tableofcontents
 
@@
 Status and trend are two of the most fundamental values to quantify in the
-management of ecological populations [e.g. @iucn2015]. However, managers are
-often faced with reconciling multiple uncertain and potentially conflicting
-estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015].
-For example, one model may suggest a population is at risk and declining in
-abundance while others may suggest it is not at risk and stable.
+management of ecological populations [e.g. @hutchings2010; @iucn2015]. However,
+managers are often faced with reconciling multiple uncertain and potentially
+conflicting estimates of status and trend [e.g. @brodziak2010; @branch2011;
+@deroba2015]. For example, one model may suggest a population is at risk and
+declining in abundance while others may suggest it is not at risk and stable.
 
@@
 conservation concern and of economic interest to fisheries [@fao2014].
-
 Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$
@@
 # Methods
-
-<!--
-Structure of methods: summary of methods
- 1. indiviudla models
- 2. ensemble models
- 3. additional covariates
- 4. data used to test and cross validation,
- 5. testing model perforance. -->
 
@@
 single-species fishery, (3) a scenario where exploitation rate increased
-continuously, and (4) a 'roller-coaster' scenario where the exploitation
+continuously, and (4) a "roller-coaster" scenario where the exploitation
 rate increased and then decreased. Process noise (recruitment variability;
@@
 years. Together, these quantities address the recent state and trend of stock
-status, which are both of management and conservation interest [e.g.
-@mace2008] TODO add fishery REF. To avoid undue influence of the end points of
-the time series on the calculated slope, we measured the slope as the
-Theil-Sen estimator of median slope [@theil1950].
+status, which are both of management and conservation interest [e.g. @hutchings2010; @iucn2015]. To avoid undue influence of the time series end points
+on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
 
@@
 estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
-represented as $\hat{b}$ for models $i$ $1$ through $n$. The subscript $j$
-indexes the simulated fisheries from $1$ through $m$. The ensemble average was
-calculated as:
+represented as $\hat{b}$ for models $i$ $1$ through $4$ (CMSY, SSCOM, COM-SIR, mPRM).
+The ensemble average for each fishery $i$ was calculated as:
 
 \begin{equation}
-\hat{\theta_j} = \frac{1}{n} \sum_{i=1}^n \hat{b}_{i,j} , \quad
-\mathrm{for}\ j  =  1,...,m.
+\hat{\theta_i} = \left( \hat{b}_{i,1} + \hat{b}_{i,2} + \hat{b}_{i,3} + \hat{b}_{i,4} \right) / 4, \quad
+\mathrm{for}\ i  =  1,...,n.
 \end{equation}
@@
 \begin{equation}
-\hat{\theta_j} = \beta_0 + \beta_1 \hat{b}_{j,0} + ... +
-\beta_{1,2} \hat{b}_{j,1} \hat{b}_{j,2} + ... +
-\epsilon_j,
+\hat{\theta_i} = \beta_0 + \beta_1 \hat{b}_{i,1} + ... +
+\beta_{1,2} \hat{b}_{i,1} \hat{b}_{i,2} + ... +
+\epsilon_i,
 \quad
 \epsilon \sim \mathrm{Normal}(0, \sigma^2),
-\quad \mathrm{for}\ j  =  1,...,m.
+\quad \mathrm{for}\ i  =  1,...,n.
 \end{equation}
@@
 trees sequentially determine what value of a predictor best splits the
-response data into two 'branches' based on a loss function [@breiman1984]. In
+response data into two branches based on a loss function [@breiman1984]. In
 random forests, a series of regression trees are built on a random subset of
@@
 ## Additional covariates
-
-<!--TODO Kristin and Mark: It might be a good idea to add a figure that
-illustrates the spectrum of ensemble/superensembles that are applied here and
-what is included (e.g., non-linear relationships, etc.). This will help the
-reader understand the progression in complexity of the ensemble approaches
-I think. Jess: probably add this below the section explaining superensembles-->
-<!-- TODO deal with removed spectral -->
 
@@
 measure proportional error, defined as $(\hat{\theta} - \theta)/\abs{\theta}$,
-where $\hat{\theta}$ and $\theta$ represent estimated and 'true' (or
+where $\hat{\theta}$ and $\theta$ represent estimated and "true" (or
 stock-assessed) mean or slope of $B/B_\mathrm{MSY}$. We calculated median
@@
 measure accuracy, and Spearman's rank-order correlation between predicted and
-'true' values to measure the ability to correctly rank populations. When
+"true" values to measure the ability to correctly rank populations. When
 testing with the RAM Legacy Stock Assessment database, we treated the
@@
 approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high
-clusters of predictions (Fig. \ref{partial-sim}). Superensembles also exploited
+clusters of predictions (Fig. \ref{partial-sim}). SSCOM contributed most strongly on its own to determining trend (Fig. \ref{partial-sim-slope}). Superensembles also exploited
 the covariance between individual model predictions. For instance, both the
@@
 average to machine learning superensembles). Our results suggest choosing a
-superensemble model that allows for non-parametric relationships, such as
+superensemble model that allows for non-linear relationships, such as
 machine learning methods. These methods provide added insight into individual
 model behavior and generally performed the best; however, even a simple average
-of predictions across multiple models may provide a more useful measure of the
-desired parameter than a single model in many cases.
+of predictions across multiple models may provide a more useful metric than a single model in many cases.
 
@@
 individual models that choose different structural model forms, explore
-contrasting but plausible ranges of parameter values, or make uncorrelated
-errors [@ali1996; @dietterich2000; @tebaldi2007]. Such individual models would
-be expected to perform well in different conditions and an ensemble model can
-exploit the best predictive performance of each superensembles. Second,
+contrasting but plausible ranges of parameter values, and make uncorrelated
+errors [@ali1996; @dietterich2000; @tebaldi2007]. We would expect such models
+to perform well in different conditions and an ensemble model can
+exploit the best predictive performance of each. Second,
 ensemble models will be most effective when they are not overfit to the
@@
 well as complex machine learning models when we trained our superensembles on
-the simulation dataset and tested them on a separate 'real' dataset (i.e. the
+the simulation dataset and tested them on a separate "real" dataset (i.e. the
 RAM Legacy Stock Assessment database, Fig. \ref{performance}b). Third, ensemble
@@
 information theoretics such as the Akaike Information Criterion (AIC) is
-a common analytical approach in fisheries and ecology [@burnham2002; @johnson2004;
+a common analytical approach in fisheries and ecology [e.g. @burnham2002; @johnson2004;
 @grueber2011]. The ensemble methods described in this paper share similarities
@@
 could further include a loss function tailored to the goals of the model, say
-placing greater weight on accuracy at low status levels than higher values.
+placing greater weight on accuracy at lower rather than higher status levels.
 Conversely, because superensembles are tailored to a specific response and loss
@@
 understanding can be gained by studying its structure. For example, when SSCOM
-estimates very low status this is likely the case, conversely when COMSIR
+estimates low status this is likely the case, conversely when COMSIR
 estimates low status, the true status is more likely to be high (Fig.
@@
 suitable representation of effort dynamics at low biomasses in SSCOM. We think
-that such investigation of the parameters of a superensemble may lead to
+that such investigation of the structure of a superensemble may lead to
 improvement in the mechanisms assumed in the individual models.
 
-Combining predictions from multiple models via superensemble methods is also
-broadly useful in other subfields of ecology and fisheries science. Predictions
-about extinction risk are widely used at the national (e.g. the US Endangered
+Combining predictions from multiple models via superensemble methods is
+broadly useful in other subfields of fisheries science and ecology in general.
+Predictions about extinction risk are widely used at the national (e.g. the US Endangered
 Species Act and the Canadian Species at Risk Act) and international [e.g. the
@@
 fitting regression models to outcomes for individual species along with
-predictors of extinction risk [e.g. @anderson2011a; @pinsky2011], or population
-dynamics models to data for individual species [e.g. @dfo2010]. Both types of
+predictors of extinction risk [e.g. @anderson2011a; @pinsky2011], or fitting population-dynamic models to data for individual species [e.g. @dfo2010]. Both types of
 models are prone to error caused by model-misspecification and therefore results
@@
 
-We thank members of Phase I of the working group 'Developing new approaches to
-global stock status assessment and fishery production potential of the seas'
+We thank members of Phase I of the working group "Developing new approaches to
+global stock status assessment and fishery production potential of the seas"
 who contributed to developing the data-limited methods and simulations used in
@@
 on an earlier version of the manuscript. We thank the Gordon and Betty Moore
-Foundation for funding the working group 'Applying data-limited stock status
-models and developing management guidance for unassessed fish stocks'.
+Foundation for funding the working group "Applying data-limited stock status
+models and developing management guidance for unassessed fish stocks".
 

### 2016-05-16 text/ms.Rmd
Copy editing and cleaning

@@
 status and trend of exploited fish populations are stock assessments, i.e.,
-population models coupled to an observation model, that incorporate all
+population models coupled to an observation model that incorporate all
 appropriate data (e.g. catches, size and age distributions, surveys, and
@@
 Here, we estimate population status and trend of exploited fish populations
-using ensembles and superensembles (collectively 'ensemble methods') of these
+using ensembles and superensembles (collectively "ensemble methods") of these
 four data-limited models. We apply four ensemble-method approaches of varying
@@
 
-<!--TODO Add in the reference to (Fig. \ref{didactic}) and reorder the figures and add
-in the figure with the partial dependence -->
+TODO ADD OVERVIEW HERE.
 
@@
 
-The individual models we seek to combine with superensembles provide time
-series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use
-superensembles to estimate any property of these time series. Here, we focus
-on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last five
-years. Together, these quantities address the recent state and trend of stock
-status, which are both of management and conservation interest [e.g. @hutchings2010; @iucn2015]. To avoid undue influence of the time series end points
-on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].
+The individual models we seek to combine with superensembles provide time series
+of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use superensembles to
+estimate any property of these time series. Here, we focus on two properties:
+the mean and slope of $B/B_\mathrm{MSY}$ in the last five years. Together, these
+quantities address the recent state and trend of stock status, which are both of
+management and conservation interest [e.g. @hutchings2010; @iucn2015]. To avoid
+undue influence of the time series end points on the calculated slope, we
+measured the slope as the Theil-Sen estimator of median slope [@theil1950].
 
-We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and
-the predictions from the individual models as predictors in our superensemble
-models. When modelling mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero ---
-we fit the superensemble models in log space and exponentiated the
-predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, which are not
-bounded at zero, we fit superensemble models on the natural untransformed
-scale.
+We used the mean or slope of $B/B_\mathrm{MSY}$ as the response variable and the
+predictions from the individual models as predictors in our superensemble models
+(Fig. \ref{didactic}a). When modelling mean $B/B_\mathrm{MSY}$ --- a ratio
+bounded at zero --- we fit the superensemble models in log space and
+exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope,
+which are not bounded at zero, we fit superensemble models on the natural
+untransformed scale.
 
-TODO: standardize notation of the following with figure 2
-
-We compared an ensemble average and three superensembles of varying
-complexity: a linear model with two-way interactions, a random forest, and a
-boosted regression tree. We describe these models as estimating
-$\hat{\theta}$, which represents either the ensemble estimated $\log
-B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$. The individual model
-estimates of $\log B/B_\mathrm{MSY}$ or slope of $B/B_\mathrm{MSY}$ are
-represented as $\hat{b}$ for models $i$ $1$ through $4$ (CMSY, SSCOM, COM-SIR, mPRM).
-The ensemble average for each fishery $i$ was calculated as:
+We compared an ensemble average and three superensembles of varying complexity:
+a linear model with two-way interactions, a random forest, and a boosted
+regression tree. We describe these models as estimating $\hat{\theta}$, which
+represents either the ensemble estimated $\log B/B_\mathrm{MSY}$ or slope of
+$B/B_\mathrm{MSY}$. The individual model estimates of $\log B/B_\mathrm{MSY}$ or
+slope of $B/B_\mathrm{MSY}$ are represented as $\hat{b}$ for models $i$ $1$
+through $4$ (CMSY, COM-SIR, SSCOM, mPRM). The ensemble average for each fishery
+$i$ was calculated as:
 
@@
 
-Once the superensemble models are built, we can apply them to new stocks that
-we'd like to know the status of. To do this we applied the individual models
-to our stocks of interest and then applied these data to our already built
-superensemble models. In this paper we applied the superensemble models to
-subsets of the simulated data as a cross-validation test and to the RAM Legacy
-Stock Assessment database to test on real stocks.
+Once the superensemble models are built, we can apply them to new stocks to
+estimate their status (Fig. \ref{didactic}b). To do this we applied the
+individual models to our stocks of interest and then applied these data to our
+already built superensemble models. In this paper we applied the superensemble
+models to subsets of the simulated data as a cross-validation test and to the
+RAM Legacy Stock Assessment database to test on real stocks.
 
@@
 had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
-\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
-approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high
-clusters of predictions (Fig. \ref{partial-sim}). SSCOM contributed most strongly on its own to determining trend (Fig. \ref{partial-sim-slope}). Superensembles also exploited
-the covariance between individual model predictions. For instance, both the
-linear model and GBM ensemble suggest that if mPRM and SSCOM predict high
-status, the true status also tends to be high (Figs \ref{lm-coefs},
+\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were approximately
+linearly related to $B/B_\mathrm{MSY}$ within the low and high clusters of
+predictions (Fig. \ref{partial-sim}). SSCOM contributed most strongly on its own
+to determining trend (Fig. \ref{partial-sim-slope}). Superensembles also
+exploited the covariance between individual model predictions. For instance,
+both the linear model and GBM ensemble suggest that if mPRM and SSCOM predict
+high status, the true status also tends to be high (Figs \ref{lm-coefs},
 \ref{partial-2d-sim}l). The addition of spectral density covariates helped the
@@
 # Discussion
-
-<!-- Dan: Lastly, it's worth noting that this approach, as laid out here is for
-helping improve potentially conflicting estimates of the same metric, B/Bmsy. In
-many cases though a fishery is going to be basing its decisions on streams of
-multiple indicators, e.g trends in SPR and B/Bmsy. While the ensemble methods
-here can help refine estimates of individual metrics, it's still worth looking
-at what alternative data streams are telling you if possible. E.g. if a purely
-catch based B/Bmsy estimate is high and has a positive slope, but mean lengths
-and SPR are decreasing, managers will still need to think about how to deal with
-these conflicting signals. So, this ensemble method, as I interpret it, is
-helpful for resolving conflicts within metric, but doesn't necesarily resolve
-conflicts across metrics (if I'm understanding these methods right) -->
 
@@
 
-<!-- TODO put back when space: or to evaluate the effect of conservation
-management actions when experimental or observational data are limited
-[@walsh2012].-->
-
 \noindent
@@
 models and developing management guidance for unassessed fish stocks".
-
-<!--Funding...-->
-
-<!--# Citation notes-->
-
-<!--
-- @breiner2015 ensemble models of species distribution models for rare species
-
-- @jones2015 ensemble models of species distribution models - globally for
-  marine biodiversity
-
-- [@greene2006] example similar to ours but with climate models (Bayesian
-  multilevel ensemble)
-
-- [@kell2007] FLR
-
-- multiple learners book [@alpaydin2010]
-
-- [@caruana2004] nice paper on ensemble models; prob of overfitting increases
-  with more models, bagging important
-
-- [@tebaldi2007] key paper: review of 'multi-model ensembles for climate projections'
-
-- famous ensemble is DEMETER: Development of a European Multi-model Ensemble
-  System for Seasonal to Interannual Prediction @hagedorn2005 is a good
-  reference
-
-- Examples of where ensemble models are shown to be better than any one:
-  @thomson2006 (public health - malaria), @cantelaube2005 (agriculture - crop
-  yield) (citations taken from @tebaldi2007)
-
-- simple averages are used very commonly in climate science - e.g. IPCC 2001
-
-- @tebaldi2007: weighting obviously makes sense, but how do we define a
-  performance metric that is based on past observations that is relevant to the
-  future?
-
-- @tebaldi2007: model independence important
-
-- for climate, weighted averages perform better than simple averages [@min2006],
-  but are those same weights applicable to the future (or in our case other
-  fisheries)?
-
-Why do ensemble methods work?
-
-- error cancellation is one but not the only reason for superiority of ensemble
-  models [@hagedorn2005]
-
-- ensemble model may be only marginally better than the best single model in
-  any given case, but we don't usually know which is the best single model
-  [@hagedorn2005]
-
-- ensemble models useful for regional climate models too; as an example,
-  @pierce2009 use 42 metrics to characterize model performance in regional
-  climate model ensembles; found that ensemble models were superior to any one
-  model --- especially when considering multiple metrics
-
-- @knutti2009: key review paper on motivation and challenges of combining
-  climate projection models; performance on testing / current data may only
-  weakly relate to future / other datasets
-
-- @murphy2004: Nature paper on ensembles of climate model simulations; weighted
-  average better performance than unweighted average
-
-- @dietterich2000: highly cited book chapter "Ensemble Methods in Machine
-  Learning"; ensembles are often much more accurate than the "individual
-  classifiers that make them up"; but components must be diverse and accurate
-
-- @dietterich2000: a main justification for ensembles: they are
-  representational --- no one model usually can contain all hypothetical
-  functional forms, but many separate models can cover more hypotheses
-
-- strong paper showing that ensemble models are most accurate when the various
-  individual models make errors in uncorrelated ways [@ali1996]
-
-- without substantial training-testing data, appropriate model weights can be
-  very hard to deduce and can cause more harm than good (compared to equal
-  weighting) [@weigel2010] (they give the example of seasonal forecasting where
-  a ton of hind cast testing can be done, vs. long-term climate) (asymmetrical
-  loss function)
-
-- [@weigel2010] optimal weights are always as good or better than equal
-  weights, but if you get the weights wrong, you can be better off just using
-  equal weights; but averaging was almost always better than any one model
-
-- @rykiel1996 "Testing ecological models: the meaning of validation" (see for
-  performance criteria, theory on model testing and assessment)
-
-model averaging AIC:
-f Chamberlin 1890).  multiple woring hypotheses
-  (e.g., Johnson and Omland 2004, Hobbs and Hilborn 2006, Burnham et al. 2011,
-  52 Grueber et al. 2011).
--->
-
-<!-- vim: set formatoptions=nroq: -->

### 2016-05-16 text/ms.Rmd
Painfully shorten abstract

@@
 
-\noindent \textbf{Abstract:} Fishery managers must often reconcile conflicting
-estimates of population status and trend. One solution is to average
-predictions using an ensemble method, but this approach ignores covariance
-between models and may not optimally reduce bias and improve accuracy.
-Superensemble models, commonly used in climate and weather forecasting, may
-provide a superior solution. Superensembles use predictions from multiple
-models as covariates in an additional model. By training this superensemble
-model using known data, we can then use it to predict unknown situations. We
-evaluated the potential for ensemble and superensemble models ('ensemble
-methods') to improve estimates of population status and trend for fisheries.
-We fit four widely applicable data-limited models that estimate stock biomass
-relative to the equilibrium biomass at maximum sustainable yield
-($B/B_\mathrm{MSY}$). We combined the models' estimates to build four ensemble
-methods: an ensemble average and three superensembles (a linear model,
-a random forest, and a boosted regression tree). We used the ensemble methods
-to predict both (i) fishery status and (ii) recent trends of
-$B/B_\mathrm{MSY}$. We trained our models on a simulated dataset of 5760
-stocks and tested them with cross-validation and against a global database of
-`r ram_stocks_n` real stock assessments. We found that ensemble methods
-substantially improved estimates of population status and trend. Random forest
-and boosted regression trees performed the best for estimating population
-status: accuracy improved
-`r mean_sim$mach_vs_ind_mare_fold`%, rank-order
-correlation between predicted and true status improved from
-`r mean_sim$ind_corr_range` to
-`r mean_sim$mach_corr_range` across models, and
-bias (median proportional error) declined from
-`r mean_sim$ind_mre_range` to
-`r mean_sim$mach_mre_range`. We found similar improvements when predicting the
-trend of status and when applying the simulation-trained superensembles to
-catch data for global fish stocks. Ensemble methods can improve estimates of
-status and trends; however, they must be tested, formed from a diverse set of
-accurate models, and built on a dataset representative of the populations to
-which they are applied. *TODO: Still need to cut ~30 words*
+\noindent \textbf{Abstract:}
+Fishery managers must often reconcile conflicting estimates of population status and trend.
+Superensemble models, commonly used in climate and weather forecasting, may provide an effective solution.
+This approach uses predictions from multiple models as covariates in an additional "superensemble" model fitted to known data.
+We evaluated the potential for ensemble averages and superensemble models ("ensemble methods") to improve estimates of population status and trend for fisheries.
+We fit four widely applicable data-limited models that estimate stock biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$).
+We combined estimates of recent fishery status and trends in $B/B_\mathrm{MSY}$ with four ensemble methods: an ensemble average and three superensembles (a linear model, random forest, and boosted regression tree).
+We trained our superensembles on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments.
+Ensemble methods substantially improved estimates of population status and trend.
+Random forest and boosted regression trees performed the best at estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements when predicting trend and when applying the simulation-trained superensembles to catch data for global fish stocks.
+Ensemble methods can improve estimates of status and trends; however, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
 
@@
 
-\noindent Keywords: data-limited fisheries, ensemble methods, multi-model
+\noindent Keywords: CMSY, data-limited fisheries, ensemble methods, multi-model
 averaging, population dynamics, sustainable resource management

### 2016-05-17 text/ms.Rmd
Add intro paragraph to methods

@@
 - [x] rerun the main simulation analysis because the equations of change for the slope
-- [ ] manually adjusted the labels in figure 4
+- [x] manually adjusted the labels in figure 4
 - [x] update the magic number get inserted to use the equation that only has the basic model
-- [ ] check figure order
+- [x] check figure order
 - [ ] check all remaining in-text todo items
@@
 
-TODO ADD OVERVIEW HERE.
+To test the ability of superensembles to improve estimates of status and trend in data-limited fish stocks, we first fit four individual assessment models to a large simulated dataset of fish stocks.
+We then built and tested the performance of superensembles using cross-validation of the simulated dataset.
+Finally, we tested superensembles built with the entire simulated dataset against a database of global fish sticks.
+We describe these steps in detail below and illustrate the general approach both with illustrations and pseudocode in Fig.
+2.
 

### 2016-05-19 text/ms.Rmd
Integrate comments from Jess and Kristin

@@
 
-\noindent \textbf{Abstract:}
-Fishery managers must often reconcile conflicting estimates of population status and trend.
-Superensemble models, commonly used in climate and weather forecasting, may provide an effective solution.
-This approach uses predictions from multiple models as covariates in an additional "superensemble" model fitted to known data.
-We evaluated the potential for ensemble averages and superensemble models ("ensemble methods") to improve estimates of population status and trend for fisheries.
-We fit four widely applicable data-limited models that estimate stock biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$).
-We combined estimates of recent fishery status and trends in $B/B_\mathrm{MSY}$ with four ensemble methods: an ensemble average and three superensembles (a linear model, random forest, and boosted regression tree).
-We trained our superensembles on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments.
-Ensemble methods substantially improved estimates of population status and trend.
-Random forest and boosted regression trees performed the best at estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
-We found similar improvements when predicting trend and when applying the simulation-trained superensembles to catch data for global fish stocks.
-Ensemble methods can improve estimates of status and trends; however, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.
+\noindent \textbf{Abstract:} Fishery managers must often reconcile conflicting
+estimates of population status and trend. Superensemble models, commonly used in
+climate and weather forecasting, may provide an effective solution. This
+approach uses predictions from multiple models as covariates in an additional
+"superensemble" model fitted to known data. We evaluated the potential for
+ensemble averages and superensemble models ("ensemble methods") to improve
+estimates of population status and trend for fisheries. We fit four widely
+applicable data-limited models that estimate stock biomass relative to the
+equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$). We
+combined these estimates of recent fishery status and trends in $B/B_\mathrm{MSY}$
+with four ensemble methods: an ensemble average and three superensembles (a
+linear model, random forest, and boosted regression tree).  We trained our
+superensembles on a simulated dataset of 5760 stocks and tested them with
+cross-validation and against a global database of `r ram_stocks_n` stock
+assessments. Ensemble methods substantially improved estimates of population
+status and trend. Random forest and boosted regression trees performed the best
+at estimating population status: accuracy improved `r
+mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and
+true status improved from `r mean_sim$ind_corr_range` to `r
+mean_sim$mach_corr_range`, and bias (median proportional error) declined from
+`r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+We found similar improvements when predicting trend and when applying the
+simulation-trained superensembles to catch data for global fish stocks.
+Superensembles can optimally leverage multiple model predictions;
+however, they must be tested, formed from a diverse set of accurate models, and
+built on a dataset representative of the populations to which they are applied.
 
@@
 Status and trend are two of the most fundamental values to quantify in the
-management of ecological populations [e.g. @hutchings2010; @iucn2015]. However,
+management of ecological populations [e.g., @hutchings2010; @iucn2015]. However,
 managers are often faced with reconciling multiple uncertain and potentially
-conflicting estimates of status and trend [e.g. @brodziak2010; @branch2011;
+conflicting estimates of status and trend [e.g., @brodziak2010; @branch2011;
 @deroba2015]. For example, one model may suggest a population is at risk and
@@
 One solution is to take the average or weighted average of several model
-predictions, i.e. an ensemble. Such ensembles are typically more accurate and
-less biased than individual model estimates and can integrate uncertainty in
+predictions, i.e., an ensemble. Such ensembles are typically more accurate and
+less biased than individual model estimates and can incorporate various types of  uncertainty, such as uncertainty in
 model structure, initial conditions, and parameter estimation
@@
 [@dietterich2000]. This approach forms the basis of many machine learning
-methods [e.g. @dietterich2000], has helped reconcile climate forecasts from
-dozens of models [e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even
+methods [e.g., @dietterich2000], has helped reconcile climate forecasts from
+dozens of models [e.g., @murphy2004; @tebaldi2007; @ipcc2013], and even
 improved early warning signs of malaria outbreaks [@thomson2006]. In ecology,
 ensemble methods are sometimes used to improve species distribution modelling
-[e.g. @araujo2007; @breiner2015] and indeed have been used to combine
-estimates of population status and trend [e.g. @brodziak2010].
+[e.g., @araujo2007; @breiner2015] and indeed have been used to combine
+estimates of population status and trend [e.g., @brodziak2010].
 
@@
 called superensemble modelling [@krishnamurti1999], is common in climate and
-weather forecasting [e.g. @yun2005; @mote2015]. The superensemble is fit to a
+weather forecasting [e.g., @yun2005; @mote2015]. The superensemble is fit to a
 training dataset where outcomes are well known and then used to predict on a
@@
 population models coupled to an observation model that incorporate all
-appropriate data (e.g. catches, size and age distributions, surveys, and
+appropriate data (e.g., catches, size and age distributions, surveys, and
 tagging information) to quantify values such as the biomass of a stock that
@@
 based on the limited data available for the majority of fish stocks: (1) a time
-series of the total biomass of catch and (2) a basic understanding of
-population productivity [e.g. @vasconcellos2005; @martell2013]. Recently,
+series of the total weight of catch and (2) a basic understanding of
+population productivity [e.g., @vasconcellos2005; @martell2013]. Recently,
 @rosenberg2014 investigated the performance of four data-limited models through
@@
 more data-rich stock-assessment output. The four models frequently disagreed
-about population status (e.g. Fig. \ref{motivate}), no one model had strong
+about population status (e.g., Fig. \ref{motivate}), no one model had strong
 performance across all fish stocks, and some models performed better than
@@
 
-To test the ability of superensembles to improve estimates of status and trend in data-limited fish stocks, we first fit four individual assessment models to a large simulated dataset of fish stocks.
-We then built and tested the performance of superensembles using cross-validation of the simulated dataset.
-Finally, we tested superensembles built with the entire simulated dataset against a database of global fish sticks.
-We describe these steps in detail below and illustrate the general approach both with illustrations and pseudocode in Fig.
-2.
+To test the ability of superensembles to improve estimates of status and trend
+in data-limited fish stocks, we first fit four individual assessment models to a
+large simulated dataset of fish stocks. We then built and tested the performance
+of superensembles using cross-validation of the simulated dataset. Finally, we
+tested superensembles built with the entire simulated dataset against a database
+of global fish stocks. We describe these steps in detail below and illustrate
+the general approach both with illustrations and pseudocode in Fig. 2.
 
@@
 * *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
-  biomass dynamics [@martell2013]. It requires a prior distribution for $r$
-  and a prior on the relative proportion of biomass at the end of the time
-  series compared to unfished biomass (depletion), which is derived from the
-  percentage of maximum catch at the end of the time series. The version of
+  biomass dynamics [@martell2013]. It requires a prior distributions on $r$ and $K$
+  as well as priors on the relative proportion of biomass at the beginning and end of the time
+  series compared to unfished biomass (depletion). The version of
   the model used in @rosenberg2014 was modified from @martell2013 to generate
@@
 rate increased and then decreased. Process noise (recruitment variability;
-i.e. unexplained variability in population dynamics) was introduced to the
+i.e., unexplained variability in population dynamics) was introduced to the
 models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$,
@@
 quantities address the recent state and trend of stock status, which are both of
-management and conservation interest [e.g. @hutchings2010; @iucn2015]. To avoid
+management and conservation interest [e.g., @hutchings2010; @iucn2015]. To avoid
 undue influence of the time series end points on the calculated slope, we
@@
 
-TODO add short description of spectral densities
-
 ## Applying the superensemble models and testing performance
 
-Once the superensemble models are built, we can apply them to new stocks to
-estimate their status (Fig. \ref{didactic}b). To do this we applied the
-individual models to our stocks of interest and then applied these data to our
+Once the superensemble models are built and trained using the simulated stocks
+(or any dataset with "known" status), we can use the superensembles to estimate
+the status of new stocks (Fig. \ref{didactic}b). To do this we applied the
+individual models to our stocks of interest (i.e., CMSY, COM-SIR, SSCOM, mPRM)
+and then used these individual model estimates of status or trend as data in our
 already built superensemble models. In this paper we applied the superensemble
-models to subsets of the simulated data as a cross-validation test and to the
-RAM Legacy Stock Assessment database to test on real stocks.
+models to subsets of the simulated data as a cross-validation test to test
+predictive performance and to the RAM Legacy Stock Assessment database to test
+predictive performance on real stocks.
 
-We used repeated three-fold cross validation to test predictive performance: we
-randomly divided the dataset into three sets, built superensemble models on
-two-thirds of the data, and evaluated predictive performance on the remaining
-third. We repeated this across each of the three splits and then repeated the
-whole procedure 50 times to account for bias that may result from any one set
-of validation splits. In the simulated dataset, there were 10 replicates of
-each unique combination of simulation parameters that differed only in
-stochastic variability. Since the dynamics of these populations were often
-similar, we grouped these stocks in the cross-validation process into either
-the training or testing split.
+We used repeated three-fold cross validation: we randomly divided the dataset
+into three sets, built superensemble models on two-thirds of the data, and
+evaluated predictive performance on the remaining third. We repeated this across
+each of the three splits and then repeated the whole procedure 50 times to
+account for bias that may result from any one set of validation splits. In the
+simulated dataset, there were 10 replicates of each unique combination of
+simulation parameters that differed only in stochastic variability. Since the
+dynamics of these populations were often similar, we grouped these stocks in the
+cross-validation process into either the training or testing split.
 
@@
 bias, precision, accuracy (a combination of bias and precision), or the ability
-to correctly rank or correlate across populations [e.g. @walther2005]. Here, we
+to correctly rank or correlate across populations [e.g., @walther2005]. Here, we
 measure proportional error, defined as $(\hat{\theta} - \theta)/\abs{\theta}$,
@@
 Applied to the simulated dataset of known stock status, the individual models
-had variable success at recovering the mean (status) and slope (trend) of
+had variable success at estimating the mean (status) and slope (trend) of
 $B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of
@@
 superensemble model that allows for non-linear relationships, such as
-machine learning methods. These methods provide added insight into individual
-model behavior and generally performed the best; however, even a simple average
-of predictions across multiple models may provide a more useful metric than a single model in many cases.
+machine learning methods; these models provided added insight into individual
+model behavior and generally performed the best.
 
@@
 well as complex machine learning models when we trained our superensembles on
-the simulation dataset and tested them on a separate "real" dataset (i.e. the
+the simulation dataset and tested them on a separate "real" dataset (i.e., the
 RAM Legacy Stock Assessment database, Fig. \ref{performance}b). Third, ensemble
@@
 information theoretics such as the Akaike Information Criterion (AIC) is
-a common analytical approach in fisheries and ecology [e.g. @burnham2002; @johnson2004;
+a common analytical approach in fisheries and ecology [e.g., @burnham2002; @johnson2004;
 @grueber2011]. The ensemble methods described in this paper share similarities
@@
 provide a general purpose tool: they do not require information theoretics
-and they can combine different types of models (e.g. parametric and
+and they can combine different types of models (e.g., parametric and
 non-parametric models or frequentist and Bayesian predictions). Furthermore,
@@
 \ref{partial-sim-slope}). Formally, fitting superensemble models to specific
-quantities of interest provides an additional calibration step [@rykiel1996].
-Individual models are typically calibrated (i.e. parameters are estimated) based
-on a single response variable; however, this model might then provide a biased
-or inaccurate estimate of other quantities of interest. Ensemble methods provide
-a final calibration step to a quantity of interest. This ensemble calibration
-could further include a loss function tailored to the goals of the model, say
-placing greater weight on accuracy at lower rather than higher status levels.
-Conversely, because superensembles are tailored to a specific response and loss
-function, superensembles force a modeller to choose an operational purpose for
-their model up front [*sensu* @dickey-collas2014].
+quantities of interest (such as the slope of $B/B_\mathrm{MSY}$) provides an additional
+calibration step [@rykiel1996]. Individual models are typically calibrated
+(i.e., parameters are estimated) based on a single response variable; however,
+this model might then provide a biased or inaccurate estimate of other
+quantities of interest. Ensemble methods provide a final calibration step to a
+quantity of interest. This ensemble calibration could include a loss function
+tailored to the goals of the model, say placing greater weight on accuracy at
+lower rather than higher status levels. Conversely, because superensembles are
+tailored to a specific response and loss function, superensembles force a
+modeller to choose an operational purpose for their model up front [*sensu*
+@dickey-collas2014].
 
@@
 broadly useful in other subfields of fisheries science and ecology in general.
-Predictions about extinction risk are widely used at the national (e.g. the US Endangered
-Species Act and the Canadian Species at Risk Act) and international [e.g. the
+Predictions about extinction risk are widely used at the national (e.g., the US Endangered
+Species Act and the Canadian Species at Risk Act) and international [e.g., the
 IUCN Red List, @iucn2015] levels. These risk assessments generally involve
 fitting regression models to outcomes for individual species along with
-predictors of extinction risk [e.g. @anderson2011a; @pinsky2011], or fitting population-dynamic models to data for individual species [e.g. @dfo2010]. Both types of
+predictors of extinction risk [e.g., @anderson2011a; @pinsky2011], or fitting population-dynamic models to data for individual species [e.g., @dfo2010]. Both types of
 models are prone to error caused by model-misspecification and therefore results
@@
 are options to account for potential model-misspecification in determination of
-species risk [e.g. coefficient averaging, @burnham2002; generalized modeling,
+species risk [e.g., coefficient averaging, @burnham2002; generalized modeling,
 @yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are

### 2016-05-19 analysis/9-output-values.R
Correct the denominator in percent change

@@ -18,8 +18,8 @@ get_performance_stats <- function(dat, digits_fold = 1, digits_raw = 2,
   ind <- dat %>% filter(ensemble == FALSE)
 
   x$mach_vs_ind_mare_fold <- c(
-    round((min(ind$mare) - max(mach$mare)) / max(mach$mare), digits_fold),
-    round((max(ind$mare) - min(mach$mare)) / min(mach$mare), digits_fold)
+    round((min(ind$mare) - max(mach$mare)) / max(ind$mare), digits_fold),
+    round((max(ind$mare) - min(mach$mare)) / min(ind$mare), digits_fold)
   ) * 100
 
 #   x$mach_vs_ind_bias <- c(

### 2016-09-27 text/ms.Rmd
Correct the denominator in percent change
All edits before first submission to Fish and Fisheries

@@
 load("values.rda") # generated by `../analysis/9-output-values.R`
-
+if (mean_sim_basic$mach_mare_range == "0.32--0.32") mean_sim_basic$mach_mare_range <- 0.32
 # Includes: mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n
@@
 linear model, random forest, and boosted regression tree).  We trained our
-superensembles on a simulated dataset of 5760 stocks and tested them with
+superensembles on 5760 simulated stocks and tested them with
 cross-validation and against a global database of `r ram_stocks_n` stock
@@
 status and trend. Random forest and boosted regression trees performed the best
-at estimating population status: accuracy improved `r
-mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and
-true status improved from `r mean_sim$ind_corr_range` to `r
-mean_sim$mach_corr_range`, and bias (median proportional error) declined from
-`r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`.
+at estimating population status: inaccuracy (median absolute proportional error) decreased from
+`r mean_sim_basic$ind_mare_range` to `r mean_sim_basic$mach_mare_range`,
+rank-order correlation between predicted and
+true status improved from `r mean_sim_basic$ind_corr_range` to
+`r mean_sim_basic$mach_corr_range`, and bias (median proportional error) declined from
+`r mean_sim_basic$ind_mre_range` to `r mean_sim_basic$mach_mre_range`.
 We found similar improvements when predicting trend and when applying the
@@
 predictions, i.e., an ensemble. Such ensembles are typically more accurate and
-less biased than individual model estimates and can incorporate various types of  uncertainty, such as uncertainty in
-model structure, initial conditions, and parameter estimation
-[@dietterich2000; @araujo2007]. Ensembles are superior to individual models
-in at least three ways: (1) statistically by averaging across models and
-therefore being less likely to pick the "wrong" model, (2) computationally by
-reducing the risk of getting stuck in a local optima, and (3)
+less biased than individual model estimates and can incorporate various types of
+uncertainty, such as uncertainty in model structure, initial conditions, and
+parameter estimation [@dietterich2000; @araujo2007]. Ensembles are superior to
+individual models in at least three ways: (1) statistically by averaging across
+models and therefore being less likely to pick the "wrong" model, (2)
+computationally by reducing the risk of getting stuck in a local optima, and (3)
 representationally by expanding the range of hypotheses explored
@@
 methods [e.g., @dietterich2000], has helped reconcile climate forecasts from
-dozens of models [e.g., @murphy2004; @tebaldi2007; @ipcc2013], and even
-improved early warning signs of malaria outbreaks [@thomson2006]. In ecology,
-ensemble methods are sometimes used to improve species distribution modelling
-[e.g., @araujo2007; @breiner2015] and indeed have been used to combine
-estimates of population status and trend [e.g., @brodziak2010].
+dozens of models [e.g., @murphy2004; @tebaldi2007; @ipcc2013], and even improved
+early warning signs of malaria outbreaks [@thomson2006]. In ecology, ensemble
+methods are sometimes used to improve species distribution modelling [e.g.,
+@araujo2007; @breiner2015] and indeed have been used to combine estimates of
+population status and trend [e.g., @brodziak2010].
 
@@
 
-* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer
-  biomass dynamics [@martell2013]. It requires a prior distributions on $r$ and $K$
-  as well as priors on the relative proportion of biomass at the beginning and end of the time
-  series compared to unfished biomass (depletion). The version of
+* *CMSY* (catch-MSY) implements a stock-reduction analysis with Schaefer biomass
+  dynamics [@martell2013]. It requires a prior distributions on $r$ and $K$ as
+  well as priors on the relative proportion of biomass at the beginning and end
+  of the time series compared to unfished biomass (depletion). The version of
   the model used in @rosenberg2014 was modified from @martell2013 to generate
@@
 any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h).
-Compared to the individual models, machine learning superensembles improved
-accuracy (median absolute proportional error) by
-`r mean_sim_basic$mach_vs_ind_mare_fold`%, increased rank-order correlation from
+Compared to the individual models, machine learning superensembles decreased
+inaccuracy (median absolute proportional error) from
+`r mean_sim_basic$ind_mare_range` to `r mean_sim_basic$mach_mare_range`,
+increased rank-order correlation from
 `r mean_sim_basic$ind_corr_range` to `r mean_sim_basic$mach_corr_range`, and reduced bias
@@
 Results were similar when predicting trend: compared to individual models,
-machine learning superensembles improved accuracy by
-`r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from
+machine learning superensembles decreased inaccuracy from
+`r slope_sim$ind_mare_range` to `r slope_sim$mach_mare_range`,
+increased rank-order correlation from
 `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced
@@
 machine learning methods; these models provided added insight into individual
-model behavior and generally performed the best.
+model behaviour and generally performed the best.
 
@@
 
-Combining predictions from multiple models via superensemble methods is
-broadly useful in other subfields of fisheries science and ecology in general.
-Predictions about extinction risk are widely used at the national (e.g., the US Endangered
-Species Act and the Canadian Species at Risk Act) and international [e.g., the
-IUCN Red List, @iucn2015] levels. These risk assessments generally involve
-fitting regression models to outcomes for individual species along with
-predictors of extinction risk [e.g., @anderson2011a; @pinsky2011], or fitting population-dynamic models to data for individual species [e.g., @dfo2010]. Both types of
-models are prone to error caused by model-misspecification and therefore results
-are sensitive to decisions about model structure [@brooks2015]. Although there
-are options to account for potential model-misspecification in determination of
-species risk [e.g., coefficient averaging, @burnham2002; generalized modeling,
-@yeakel2011; or semi-parametric methods, @thorson2014], ensemble methods are
-a relatively simple way to combine predictions in a transparent manner. Beyond
-estimates of status and trend, ensemble methods could be used, for example, to
-increase the robustness of spatial predictions when designing networks of
-protected areas [@rassweiler2014] or to forecast potential spatial shifts in
-species distribution given climate impacts [@harsch2014].
+Combining predictions from multiple models via superensemble methods is broadly
+useful in other subfields of fisheries science and ecology in general.
+Predictions about extinction risk are widely used at the national (e.g., the US
+Endangered Species Act and the Canadian Species at Risk Act) and international
+[e.g., the IUCN Red List, @iucn2015] levels. These risk assessments generally
+involve fitting regression models to outcomes for individual species along with
+predictors of extinction risk [e.g., @anderson2011a; @pinsky2011], or fitting
+population-dynamic models to data for individual species [e.g., @dfo2010]. Both
+types of models are prone to error caused by model-misspecification and
+therefore results are sensitive to decisions about model structure
+[@brooks2015]. Although there are options to account for potential
+model-misspecification in determination of species risk [e.g., coefficient
+averaging, @burnham2002; generalized modeling, @yeakel2011; or semi-parametric
+methods, @thorson2014], ensemble methods are a relatively simple way to combine
+predictions in a transparent manner. Beyond estimates of status and trend,
+ensemble methods could be used, for example, to increase the robustness of
+spatial predictions when designing networks of protected areas
+[@rassweiler2014] or to forecast potential spatial shifts in species
+distribution given climate impacts [@harsch2014].
 
@@
 
-# Acknowledgments
+# Acknowledgements
 
@@
 comments during the development of this project, and R.D. Methot for comments
-on an earlier version of the manuscript. We thank the Gordon and Betty Moore
-Foundation for funding the working group "Applying data-limited stock status
-models and developing management guidance for unassessed fish stocks".
+on an earlier version of the manuscript. A.A. Rosenberg is president of MRAG
+Americas Inc, a consulting company specializing in fisheries science. We thank
+the Gordon and Betty Moore Foundation for funding the working group "Applying
+data-limited stock status models and developing management guidance for
+unassessed fish stocks".

### 2016-09-28 text/ms.Rmd
Edits dealing with reviewer comments

@@
 load("values.rda") # generated by `../analysis/9-output-values.R`
-if (mean_sim_basic$mach_mare_range == "0.32--0.32") mean_sim_basic$mach_mare_range <- 0.32
+# if (mean_sim_basic$mach_mare_range == "0.32--0.32") mean_sim_basic$mach_mare_range <- 0.32
 # Includes: mean_sim, mean_ram, slope_sim, auc_sim, ram_stocks_n
@@
 based on the RAM Legacy Stock Assessment Database [@ricard2012]. @rosenberg2014
-provide a full background on these four methods and code to fit all the models
+provide a full background on these four methods \R{C} 
+(<http://www.fao.org/docrep/019/i3491e/i3491e00.htm>, last accessed 2016-09-27) 
+and code to fit all the models
 is available in an accompanying package `datalimited` for the statistical
-software R [@r2015] (<https://github.com/datalimited/datalimited>) (*Private while in
-review*). In summary:
+software R [@r2015] <https://github.com/datalimited/datalimited> 
+(last accessed 2016-09-27). In summary:
 
@@
 We first developed and tested ensemble methods on a fully factorial simulated
-dataset of fisheries [@rosenberg2014] with known status. These simulated data
-included the following scenarios: three fish life histories: small pelagic,
-demersal, and large pelagic; three levels of initial biomass depletion
+dataset of fisheries with known status [@rosenberg2014]. 
+Briefly, these simulations were implemented 
+with the FLR packages [@kell2007] 
+for the statistical software R, and, in particular, the FLBRP package.
+The framework takes a series of 
+life-history parameters and fishery characteristics
+to generate population projections
+and resulting catch timeseries.
+Life-history values (e.g. mean asymptotic length)
+for three fish life histories (small pelagic,
+demersal, and large pelagic) 
+were translated into a complete set of parameters for
+a von Bertalanffy growth model,
+a maturity ogive,
+natural mortality, 
+a selectivity function,
+and a Beverton-Holt stock-recruitment function\R{A4}
+using the life-history relationships derived in @gislason2008.
+Fishing scenarios included three levels of initial biomass depletion
 compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying
-capacity; and four exploitation settings: (1) a constant exploitation rate,
+capacity; and four exploitation patterns: (1) a constant exploitation rate,
 (2) an exploitation rate coupled with biomass to mimic an open-access
@@
 of factors adding stochastic draws of recruitment and catch-recording
-variability each time to generate a total of 5760 stocks. The simulation
-models were built in the `FLR` packages [@kell2007,
-<http://flr-project.org>] for the statistical software R. Code to
-generate the simulations is available at
-<https://github.com/iagomosqueira/stockStatusFAO>.
+variability each time to generate a total of 5760 stocks. 
+Code to generate the simulations is available at
+<https://github.com/iagomosqueira/stockStatusFAO> (last accessed 2016-09-27).
 
@@
 densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year
-cycles. We include the results of adding these additional covariates in the
+cycles. For the linear model superensemble, we incorporated the two spectral
+covariates (S1, S2) along with all second-order interactions as:\R{D}
+
+\begin{equation}
+\hat{\theta_i} = \beta_0 + \beta_1 \hat{b}_{i,1} + ... +
+\beta_{1,2} \hat{b}_{i,1} \hat{b}_{i,2} + ... +
+\beta_{S1} S1_i + 
+\beta_{S2} S2_i + 
+\beta_{1,S2} \hat{b}_{i,1} S2_i + ... +
+\epsilon_i,
+\end{equation}
+
+\noindent
+with $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and for
+simulated fisheries $i$ 1 through $n$.
+We include the results of adding these additional covariates in the
 supplementary materials.
@@
 both for the individual and superensemble models. An alternative would be to
-fit superensemble models to subsets of the individual models that did converge,
+fit separate superensemble models to subsets of the individual models that did converge,
 but for simplicity we will only use superensemble models fitted to all four
-individual models.
+individual models.\R{A6}
 
@@
 (for the individual model and superensemble) was separate from the dataset with
-which it was tested.
+which it was tested. This meant that, for each iteration of cross validation, 
+we split the RAM database into three, fit the mPRM model to two-thirds of
+the RAM database, fit a superensemble with this version of mPRM, and then 
+tested the performance of the superensemble on the third of the RAM database
+we had withheld.\R{A7}
 
@@
 each of the four data-limited models and the ensembles. Code to reproduce our
-analysis is available at <https://github.com/datalimited/ensembles> (*Private while in
-review*).
+analysis is available at <https://github.com/datalimited/ensembles> (last accessed 2016-09-27).
 
@@
 
-Ensemble methods, and in particular the machine learning superensemble models
+Ensemble methods, and in particular the machine learning superensemble models\R{A8}
 (random forest and GBM), generally improved estimates of status and trend over
@@
 high status, the true status also tends to be high (Figs \ref{lm-coefs},
-\ref{partial-2d-sim}l). The addition of spectral density covariates helped the
+\ref{partial-2d-sim}f). \R{A9}The addition of spectral density covariates helped the
 superensemble models correctly predict higher status values (Fig.
@@
 quantities of interest (such as the slope of $B/B_\mathrm{MSY}$) provides an additional
-calibration step [@rykiel1996]. Individual models are typically calibrated
-(i.e., parameters are estimated) based on a single response variable; however,
-this model might then provide a biased or inaccurate estimate of other
-quantities of interest. Ensemble methods provide a final calibration step to a
-quantity of interest. This ensemble calibration could include a loss function
+calibration step to a quantity of interest [@rykiel1996].\R{A10}
+This ensemble calibration could include a loss function
 tailored to the goals of the model, say placing greater weight on accuracy at

### 2016-11-04 text/ms.Rmd
Update paper to deal with rest of reviewer comments

@@
 superensemble regression fit to observed data. Their superensemble was
-considerably more accurate than any individual prediction or weighted average
-of predictions.
+considerably more accurate than any individual prediction or an average
+of the predictions.
 
@@
 using the life-history relationships derived in @gislason2008.
+
 Fishing scenarios included three levels of initial biomass depletion
@@
 correlation of $0.6$. The simulation also included a scenario with $N(0,
-0.2^2)$ measurement error around log(catch) and one scenario without
+0.2^2)$ measurement error around log catch  and one scenario without
 measurement error. @rosenberg2014 ran ten iterations for each combination
@@
 cycles. For the linear model superensemble, we incorporated the two spectral
-covariates (S1, S2) along with all second-order interactions as:\R{D}
+covariates ($S1$, $S2$) along with all second-order interactions as:\R{D}
 
@@
 \beta_{S2} S2_i + 
+\beta_{S1,S2} S1_i S2_i + 
 \beta_{1,S2} \hat{b}_{i,1} S2_i + ... +
@@
 based on version 2.5. After removing stocks for which at least one of the
-individual models described below did not converge (121), this database
+individual models did not converge (121), this database
 included `r ram_stocks_n` stocks. We removed these stocks for all methods ---
@@
 fit separate superensemble models to subsets of the individual models that did converge,
-but for simplicity we will only use superensemble models fitted to all four
+but for simplicity we only used superensemble models fitted to all four
 individual models.\R{A6}
@@
 
+<!-- It is often important to provide measures of uncertainty with quantitative -->
+<!-- estimates. --> 
+\R{B}
+We illustrated that superensembles can improve point
+estimates of status and trends in status; however, there is no reason why
+superensembles cannot also be used to
+provide measures of uncertainty around those point estimates. 
+First, the same approaches to deriving measures of uncertainty from any regression model are available to a superensemble. 
+For example, likelihood profile confidence intervals or Bayesian credible intervals 
+are available for superensembles fit via maximum likelihood or Bayesian procedures.
+Measures of predictive uncertainty can be generated for machine learning
+methods such as random forests or GBMs using 
+bootstrap procedures [e.g. @hastie2009; @finnegan2015].
+Furthermore, uncertainty from the component models could 
+be included in superensembles.
+These superensembles could be fit using any errors-in-variables or
+measurement-error modelling approach [e.g. @carroll2006]
+
 Multi-model inference in the form of coefficient averaging weighted by
@@
 tailored to a specific response and loss function, superensembles force a
-modeller to choose an operational purpose for their model up front [*sensu*
-@dickey-collas2014].
+modeller to choose an operational purpose for their model upfront [*sensu*
+@dickey-collas2014]. For instance, one could have an ensemble 
+estimate of $B$ and an ensemble estimate of $B_0$, but their
+ratio may not be the same as an ensemble estimate of $B/B_0$.
+A modeller might therefore choose to focus on $B/B_0$, 
+which provides a unitless ratio, is easier to compare across stocks, 
+and the ratio is often a more stable estimate across models [@deroba2015].
 
@@
 that such investigation of the structure of a superensemble may lead to
-improvement in the mechanisms assumed in the individual models.
+improvement in the mechanisms assumed in individual models.
 
@@
 
+\R{A}Superensembles can help address some longstanding
+issues in fisheries science.
+For example, superensembles assist by not needing to decide on one model ---
+instead of deciding on dome versus asymptotic fisheries 
+selectivity [e.g., @sampson2012], 
+or on whether to fix or estimate natural mortality [e.g., @johnson2014],
+superensembles can use multiple models to draw inference.
+Furthermore, the relative contributions of individual models can help tease apart 
+the conditions under which various model assumptions 
+result in the most accurate predictions.
+Finally, superensembles can be used to directly 
+estimate other quantities of interest in fisheries science. 
+For instance, superensembles could be used to estimate 
+fishing mortality compared to fishing mortality at MSY ($F/F_\mathrm{MSY}$)
+or a fisheries selectivity curve parameter.
+However, superensembles are not a panacea 
+and are ultimately limited by the quality, breadth, and representativeness
+of simulated or trusted data to which superensembles are calibrated.
+
 \noindent
@@
 comments during the development of this project, and R.D. Methot for comments
-on an earlier version of the manuscript. A.A. Rosenberg is president of MRAG
-Americas Inc, a consulting company specializing in fisheries science. We thank
+on an earlier version of the manuscript. We thank
 the Gordon and Betty Moore Foundation for funding the working group "Applying

### 2016-11-04 text/ms.Rmd
Copyediting

@@
 
-\R{A}Superensembles can help address some longstanding
+\R{A}Superensembles may be able to assist with some longstanding
 issues in fisheries science.
@@
 and are ultimately limited by the quality, breadth, and representativeness
-of simulated or trusted data to which superensembles are calibrated.
+of simulated or trusted data to which they are calibrated.
 

### 2016-11-04 text/ms.Rmd
More copyediting

@@
 We illustrated that superensembles can improve point
-estimates of status and trends in status; however, there is no reason why
+estimates of population status and trends in status; however, there is no reason why
 superensembles cannot also be used to
 provide measures of uncertainty around those point estimates. 
-First, the same approaches to deriving measures of uncertainty from any regression model are available to a superensemble. 
+The same approaches to deriving measures of uncertainty from any regression model are available to a superensemble. 
 For example, likelihood profile confidence intervals or Bayesian credible intervals 
-are available for superensembles fit via maximum likelihood or Bayesian procedures.
+are available for superensembles fit via 
+maximum likelihood or Bayesian procedures, respectively.
 Measures of predictive uncertainty can be generated for machine learning
@@
 useful in other subfields of fisheries science and ecology in general.
-Predictions about extinction risk are widely used at the national (e.g., the US
+\R{A}In fisheries science, superensembles provide an 
+additional tool to assist with some longstanding issues.
+For example, superensembles assist since modelers need not 
+decide on one model ---
+instead of deciding on dome versus asymptotic fisheries 
+selectivity [e.g., @sampson2012], 
+or on whether to fix or estimate natural mortality [e.g., @johnson2014],
+superensembles can use multiple models to draw inference.
+Furthermore, the relative contributions of individual models can help tease apart 
+the conditions under which various model assumptions 
+result in the most accurate predictions.
+Finally, superensembles can be used to directly 
+estimate other quantities of interest in fisheries science. 
+For instance, superensembles could help assess overfishing by estimating 
+fishing mortality compared to fishing mortality at MSY ($F/F_\mathrm{MSY}$)
+or be trained to estimate natural mortality.
+
+More broadly, in ecology 
+predictions about extinction risk are widely used at national (e.g., the US
 Endangered Species Act and the Canadian Species at Risk Act) and international
@@
 distribution given climate impacts [@harsch2014].
-
-\R{A}Superensembles may be able to assist with some longstanding
-issues in fisheries science.
-For example, superensembles assist by not needing to decide on one model ---
-instead of deciding on dome versus asymptotic fisheries 
-selectivity [e.g., @sampson2012], 
-or on whether to fix or estimate natural mortality [e.g., @johnson2014],
-superensembles can use multiple models to draw inference.
-Furthermore, the relative contributions of individual models can help tease apart 
-the conditions under which various model assumptions 
-result in the most accurate predictions.
-Finally, superensembles can be used to directly 
-estimate other quantities of interest in fisheries science. 
-For instance, superensembles could be used to estimate 
-fishing mortality compared to fishing mortality at MSY ($F/F_\mathrm{MSY}$)
-or a fisheries selectivity curve parameter.
-However, superensembles are not a panacea 
+In any case, superensembles are not a panacea 
 and are ultimately limited by the quality, breadth, and representativeness

### 2016-11-04 text/ms.Rmd
Tweaks to main text

@@
 
+\noindent \R{A5}For this illustrative example 
+we chose this level of model complexity *a priori*
+but a modeller could apply model selection via information-theoretic
+or cross-validation approaches.
+
 Our two machine learning superensemble models, a random forest and a
@@
 modeller to choose an operational purpose for their model upfront [*sensu*
-@dickey-collas2014]. For instance, one could have an ensemble 
+@dickey-collas2014]. \R{E}For instance, one could have an ensemble 
 estimate of $B$ and an ensemble estimate of $B_0$, but their

### 2016-11-04 analysis/5-simulation-ensembles.R
Add AIC check of model complexity

@@ -131,6 +131,14 @@ nvar <- 6L
 
 # -------------------
 
+mlm <- lm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
+  spec_freq_0.05 + spec_freq_0.2, data = d_mean)
+mlm2 <- lm(log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
+ spec_freq_0.05 + spec_freq_0.2)^2, data = d_mean)
+mlm3 <- lm(log(bbmsy_true_mean) ~ (CMSY + COMSIR + mPRM + SSCOM +
+ spec_freq_0.05 + spec_freq_0.2)^3, data = d_mean)
+bbmle::AICtab(mlm, mlm2, mlm3)
+
 m <- gbm::gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
     spec_freq_0.05 + spec_freq_0.2, distribution = "gaussian",
   data = d_mean, n.trees = 2000L, interaction.depth = 6, shrinkage = 0.01)

### 2016-11-07 text/ms.Rmd
Add a ,

@@
 
-More broadly, in ecology 
+More broadly, in ecology, 
 predictions about extinction risk are widely used at national (e.g., the US

### 2016-11-08 text/ms.Rmd
Update addresses
Copyediting

@@
 
-\noindent ^10^European Commission, Joint Research Centre, Institute for the
-Protection and Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi
-2749, 21027 Ispra (VA), Italy
+\noindent ^10^European Commission, DG Joint Research Centre, Directorate D --
+Sustainable Resources, Unit D.02 Water and Marine Resources, Via E. Fermi 2749,
+21027 Ispra VA, Italy
 
@@
 additional tool to assist with some longstanding issues.
-For example, superensembles assist since modelers need not 
+For example, superensembles are helpful since modelers need not 
 decide on one model ---

### 2016-11-08 text/ms.Rmd
Tweak links

@@
 provide a full background on these four methods \R{C} 
-(<http://www.fao.org/docrep/019/i3491e/i3491e00.htm>, last accessed 2016-09-27) 
+(<http://www.fao.org/docrep/019/i3491e/i3491e00.htm>, last accessed 2016-11-08) 
 and code to fit all the models
@@
 software R [@r2015] <https://github.com/datalimited/datalimited> 
-(last accessed 2016-09-27). In summary:
+(last accessed 2016-11-08). In summary:
 
@@
 Code to generate the simulations is available at
-<https://github.com/iagomosqueira/stockStatusFAO> (last accessed 2016-09-27).
+<https://github.com/datalimited/stocksims> (last accessed 2016-11-08).
 
@@
 each of the four data-limited models and the ensembles. Code to reproduce our
-analysis is available at <https://github.com/datalimited/ensembles> (last accessed 2016-09-27).
+analysis is available at <https://github.com/datalimited/ensembles> (last accessed 2016-11-08).
 

### 2016-12-22 analysis/0-ensemble-functions.R
Clean up analysis; make sure reproducible

@@ -98,11 +98,11 @@ cross_val_ensembles <- function(.n, dat, fraction_train = 0.5,
 }
 
 add_label <- function(xfrac, yfrac, label, pos = 4, add_bg = FALSE, ...) {
-# xfrac The fraction over from the left side.
-# yfrac The fraction down from the top.
-# label The text to label with.
-# pos Position to pass to text()
-# ... Anything extra to pass to text(), e.g. cex, col.
+  # xfrac The fraction over from the left side.
+  # yfrac The fraction down from the top.
+  # label The text to label with.
+  # pos Position to pass to text()
+  # ... Anything extra to pass to text(), e.g. cex, col.
   u_ <- par("usr")
   width <- u_[2] - u_[1]
   height <- u_[4] - u_[3]
@@ -129,13 +129,10 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   par(mfrow = c(rows, 4), mgp = c(1.5, 0.5, 0), las = 1, tck = -0.03,
     oma = oma, cex = 0.8, mar = c(0, 0, 0.27, 0),
     xaxs = "i", yaxs = "i", col.axis = "grey60", col.lab = "grey50")
-  #par(family="serif")
 
   hexcol1 <- RColorBrewer::brewer.pal(9, "Blues")
-  #hexcol1 <- c("#FFFFFF", "#6A4A3C")
   hexcol2 <- RColorBrewer::brewer.pal(9, "Oranges")
   hexcol2 <- RColorBrewer::brewer.pal(9, "YlOrRd")
-  #hexcol2 <- c("#FFFFFF", "#CC333F")
   hexcol3 <- RColorBrewer::brewer.pal(9, "Greens")
   hexcol4 <- RColorBrewer::brewer.pal(9, "Purples")
   hexcol_ensemble <- RColorBrewer::brewer.pal(9, "Greys")
@@ -145,31 +142,7 @@ plot_hex_fig <- function(dat, xbins = 100L, xlab = expression(B/B[MSY]),
   hexcol3 <- hexcol2
   hexcol4 <- hexcol2
   library("RColorBrewer")
-  # hexcol1 <- c("#F9F9F9", brewer.pal(4, "Dark2")[1])
-  # hexcol2 <- c("#F9F9F9", brewer.pal(4, "Dark2")[2])
-  # hexcol3 <- c("#F9F9F9", brewer.pal(4, "Dark2")[3])
[…]

### 2016-12-22 analysis/3-load-sim-data.R

@@ -2,12 +2,10 @@ library("dplyr")
 
 # Bring in all results exept CMSY:
 # TODO: local file path:
-load("../../../Dropbox/FisheriesWorkingGroupPhaseII/Decision_trees/CMSY_COMSIR_SSCOM_COSTELLO_STO_BATCH_ALL_RESULTS 2013-06-24 .Rdata")
+d1 <- readRDS("raw-data/batch1result.rds")
 
 # bringing in corrected fits below
-d1 <- dplyr::filter(batch1results, method_id != "CMSY")
-
-rm(batch1results)
+d1 <- dplyr::filter(d1, method_id != "CMSY")
 
 d1$stock_id <- as.character(d1$stock_id)
 d1$ID <- as.numeric(as.character(d1$ID))
@@ -19,7 +17,6 @@ d1$seed <- as.numeric(as.character(d1$seed))
 d1$n_iterations <- as.numeric(as.character(d1$n_iterations))
 d1$effective_sample_size <- as.numeric(as.character(d1$effective_sample_size))
 
-# TODO: data only local:
 load("raw-data/cmsy_fits_22052015.RData")
 d2 <- cmsynewfits
 rm(cmsynewfits)

### 2016-12-22 analysis/5-simulation-ensembles.R

@@ -23,7 +23,7 @@ dsim <- dsim %>%
 
 # summarise the mean and slope in the last N years:
 library("doParallel")
-registerDoParallel(cores = 4)
+registerDoParallel(cores = parallel::detectCores())
 
 dsim_sum <- plyr::ddply(dsim, c("stockid", "method", "iter"),
   .parallel = TRUE, .fun = mean_slope_bbmsy)
@@ -98,39 +98,6 @@ nvar <- 6L
 # plot(mm_gbm, metric = "Rsquared")
 # dev.off()
 
-# -----------------
-# # try and avoid positive bias at low bbmsy with weights:
-# library(randomForest)
-# library(gbm)
-# m1 <- randomForest(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
-#     spec_freq_0.05 + spec_freq_0.2, data = d_mean)
-# d_mean$rf1 <- exp(predict(m1, n.trees = 1000))
-# ggplot(d_mean, aes(bbmsy_true_mean, rf1)) + geom_point() +
-#   geom_abline(slope = 1, intercept = 0, col = "red")
-#
-# m2 <- gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
-#     spec_freq_0.05 + spec_freq_0.2, data = d_mean, distribution = "gaussian",
-#   n.trees = 1000L, interaction.depth = 4, shrinkage = 0.01)
-# d_mean$gbm2 <- exp(predict(m2, n.trees = 1000))
-# ggplot(d_mean, aes(bbmsy_true_mean, gbm2)) + geom_point() +
-#   geom_abline(slope = 1, intercept = 0, col = "red")
-#
-# # try with weights inverse to true b/bmsy
-# m2_weights <- gbm(log(bbmsy_true_mean) ~ CMSY + COMSIR + mPRM + SSCOM +
-#     spec_freq_0.05 + spec_freq_0.2, data = d_mean, distribution = "gaussian",
-#   n.trees = 1000L, interaction.depth = 4, shrinkage = 0.01,
-#   weights = 1/d_mean$bbmsy_true_mean)
-# d_mean$gbm2_weights <- exp(predict(m2_weights, n.trees = 1000))
-# ggplot(d_mean, aes(bbmsy_true_mean, gbm2_weights)) + geom_point() +
[…]

### 2016-12-22 analysis/8-main-figs.R

@@ -21,7 +21,8 @@ d_sim <- readRDS("generated-data/cv_sim_long.rds")
 d_sim <- suppressWarnings(inner_join(d_sim, clean_names))
 
 get_performance <- function(validation_data) {
-  d_sim_perf <- validation_data %>% group_by(test_iter, type, clean_method, order, label_fudge_x, label_fudge_y) %>%
+  d_sim_perf <- validation_data %>% group_by(test_iter, type, clean_method,
+    order, label_fudge_x, label_fudge_y) %>%
     summarise(
       mare = median(abs(re)),
       mre = median(re),
@@ -91,7 +92,7 @@ re_ram_sum_long$label_fudge_y <- 0
 
 get_roc <- function(true, est) {
   library("pROC") # required or pROC:: will generate errors
-  y <- pROC::roc(response = true, predictor = est, 5)
+  y <- pROC::roc(response = true, predictor = est)
   data.frame(sens = y$sensitivities, spec = y$specificities,
     auc = as.numeric(y$auc))
 }
@@ -100,9 +101,14 @@ get_roc <- function(true, est) {
 # we'll use 0.5 B/Bmsy as the treshold as in US overfished definition
 d_sim_mean <- d_sim_basic %>% filter(type == "mean") %>%
   mutate(above1 = ifelse(bbmsy_true > 0.5, 1, 0))
-rocs_sim_mean <- d_sim_mean %>% group_by(clean_method, order) %>%
-  do({get_roc(true = .$above1, est = .$bbmsy_est)}) %>%
-  as.data.frame
+
+library("doParallel")
+registerDoParallel(cores = parallel::detectCores())
+rocs_sim_mean <- plyr::ddply(d_sim_mean, c("clean_method", "order"), function(x) {
+  get_roc(true = x$above1, est = x$bbmsy_est)
+}, .parallel = TRUE)
+
+
 rocs_sim_mean$Ensemble <- ifelse(grepl("ensemble", rocs_sim_mean$clean_method, ignore.case = TRUE),
   "(a) Ensemble", "(b) Individual")
 auc_sim_mean <- rocs_sim_mean %>%
@@ -144,7 +150,6 @@ make_roc_curves <- function(dat) {
 ## Figure making:
 
 ## ROC curves:
[…]

### 2016-12-22 analysis/README.R

@@ -1,0 +1,18 @@
+install.packages(c("mblm", "randomForest", "dplyr", "plyr", "ggplot2",
+  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC"))
+devtools::install_github("datalimited/datalimited")
+
+# start in the root "ensembles" folder, then:
+setwd("analysis")
+source("0-ensemble-functions.R")
+source("1-read-old-ram-fits.R")
+source("2-merge-ram-fits.R")
+source("3-load-sim-data.R")
+source("4-add-spectral.R")
+source("5-simulation-ensembles.R")
+source("6-sim-to-ram.R")
+source("7-plot-sims.R")
+source("8-main-figs.R")
+source("8-motivate-fig.R")
+source("9-output-values.R")
+

### 2016-12-22 analysis/3-load-sim-data.R
Delete old data

@@ -1,7 +1,6 @@
 library("dplyr")
 
 # Bring in all results exept CMSY:
-# TODO: local file path:
 d1 <- readRDS("raw-data/batch1result.rds")
 
 # bringing in corrected fits below

### 2016-12-22 analysis/README.R
Add datalimited dependencies

@@ -1,5 +1,6 @@
 install.packages(c("mblm", "randomForest", "dplyr", "plyr", "ggplot2",
-  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC"))
+  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC",
+  "Rcpp", "R2jags"))
 devtools::install_github("datalimited/datalimited")
 
 # start in the root "ensembles" folder, then:

### 2016-12-22 analysis/README.R
Add árm'package

@@ -1,6 +1,6 @@
 install.packages(c("mblm", "randomForest", "dplyr", "plyr", "ggplot2",
   "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC",
-  "Rcpp", "R2jags"))
+  "Rcpp", "R2jags", "arm"))
 devtools::install_github("datalimited/datalimited")
 
 # start in the root "ensembles" folder, then:

### 2016-12-22 analysis/README.R
Add bbmle

@@ -1,6 +1,6 @@
 install.packages(c("mblm", "randomForest", "dplyr", "plyr", "ggplot2",
   "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC",
-  "Rcpp", "R2jags", "arm"))
+  "Rcpp", "R2jags", "arm", "bbmle"))
 devtools::install_github("datalimited/datalimited")
 
 # start in the root "ensembles" folder, then:

### 2016-12-22 analysis/6-sim-to-ram.R
No parallel on Windows

@@ -185,8 +185,9 @@ cv_ensemble_ram <- function(nfold = 3L, .n = 1L) {
 
 library("doParallel")
 registerDoParallel(cores = parallel::detectCores())
+.parallel <- ifelse(Sys.info()[["sysname"]] == "Windows", FALSE, TRUE)
 qq <- plyr::ldply(seq_len(100L), function(i) cv_ensemble_ram(nfold = 3L, .n = i),
-  .parallel = TRUE)
+  .parallel = .parallel)
 
 d_mean_long <- qq %>%
   select(-spec_freq_0.05, -spec_freq_0.2) %>%

### 2016-12-22 analysis/8-main-figs.R
Really - no parallel on Windows
(avoiding needed foreach exports)

@@ -104,9 +104,10 @@ d_sim_mean <- d_sim_basic %>% filter(type == "mean") %>%
 
 library("doParallel")
 registerDoParallel(cores = parallel::detectCores())
+.parallel <- ifelse(Sys.info()[["sysname"]] == "Windows", FALSE, TRUE)
 rocs_sim_mean <- plyr::ddply(d_sim_mean, c("clean_method", "order"), function(x) {
   get_roc(true = x$above1, est = x$bbmsy_est)
-}, .parallel = TRUE)
+}, .parallel = .parallel)
 
 
 rocs_sim_mean$Ensemble <- ifelse(grepl("ensemble", rocs_sim_mean$clean_method, ignore.case = TRUE),

### 2016-12-23 analysis/8-motivate-fig.R
Remove more dropbox links

@@ -8,52 +8,49 @@ ram <- read.csv("raw-data/RAM_bmsy_Ctousev4.csv", stringsAsFactors=FALSE) %>%
   select(tsyear, Bbmsy_toUse) %>%
   rename(yr = tsyear, b_bmsy = Bbmsy_toUse)
 
-load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/CMSY/cmsy_rlegacy_results_table_v0.RData")
-cmsy <- cmsy.rlegacy.df0 %>%
-  filter(stock_id == "SBWHITACIR") %>%
-  select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
-  mutate(method = "CMSY")
+# load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/CMSY/cmsy_rlegacy_results_table_v0# .RData")
+# cmsy <- cmsy.rlegacy.df0 %>%
+#   filter(stock_id == "SBWHITACIR") %>%
+#   select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
+#   mutate(method = "CMSY")
+# saveRDS(cmsy, file = "generated-data/SBWHITACIR-cmsy.rds")
+cmsy <- readRDS("generated-data/SBWHITACIR-cmsy.rds")
 
-load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/COMSIR/comsir_rlegacy_results_table_v0.RData")
-comsir <- comsir.rlegacy.df0 %>%
-  filter(stock_id == "SBWHITACIR") %>%
-  select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
-  mutate(method = "COM-SIR")
+# load("~/Dropbox/FisheriesWorkingGroupPhaseII/RAM_Legacy_fits/COMSIR/comsir_rlegacy_results_table_v0# .RData")
+# comsir <- comsir.rlegacy.df0 %>%
+#   filter(stock_id == "SBWHITACIR") %>%
+#   select(yr, b_bmsyiq25, b_bmsy, b_bmsyiq75) %>%
+#   mutate(method = "COM-SIR")
+# saveRDS(comsir, file = "generated-data/SBWHITACIR-comsir.rds")
+comsir <- readRDS("generated-data/SBWHITACIR-comsir.rds")
 
[…]

### 2016-12-23 analysis/README.R
Add RColorBrewer

@@ -1,5 +1,6 @@
 install.packages(c("mblm", "randomForest", "dplyr", "plyr", "ggplot2",
-  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC"))
+  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC",
+  "RColorBrewer"))
 devtools::install_github("datalimited/datalimited")
 
 # start in the root "ensembles" folder, then:

### 2016-12-23 README.Rmd
Update documentation

---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# Improving estimates of population status and trend with superensemble models

This repository holds the analysis for:
Anderson, S. C., A. B. Cooper, O. P. Jensen, C. Minto, J. T. Thorson, J. C. Walsh, M. Dickey-Collas, K. M. Kleisner, C. Longo, G. C. Osio, D. Ovando, I. Mosqueira, A. A. Rosenberg, and E. R. Selig. 2017. Improving estimates of population status and trend with superensemble models. Fish and Fisheries. In press.

To recreate the analysis, source the `.R` files in the `analysis` folder in order. Or run the following in R:

```{r}
packages <- c("mblm", "randomForest", "plyr", "dplyr", "ggplot2",
  "reshape2", "gbm", "devtools", "doParallel", "hexbin", "pROC",
  "RColorBrewer", "Rcpp", "R2jags", "arm", "bbmle")
```

```{r, eval=FALSE}
install.packages(packages)
devtools::install_github("datalimited/datalimited")

# start in the root "ensembles" folder, then:
setwd("analysis")
source("0-ensemble-functions.R")
source("1-read-old-ram-fits.R")
source("2-merge-ram-fits.R")
source("3-load-sim-data.R")
source("4-add-spectral.R")
source("5-simulation-ensembles.R")
source("6-sim-to-ram.R")
source("7-plot-sims.R")
source("8-main-figs.R")
source("8-motivate-fig.R")
source("9-output-values.R")
```

Or run `make` on the command line from the `analysis` folder:

```sh
cd analysis
make
```

Generated data ends up in `generated-data` and figures in `figs`. Intermediate versions of some raw data have been cached in the `generated-data` folder so that the scripts will run from a freshly cloned version of this repository. (Some of the raw data files are too large for Git.)

## Package versions used:

```{r, echo=FALSE}
packages <- c("datalimited", packages)
x <- suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE))
x <- devtools::session_info()
dplyr::filter(x$packages, package %in% packages) %>%
  dplyr::arrange(package) %>%
  dplyr::select(-source, -`*`) %>%
  knitr::kable()
```

### 2016-12-23 README.Rmd
Remove text folder because of publisher requirements

@@
 
-Generated data ends up in `generated-data` and figures in `figs`.
-Intermediate versions of some raw data have been cached in the `generated-data` folder so that the scripts will run from a freshly cloned version of this repository.
+Generated data ends up in `analysis/generated-data` and figures in `figs`.
+Intermediate versions of some raw data have been cached in the `analysis/generated-data` folder so that the scripts will run from a freshly cloned version of this repository.
 (Some of the raw data files are too large for Git.)

### 2016-12-23 text/pseudocode.R

@@ -1,23 +1,0 @@
-# Fit models to known data:
-cmsy_status <- cmsy(...)
-comsir_status <- comsir(...)
-# Combine and reshape output into wide format (not shown)
-head(data_known)
-# > cmsy_status comsir_status known_status
-# > 1.4         1.3           1.2
-# > ...         ...           ...
-# Build super ensemble model:
-ens_model <- lm(known_status ~ cmsy_status + comsir_status,
-  data = data_known)
-
-# Fit models to data of interest:
-cmsy_status <- cmsy(...)
-comsir_status <- comsir(...)
-# Combine and reshape output into wide format (not shown)
-head(data_unknown)
-# > cmsy_status comsir_status ...
-# > 0.9         1.1
-# > ...         ...
-# Predict status using super ensemble model:
-predict(model, newdata = data_unknown)
-# > 1.06  0.98  0.85 ...

### 2016-12-23 analysis/test-mprm.R
Delete old files

@@ -1,36 +1,0 @@
-# compare the datalimited::prm function to the original fits
-library("datalimited")
-library("dplyr")
-library("ggplot2")
-
-d <- dplyr::inner_join(datalimited::ram_ts, datalimited::spp_categories,
-  by = "scientificname")
-ram_prm_dat <- plyr::ddply(d, "stockid", function(x) {
-  datalimited::format_prm(year = x$year, catch = x$catch,
-    bbmsy = x$bbmsy_ram,
-    species_cat = x$spp_category[1L])
-})
-
-m <- datalimited::fit_prm(ram_prm_dat)
-d <- ram_prm_dat
-d <- cbind(d, datalimited::predict_prm(d, ci = TRUE))
-d <- dplyr::rename(d, package = fit)
-
-ram <- readRDS("generated-data/ram_fits.rds")
-qq <- subset(ram, method == "Costello")
-q <- dplyr::inner_join(select(qq, stockid, tsyear, b_bmsy_est),
-  select(d, stockid, year, package) %>% rename(tsyear = year))
-qlong <- reshape2::melt(q, id.vars = c("stockid", "tsyear"))
-
-ggplot(dplyr::filter(qlong, stockid %in% unique(qlong$stockid)[1:25]),
-  aes(tsyear, value)) + facet_wrap(~stockid) +
-  geom_line(aes(colour = variable))
-
-# CIs:
-ggplot(dplyr::filter(d, stockid %in% unique(d$stockid)[1:10]),
-  aes(year, package)) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#FF000040") +
-  geom_line() +
-  ylab(expression(B/B[MSY])) +
-  facet_wrap(~stockid)
-

### 2017-07-18 analysis/99-mprm-test.R
Add mprm test

@@ -1,0 +1,43 @@
+# compare the datalimited::prm function to the original fits
+# start with setwd("analysis") from the ensembles root folder
+
+library("datalimited")
+library("dplyr")
+library("ggplot2")
+
+d <- dplyr::inner_join(datalimited::ram_ts, datalimited::spp_categories,
+  by = "scientificname")
+ram_prm_dat <- plyr::ddply(d, "stockid", function(x) {
+  datalimited::format_prm(year = x$year, catch = x$catch,
+    bbmsy = x$bbmsy_ram,
+    species_cat = x$spp_category[1L])
+})
+
+m <- datalimited::fit_prm(ram_prm_dat)
+d <- ram_prm_dat
+d <- cbind(select(d, stockid), datalimited::predict_prm(d, ci = TRUE))
+d <- dplyr::rename(d, package = bbmsy_q50)
+
+ram <- readRDS("generated-data/ram_fits.rds") # original fits
+qq <- subset(ram, method == "Costello")
+q <- dplyr::inner_join(select(qq, stockid, tsyear, b_bmsy_est),
+  select(d, stockid, year, package) %>% rename(tsyear = year))
+qlong <- reshape2::melt(q, id.vars = c("stockid", "tsyear"))
+
+set.seed(99)
+ids <- sample(unique(qlong$stockid), size = 24)
+ggplot(dplyr::filter(qlong, stockid %in% ids),
+  aes(tsyear, value)) + facet_wrap(~stockid) +
+  geom_line(aes(colour = variable))
+
+ggplot(q, aes(b_bmsy_est, package)) + geom_point(alpha = 0.1) +
+  coord_fixed() +
+  geom_abline(intercept = 0, slope = 1, colour = "red")
+
+## CIs:
+# ggplot(dplyr::filter(d, stockid %in% unique(d$stockid)[1:10]),
+#   aes(year, package)) +
+#   geom_ribbon(aes(ymin = bbmsy_q2.5, ymax = bbmsy_q97.5), fill = "red", alpha = 0.3) +
+#   geom_line() +
+#   ylab(expression(B/B[MSY])) +
+#   facet_wrap(~stockid)

### 2017-07-18 analysis/99-ohi.Rmd
Add OHI ROC analysis

---
title: "ROC for OHI thresholds"
author: "Sean Anderson"
date: '2017-07-18'
output: html_document
---

Let's read in the data, source some functions, and manipulate the data so we can create some ROC curves at relevant thresholds for the OHI.

```{r}
library(dplyr)
library(ggplot2)
source("0-ensemble-functions.R")

# bring in the simulated data and shape it for plotting:
clean_names <- dplyr::data_frame(
  method = c("CMSY", "COMSIR", "mPRM", "SSCOM",
    "gbm_ensemble", "rf_ensemble", "lm_ensemble", "mean_ensemble"),
  clean_method = c("CMSY", "COM-SIR", "mPRM", "SSCOM",
    "GBM Superensemble", "RF Superensemble", "LM Superensemble", "Mean Ensemble"),
  order = c(1, 2, 4, 3, 8, 7, 6, 5),
  label_fudge_x = c(
    0, -0.03,0,-0.04,
    0, 0,0,-0.05),
  label_fudge_y = c(
    0, -0.02,0.03,0.04,
    0, 0.02,0,-0.02))

d_sim <- readRDS("generated-data/cv_sim_long.rds")
d_sim <- suppressWarnings(inner_join(d_sim, clean_names))

get_performance <- function(validation_data) {
  d_sim_perf <- validation_data %>% group_by(test_iter, type, clean_method,
    order, label_fudge_x, label_fudge_y) %>%
    summarise(
      mare = median(abs(re)),
      mre = median(re),
      corr = cor(bbmsy_true_trans, bbmsy_est_trans, method = "spearman",
        use = "pairwise.complete.obs"))

  d_sim_perf$clean_method <- as.factor(d_sim_perf$clean_method)
  d_sim_perf$clean_method <- reorder(d_sim_perf$clean_method, d_sim_perf$order)

  d_sim_perf_slope <- filter(d_sim_perf, type == "slope")
  d_sim_perf <- filter(d_sim_perf, type == "mean")

  d_sim_perf_long <- reshape2::melt(d_sim_perf,
    id.vars = c("test_iter", "clean_method", "type", "order", 
      "label_fudge_x", "label_fudge_y"))
  d_sim_perf_summ <- d_sim_perf_long %>%
    group_by(clean_method, order, variable, label_fudge_x, label_fudge_y) %>%
    summarise(
      m = median(value),
      l = quantile(value, 0.25),
      u = quantile(value, 0.75)) %>%
    as.data.frame
  d_sim_perf_summ$text_col <- ifelse(grepl("ensemble", 
    d_sim_perf_summ$clean_method, ignore.case = TRUE),
    "grey20", "grey50")
  d_sim_perf_wide_temp <- reshape2::dcast(d_sim_perf_summ, 
    clean_method + order ~ variable,
    value.var = "m")
  d_sim_perf_wide_temp
}
d_sim_perf_wide <- get_performance(d_sim)
saveRDS(d_sim_perf_wide, "generated-data/d_sim_perf_wide.rds")

# Format basic simulation ensemble without covariates for plotting:
d_sim_basic <- readRDS("generated-data/cv_sim_mean_basic_long.rds")
d_sim_basic <- suppressWarnings(inner_join(d_sim_basic, clean_names))
d_sim_perf_wide_basic <- get_performance(d_sim_basic)
saveRDS(d_sim_perf_wide_basic, "generated-data/d_sim_perf_wide_basic.rds")
d_sim_basic$bbmsy_est[d_sim_basic$bbmsy_est > 10] <- NA
d_sim_basic <- na.omit(d_sim_basic)
```

Get ROC and AUC for models applied to simulation data with b/bmsy as the response. We'll use 0.8 B/Bmsy as the threshold to start with.

```{r}
library("doParallel")
registerDoParallel(cores = parallel::detectCores())
.parallel <- ifelse(Sys.info()[["sysname"]] == "Windows", FALSE, TRUE)

d_sim_mean <- d_sim_basic %>% filter(type == "mean") %>%
  mutate(above_0.8 = ifelse(bbmsy_true > 0.8, 1, 0)) %>% 
  mutate(above_1.45 = ifelse(bbmsy_true > 1.45, 1, 0)) %>% 
  mutate(below_1.45_above_0.8 = ifelse(bbmsy_true < 1.45 & bbmsy_true > 0.8, 1, 0)) %>% 
  mutate(clean_method = paste(order, clean_method, sep = "-"))

d_sim_mean$Ensemble <- ifelse(grepl("ensemble", 
  d_sim_mean$clean_method, ignore.case = TRUE),
  "(a) Ensemble", "(b) Individual")

get_roc <- function(true, est) {
  library("pROC") # required or pROC:: will generate errors
  y <- pROC::roc(response = true, predictor = est)
  data.frame(sens = y$sensitivities, spec = y$specificities,
    auc = as.numeric(y$auc))
}

rocs_sim_mean_0.8 <- plyr::ddply(d_sim_mean, c("clean_method", "Ensemble"), 
  function(x) {
    get_roc(true = x$above_0.8, est = x$bbmsy_est)
  }, .parallel = .parallel)

rocs_sim_mean_1.45 <- plyr::ddply(d_sim_mean, c("clean_method", "Ensemble"), 
  function(x) {
    get_roc(true = x$above_1.45, est = x$bbmsy_est)
  }, .parallel = .parallel)

rocs_sim_mean_0.8_1.45 <- plyr::ddply(d_sim_mean, c("clean_method", "Ensemble"), 
  function(x) {
    get_roc(true = x$below_1.45_above_0.8, est = x$bbmsy_est)
  }, .parallel = .parallel)

saveRDS(rocs_sim_mean_0.8, file = "generated-data/rocs_sim_mean_0.8.rds")
saveRDS(rocs_sim_mean_1.45, file = "generated-data/rocs_sim_mean_1.45.rds")
saveRDS(rocs_sim_mean_0.8_1.45, file = "generated-data/rocs_sim_mean_0.8_1.45.rds")

# auc_sim_mean <- rocs_sim_mean %>%
#   group_by(clean_method) %>%
#   summarise(auc = auc[1])
```

Now make the curves:

```{r}
make_roc_curves <- function(dat) {
  p <- ggplot(dat, aes(spec, sens, colour = clean_method, group = clean_method)) +
    geom_line() + coord_equal() +
    geom_abline(intercept = 1, slope = 1, lty = 2, col = "darkgrey") +
    xlim(1, 0) +
    xlab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as < threshold)) +
    ylab(expression(Rate~of~correctly~categorizing~B/B[MSY]~as > threshold)) +
    theme_bw() +
    theme(plot.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()) + facet_wrap(~Ensemble) +
    scale_colour_brewer(palette="Spectral", guide = guide_legend(title = "Model"))
  p
}
p <- make_roc_curves(rocs_sim_mean_0.8)
ggsave("../figs/ohi-roc-sim-0.8.png", width = 8, height = 5)

p <- make_roc_curves(rocs_sim_mean_1.45)
ggsave("../figs/ohi-roc-sim-1.45.png", width = 8, height = 5)

p <- make_roc_curves(rocs_sim_mean_0.8_1.45)
ggsave("../figs/ohi-roc-sim-0.8-1.45.png", width = 8, height = 5)
```

### 2017-07-18 analysis/99-ohi.Rmd
Expand documentation on OHI ROC

@@
 Let's read in the data, source some functions, and manipulate the data so we can create some ROC curves at relevant thresholds for the OHI.
+The data we are working with is the cross validated (out-of-sample) simulation data.
 
 ```{r}
+setwd("analysis")
 library(dplyr)
@@
 
-Get ROC and AUC for models applied to simulation data with b/bmsy as the response. We'll use 0.8 B/Bmsy as the threshold to start with.
+Get ROC and AUC for models applied to simulation data with b/bmsy as the response.
+We'll test 3 threshold levels.
+0.8, 1.45, and > 0.8 but < 1.45.
+The last one checks whether the model correctly put b/bmsy between 0.8 and 1.45 vs.
+above or below.
 
@@
 saveRDS(rocs_sim_mean_0.8_1.45, file = "generated-data/rocs_sim_mean_0.8_1.45.rds")
+```
 
-# auc_sim_mean <- rocs_sim_mean %>%
-#   group_by(clean_method) %>%
-#   summarise(auc = auc[1])
+Let's read in the model output and calculate the AUC values (area under the curve). 0.5 is as good as a coin flip. 1.00 is perfect.
+
+```{r}
+rocs_sim_mean_0.8 <- readRDS("generated-data/rocs_sim_mean_0.8.rds")
+rocs_sim_mean_1.45 <- readRDS("generated-data/rocs_sim_mean_1.45.rds")
+rocs_sim_mean_0.8_1.45 <- readRDS("generated-data/rocs_sim_mean_0.8_1.45.rds")
+
+rocs_sim_mean_0.8 %>%
+  group_by(clean_method) %>%
+  summarise(auc = round(auc[1], 2)) %>% 
+  arrange(-auc)
+
+rocs_sim_mean_1.45 %>%
+  group_by(clean_method) %>%
+  summarise(auc = round(auc[1], 2)) %>% 
+  arrange(-auc)
+
+rocs_sim_mean_0.8_1.45 %>%
+  group_by(clean_method) %>%
+  summarise(auc = round(auc[1], 2)) %>% 
+  arrange(-auc)
 ```
